### 2019-09-19 draft.tex
first commit

\documentclass{comjnl}

\usepackage{amsmath}

\usepackage{pagecolor}
%\copyrightyear{2009} \vol{00} \issue{0} \DOI{000}

% random
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{nicefrac}       % compact symbols for 1/2, etc.

% double column box
\usepackage{cuted,tcolorbox,lipsum}

% code embeds
\usepackage{listings}
\usepackage{color}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
  language=Go,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

\begin{document}
\pagecolor{white}

\title[An event-sourced and distributed database with IPLD and materialized views]{An event-sourced and distributed database with IPLD and materialized views}
\author{Pick}
\author{Farmer}
\author{Sutula}
\author{Hill}
\affiliation{www.textile.io} \email{contact@textile.io}

\shortauthors{www.textile.io}

\revised{18 September 2019}

%\category{C.2}{Computer Communication Networks}{Computer Networks}
%\category{C.4}{Performance of Systems}{Analytical Models}
%\category{G.3}{Stochastic Processes}{Queueing Systems}
%\terms{Internet Technologies, E-Commerce}
\keywords{Databases; Event Sourcing; IPFS; IPLD}

\begin{abstract}
As the Internet expands, the division between a person's digital and physical existence continues to blur. An emerging issue of concern is that the digital part of a person's life is mainly being captured by apps and stored away from the person’s control by the companies building those apps. An alternative, called user-siloed data is an exciting area of research that if implemented successfully, could reverse the flow of value derived from personal data, from apps to users. In this paper, we investigate the data formats, and storage and transfer protocols necessary to build a system for user ownership of large-scale digital datasets. The aim of the proposed system is to help power a new generation of web technologies. Our solution combines a novel use of event sourcing with Interplanetary Linked Data (IPLD) to provide a distributed, scalable, and flexible database solution for decentralized applications.
\end{abstract}

\maketitle

\section{Introduction}

Compared to their predecessors, modern cloud-based apps and services provide an extremely high level of convenience. Users can entirely forget about the risk of data loss and enjoy seamless access to their apps across multiple devices. This convenience is now expected, but has come at the cost of additional consequences for users. One such consequence is that many of the same development patterns that bring convenience (e.g., single sign-on, minimal encryption, centralized servers, and databases) also enable or even require data hoarding by the apps. While collecting large amounts of user’s data can create value for the companies building apps (e.g., app telemetry, predictive tools, or even new revenue streams), that value flows mostly in one direction; Apps collect, process and benefit from a user's private data, but users rarely have access to the original data or the new insights that come from it. Additionally,  the data may not be readily accessible to other apps and services. This is app-siloed data.

While companies collecting user data may not be itself a significant problem, it has been shown that over time company incentives often shift from providing value to users to extracting value from users [ref Dixon]. When this incentive shift happens for companies that have also been hoarding user data, that data may be a new source of value or even revenue for the company. It may be possible to stop this trend through extreme privacy measures, government intervention/legislation, or by boycotting companies that collect any form of data. Ideally, there is an alternative approach that allows individuals to capture the value of their data and still allow developers to build new interfaces and experience on top of this data. This approach is called user-siloed data and it fundamentally separates apps from their user’s data.

One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data. Other projects have identified app silo-ed data as a problem [ref], and some have identified user-siloed data as a solution [ref Solid]. However, none so far have provided a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data. This is the key to building great applications on user-siloed data.

In this paper, we study existing technologies that could be used to build a network for user-siloed data. We outline six challenge areas: flexible data format, efficient synchronization, conflict resolution, access-control, scalable storage, and network communication. Based on our findings, we propose a novel architecture for event sourcing with Interplanetary Linked Data (IPLD), designed to store, share, and host user-sioled datasets at scale. Our proposed design leverages new and existing protocols to solve major challenges involved with building a secure and distributed network for user data while at the same time providing the flexibility and scalability required by today's applications.

\section{Background}
\label{sec:Background}

We now describe some of the technologies and concepts motivating the design of our novel distributed database system. We highlight some of the advantages and lessons that can be learned from event-sourcing and discuss drawbacks to using these approaches in distributed systems. We provide an overview of some important technologies related to IPFS that make it possible to rethink event sourcing in a decentralized network. Finally, we cover challenges of security and access control on an open and decentralized network and discuss how they are used in popular databases external to IPFS.

\subsection{Data Synchronization}

\subsubsection{CQRS, Event Sourcing, and Logs}

When developing large-scale software systems, it is common to store data in a relational database management system (RDBMS). To model realistic systems, this type of framework often requires complex techniques for mapping data between domain models and database tables, where the same data model is used to both query and update a database. An powerful alternative is to use a set of append only logs to model the state of an object simply by applying its change sequence in the correct order. This concept can be expressed succinctly by the state machine approach [ref]: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state. This is a powerful concept baked into a simple structure, and is at the heart of many distributed database systems [ref].

\begin{definition} (Logs or Append-only log). A log is a registry of database transactions that is read sequentially (normally ordered by time) from beginning to end. In distributed systems, logs are often treated as append-only, where changes or updates can only be added to the set and never removed.  \end{definition}

In modern applications it is critical to have reliable mechanisms for publishing updates and events (i.e., to support event-driven architectures), scalability (optimized write and read operations), forward-compatible application updates (e.g., code changes, retroactive events, and others), auditing systems, etc. To support such requirements, developers have begun to utilize event sourcing and command query responsibility segregation (CQRS) patterns [ref], relying on append only logs to support immutable state histories. Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate event sourcing and CQRS-based applications, including Event Store [ref], Apache Kafaka [ref] and Samza [ref], among others [see ref].

\begin{definition} (Command query responsibility segregation). Command query responsibility segregation or CQRS is a design pattern whereby reads and writes are separated into different models, using commands to write data, and queries to read data [ref]. \end{definition}

\begin{definition} (Event sourcing). Event sourcing (ES) is a design pattern for persisting the state of an application as an append-only log of state-changing events. \end{definition}

ES is particularly useful in contexts where system components are distributed or decentralized, such as in decentralized applications (DApps). This is because the design of systems that use event sourcing often necessitate infrastructure for brokering messages between loosely coupled software components and services. This feature will become increasingly important as we outline background concepts related to data synchronization in the following sections.

A key principal of ES and append only logs is that all changes to application state are stored as a sequence of events. Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates [ref]. The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see section  \ref{sec:viewsprojections}). In other words, with minimal conformity, a single log can model multiple application states [ref].

\subsubsection{Views \& Projections} \label{sec:viewsprojections}

\begin{definition}  (View). A (typically highly de-normalized) read-only model of the data. Views are tailored to the interfaces and display requirements of the application, which helps to maximize both display and query performance. Views that are backed by a database or filesystem-optimized access are referred to as materialized views. \end{definition}

\begin{definition} (Projection). An event handler and corresponding reducer/fold function used to build and maintain a view from a set of (filtered) events. While projections may lead to the generation of new events, their reducer should be a pure function (see [ref] and [ref] for examples from existing systems). \end{definition}

In CQRS and ES, the separation of write operations from read operations is a powerful concept. It allows developers to define views into the underlying data that are best suited for the use case or user interface they are building. Multiple views can be built from the same underlying event log, and they can be quite different from one another.

Views themselves are enabled by projections. Projections can be thought of as transformations or reducers that are applied to each event in a stream of events. They update the data backing the views, be this in memory, or persisted to a database. In a distributed setting, it may be necessary for projections to define and operate as eventually consistent data structures, to ensure all peers operating on the same stream of events have a consistent representation of the data.

\subsubsection{Eventual Consistency}

The CAP theorem [ref, ref] states that a distributed database can guarantee only two of the following three promises at the same time: consistency (i.e., that every read receives the most recent write or an error), availability (i.e., that every request receives a (possibly out-of-date) non-error response, and partition tolerance (i.e., that the system continues to operate despite an arbitrary number of messages being dropped (or delayed) by the network). As such, many distributed systems are now designed to provide availability and partition tolerance by trading consistency for eventual consistency. Eventual consistency allows state replicas to diverge temporarily, but eventually arrive back to the same state again. While an active area of research, designing systems with provable eventual consistency guarantees remains challenging [ref needed].

\begin{definition} (CRDT). A conflict-free replicated data type (CRDT) assures eventual consistency through optimistic replication (i.e. all new updates are allowed) and eventual merging. CRDTs rely on data structures that are mathematically guaranteed to resolve concurrent updates the same way regardless of the order those events were received. \end{definition} \label{def:crdt}

How a system provides eventual consistency is often decided based on the intended use of the system. Two well-documented categories of solutions include logs (see Definition X.x.x), sequences of deterministic changes, and CRDTs (see Definition XXX). Logs work best when there is only a single-writer or complete event causality is known. In many distributed systems with multiple log writers, a minimum requirement for synchronization through logs is that the essential order of events is respected and can be determined [3, 8]. For these cases, logical clocks are a useful tool for eventual consistency and total ordering [ref]. However, some scenarios (e.g., temporarily missing events, or ambiguous order) can force a replica into a state that cannot be later resolved without costly recalculation. In specific cases, CRDTs can provide an alternative to log-based consensus (see below).

\subsubsection{Logical Clocks}

In a distributed system, where multiple machines, each with an independent clock, are creating events, local timestamps can’t be used to construct global event causality. Machine clocks are never perfectly synchronized [1], meaning that one machine's concept of "now" is not necessarily the same as another machine's. Machine speed, network speed, and other factors compound the issue. For this reason, simple wall-clock time does not provide a sufficient notion of order in a distributed system. Alternatives to wall-clock time exist to help distributed systems achieve eventual consistency. Examples include various logical clocks (Lamport [ref below], Schwartz [ref below], and Bloom [ref below], Hybrid variants [ref below], etc), which use counter-based time-stamping to provide partial ordering.

Cryptographically linked events can also represent a clock. One such example is called the Merkle-Clock [ref below]. The Merkle-Clock relies on properties of a Merkle-DAG to provide strict partial ordering between events, which has limitations [ref section 4.3 of the merkle paper]:

\begin{quote}
Merkle-Clocks represent a strict partial order of events. Not all events in the system can be compared and ordered. For example, when having multiple heads, the Merkle-Clock cannot say which of the events happened before.
\end{quote}

See Section \ref{sec:merkleclocks}.

\subsubsection{Conflict-Free Replicated Data Types}

In distributed models that use clock-based ordering as described above, a replica can arrive in a state that will require conflict resolution through consensus or rollback. Those operations can be expensive, having adverse effects on the scalability of distributed systems, especially ones attempting to synchronize an application state.

CRDTs (see \ref{def:crdt}) are one way to achieve strong eventual consistency, where once all replicas have received the same events, they will arrive at the same final state, regardless of ordering. A review of the types of possible CRDTs is beyond the scope of this paper, however, it is important to note their role in eventually consistent systems and how they relate to clock-based event ordering. See for example [Merkle Clock ref] and [ref] for informative reviews of these types of data structures.

Whether a system (e.g., an app) uses a CRDT or a clock-based sequence of events is entirely dependent on the use-case and final data model. While CRDTs may seem superior (and are currently a popular choice among decentralized systems), it is not possible to model every system as a CRDT. Additionally, the simplicity of clock-based sequencing often makes it easier to leverage in distributed systems where data conflicts will only rarely arise. Lastly, logs and CRDTs are not mutually exclusive and can be used together or as different stages of a larger system.

\subsection{Content-based addressing}

Internet application architecture today is often designed as a system of clients (users) communicating to endpoints (hosts). Communication between clients and endpoints (servers) happens via the TCP/IP protocol stack and therefore largely relies on a mechanism referred to as location-based addressing. Location-based addressing, where the client makes a request that is routed to a specific endpoint based on prior knowledge (e.g., the domain name or IP address), works relatively well for many use-cases. However, there are many reasons why addressing content by location is problematic, such as duplication of storage, inefficient use of bandwidth, link rot, centralized control, and authentication issues. An alternative to location addressing, called content addressing may provide a solution to those problems. Content addressing is where the content itself is used to create an address and retrieve that content from a network [2].

\subsubsection{IPFS \& Content-based addressing}

There are a number of systems that utilize content addressing to access content and information [list some as refs], and there is an active body of literature covering its design and implementation [ref]. The Interplanetary File System (IPFS) — which is a set of protocols to create a content-addressed, peer-to-peer filesystem [1] — is one such system. In IPFS, the address for any content is determined based on a cryptographic hash of the content itself. In practice, the IPFS Content IDentifier (CID) is a multihash, which is a self-describing “protocol for differentiating outputs from various well-established cryptographic hash functions, addressing size + encoding considerations” [ref]. That addressing system confers several benefits to the network, including tamper resistance (we can be confident that a given piece of content has not been modified en route if its hash matches what we were expected/requested) and de-duplication (because the same content from different peers will produce the same hash address). Additionally, IPFS content addresses are immutable and universally unique.

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{The crypographic hash of content is used to make a request to the network of IPFS peers. Using the built-in routing machanisms and the distributed hash table, a peer(s) hosting the requested content is identified and content is returned. }
  \label{fig:contentaddressing}
\end{figure}

While content addressing doesn’t tell you how to get a file, IPFS (via libp2p [ref]) provides a system for moving files across the network.  On the IPFS network, a client who wants specific content requests the CID from the network of IPFS hosts. The client's request is routed to the first host capable of fulfilling the request — i.e., the first host that is actively storing the content behind the given CID. The IPFS network can be seen as a distributed file system, with all of the benefits [ref] that come with this type of file system design.

\subsubsection{IPLD}

As discussed above, IPFS uses the cryptographic hash of a given piece of content to define its content address (see [ref] for details on this process). However, in order to provide standards for accessing content-addressable data (on the web or elsewhere), it is necessary to define a common format or specification. In IPFS (and other projects [ref]), this common data format is called Interplanetary Linked Data (IPLD) [ref]. As the name suggests, IPLD is based on principals of linked data [ref, ref] with the added capabilities of a content-addressing storage network.

IPLD is used to publish linked data (subject-predicate-object triples in the linked data world [ref]) spread across different hosts, and where everything (entities, predicates, data sources) uses content addresses as unique identifiers [ref]. To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph. This allows all hash-linked data structures to be treated using a unified data model, analogous to linked data in the Semantic Web sense [ref]. In practice, IPLD is represented as objects, each with `Data` and `Links` fields, where `Data` can be a small blob of unstructured, arbitrary binary data, and `Links` is an array of links to other IPLD objects.

\subsubsection{Merkle-Clocks} \label{sec:merkleclocks}

A Merkle-Clock is a Merkle-DAG that represents a sequence of events. In other words, a Merkle-Clock is an append-only log [ref and ipfs-log]. When implemented on IPFS (or an equivalent network where content can be cryptographically addressed and fetched), Merkle-Clocks provide a number of benefits for data synchronization between replicas [ref merkle paper section 4.3]:

\begin{enumerate}
\item \label{Perf1}Broadcasting the Merkle-Clock requires broadcasting only the current root CID. The whole Clock is unambiguously identified by the CID of its root and its full DAG can be walked down from it as needed.
\item \label{Perf2}The immutable nature of a Merkle-DAG allows every replica to perform quick comparisons and fetch only those nodes that it does not already have.
\item \label{Perf3} Merkle-DAG nodes are self-verified and immune to corruption and tampering. They can be fetched from any source willing to provide them, trusted or not.
\item \label{Perf4} Identical nodes are de-duplicated by design: there can only be one unique representation for every event.
\end{enumerate}

On the downside, and as seen above (section 2.2.4.), a Merkle-Clock cannot order divergent heads (or roots):

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{Divergent heads in a multi-writer Merkle-DAG }
  \label{fig:merkledag}
\end{figure}

In the figure above, two replicas (top row, bottom row) are attempting to write (left to right) events to the same Merkle-Clock. After the first replica writes event A, the second writes event A’ and properly links to A. At that point, the two replicas stop receiving events from one another. To a third replica that does continue to receive events, there would now be two independent heads, 1 and 1’. For the third replica, resolving these two logs of events may be costly (many updates happened since the last common node) or impossible (parts of the chain may not be available on the network). See section 2.2.3. Eventual Consistency.

In order to reduce the likelihood of divergent heads, all replicas should be perfectly connected and be able to fetch all events and linkages in the Merkle-Clock. On real networks with many often offline replicas (mobile and IoT devices, laptops, etc.), these conditions are rarely met.

Based on these observations, using a single Merkle-Clock to synchronize replicas can be problematic.

\subsection{Networking}

So far we have primarily discussed the mechanics of creating or linking content in a series of updates. Now we will overview some common networking tools for connecting distributed peers who aim to maintain replicas of a shared database. This could be any decentralized network of interacting entities (e.g., cloud servers, IoT devices, botnets, sensor networks, mobile apps, etc) collectively updating a shared state. IPFS contains a collection of protocols and systems to help address the networking needs required by different use-cases. That is to say no matter what type of device we are talking about — be it a phone, desktop computer, browser, or Internet-enabled appliance — it should be able to communicate with other devices. However, each of the networking approaches described below comes with strengths and weaknesses when used to synchronize data.

\subsubsection{Libp2p}

The libp2p project provides a robust protocol communication stack. IPFS and a growing list of other projects [ref] are building on top of libp2p. Libp2p solves a number of challenges that are distinct to peer-to-peer (p2p) networks. A comprehensive coverage of networking issues in P2P systems is out of the scope for this paper, however, some core challenges that libp2p helps to address include NAT traversal, peer discovery and handshake protocols, and even encryption and transport security — libp2p supports both unencrypted (e.g. TCP, UDP) and encrypted protocols (e.g. TLS, Noise) — among others. Libp2p uses the concept of a multiaddress to address peers on a network, which essentially models network addresses as arbitrary encapsulations of protocols [ref, ref]. In addition to “transport layer” modules, libp2p provides several tools for sharing and/or disseminating data over a p2p network.

\subsubsection{Pubsub}

One of the most commonly used p2p distribution layers built on libp2p, is its Pubsub (or publish-subscribe) system. Pubsub is a standard messaging pattern where the publishers don’t know who, if anyone, will subscribe to a given topic. Publishers send messages on a given topic or category, and Subscribers receive only messages on a give topic they are subscribed to. Libp2p’s pubsub module can be configured to utilize a floodsub protocol — which floods the network with messages, and peers are required to ignore messages they are not interested in — or gossipsub — which is a proximity-aware epidemic pubsub, where peers communicate with proximal peers, and messages can be routed more efficiently. In those implementations, there is a benefit to using pubsub in that no direct connection between publishers and subscribers is required.

Another benefit to using pubsub is the ability to publish topical sequences of updates to multiple recipients. Like libp2p, encryption is a separate concern and often added in steps prior to data transmission. However, like libp2p, pubsub doesn’t offer any simple solutions for transferring encryption keys (beyond public keys), synchronizing datasets across peers (i.e. they aren’t databases), or enforcing any measures for access control (e.g. anyone subscribed to a topic can also author updates on that topic). To solve some of those challenges, some systems introduce message echoing and other partial solutions. However, it makes more sense to use pubsub and libp2p as building blocks in systems that can effectively solve these issues, by choosing multi-modal communication strategies or leveraging tools such as deferred routing (e.g. inboxing) for greater tolerance of missed messages.

\subsubsection{IPNS}

Pubsub and libp2p have so far dealt with the push transfer of data, but IPFS also offers a useful technology for hosting pull/request based data endpoints called, IPNS. IPNS aims to address the challenge of mutable data within IPFS. IPNS relies on a global namespace (shared by participating IPFS peers) based on Public Key Infrastructure. By using IPNS, a content creator generates a new address in the global namespace and points that address to an endpoint (e.g. a CID). Using their private key, a content creator can update what static route the IPNS address refers to. IPNS isn’t only useful for creating static addresses that can point to content addresses, IPNS is compatible with a naming system external from IPFS, such as DNS, onion, or bit addresses. However, many use-cases that require highly mutable data, require rapid availability of updates, or want flexible multi-party access control may not yet be suitable for IPNS. Taken together, libp2p, pubsub, IPNS, and IPFS more generally provide a useful toolkit for building robust abstractions to deliver fast, scalable, data synchronization in the decentralized network.

\subsection{Data Access \& Control}

\subsubsection{Identity}

IPFS is an implementation of public key infrastructure, where every node on the IPFS network has a key-pair. In addition to using the key-pair for secure communication between nodes, IPFS also uses the key-pair as the basis for identity. Specifically, when a new IPFS node is created, a new key-pair is generated, and then the public key is transformed into the nodes Peer ID.

\subsubsection{Agent-centric security}

Agent-centric security refers to the practice of maintaining data integrity without leveraging a central or blockchain-based consensus. The general approach is just to let the reader enforce permissions and perform validations, not the writer or some central authority. Agent-centric security is possible if the reader can reference local-only, tamper-free code or if the state can be used to determine whether a given operation (e.g. delete data) is permitted. Many decentralized networks like Secure Scuttlebutt [1] and Holochain [2] make use of agent-centric security. Each of these systems leverage cryptographic signatures to validate peer identities and messages.

\subsubsection{Access control}

All filesystems and databases have some notion of "access control." Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects [ref]. An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object.

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{Example Access Control List.}
  \label{fig:boat1}
\end{figure}

Using ACLs in systems where identity is derived from various configurations of public-key infrastructure has been around for some time [1, 2]. Still, many existing database and communication protocols built on IPFS to date lack support for an ACL or only have primitive ACL support. Where ACLs are missing, many systems use cryptographic primitives like signature schemes or enable encryption without any role-based configuration. Even more, many systems deploy an all-or-none security model, where those with access to a database have complete access, including write capabilities. ACLs should be mutable over time and permission to modify an ACL should also be recorded in an ACL.

Event-driven systems (e.g., event sourcing) often make use of ACLs with some distinct properties. The ACL of an event-driven system is usually a list of access rules built from a series of events. For example, the two events, "grant Bob write access" and "revoke read access from Alice" would together result in a final ACL state where, Bob has read and write access, but Alice does not (see Views \& Projections in Data Synchronization).

\section{The Threads protocol}

We propose Threads, a protocol and decentralized database that runs on IPFS meant to help decouple apps from user-data. The Threads protocol is designed based on an underlying, CQRS-inspired event-sourcing system for synchronizing data across collaborating peers on a network. Threads offer data ownership and a multi-role data access architecture where owners can set independent permissions for writing, reading, and following data updates. Threads differ from previous solutions by extending on the multiaddress addressing scheme to allow PULL based replica synchronization in addition to more common PUSH based synchronization seen in decentralized protocols. The flexible event-based structure enables client applications to model advanced states through aggregates, views, and custom CRDTs.

Threads are topic-based collections of single-writer logs. Taken together, these logs represent the current “state” of an object or dataset. The basic units of Threads, namely Logs and Events, provide a framework for users to create, store, and transmit data in a P2P distributed network. By structuring the underlying architecture in specific ways, this framework can be deployed to solve many of the problems discussed above.

\subsection{ Event Logs}

\subsubsection{Single-writer Event Logs}

In multi-writer systems, determining causal order in all replicas is particularity challenging. As previously discussed (section 2.2.4), a solution based on a Merkle-Clock is only partially ordered. That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy [ref]:

\begin{quote}
	A total order can be useful ... and could be obtained, for example, by considering concurrent events to be equal. Similarly, a strict total order could be built by sorting concurrent events by the CID or their nodes or by any other arbitrary user-defined strategy based on additional information attached to the clock nodes (data-layer conflict resolution).
\end{quote}

In many cases, the total order of an event sequence is meaningful, and this type of deterministic merge strategy in not sufficient. A git merge highlights one such case, in which additional information is sometimes needed in order to resolve a line conflict. Ledger based transactions present a similar challenge where the absolute order of transactions between multiple peers really does matter.

Our solution to deal with divergent Merkle-Clocks is to institute a single-writer rule: A log can only be updated by a single replica or identity. An Event Log is a single-writer Merkle-Clock that can be totally ordered. Separate Event Logs can be composed into advanced structures, including CRDTs [Ref Enes].

Practically speaking, for any given Log, Events are authored by a single IPFS Peer, or Writer.

\begin{definition}
(Writer). The single IPFS Peer capable of writing to an Event Log.
\end{definition}

This single-writer setup is a core feature of Logs, and provides properties unique to the Threads protocol. For clarity, we can similarly define a Reader as any other Peer capable of reading a Log.

\begin{definition}
(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log’s Read key (see section 3.1.3 Keys \& Encryption).
\end{definition}

As seen above in section 2.3.4., a Merkle-Clock is simply a Merkle-DAG of Events:

\begin{definition}
 (Event). An single node in a Merkle-Clock, stored on IPFS.
\end{definition}

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{A single-writer Merkle-Clock (Event Log).}
  \label{fig:boat1}
\end{figure}

Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure X.X.) enabling log verification by readers (see section 3.1.3. Keys \& Encryption).

At a minimum, a node must link to its most immediate ancestor. However, links to older ancestors are often included as well to improve concurrency during traversal and verification [ref].

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{This is really a code embed.}
  \label{fig:boat1}
\end{figure}

\begin{lstlisting}
// Node is the most basic component of a log.
// Note: In practice, this is encrypted with the Follow key.
type Node interface {
    ipld.Node

    // Event is the Log update.
    Event() Event

    // Refs are node linkages.
    Refs() []cid.Cid

    // Sig is a cryptographic signature of Event and Refs
    // created with the Log's private key.
    Sig() []byte
}

// Event represents the content of an update.
// Note: In practice, this is encrypted with the Read key.
type Event interface {
    ipld.Node

    // Header provides a means to store a timestamp
    // and a key needed for decryption.
    Header() EventHeader

    // Body contains the content of an update.
    // In practice, this is encrypted with the Header key
    // or the recipient's public key.
    Body() ipld.Node

    // Decrypt is a helper function that decrypts Body
    // with a key in Header.
    Decrypt() (ipld.Node, error)
}

// EventHeader contains Event metadata.
type EventHeader interface {
    ipld.Node

    // Time is the wall-clock time at which the Event
    // was created.
    Time() int

    // Key is an optional single-use symmetric key
    // used to encrypt Body.
    Key() []byte
}
\end{lstlisting}

As seen above, an Event’s actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.

\subsubsection{Multi-addressed Event Logs}

Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses [ref]. Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (see inset). In practice, to reach a Log via it’s IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (see inset).

Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table [ref] (DHT). Instead, they are collected from Log Events (see section 3.3 Use Cases). This is in contrast to mutable data via IPNS for example, which requires querying the network (DHT) for updates. Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed like topology. Log addresses are recorded in an address book, similar to IPFS Peer address book:

\begin{lstlisting}
// AddrBook stores log addresses
type AddrBook interface {
   AddAddr(thread.ID, peer.ID, ma.Multiaddr, time.Duration)
   AddAddrs(thread.ID, peer.ID, []ma.Multiaddr, time.Duration)
   SetAddr(thread.ID, peer.ID, ma.Multiaddr, time.Duration)
   SetAddrs(thread.ID, peer.ID, []ma.Multiaddr, time.Duration)
   UpdateAddrs(t thread.ID, id peer.ID, oldTTL time.Duration, newTTL time.Duration)
   Addrs(thread.ID, peer.ID) []ma.Multiaddr
   ClearAddrs(thread.ID, peer.ID)
}
\end{lstlisting}

Addresses can expire by specifying a time-to-live (TTL) value when adding or updating them in the address book, which allows for unresponsive addresses to eventually be removed.

Modern, real-world networks consist of many mobile or otherwise sparsely connected computers (Peers). Therefore, datasets distributed across such networks can be thought of as highly partitioned. To ensure updates are available between mostly offline or otherwise disconnected Peers (like mobile devices), Textile Logs are designed with a built-in replication or follower mechanism.

\begin{definition}Follower). Log Writers can designate other IPFS Peers to “follow” a Log, potentially replicating and/or republishing Events. A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (see section 3.1.3 Keys \& Encryption), but is not able to read the Log’s contents. Followers should be server-based, i.e., always online and behind a public IP address.\end{definition}

Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log (see inset).

In practice, Writers are solely responsible for announcing their Log’s addresses. This ensures a conflict-free address list without additional complexity. Some Followers may be in the business of replicating Logs (see section 5.1. Textile: The Thread \& Bot Network), in which case Writers will announce the additional Log address to Readers. This allows them to pull (or subscribe to push-based) Events from the Follower’s Log address when the Writer is offline or unreachable.

\begin{strip}
\begin{tcolorbox}[title=Log Multiaddress]
"/ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab"
\end{tcolorbox}
\end{strip}

\begin{strip}
\begin{tcolorbox}[title=Log IPEL Multiaddress]
"/ip4/127.0.0.1/tcp/1234/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab"
\end{tcolorbox}
\end{strip}

\begin{strip}
\begin{tcolorbox}[title=Log Multiple Multiaddress]
[\\
  "/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab",\\
  "/p2p/12D..dJT6nXY/ipel/12D..bCcW8ab" // Follower\\
]
\end{tcolorbox}
\end{strip}

\subsubsection{Keys \& Encryption}

Textile Logs are designed to be shared, composed, and layered into datasets. Therefore, Logs are encrypted by default in a manner that enables access control (see section 3.4. Access Control) and the Follower mechanism discussed in the previous section.

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{The three layers of Log Event encryption.}
  \label{fig:boat1}
\end{figure}

\begin{definition}
(Identity Key). Every Log requires an asymmetric key-pair that determines ownership and identity. The private key is used to sign each Event added to the Log, so down-stream processes can verify the Log’s authenticity. Like IPFS peers, the public key of the Log is used as an identifier (Log ID).
\end{definition}

The body, or content of an event, is encrypted by a Content Key. Content keys are generated per-content and never reused. The Content key is distributed directly in the header of the Event Block. We define the Content Key as follows,

\begin{definition}
(Content Key). The Content key is a variable-format key used to encrypt the body (content) of an event. This key can be symmetric, asymmetric, or possibly non-existent in cases where encryption is not needed. One of two common encryption choices will typically be used per event, 
\begin{enumerate}
\item \label{Perf1}When broadcasting events to many possible recipients, a single-use symmetric key is generated per unique content body.
\item \label{Perf2}When sending events to specific recipients, the recipient's public key can be used to restrict access from all others.
\end{enumerate}
\end{definition}

If a single-use symmetric key is used for the Content key, it is necessary to distribute each new key to users by including it in the header of the Event Block. Therefore, the Event Block itself is further encrypted using a Read key. The Read key is not distributed within the Log itself but is distributed to any peers to grant them access to the content of the Log.

\begin{definition}
 (Read Key). The Read key is a symmetric key created by the Log owner and used to encrypt the Content Key in each event.
\end{definition}

Finally, the encrypted Event Block, its signature, and the IPLD linkage(s) from an Event to its antecedents are encrypted together using a Follow key. Follow keys allow Logs to be followed by peers on the network who do not have access to any content within the event, and can only see signatures and linkage(s) between Events.

\begin{definition}
(Follow Key). The Follow key is a symmetric key created by the Log owner and used to encrypt the entire event payload before adding the event to the Log.
\end{definition}

Much like the Log address book, Log keys are stored in a key book:

\begin{lstlisting}
// KeyBook stores log keys
type KeyBook interface {
   PubKey(thread.ID, peer.ID) ic.PubKey
   AddPubKey(thread.ID, peer.ID, ic.PubKey) error
   PrivKey(thread.ID, peer.ID) ic.PrivKey
   AddPrivKey(thread.ID, peer.ID, ic.PrivKey) error
   ReadKey(thread.ID, peer.ID) []byte
   AddReadKey(thread.ID, peer.ID, []byte) error
   FollowKey(thread.ID, peer.ID) []byte
   AddFollowKey(thread.ID, peer.ID, []byte) error
}
\end{lstlisting}

\subsection{Threads}

In Textile, the interface to Logs is managed as a Thread. As mentioned previously, a Thread is a collection of Logs on a given topic. Threads are an event-sourced, distributed database, and can be used to maintain a single, collaboratively edited, watched, or hosted dataset across multiple Peers. Threads provide the mechanism to combine multiple Logs from individual authors into singular shared states through the use of either cross-Log sequencing (e.g. using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock [ref]) or a CRDT (see section xx).

\subsubsection{Identity}

A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pub/Sub-based synchronization. TIDs are defined with the following format:

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{The thread id formula.}
  \label{fig:boat1}
\end{figure}

TIDs share some similarities with UUIDs [ref] (version and variant) and IPFS-based CIDs and are multibase encoded [ref] for maximum forward-compatibility.

\begin{definition}
(Multibase Prefix). The encoding type used by the multibase encoder. 1 byte.
\end{definition}

Base32 encoding is used by default. However, any multibase-supported string encoding may be used.

\begin{definition}
(Version). ID format version. 8 bytes max. This allows future version to be backwards-compatible.
\end{definition}

\begin{definition}
(Variant). Used to specify thread-level expectations, like access-control. 8 bytes max. See section 3.2.2 for more about variants.
\end{definition}

\begin{definition}
(Random Number). A random number of a user-specified length. 16 bytes or more.
\end{definition}

\subsubsection{Variants}

Certain ID variants may be more appropriate than the others in specific use cases. For example, Textile provides an access-controlled Thread variant, which supports various collaborative structures — e.g., social media feeds, shared documents, blogs, photo albums, etc.

\begin{definition}
(Raw). This variant declares that consumers are not expected to make additional assumptions. This is the default variant. 
\end{definition}

Example (V1, 128 bit): `bafkxd5bjgi6k4zivuoyxo4ua4mzyy`

\begin{definition}
(Access-Controlled): This variant declares that consumers should assume an access control list is composable from Log Events. The ACL represents a permissions rule set that must be applied when reading data (see section 3.4.2. Access Control).
\end{definition}

Example (V1, 256 bit): `bafyoiobghzefwlidfrwkqmzz2ka66zgmdmgeobw2mimktr5jivsavya`

\subsubsection{Log Synchronization}

Log Writers, Readers, and Followers synchronize the state of their Logs by sending and receiving Events. Inspired by Git [ref], a reference to the latest Event in a Log is referred to as the Head (or sometimes the root). When a new Event is received, Readers and Followers simply advance their Head reference.

Regardless of the network protocol, Events are transported between Peers in a standardized Event Envelope:

\begin{definition}
(Event Envelope). An over-the-wire message containing an Event and the sender’s signature of the Event.
\end{definition}

A new Thread is created by generating a TID and Log. The Log’s creator is the Writer, meaning it has possession of the Log’s Identity, Read, and Follow keys. All of these keys are needed to compose Events. At this point, the Thread only exists on the Writer’s machine. Whether for collaboration, reading, or following, the process of sharing a Thread with other Peers starts by authoring a special Event called an Invite, which contains a set of keys from all of the Thread’s Logs, called a Key Set.

\begin{definition}
(Invite). An Event containing a mapping of Log IDs to Key Sets, which can be used to join a Thread. Threads backed by an access control list (see section 3.4.2. Access Control) will also include the current ACL for the Thread in an Invite. This enables Peers to invite others to only read or follow a Thread, instead of becoming a full-on collaborator, i.e., a new Log Writer.
\end{definition}

\begin{definition}
(Key Set). A set of keys for a Log. Depending on the context, a Key Set may contain the Follow and Read key, or just the Follow key. Encrypted with the recipient’s public key.
\end{definition}

The Invite is authored in the sender’s Log. Because the recipient does not yet have this Log’s Key Set, the Event is encrypted with the recipient’s public key. If the recipient accepts the Invite, they will author another special Event called a Join in a new Log of their own.

\begin{definition}
(Join). An Event containing an invitee’s new Log ID and Key Set, encrypted with the Key Set of the inviting Peer’s Log.
\end{definition}

For a Join to be successful, all Log Writers must receive a copy of the new Key Set so they can properly handle future Events in the new Log. Instead of encrypting a Join with the public key of each existing Writer, we can encrypt a single Join with the Key Set of the inviting Peer’s Log, which the other Writers also have.

Once a Peer has accepted an Invite, it will receive new Events from Log Writers. In cases where the invitee becomes a collaborator, i.e., a Writer, it is also responsible for sending its own Events out to the network.

Sending

Sending is performed in multiple phases because, invariably, some Thread participants will be offline or unresponsive.

\begin{enumerate}
\item \label{Perf1}New Events are pushed directly to the Thread’s other Log Writers.
\item \label{Perf2}New Events are pushed directly to the target Log’s Follower(s), who may not maintain their own Log.
\item \label{Perf3} New Events are published over Libp2p’s gossip-based Pub/Sub infrastructure using TID as a topic, which provides potentially unknown Readers or Followers with an opportunity to consume Events in real-time.
\end{enumerate}

Step 2 above allows for additional push mechanisms, as followers with public IP addresses become relays:

\begin{enumerate}
\item \label{Perf1}New Events may be pushed to web-based participants over a WebSocket.
\item \label{Perf2}New Events may be pushed to the Thread’s other Log Writers via federated notification services like Apple Push Notification Service (APNS) [ref], Google Cloud Messaging (GCM) [ref], Firebase Cloud Messaging (FCM) [ref], and/or Windows Notification Service (WNS) [ref].
\item \label{Perf3} New Events may trigger web-hooks [ref] enabling complex workflows (see Section X below).
\end{enumerate}

Receiving

Similar to sending, there are multiple paths to receiving new Events that together maximize connectivity between Peers who are often offline or unreachable.

\begin{enumerate}
\item \label{Perf1}Log Writers can receive Events directly from the author.
\item \label{Perf2} Events can be pulled from replicating Followers via HTTP, libp2p, RSS, Atom, etc.
    a. In conjunction with push over WebSockets (2a above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events.
\item \label{Perf3} Writers and readers can receive new Events via a Pub/Sub subscription at the TID.
\end{enumerate}

<insert diagram of pulling from a replicating Follower)

\subsubsection{Log Replication}

Replication of Thread data can happen by duplicating/syncing Log data to other Peers.

The notion of the Follow key (see section 3.1.3 Keys \& Encryption) makes duplicating all Log Events trivial. This allows any Peer in the network to be granted the responsibility of replicating data from another Peer without having read access to the data contained within the Log entries. This type of Log replication can act as a data backup mechanism. It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (see section 3.2.3 Log Synchronization).

\subsection{Thread Views} \label{sec:threadviews}

The use of well-established design patterns from the CQRS+ES (and related) literature means the Threads protocol can provide the foundation for many higher-level concepts and technologies. For instance, the public application programming interface (API) for a Thread provides several features that you would expect when operating on a dataset or table within a database. Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more.

Like many CQRS-based systems, the “public” API for Threads revolves around the concept of Views (see background section xxx), which are used to optimize queries on the local event store. A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store. In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be materialized to local persistent storage). This simple subscription system provides a flexible framework for building complex View-driven application logic.

\subsubsection{Default Views}

The core Textile Threads interfaces are also exposed via interfaces compatible with existing datastores and/or communications systems. These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability. In the remainder of this section, we highlight a number of planned database abstractions to provide further utility to developers interested in building on Textile’s Threads protocol.\\

\textbf{Databases} \\

A key/value store built on top of Threads can map database `Put` operations into internal `KeyValueUpdated` events. These Events would then operate on a map-like view model, via an internal “projection“ (reducer function). The key/value view could be persisted or kept in-memory, and would be directly queried via a corresponding `Get` operation. Because Textile uses ES patterns, a `Del` operation would simply lead to an internal `KeyValueDeleted` event, marking the key/value pair for removal from the key/value materialized view.

\begin{lstlisting}
type TextileKVStore interface {
  Put(key string, value interface{}) error
  Get(key string) (interface{}, error)
  Del(key string) error
}
\end{lstlisting}

Similar abstractions could (and will) be used to implement additional database types and functions. The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, but exposes the underlying event sourced data for customizing as needed. Developers should not have to learn a whole new set of terms and tools to take advantage of Textile’s Threads capabilities. Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents:

\begin{lstlisting}
type TextileDocumentStore interface {
  Put(doc Inedexable) error
  Get(key string) (Indexable, error)
  Del(key string) error
  Query(query Query) ([]Indexable, error)
}
\end{lstlisting}

Where `Indexable` can be satisfied by any structure with a  `key` field and and `Query` is taken from the `go-datastore` interface library [ref] or similar. Tables, feeds, counters, and other simple stores can also be built on Threads, and may require ORM-based solutions to support persistence to certain backing stores. Each database style will be implemented as a standalone software library, allowing application developers to pick and choose the solution most useful to the application at hand. Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these “example“ libraries.\\

\textbf{CRDTs} \\

Eventually consistent, CRDT-based stores can also be implemented on top of Threads. CRDT-based stores are particularly useful for managing Views of a document in a multi-peer collaborative editing environment (like GoogleDocs [ref] or similar). For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype [ref] that applies updates on a JSON document View model. Libraries such as Automerge [ref] provide useful examples of reducer function that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available. A practical example of using a JSON CRDT in Textile is given in section 3.4, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.\\

\textbf{Application State} \\

For application developers familiar with common state management issues, tools such as Redux provide an opinionated framework for making state mutations predictable by imposing certain restrictions on how and when updates can happen. In fact, Redux builds on concepts from CRQS and ES itself [ref], and combines them with three core principals [ref]: 1) that the entire “state“ of an application should be stored in a single “store“, 2) where the only way to change the state is to emit an “action“ (or event), which 3) transforms the application state using a “pure“ (i.e., zero side effects) reducer function (projection). In practice, Redux doesn't care how a developer persists the application state, which means a Thread’s event store and read-only materialized views are excellent candidates. This focus on unidirectional data flow in perfectly in line with architecture proposed in this paper, and as such, a compatible Redux API will be built on top of Textile Threads.

\subsection{Thread Extensions}

At an abstract level, the Textile protocol provides a highly distributed framework for building shared, offline first, datastores that are fault tolerant, eventually consistent, and scalable. Any internal implementation details of a compliant Threads “client“ may use any number of well-established design patterns from the CQRS+ES (and related) literature to “extend“ the Threads protocol with additional features and controls. Some examples of extensions that are included by default in Textile’s Threads implementation are outlined in this section to provide some indication of the extensibility of Threads.

\subsubsection{Snapshots and Compaction}

Snapshots are simply the current state of a View at a given point in time. They can be used to rebuild the state of a view without having to query and re-play all previous events. When a Snapshot is available, a Thread Peer can rebuild the state of a given View by replaying any events generated since the latest Snapshot (with a corresponding reduction in the total number of events that need to be processed) using the View’s Reducer function. Multiple peers processing the same Log could create a Snapshot every 1000 events and be guaranteed to create the exact same Snapshot because each Peer’s Event counts are identical.

In practice, Snapshots are written to their own Event log and stored locally. They can potentially be synced (see section 3.2.4) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time). They can similarly be used for initializing state during recovery.

Compaction can similarly speed up re-hydration of state, and is useful when only the latest Event per Event type is required. In addition to being used as a starting point for initialization and recovery, compacted logs can be used to replace the Log from which they were created in order to free up local disk space. In Textile, compaction becomes much easier because a log only has one writer and conflicts are not possible. In practice, Log compaction is a local-only operation (i.e., other Peers do not need to be aware that Compaction was performed).

\subsubsection{Access Control}

One of the most important properties of a shared data model is the ability to apply access control rules. There are two forms of access control possible in Threads, Sequence-level ACLs and Thread-level ACLs. Thread-level access control lists (ACLs) allow creators to specify who can follow, read, write, and delete Thread data. Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition XXX) basis. Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see section 3.3.2 for details). ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible). First, ACLs are fully mutable, allowing developers to create advanced rules for collaboration with any combination of readers, writers, and followers. Second, because ACLs are JSON documents, ACLs can specify their own editing rules (i.e. allowing multiple Thread participants to modify the ACL) in a self-referencing way.

\begin{definition}
(Sequence). An Event Sequence is a series of ordered events referring to a specific entity or object. For example, an ACL JSON document is a single entity made up of a sequence of Thread Events. A Sequence might have a unique UUID (.e.g, id =bafykrq5i25vd64ghamtgus6lue74k) as in the example below.
\end{definition}

fixidexampleabove

Textile’s Threads includes ACL management tooling based on a Role-based access control [ref] pattern, wherein individuals or groups are assigned roles which carry specific permissions. Roles can be added and removed as needed. Textile ACLs can make use of five distinct roles: No-access, Follow, Read, Write, and Delete.

\begin{definition}
(No-access). No access is permitted. This is the default role.
\end{definition}

\begin{definition}
(Follow). Access to Log Follow Keys is permitted. Members of this role are able to verify Events and follow linkages. The Follow role is used to designate a “follower” peer for offline replication and/or backup.
\end{definition}

\begin{definition}
(Read). Access to Log Read Keys is permitted in addition to Follow Keys. Members of this role are able to read Log Event payloads.
\end{definition}

\begin{definition}
(Write). Members of this role are able to author new Events, which also implies access to Log Follow and Read Keys. At the Thread-level, this means authoring a Log. At the document-level, the Write role means that Events in this Log are able to target a particular document.
\end{definition}

\begin{definition}
(Delete). Members of this role are able to delete Events, which implies access to Log Follow Keys. In practice, this means marking an older Event as “deleted”.
\end{definition}

Below is a typical Thread-level ACL JSON document (where id is a unique id), which could be persisted to a local document store as part of a materialized view.

todo: fixidexampleabove

todo: change code type below

\begin{lstlisting}
{
  "_id": "bafykrq5i25vd64ghamtgus6lue74k", // TID
  "default": "no-access",
  "peers": {
    "12D..dwaA6Qe": ["write", "delete"],
    "12D..dJT6nXY": ["follow"],
    "12D..P2c6ifo": ["read"],
  }
}
\end{lstlisting}

The `default` key states the default role for all network peers. The `peers` map is where roles are delegated to specific peers. Here, `12D..dwaA6Qe` is likely the owner, `12D..dJT6nXY` is a designated follower, and `12D..P2c6ifo` has been given read access.

 A Thread-level ACL has it’s own document ACL, which also applies to all other document ACLs:

\begin{lstlisting}
{
  "_id": "bafykrq5i25vd64ghamtgus6lue74k-acl", // TID-acl
  "default": "no-access",
  "peers": {
    "12D..dwaA6Qe": ["write", "delete"],
  }
}
\end{lstlisting}

This means that only `12D..dwaA6Qe` is able to alter the access-control list.

\section{Conclusion}

In this paper, we described the challenges and considerations when attempting to create a protocol suitable for large-scale data storage, synchronization, and use in a distributed system. We identified six requirements for enabling user-siloed data: flexible data formats, efficient synchronization, conflict resolution, access-control, scalable storage, and network communication. We presented a novel solution to satisfy each of these six requirements. We have introduced a solution that extends on IPFS and prior research done by Textile and others, which we term Threads. Threads are a novel data architecture that builds upon a collection of protocols to deliver a scalable and robust storage system for end-user data.

We show that the flexible core structure of single-writer append-only logs can be used to compose higher-order structures such as Threads, Views, and/or CRDTs. In particular, we show that through the design of specific default Views, we can support important features such as access control lists. The Threads protocol described here is flexible enough to derive numerous specific database types (e.g. key/value stores, document stores, relational stores, etc) and model an unlimited number of applications states. The cryptography used throughout Threads will help shift the data ownership model from apps to users.

TABLE

\subsection{Future Work}

The research and development of Textile Threads has highlighted several additional areas of work that would lead to increased benefits for users and developers. In particular, we have highlighted network services and security enhancements as core future work. In the following two sections, we briefly outline planned future work in these critical areas.

\subsubsection{Enhanced Log Security}

Threads change the relationship between a user, their data, and the services they connect with that data. The nested, or multi-layered, encryption combined with powerful ACL capabilities create new opportunities to build distributed services, or Bots, on a network of Textile peers. Based on the Follow key now available in Threads, Bots would be able to relay, replicate, or store data that is synchronized via real-time updates in a trust-less way [ref]. Bots can additionally enhance the IPFS network by providing a framework to build and deploy many kinds of trust-less or trusted services. Examples include simple archival, caching and republishing Bots. Other examples include payment, re-encryption, or bridges to Web 2.0 services to offer decentralized access to Web 2.0.

\subsubsection{Textile: The Thread \& Bot Network}

The use of a single Read and Follow key for an entire Log means that, should either of these keys be leaked via malicious (or other/accidental) means, there is no way to prevent a Peer with the leaked keys from listening to Events or traversing the Log history. Potential solutions currently being explored by Textile developers include key rotation at specific Event offsets [ref], and/or incorporating the Double Ratchet Algorithm [ref] for forward secrecy [ref].

\section{Latex Examples}

\begin{figure}
  \includegraphics[width=\linewidth]{boat.jpg}
  \caption{A boat.}
  \label{fig:boat1}
\end{figure}

\begin{quotation}
A quote
\end{quotation}

Here is a citation models~\cite{gelenbe06}. The approach constructs

here is a ref provides performance measures of interest to the SB and to the
seller, is discussed in Section \ref{Performance} where we first
discuss how the SB can behave in order to optimise outcomes that
are in its best interest, and provide  numerical examples to
illustrate the approach and the model predictions. We then explore
how the SB can try to achieve balance and compete with the other
bidders in Section \ref{KeepUp}. Finally Section
\ref{Price-Dependent} generalises the analysis to the case where
the bidding rates depend on the current price attained in the
auction. Conclusions are drawn in Section \ref{Conclusions} where
we also suggest further work.

\section{Conclusions} \label{Conclusions}
In this paper we have considered auctions in which bidders make
offers that are sequentially increasing in value by a unit price
in order to minimally surpass the previous highest bid, and
modelled them as discrete state-space random processes in
continuous time. Analytical solutions are obtained and measures
that are of interest to the SB are derived.

The measures that can be computed in this way include the SB's
probability of winning the auction, its expected savings with
respect to the maximum sum it is willing to pay, and the average
time that the SB spends before it can make a purchase. An
extension of the model that incorporates price-dependent
behaviours of the agents has also been presented.

The model allows us to quantitatively characterise intuitive and
useful trade-offs between improving the SB's chances of buying a
good quickly, and the price that it has to pay, in the presence of
different levels of competition from the other bidders.

There are interesting extensions and applications of these models
that can be considered, such as the behaviour of bidders and
sellers that may have time constraints for making a purchase, and
the possibility of the SB's moving among different auctions so as
to optimise measures which represent its self-interest. Another
interesting area of study may be to examine bidders who are
``rich'' and are willing to drive away rivals at any cost, and who
may create different auction environments for bidders that have
significantly different levels of wealth. Yet another area of
interest concerns auctions where items are sold in batches of
varying sizes, with prices which depend on the number of items
that are being bought.

\ack{This research was undertaken as part of the ALADDIN
(Autonomous Learning Agents for Decentralised Data and Information
Networks) project and is jointly funded by a BAE Systems and EPSRC
(Engineering and Physical Research Council) strategic partnership
(EP/C548051/1).}

\nocite{*}

\bibliographystyle{compj}
% \bibliography{ModellingBidders}
\input{sample.bbl}

\end{document}

### 2019-09-19 draft.tex
links a bit more; fixes a few things
Signed-off-by: andrew <andrew@textile.io>

@@
 In distributed systems, logs are often treated as append-only, where changes or updates can only be added to the set and never removed.
-\end{definition}
+\end{definition} \label{def:Logs}
 
@@
 Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates [ref].
-The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see section  \ref{sec:viewsprojections}).
+The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see section  \ref{sec:ViewsProjections}).
 In other words, with minimal conformity, a single log can model multiple application states [ref].
 
-\subsubsection{Views \& Projections} \label{sec:viewsprojections}
+\subsubsection{Views \& Projections} \label{sec:ViewsProjections}
 
@@
 
-\subsubsection{Eventual Consistency}
+\subsubsection{Eventual Consistency} \label{sec:EventualConsistency}
 
@@
 CRDTs rely on data structures that are mathematically guaranteed to resolve concurrent updates the same way regardless of the order those events were received.
-\end{definition} \label{def:crdt}
+\end{definition} \label{def:CRDT}
 
 How a system provides eventual consistency is often decided based on the intended use of the system.
-Two well-documented categories of solutions include logs (see Definition X.x.x), sequences of deterministic changes, and CRDTs (see Definition XXX).
+Two well-documented categories of solutions include logs (see Definition  \ref{def:Logs}), sequences of deterministic changes, and CRDTs (see Definition section  \ref{def:CRDT}).
 Logs work best when there is only a single-writer or complete event causality is known.
@@
 
-CRDTs (see \ref{def:crdt}) are one way to achieve strong eventual consistency, where once all replicas have received the same events, they will arrive at the same final state, regardless of ordering.
+CRDTs (see Definition \ref{def:CRDT}) are one way to achieve strong eventual consistency, where once all replicas have received the same events, they will arrive at the same final state, regardless of ordering.
 A review of the types of possible CRDTs is beyond the scope of this paper, however, it is important to note their role in eventually consistent systems and how they relate to clock-based event ordering.
@@
 For the third replica, resolving these two logs of events may be costly (many updates happened since the last common node) or impossible (parts of the chain may not be available on the network).
-See section 2.2.3.
-Eventual Consistency.
+See Section  \ref{sec:EventualConsistency}.
 
@@
 The ACL of an event-driven system is usually a list of access rules built from a series of events.
-For example, the two events, "grant Bob write access" and "revoke read access from Alice" would together result in a final ACL state where, Bob has read and write access, but Alice does not (see Views \& Projections in Data Synchronization).
+For example, the two events, "grant Bob write access" and "revoke read access from Alice" would together result in a final ACL state where, Bob has read and write access, but Alice does not (see Section  \ref{sec:ViewsProjections}).
 
@@
 \begin{definition}
-(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log’s Read key (see section 3.1.3 Keys \& Encryption).
+(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log’s Read key (see Section  \ref{sec:KeysEncryption}).
 \end{definition}
@@
 
-Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure X.X.) enabling log verification by readers (see section 3.1.3. Keys \& Encryption).
+Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure X.X.) enabling log verification by readers (see Section  \ref{sec:KeysEncryption}).
 
@@
 Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table [ref] (DHT).
-Instead, they are collected from Log Events (see section 3.3 Use Cases).
+Instead, they are collected from Log Events (see Section  \ref{sec:UseCases}).
 This is in contrast to mutable data via IPNS for example, which requires querying the network (DHT) for updates.
@@
 Log Writers can designate other IPFS Peers to “follow” a Log, potentially replicating and/or republishing Events.
-A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (see section 3.1.3 Keys \& Encryption), but is not able to read the Log’s contents.
+A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (see Section  \ref{sec:KeysEncryption}), but is not able to read the Log’s contents.
 Followers should be server-based, i.e., always online and behind a public IP address.\end{definition}
@@
 This ensures a conflict-free address list without additional complexity.
-Some Followers may be in the business of replicating Logs (see section 5.1.
-Textile: The Thread \& Bot Network), in which case Writers will announce the additional Log address to Readers.
+Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
 This allows them to pull (or subscribe to push-based) Events from the Follower’s Log address when the Writer is offline or unreachable.
@@
 
-\subsubsection{Keys \& Encryption}
+\subsubsection{Keys \& Encryption} \label{sec:KeysEncryption}
 
 Textile Logs are designed to be shared, composed, and layered into datasets.
-Therefore, Logs are encrypted by default in a manner that enables access control (see section 3.4.
-Access Control) and the Follower mechanism discussed in the previous section.
+Therefore, Logs are encrypted by default in a manner that enables access control (see Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
 
@@
 (Access-Controlled): This variant declares that consumers should assume an access control list is composable from Log Events.
-The ACL represents a permissions rule set that must be applied when reading data (see section 3.4.2.
-Access Control).
+The ACL represents a permissions rule set that must be applied when reading data (see Section  \ref{sec:AccessControl}).
 \end{definition}
@@
 
-\subsubsection{Log Synchronization}
+\subsubsection{Log Synchronization} \label{sec:LogSync}
 
@@
 An Event containing a mapping of Log IDs to Key Sets, which can be used to join a Thread.
-Threads backed by an access control list (see section 3.4.2.
-Access Control) will also include the current ACL for the Thread in an Invite.
+Threads backed by an access control list (see Section  \ref{sec:AccessControl}) will also include the current ACL for the Thread in an Invite.
 This enables Peers to invite others to only read or follow a Thread, instead of becoming a full-on collaborator, i.e., a new Log Writer.
@@
 Once a Peer has accepted an Invite, it will receive new Events from Log Writers.
-In cases where the invitee becomes a collaborator, i.e., a Writer, it is also responsible for sending its own Events out to the network.
-
-Sending
+In cases where the invitee becomes a collaborator, i.e., a Writer, it is also responsible for sending its own Events out to the network.\\
+
+\textbf{Sending} \\
 
@@
 \item \label{Perf2}New Events may be pushed to the Thread’s other Log Writers via federated notification services like Apple Push Notification Service (APNS) [ref], Google Cloud Messaging (GCM) [ref], Firebase Cloud Messaging (FCM) [ref], and/or Windows Notification Service (WNS) [ref].
-\item \label{Perf3} New Events may trigger web-hooks [ref] enabling complex workflows (see Section X below).
+\item \label{Perf3} New Events may trigger web-hooks [ref] enabling complex workflows (see Section X below).\\
 \end{enumerate}
 
-Receiving
+\textbf{Receiving} \\
 
@@
 
-The notion of the Follow key (see section 3.1.3 Keys \& Encryption) makes duplicating all Log Events trivial.
+The notion of the Follow key (see Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial.
 This allows any Peer in the network to be granted the responsibility of replicating data from another Peer without having read access to the data contained within the Log entries.
 This type of Log replication can act as a data backup mechanism.
-It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (see section 3.2.3 Log Synchronization).
+It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (see section  \ref{sec:LogSync}).
 
@@
 
-\subsubsection{Access Control}
+\subsubsection{Access Control} \label{sec:AccessControl}
 
@@
 % \bibliography{ModellingBidders}
-\input{sample.bbl}
+\input{draft.bbl}
 

### 2019-09-19 draft.tex
couple more linkages
Signed-off-by: andrew <andrew@textile.io>
remove placeholder
Signed-off-by: andrew <andrew@textile.io>
more fiddling
Signed-off-by: andrew <andrew@textile.io>

@@
 
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
-  \caption{This is really a code embed.}
-  \label{fig:boat1}
-\end{figure}
-
 \begin{lstlisting}
@@
 }
-
+\end{lstlisting}
+
+\begin{lstlisting}
 // Event represents the content of an update.
@@
 }
-
+\end{lstlisting}
+\begin{lstlisting}
 // EventHeader contains Event metadata.
@@
 Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses [ref].
-Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (see inset).
-In practice, to reach a Log via it’s IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (see inset).
+Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses:\\
+
+\tiny
+\textbf\textit /ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab\\
+\normalsize
+
+In practice, to reach a Log via it’s IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress:\\
+
+\tiny
+\textbf\textit /ip4/127.0.0.1/tcp/1234/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab\\
+\normalsize
 
@@
 
-Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log (see inset).
+Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log:
+
+\begin{lstlisting}
+[
+  "/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab",
+  "/p2p/12D..dJT6nXY/ipel/12D..bCcW8ab" // Follower
+]
+\end{lstlisting}
 
@@
 This allows them to pull (or subscribe to push-based) Events from the Follower’s Log address when the Writer is offline or unreachable.
-
-\begin{strip}
-\begin{tcolorbox}[title=Log Multiaddress]
-"/ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab"
-\end{tcolorbox}
-\end{strip}
-
-\begin{strip}
-\begin{tcolorbox}[title=Log IPEL Multiaddress]
-"/ip4/127.0.0.1/tcp/1234/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab"
-\end{tcolorbox}
-\end{strip}
-
-\begin{strip}
-\begin{tcolorbox}[title=Log Multiple Multiaddress]
-[\\
-  "/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab",\\
-  "/p2p/12D..dJT6nXY/ipel/12D..bCcW8ab" // Follower\\
-]
-\end{tcolorbox}
-\end{strip}
 

### 2019-09-19 draft.tex
try to stop code from wrapping screens
figured out full width images
Signed-off-by: andrew <andrew@textile.io>
full width code example
Signed-off-by: andrew <andrew@textile.io>
figure embed, code blocks etc
Signed-off-by: andrew <andrew@textile.io>

@@
 %\copyrightyear{2009} \vol{00} \issue{0} \DOI{000}
+
+% for full width figures\
+\usepackage{lipsum}% http://ctan.org/pkg/lipsum
+\usepackage{graphicx}% http://ctan.org/pkg/graphicx
 
@@
 
+% color the json
+\lstdefinelanguage{json}{
+  basicstyle={\small\ttfamily},
+  numbers=none,
+    numberstyle=\scriptsize,
+    stepnumber=1,
+    numbersep=8pt,
+    showstringspaces=false,
+    breaklines=true,
+    frame=lines,
+    backgroundcolor=\color{white},
+    literate=
+     *{0}{{{\color{black}0}}}{1}
+      {1}{{{\color{black}1}}}{1}
+      {2}{{{\color{black}2}}}{1}
+      {3}{{{\color{black}3}}}{1}
+      {4}{{{\color{black}4}}}{1}
+      {5}{{{\color{black}5}}}{1}
+      {6}{{{\color{black}6}}}{1}
+      {7}{{{\color{black}7}}}{1}
+      {8}{{{\color{black}8}}}{1}
+      {9}{{{\color{black}9}}}{1}
+      {:}{{{\color{gray}{:}}}}{1}
+      {\{}{{{\color{gray}{\{}}}}{1}
+      {\}}{{{\color{gray}{\}}}}}{1}
+      {[}{{{\color{gray}{[}}}}{1}
+      {"}{{{\color{gray}{"}}}}{1}
+      {,}{{{\color{gray}{,}}}}{1}
+      {]}{{{\color{gray}{]}}}}{1},
+}
+
 \begin{document}
@@
 
+TODO: Ref to appendix
+
+As seen above, an Event’s actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
+
+\subsubsection{Multi-addressed Event Logs}
+
+\begin{figure*}
+\centering
+\begin{minipage}{0.6\textwidth}
 \begin{lstlisting}
-// Node is the most basic component of a log.
-// Note: In practice, this is encrypted with the Follow key.
-type Node interface {
-    ipld.Node
-
-    // Event is the Log update.
-    Event() Event
-
-    // Refs are node linkages.
-    Refs() []cid.Cid
-
-    // Sig is a cryptographic signature of Event and Refs
-    // created with the Log's private key.
-    Sig() []byte
-}
+// Log multiaddress
+ipel := "/ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab"
+
+// Encapsulated multiaddress
+address := "/ip4/127.0.0.1/tcp/1234/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab"
+
+// Address book
+[
+  "/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab",
+  "/p2p/12D..dJT6nXY/ipel/12D..bCcW8ab" // Follower
+]
 \end{lstlisting}
-
-\begin{lstlisting}
-// Event represents the content of an update.
-// Note: In practice, this is encrypted with the Read key.
-type Event interface {
-    ipld.Node
-
-    // Header provides a means to store a timestamp
-    // and a key needed for decryption.
-    Header() EventHeader
-
-    // Body contains the content of an update.
-    // In practice, this is encrypted with the Header key
-    // or the recipient's public key.
-    Body() ipld.Node
-
-    // Decrypt is a helper function that decrypts Body
-    // with a key in Header.
-    Decrypt() (ipld.Node, error)
-}
-\end{lstlisting}
-\begin{lstlisting}
-// EventHeader contains Event metadata.
-type EventHeader interface {
-    ipld.Node
-
-    // Time is the wall-clock time at which the Event
-    // was created.
-    Time() int
-
-    // Key is an optional single-use symmetric key
-    // used to encrypt Body.
-    Key() []byte
-}
-\end{lstlisting}
-
-As seen above, an Event’s actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
-
-\subsubsection{Multi-addressed Event Logs}
+\end{minipage}
+  \caption{The Log Multiaddress.}
+  \label{fig:boat1}
+\end{figure*}
 
@@
 
-\tiny
-\textbf\textit /ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab\\
-\normalsize
-
-In practice, to reach a Log via it’s IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress:\\
-
-\tiny
-\textbf\textit /ip4/127.0.0.1/tcp/1234/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab\\
-\normalsize
+TODO: ref to embed
+
+In practice, to reach a Log via it’s IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress:
+
+TODO: ref to embed
 
@@
 
+\begin{minipage}{.45\textwidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\end{minipage}
 
@@
 
-Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log:
-
-\begin{lstlisting}
-[
-  "/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab",
-  "/p2p/12D..dJT6nXY/ipel/12D..bCcW8ab" // Follower
-]
-\end{lstlisting}
+Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log
+
+TODO: ref to embed
 
@@
 
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
+\begin{figure*}
+\centering
+\begin{minipage}{0.6\textwidth}
+  \includegraphics[width=\linewidth]{figures/Event_Log_With_Encryption.png}
   \caption{The three layers of Log Event encryption.}
-  \label{fig:boat1}
-\end{figure}
+  \label{fig:LogEncryption}
+  \end{minipage}
+\end{figure*}
 
@@
 
-\begin{lstlisting}
+\begin{minipage}{.45\textwidth}
+\begin{lstlisting}[language=json,firstnumber=1]
 {
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=json,firstnumber=1]
 {
@@
 \end{quotation}
+
+\begin{figure*}
+  \includegraphics[width=\textwidth,height=4cm]{boat.jpg}
+  \caption{This is a full width boat.}
+  \label{fig:boat1}
+\end{figure*}
 
@@
 we also suggest further work.
+
+\begin{figure*}
+
+\begin{lstlisting}
+// Node is the most basic component of a log.
+// Note: In practice, this is encrypted with the Follow key.
+type Node interface {
+    ipld.Node
+
+    // Event is the Log update.
+    Event() Event
+
+    // Refs are node linkages.
+    Refs() []cid.Cid
+
+    // Sig is a cryptographic signature of Event and Refs
+    // created with the Log's private key.
+    Sig() []byte
+}
+
+// Event represents the content of an update.
+// Note: In practice, this is encrypted with the Read key.
+type Event interface {
+    ipld.Node
+
+    // Header provides a means to store a timestamp
+    // and a key needed for decryption.
+    Header() EventHeader
+
+    // Body contains the content of an update.
+    // In practice, this is encrypted with the Header key
+    // or the recipient's public key.
+    Body() ipld.Node
+
+    // Decrypt is a helper function that decrypts Body
+    // with a key in Header.
+    Decrypt() (ipld.Node, error)
+}
+
+// EventHeader contains Event metadata.
+type EventHeader interface {
+    ipld.Node
+
+    // Time is the wall-clock time at which the Event
+    // was created.
+    Time() int
+
+    // Key is an optional single-use symmetric key
+    // used to encrypt Body.
+    Key() []byte
+}
+\end{lstlisting}
+  \caption{This is a full width code block.}
+
+\end{figure*}
 
@@
 % \bibliography{ModellingBidders}
-\input{draft.bbl}
+\input{whitepaper.bbl}
 

### 2019-09-19 draft.tex
lots of figure layout changes; added "examples" type
Signed-off-by: andrew <andrew@textile.io>

@@
 
+\usepackage{trivfloat}
+\trivfloat{example}
+
 \begin{document}
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
+  \includegraphics[width=\linewidth]{figures/Hash_Request.png}
   \caption{The crypographic hash of content is used to make a request to the network of IPFS peers.
@@
 
-\subsubsection{Multi-addressed Event Logs}
-
-\begin{figure*}
+\begin{example*}
 \centering
-\begin{minipage}{0.6\textwidth}
+\begin{minipage}{0.7\textwidth}
 \begin{lstlisting}
@@
   \caption{The Log Multiaddress.}
-  \label{fig:boat1}
-\end{figure*}
+  \label{ex:Multiaddress}
+\end{example*}
+
+\subsubsection{Multi-addressed Event Logs}
 
 Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses [ref].
-Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses:\\
-
-TODO: ref to embed
-
-In practice, to reach a Log via it’s IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress:
-
-TODO: ref to embed
-
-Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table [ref] (DHT).
-Instead, they are collected from Log Events (see Section  \ref{sec:UseCases}).
-This is in contrast to mutable data via IPNS for example, which requires querying the network (DHT) for updates.
-Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed like topology.
-Log addresses are recorded in an address book, similar to IPFS Peer address book:
-
-\begin{minipage}{.45\textwidth}
+Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (see Example \ref{ex:Multiaddress}).
+
+In practice, to reach a Log via it’s IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (see Example \ref{ex:Multiaddress}).
+
+\begin{example}
 \begin{lstlisting}
-// AddrBook stores log addresses
 type AddrBook interface {
@@
 \end{lstlisting}
-\end{minipage}
+\caption{AddrBook storing log addresses.}
+\end{example} \label{ex:AddrBook}
+
+Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table [ref] (DHT).
+Instead, they are collected from Log Events (see Section  \ref{sec:UseCases}).
+This is in contrast to mutable data via IPNS for example, which requires querying the network (DHT) for updates.
+Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed like topology.
+Log addresses are recorded in an address book, similar to IPFS Peer address book:
 
@@
 \subsubsection{Keys \& Encryption} \label{sec:KeysEncryption}
-
-Textile Logs are designed to be shared, composed, and layered into datasets.
-Therefore, Logs are encrypted by default in a manner that enables access control (see Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
 
@@
 
-\begin{definition}
-(Identity Key).
-Every Log requires an asymmetric key-pair that determines ownership and identity.
-The private key is used to sign each Event added to the Log, so down-stream processes can verify the Log’s authenticity.
-Like IPFS peers, the public key of the Log is used as an identifier (Log ID).
-\end{definition}
-
-The body, or content of an event, is encrypted by a Content Key.
-Content keys are generated per-content and never reused.
-The Content key is distributed directly in the header of the Event Block.
-We define the Content Key as follows,
-
-\begin{definition}
-(Content Key).
-The Content key is a variable-format key used to encrypt the body (content) of an event.
-This key can be symmetric, asymmetric, or possibly non-existent in cases where encryption is not needed.
-One of two common encryption choices will typically be used per event,
-\begin{enumerate}
-\item \label{Perf1}When broadcasting events to many possible recipients, a single-use symmetric key is generated per unique content body.
-\item \label{Perf2}When sending events to specific recipients, the recipient's public key can be used to restrict access from all others.
-\end{enumerate}
-\end{definition}
-
-If a single-use symmetric key is used for the Content key, it is necessary to distribute each new key to users by including it in the header of the Event Block.
-Therefore, the Event Block itself is further encrypted using a Read key.
-The Read key is not distributed within the Log itself but is distributed to any peers to grant them access to the content of the Log.
-
-\begin{definition}
- (Read Key). The Read key is a symmetric key created by the Log owner and used to encrypt the Content Key in each event.
-\end{definition}
-
-Finally, the encrypted Event Block, its signature, and the IPLD linkage(s) from an Event to its antecedents are encrypted together using a Follow key.
-Follow keys allow Logs to be followed by peers on the network who do not have access to any content within the event, and can only see signatures and linkage(s) between Events.
-
-\begin{definition}
-(Follow Key). The Follow key is a symmetric key created by the Log owner and used to encrypt the entire event payload before adding the event to the Log.
-\end{definition}
-
-Much like the Log address book, Log keys are stored in a key book:
-
-\begin{minipage}{.45\textwidth}
+\begin{example}
 \begin{lstlisting}
@@
 \end{lstlisting}
-\end{minipage}
+\caption{A KeyBook storing log keys.}
+\end{example} \label{ex:KeyBook}
+
+Textile Logs are designed to be shared, composed, and layered into datasets.
+Therefore, Logs are encrypted by default in a manner that enables access control (see Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
+
+\begin{definition}
+(Identity Key).
+Every Log requires an asymmetric key-pair that determines ownership and identity.
+The private key is used to sign each Event added to the Log, so down-stream processes can verify the Log’s authenticity.
+Like IPFS peers, the public key of the Log is used as an identifier (Log ID).
+\end{definition}
+
+The body, or content of an event, is encrypted by a Content Key.
+Content keys are generated per-content and never reused.
+The Content key is distributed directly in the header of the Event Block.
+We define the Content Key as follows,
+
+\begin{definition}
+(Content Key).
+The Content key is a variable-format key used to encrypt the body (content) of an event.
+This key can be symmetric, asymmetric, or possibly non-existent in cases where encryption is not needed.
+One of two common encryption choices will typically be used per event,
+\begin{enumerate}
+\item \label{Perf1}When broadcasting events to many possible recipients, a single-use symmetric key is generated per unique content body.
+\item \label{Perf2}When sending events to specific recipients, the recipient's public key can be used to restrict access from all others.
+\end{enumerate}
+\end{definition}
+
+If a single-use symmetric key is used for the Content key, it is necessary to distribute each new key to users by including it in the header of the Event Block.
+Therefore, the Event Block itself is further encrypted using a Read key.
+The Read key is not distributed within the Log itself but is distributed to any peers to grant them access to the content of the Log.
+
+\begin{definition}
+ (Read Key). The Read key is a symmetric key created by the Log owner and used to encrypt the Content Key in each event.
+\end{definition}
+
+Finally, the encrypted Event Block, its signature, and the IPLD linkage(s) from an Event to its antecedents are encrypted together using a Follow key.
+Follow keys allow Logs to be followed by peers on the network who do not have access to any content within the event, and can only see signatures and linkage(s) between Events.
+
+\begin{definition}
+(Follow Key). The Follow key is a symmetric key created by the Log owner and used to encrypt the entire event payload before adding the event to the Log.
+\end{definition}
+
+Much like the Log address book, Log keys are stored in a key book (see Example \ref{ex:KeyBook}).
 
@@
 
+\begin{example}
 \begin{minipage}{.45\textwidth}
@@
 \end{minipage}
+\caption{The Key-Value store interface.}
+\end{example} \label{ex:KVStore}
 
@@
 \end{minipage}
+\caption{The Document store interface.}
+\end{example} \label{ex:AddrBook}
 
@@
 \end{lstlisting}
+\caption{blahblah}
 \end{minipage}
+\end{example}
 
@@
 
+\begin{example}
 \begin{lstlisting}[language=json,firstnumber=1]
@@
 \end{lstlisting}
+\caption{Blahblah}
+\end{example}
 
@@
 
-\begin{figure*}
+\begin{example*}
 
@@
 
-\end{figure*}
+\end{example*}
 

### 2019-09-19 draft.tex
remove inline
Signed-off-by: andrew <andrew@textile.io>

@@
 \begin{lstlisting}
-// KeyBook stores log keys
 type KeyBook interface {

### 2019-09-19 draft.tex
rm var names
Signed-off-by: andrew <andrew@textile.io>

@@
 // Log multiaddress
-ipel := "/ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab"
+ "/ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab"
 
 // Encapsulated multiaddress
-address := "/ip4/127.0.0.1/tcp/1234/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab"
+ "/ip4/127.0.0.1/tcp/1234/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab"
 

### 2019-09-19 draft.tex
updates
Signed-off-by: andrew <andrew@textile.io>

@@
 An emerging issue of concern is that the digital part of a person's life is mainly being captured by apps and stored away from the person’s control by the companies building those apps.
-An alternative, called user-siloed data is an exciting area of research that if implemented successfully, could reverse the flow of value derived from personal data, from apps to users.
+An alternative, called user-siloed data is one area of research that if implemented successfully, could reverse the flow of value derived from personal data, from apps to users.
 In this paper, we investigate the data formats, and storage and transfer protocols necessary to build a system for user ownership of large-scale digital datasets.

### 2019-09-19 draft.tex
more
Signed-off-by: andrew <andrew@textile.io>
fix block
Signed-off-by: andrew <andrew@textile.io>

@@
 \begin{minipage}{0.7\textwidth}
-\begin{lstlisting}
-// Log multiaddress
- "/ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab"
-
-// Encapsulated multiaddress
- "/ip4/127.0.0.1/tcp/1234/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab"
-
-// Address book
+\begin{lstlisting}[language=json,firstnumber=1]
+# Log multiaddress
+ /ipel/12D3KooWC2zyCVron7AA34N6oKNtaXaZB51feG9rBkr7QbCcW8ab
+
+# Encapsulated multiaddress
+ /ip4/127.0.0.1/tcp/1234/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab
+
+# Address book
 [
-  "/p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab",
-  "/p2p/12D..dJT6nXY/ipel/12D..bCcW8ab" // Follower
+  /p2p/12D..dwaA6Qe/ipel/12D..bCcW8ab,
+  /p2p/12D..dJT6nXY/ipel/12D..bCcW8ab # Follower
 ]
@@
 
-Example (V1, 256 bit): `bafyoiobghzefwlidfrwkqmzz2ka66zgmdmgeobw2mimktr5jivsavya`
+\begin{example}
+\begin{lstlisting}
+"bafyoiobghzefwlidfrwkqmzz2ka66zgmdmgeobw2mimktr5jivsavya"
+\end{lstlisting}
+\caption{V1, 256 bit.}
+\end{example} \label{ex:v1256}
 

### 2019-09-19 draft.tex
updated gitignore to avoid some conflicts; cleaned up some within docs refs
Signed-off-by: andrew <andrew@textile.io>

@@
 Threads provide the mechanism to combine multiple Logs from individual authors into singular shared states through the use of either cross-Log sequencing (e.g.
-using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock [ref]) or a CRDT (see section xx).
+using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock [ref]) or a CRDT.
 
@@
 \begin{definition}
-(Random Number). A random number of a user-specified length. 16 bytes or more.
+(Random Number). A random number of a user-specified length. 16 bytes or more  (see Example \ref{ex:Identity}).
 \end{definition}
@@
 \begin{definition}
-(Raw). This variant declares that consumers are not expected to make additional assumptions. This is the default variant. 
-\end{definition}
-
-Example (V1, 128 bit): `bafkxd5bjgi6k4zivuoyxo4ua4mzyy`
+(Raw). This variant declares that consumers are not expected to make additional assumptions. This is the default variant.\\
+
+See Example \ref{ex:Identity}).
+\end{definition}
 
@@
 (Access-Controlled): This variant declares that consumers should assume an access control list is composable from Log Events.
-The ACL represents a permissions rule set that must be applied when reading data (see Section  \ref{sec:AccessControl}).
-\end{definition}
-
-\begin{example}
-\begin{lstlisting}
-"bafyoiobghzefwlidfrwkqmzz2ka66zgmdmgeobw2mimktr5jivsavya"
+The ACL represents a permissions rule set that must be applied when reading data (see Section  \ref{sec:AccessControl}).\\
+
+See Example \ref{ex:Identity}).
+\end{definition}
+
+\begin{example*}
+\centering
+\begin{minipage}{0.6\textwidth}
+\begin{lstlisting}[language=json,firstnumber=1]
+# Raw identity. V1, 128 bit
+bafkxd5bjgi6k4zivuoyxo4ua4mzyy
+
+# ACL enabled identity. V1, 256 bit.
+bafyoiobghzefwlidfrwkqmzz2ka66zgmdmgeobw2mimktr5jivsavya
 \end{lstlisting}
-\caption{V1, 256 bit.}
-\end{example} \label{ex:v1256}
+\end{minipage}
+  \caption{Identity variants.}
+  \label{ex:Identity}
+\end{example*}
 
@@
 For example, an ACL JSON document is a single entity made up of a sequence of Thread Events.
-A Sequence might have a unique UUID (.e.g, id =bafykrq5i25vd64ghamtgus6lue74k) as in the example below.
-\end{definition}
-
-fixidexampleabove
+A Sequence might have a unique UUID (see Example \ref{ex:SequenceID}) as in the example below.
+\end{definition}
+
+\begin{example}
+\centering
+\begin{lstlisting}[language=json,firstnumber=1]
+# UUID
+bafykrq5i25vd64ghamtgus6lue74k
+\end{lstlisting}
+  \caption{Sequence ID.}
+  \label{ex:SequenceID}
+\end{example}
 

### 2019-09-19 draft.tex
adds another secref
Signed-off-by: andrew <andrew@textile.io>

@@
 
-Like many CQRS-based systems, the “public” API for Threads revolves around the concept of Views (see background section xxx), which are used to optimize queries on the local event store.
+Like many CQRS-based systems, the “public” API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store.
 A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store.
@@
 In practice, Snapshots are written to their own Event log and stored locally.
-They can potentially be synced (see section 3.2.4) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time).
+They can potentially be synced (see Section \ref{sec:LogSync}) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time).
 They can similarly be used for initializing state during recovery.

### 2019-09-20 draft.tex

@@
+% !TEX encoding = UTF-8 Unicode
 \documentclass{comjnl}
@@
 As the Internet expands, the division between a person's digital and physical existence continues to blur.
-An emerging issue of concern is that the digital part of a person's life is mainly being captured by apps and stored away from the person’s control by the companies building those apps.
+An emerging issue of concern is that the digital part of a person's life is mainly being captured by apps and stored away from the person's control by the companies building those apps.
 An alternative, called user-siloed data is one area of research that if implemented successfully, could reverse the flow of value derived from personal data, from apps to users.
@@
 One such consequence is that many of the same development patterns that bring convenience (e.g., single sign-on, minimal encryption, centralized servers, and databases) also enable or even require data hoarding by the apps.
-While collecting large amounts of user’s data can create value for the companies building apps (e.g., app telemetry, predictive tools, or even new revenue streams), that value flows mostly in one direction; Apps collect, process and benefit from a user's private data, but users rarely have access to the original data or the new insights that come from it.
+While collecting large amounts of users' data can create value for the companies building apps (e.g., app telemetry, predictive tools, or even new revenue streams), that value flows mostly in one direction; Apps collect, process and benefit from a user's private data, but users rarely have access to the original data or the new insights that come from it.
 Additionally,  the data may not be readily accessible to other apps and services.
@@
 
-While companies collecting user data may not be itself a significant problem, it has been shown that over time company incentives often shift from providing value to users to extracting value from users [ref Dixon].
+While companies collecting user data may not be itself a significant problem, it has been shown that over time company incentives often shift from providing value to users to extracting value from users \cite{dixonWhyDecentralizationMatters2018}.
 When this incentive shift happens for companies that have also been hoarding user data, that data may be a new source of value or even revenue for the company.
@@
 Ideally, there is an alternative approach that allows individuals to capture the value of their data and still allow developers to build new interfaces and experience on top of this data.
-This approach is called user-siloed data and it fundamentally separates apps from their user’s data.
+This approach is called user-siloed data and it fundamentally separates apps from their user's data.
 
 One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data.
-Other projects have identified app silo-ed data as a problem [ref], and some have identified user-siloed data as a solution [ref Solid].
+Other projects have identified app silo-ed data as a problem [ref], and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}.
 However, none so far have provided a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data.
@@
 To model realistic systems, this type of framework often requires complex techniques for mapping data between domain models and database tables, where the same data model is used to both query and update a database.
-An powerful alternative is to use a set of append only logs to model the state of an object simply by applying its change sequence in the correct order.
-This concept can be expressed succinctly by the state machine approach [ref]: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state.
-This is a powerful concept baked into a simple structure, and is at the heart of many distributed database systems [ref].
+A powerful alternative is to use a set of append only logs to model the state of an object simply by applying its change sequence in the correct order.
+This concept can be expressed succinctly by the state machine approach \cite{schneiderImplementingFaulttolerantServices1990}: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state.
+This is a powerful concept baked into a simple structure, and is at the heart of many distributed database systems \cite{LogWhatEvery}.
 
@@
 In modern applications it is critical to have reliable mechanisms for publishing updates and events (i.e., to support event-driven architectures), scalability (optimized write and read operations), forward-compatible application updates (e.g., code changes, retroactive events, and others), auditing systems, etc.
-To support such requirements, developers have begun to utilize event sourcing and command query responsibility segregation (CQRS) patterns [ref], relying on append only logs to support immutable state histories.
-Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate event sourcing and CQRS-based applications, including Event Store [ref], Apache Kafaka [ref] and Samza [ref], among others [see ref].
+To support such requirements, developers have begun to utilize event sourcing and command query responsibility segregation (CQRS) patterns \cite{bettsExploringCQRSEvent2013}, relying on append only logs to support immutable state histories.
+Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate event sourcing and CQRS-based applications, including Event Store \cite{EventStore}, Apache Kafaka \cite{ApacheKafka} and Samza \cite{Samza}, among others (see \cite{kleppmannDesigningDataintensiveApplications2017}).
 
@@
 A key principal of ES and append only logs is that all changes to application state are stored as a sequence of events.
-Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates [ref].
+Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates \cite{fowlerEventSourcing}.
 The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see section  \ref{sec:ViewsProjections}).
-In other words, with minimal conformity, a single log can model multiple application states [ref].
+In other words, with minimal conformity, a single log can model multiple application states \cite{microsoftcorporationAzureApplicationArchitecture}.
 
@@
 An event handler and corresponding reducer/fold function used to build and maintain a view from a set of (filtered) events.
-While projections may lead to the generation of new events, their reducer should be a pure function (see [ref] and [ref] for examples from existing systems).
+While projections may lead to the generation of new events, their reducer should be a pure function (see \cite{eventsouceEventSourcingBasics} and \cite{redux.js.orgGettingStartedRedux} for examples from existing systems).
 \end{definition}
@@
 
-Views themselves are enabled by projections.
+Views themselves are enabled by projections\footnote{Terminology in this section may differ from some other examples of ES and CQRS patterns, but reflects the underlying architecture and designs that the Textile team will elaborate on in \ref{sec:ThreadsProtocol}}.
 Projections can be thought of as transformations or reducers that are applied to each event in a stream of events.
@@
 
-The CAP theorem [ref, ref] states that a distributed database can guarantee only two of the following three promises at the same time: consistency (i.e., that every read receives the most recent write or an error), availability (i.e., that every request receives a (possibly out-of-date) non-error response, and partition tolerance (i.e., that the system continues to operate despite an arbitrary number of messages being dropped (or delayed) by the network).
+The CAP theorem \cite{brewerRobustRobustDistributed2000,gilbertBrewerConjectureFeasibility2002a} states that a distributed database can guarantee only two of the following three promises at the same time: consistency (i.e., that every read receives the most recent write or an error), availability (i.e., that every request receives a (possibly out-of-date) non-error response, and partition tolerance (i.e., that the system continues to operate despite an arbitrary number of messages being dropped (or delayed) by the network).
 As such, many distributed systems are now designed to provide availability and partition tolerance by trading consistency for eventual consistency.
 Eventual consistency allows state replicas to diverge temporarily, but eventually arrive back to the same state again.
-While an active area of research, designing systems with provable eventual consistency guarantees remains challenging [ref needed].
+While an active area of research, designing systems with provable eventual consistency guarantees remains challenging \cite{shapiroComprehensiveStudyConvergent2011a,almeidaDeltaStateReplicated2018}.
 
@@
 Logs work best when there is only a single-writer or complete event causality is known.
-In many distributed systems with multiple log writers, a minimum requirement for synchronization through logs is that the essential order of events is respected and can be determined [3, 8].
-For these cases, logical clocks are a useful tool for eventual consistency and total ordering [ref].
+In many distributed systems with multiple log writers, a minimum requirement for synchronization through logs is that the essential order of events is respected and can be determined \cite{schwarzDetectingCausalRelationships1994,katzInterleavingSetTemporal1990}.
+For these cases, logical clocks are a useful tool for eventual consistency and total ordering \cite{kulkarniLogicalPhysicalClocks2014}.
 However, some scenarios (e.g., temporarily missing events, or ambiguous order) can force a replica into a state that cannot be later resolved without costly recalculation.
@@
 
-\subsubsection{Logical Clocks}
-
-In a distributed system, where multiple machines, each with an independent clock, are creating events, local timestamps can’t be used to construct global event causality.
-Machine clocks are never perfectly synchronized [1], meaning that one machine's concept of "now" is not necessarily the same as another machine's.
+\subsubsection{Logical Clocks}\label{sec:LogicalClocks}
+
+In a distributed system, where multiple machines, each with an independent clock, are creating events, local timestamps can't be used to construct global event causality.
+Machine clocks are never perfectly synchronized \cite{lamportTimeClocksOrdering1978}, meaning that one machine's concept of "now" is not necessarily the same as another machine's.
 Machine speed, network speed, and other factors compound the issue.
@@
 Alternatives to wall-clock time exist to help distributed systems achieve eventual consistency.
-Examples include various logical clocks (Lamport [ref below], Schwartz [ref below], and Bloom [ref below], Hybrid variants [ref below], etc), which use counter-based time-stamping to provide partial ordering.
+Examples include various logical clocks (Lamport \cite{lamportTimeClocksOrdering1978} Schwartz \cite{schwarzDetectingCausalRelationships1994}, and Bloom \cite{ramabajaBloomClock2019}, Hybrid variants \cite{kulkarniLogicalPhysicalClocks2014}, etc), which use counter-based time-stamping to provide partial ordering.
 
 Cryptographically linked events can also represent a clock.
-One such example is called the Merkle-Clock [ref below].
-The Merkle-Clock relies on properties of a Merkle-DAG to provide strict partial ordering between events, which has limitations [ref section 4.3 of the merkle paper]:
+One such example is called the Merkle-Clock \cite{sanjuanMerkleCRDTs2019}.
+The Merkle-Clock relies on properties of a Merkle-DAG to provide strict partial ordering between events, which has limitations \cite[sec.
+4.3]{sanjuanMerkleCRDTs2019}:
 
@@
 
-CRDTs (see Definition \ref{def:CRDT}) are one way to achieve strong eventual consistency, where once all replicas have received the same events, they will arrive at the same final state, regardless of ordering.
+CRDTs (see Definition \ref{def:CRDT}) are one way to achieve strong eventual consistency, where once all replicas have received the same events, they will arrive at the same final state, \emph{regardless of ordering}\footnote{Though non-commutative CRDTs may require a specific ordering of events in certain cases \cite{sanjuanMerkleCRDTs2019}}.
 A review of the types of possible CRDTs is beyond the scope of this paper, however, it is important to note their role in eventually consistent systems and how they relate to clock-based event ordering.
-See for example [Merkle Clock ref] and [ref] for informative reviews of these types of data structures.
+See for example \cite{sanjuanMerkleCRDTs2019} and [MISSING] for informative reviews of these types of data structures.
 
@@
 An alternative to location addressing, called content addressing may provide a solution to those problems.
-Content addressing is where the content itself is used to create an address and retrieve that content from a network [2].
+Content addressing is where the content itself is used to create an address and retrieve that content from a network \cite{mortContentBasedAddressing2012}.
 
@@
 
-There are a number of systems that utilize content addressing to access content and information [list some as refs], and there is an active body of literature covering its design and implementation [ref].
-The Interplanetary File System (IPFS) — which is a set of protocols to create a content-addressed, peer-to-peer filesystem [1] — is one such system.
+There are a number of systems that utilize content addressing to access content and information [LIST SOME], and there is an active body of literature covering its design and implementation [MISSING].
+The Interplanetary File System (IPFS) --- which is a set of protocols to create a content-addressed, peer-to-peer filesystem \cite{benetIpfscontentAddressedVersioned2014} --- is one such system.
 In IPFS, the address for any content is determined based on a cryptographic hash of the content itself.
-In practice, the IPFS Content IDentifier (CID) is a multihash, which is a self-describing “protocol for differentiating outputs from various well-established cryptographic hash functions, addressing size + encoding considerations” [ref].
+In practice, the IPFS Content IDentifier (CID) is a multihash, which is a self-describing ``protocol for differentiating outputs from various well-established cryptographic hash functions, addressing size + encoding considerations\footnote{https://multiformats.io/}''.
 That addressing system confers several benefits to the network, including tamper resistance (we can be confident that a given piece of content has not been modified en route if its hash matches what we were expected/requested) and de-duplication (because the same content from different peers will produce the same hash address).
@@
 
-While content addressing doesn’t tell you how to get a file, IPFS (via libp2p [ref]) provides a system for moving files across the network.
+While content addressing doesn't tell you how to get a file, IPFS (via libp2p\footnote{https://libp2p.io/}) provides a system for moving files across the network.
 On the IPFS network, a client who wants specific content requests the CID from the network of IPFS hosts.
-The client's request is routed to the first host capable of fulfilling the request — i.e., the first host that is actively storing the content behind the given CID.
-The IPFS network can be seen as a distributed file system, with all of the benefits [ref] that come with this type of file system design.
+The client's request is routed to the first host capable of fulfilling the request (i.e., the first host that is actively storing the content behind the given CID).
+The IPFS network can be seen as a distributed file system, with all of the benefits that come with this type of file system design.
 
@@
 
-As discussed above, IPFS uses the cryptographic hash of a given piece of content to define its content address (see [ref] for details on this process).
+As discussed above, IPFS uses the cryptographic hash of a given piece of content to define its content address (see \cite{benetIpfscontentAddressedVersioned2014} for details on this process).
 However, in order to provide standards for accessing content-addressable data (on the web or elsewhere), it is necessary to define a common format or specification.
-In IPFS (and other projects [ref]), this common data format is called Interplanetary Linked Data (IPLD) [ref].
-As the name suggests, IPLD is based on principals of linked data [ref, ref] with the added capabilities of a content-addressing storage network.
-
-IPLD is used to publish linked data (subject-predicate-object triples in the linked data world [ref]) spread across different hosts, and where everything (entities, predicates, data sources) uses content addresses as unique identifiers [ref].
-To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph.
-This allows all hash-linked data structures to be treated using a unified data model, analogous to linked data in the Semantic Web sense [ref].
-In practice, IPLD is represented as objects, each with `Data` and `Links` fields, where `Data` can be a small blob of unstructured, arbitrary binary data, and `Links` is an array of links to other IPLD objects.
+In IPFS (and others \cite[e.g.,]{protocollabsFilecoinDecentralizedStorage2017} MORE NEEDED), this common data format is called Interplanetary Linked Data (IPLD)\footnote{https://ipld.io/}.
+As the name suggests, IPLD is based on principals of linked data \cite{timberners-leeLinkedData2009,bizerLinkedDataStory2011} with the added capabilities of a content-addressing storage network.
+
+IPLD is used to publish linked data (subject-predicate-object triples in the linked data world \cite{heathLinkedDataEvolving2011}) spread across different hosts, and where everything (entities, predicates, data sources) uses content addresses as unique identifiers.
+To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph\footnote{For reference, the Bitcoin Blockchain is a DAG, so is a Git version history.}.
+This allows all hash-linked data structures to be treated using a unified data model, analogous to linked data in the Semantic Web sense \cite{brendanobrienDeterministicQueryingDistributed2017}.
+In practice, IPLD is represented as objects, each with \texttt{Data} and \texttt{Links} fields, where \texttt{Data} can be a small blob of unstructured, arbitrary binary data, and \texttt{Links} is an array of links to other IPLD objects.
 
@@
 A Merkle-Clock is a Merkle-DAG that represents a sequence of events.
-In other words, a Merkle-Clock is an append-only log [ref and ipfs-log].
-When implemented on IPFS (or an equivalent network where content can be cryptographically addressed and fetched), Merkle-Clocks provide a number of benefits for data synchronization between replicas [ref merkle paper section 4.3]:
+In other words, a Merkle-Clock is an append-only log \cite{sanjuanMerkleCRDTs2019}.
+When implemented on IPFS (or an equivalent network where content can be cryptographically addressed and fetched), Merkle-Clocks provide a number of benefits for data synchronization between replicas \cite[sec.
+4.3]{sanjuanMerkleCRDTs2019}:
 
@@
 
-On the downside, and as seen above (section 2.2.4.), a Merkle-Clock cannot order divergent heads (or roots):
+On the downside, and as seen above (section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots):
 
@@
 In the figure above, two replicas (top row, bottom row) are attempting to write (left to right) events to the same Merkle-Clock.
-After the first replica writes event A, the second writes event A’ and properly links to A.
+After the first replica writes event A, the second writes event A' and properly links to A.
 At that point, the two replicas stop receiving events from one another.
-To a third replica that does continue to receive events, there would now be two independent heads, 1 and 1’.
+To a third replica that does continue to receive events, there would now be two independent heads, 1 and 1'.
 For the third replica, resolving these two logs of events may be costly (many updates happened since the last common node) or impossible (parts of the chain may not be available on the network).
@@
 On real networks with many often offline replicas (mobile and IoT devices, laptops, etc.), these conditions are rarely met.
-
 Based on these observations, using a single Merkle-Clock to synchronize replicas can be problematic.
@@
 IPFS contains a collection of protocols and systems to help address the networking needs required by different use-cases.
-That is to say no matter what type of device we are talking about — be it a phone, desktop computer, browser, or Internet-enabled appliance — it should be able to communicate with other devices.
+That is to say no matter what type of device we are talking about --- be it a phone, desktop computer, browser, or Internet-enabled appliance --- it should be able to communicate with other devices.
 However, each of the networking approaches described below comes with strengths and weaknesses when used to synchronize data.
@@
 The libp2p project provides a robust protocol communication stack.
-IPFS and a growing list of other projects [ref] are building on top of libp2p.
+IPFS and a growing list of other projects [MISSING] are building on top of libp2p.
 Libp2p solves a number of challenges that are distinct to peer-to-peer (p2p) networks.
-A comprehensive coverage of networking issues in P2P systems is out of the scope for this paper, however, some core challenges that libp2p helps to address include NAT traversal, peer discovery and handshake protocols, and even encryption and transport security — libp2p supports both unencrypted (e.g.
+A comprehensive coverage of networking issues in P2P systems is out of the scope for this paper, however, some core challenges that libp2p helps to address include NAT traversal, peer discovery and handshake protocols, and even encryption and transport security --- libp2p supports both unencrypted (e.g.
 TCP, UDP) and encrypted protocols (e.g.
-TLS, Noise) — among others.
+TLS, Noise) --- among others.
 Libp2p uses the concept of a multiaddress to address peers on a network, which essentially models network addresses as arbitrary encapsulations of protocols [ref, ref].
-In addition to “transport layer” modules, libp2p provides several tools for sharing and/or disseminating data over a p2p network.
+In addition to ``transport layer'' modules, libp2p provides several tools for sharing and/or disseminating data over a p2p network.
 
@@
 One of the most commonly used p2p distribution layers built on libp2p, is its Pubsub (or publish-subscribe) system.
-Pubsub is a standard messaging pattern where the publishers don’t know who, if anyone, will subscribe to a given topic.
+Pubsub is a standard messaging pattern where the publishers don't know who, if anyone, will subscribe to a given topic.
 Publishers send messages on a given topic or category, and Subscribers receive only messages on a give topic they are subscribed to.
-Libp2p’s pubsub module can be configured to utilize a floodsub protocol — which floods the network with messages, and peers are required to ignore messages they are not interested in — or gossipsub — which is a proximity-aware epidemic pubsub, where peers communicate with proximal peers, and messages can be routed more efficiently.
+Libp2p's pubsub module can be configured to utilize a floodsub protocol --- which floods the network with messages, and peers are required to ignore messages they are not interested in --- or gossipsub --- which is a proximity-aware epidemic pubsub, where peers communicate with proximal peers, and messages can be routed more efficiently.
 In those implementations, there is a benefit to using pubsub in that no direct connection between publishers and subscribers is required.
@@
 Like libp2p, encryption is a separate concern and often added in steps prior to data transmission.
-However, like libp2p, pubsub doesn’t offer any simple solutions for transferring encryption keys (beyond public keys), synchronizing datasets across peers (i.e.
-they aren’t databases), or enforcing any measures for access control (e.g.
+However, like libp2p, pubsub doesn't offer any simple solutions for transferring encryption keys (beyond public keys), synchronizing datasets across peers (i.e.
+they aren't databases), or enforcing any measures for access control (e.g.
 anyone subscribed to a topic can also author updates on that topic).
@@
 Using their private key, a content creator can update what static route the IPNS address refers to.
-IPNS isn’t only useful for creating static addresses that can point to content addresses, IPNS is compatible with a naming system external from IPFS, such as DNS, onion, or bit addresses.
+IPNS isn't only useful for creating static addresses that can point to content addresses, IPNS is compatible with a naming system external from IPFS, such as DNS, onion, or bit addresses.
 However, many use-cases that require highly mutable data, require rapid availability of updates, or want flexible multi-party access control may not yet be suitable for IPNS.
@@
 
-\section{The Threads protocol}
+\section{The Threads protocol}\label{sec:ThreadsProtocol}
 
@@
 Threads are topic-based collections of single-writer logs.
-Taken together, these logs represent the current “state” of an object or dataset.
+Taken together, these logs represent the current ``state'' of an object or dataset.
 The basic units of Threads, namely Logs and Events, provide a framework for users to create, store, and transmit data in a P2P distributed network.
@@
 \begin{quote}
-	A total order can be useful ...
-and could be obtained, for example, by considering concurrent events to be equal.
+	A total order can be useful \dots and could be obtained, for example, by considering concurrent events to be equal.
 Similarly, a strict total order could be built by sorting concurrent events by the CID or their nodes or by any other arbitrary user-defined strategy based on additional information attached to the clock nodes (data-layer conflict resolution).
@@
 \begin{definition}
-(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log’s Read key (see Section  \ref{sec:KeysEncryption}).
+(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log's Read key (see Section  \ref{sec:KeysEncryption}).
 \end{definition}
@@
 
-As seen above, an Event’s actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
+As seen above, an Event's actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
 
@@
 
-In practice, to reach a Log via it’s IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (see Example \ref{ex:Multiaddress}).
+In practice, to reach a Log via it's IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (see Example \ref{ex:Multiaddress}).
 
@@
 \begin{definition}Follower).
-Log Writers can designate other IPFS Peers to “follow” a Log, potentially replicating and/or republishing Events.
-A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (see Section  \ref{sec:KeysEncryption}), but is not able to read the Log’s contents.
-Followers should be server-based, i.e., always online and behind a public IP address.\end{definition}
+Log Writers can designate other IPFS Peers to ``follow'' a Log, potentially replicating and/or republishing Events.
+A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (see Section  \ref{sec:KeysEncryption}), but is not able to read the Log's contents.
+Followers should be server-based --- i.e., always online and behind a public IP address.\end{definition}
 
@@
 
-In practice, Writers are solely responsible for announcing their Log’s addresses.
+In practice, Writers are solely responsible for announcing their Log's addresses.
 This ensures a conflict-free address list without additional complexity.
 Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
-This allows them to pull (or subscribe to push-based) Events from the Follower’s Log address when the Writer is offline or unreachable.
+This allows them to pull (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable.
 
@@
 Every Log requires an asymmetric key-pair that determines ownership and identity.
-The private key is used to sign each Event added to the Log, so down-stream processes can verify the Log’s authenticity.
+The private key is used to sign each Event added to the Log, so down-stream processes can verify the Log's authenticity.
 Like IPFS peers, the public key of the Log is used as an identifier (Log ID).
@@
 Certain ID variants may be more appropriate than the others in specific use cases.
-For example, Textile provides an access-controlled Thread variant, which supports various collaborative structures — e.g., social media feeds, shared documents, blogs, photo albums, etc.
+For example, Textile provides an access-controlled Thread variant, which supports various collaborative structures --- e.g., social media feeds, shared documents, blogs, photo albums, etc.
 
@@
 \begin{definition}
-(Event Envelope). An over-the-wire message containing an Event and the sender’s signature of the Event.
+(Event Envelope). An over-the-wire message containing an Event and the sender's signature of the Event.
 \end{definition}
@@
 A new Thread is created by generating a TID and Log.
-The Log’s creator is the Writer, meaning it has possession of the Log’s Identity, Read, and Follow keys.
+The Log's creator is the Writer, meaning it has possession of the Log's Identity, Read, and Follow keys.
 All of these keys are needed to compose Events.
-At this point, the Thread only exists on the Writer’s machine.
-Whether for collaboration, reading, or following, the process of sharing a Thread with other Peers starts by authoring a special Event called an Invite, which contains a set of keys from all of the Thread’s Logs, called a Key Set.
+At this point, the Thread only exists on the Writer's machine.
+Whether for collaboration, reading, or following, the process of sharing a Thread with other Peers starts by authoring a special Event called an Invite, which contains a set of keys from all of the Thread's Logs, called a Key Set.
 
@@
 \begin{definition}
-(Key Set). A set of keys for a Log. Depending on the context, a Key Set may contain the Follow and Read key, or just the Follow key. Encrypted with the recipient’s public key.
-\end{definition}
-
-The Invite is authored in the sender’s Log.
-Because the recipient does not yet have this Log’s Key Set, the Event is encrypted with the recipient’s public key.
+(Key Set). A set of keys for a Log. Depending on the context, a Key Set may contain the Follow and Read key, or just the Follow key. Encrypted with the recipient's public key.
+\end{definition}
+
+The Invite is authored in the sender's Log.
+Because the recipient does not yet have this Log's Key Set, the Event is encrypted with the recipient's public key.
 If the recipient accepts the Invite, they will author another special Event called a Join in a new Log of their own.
@@
 \begin{definition}
-(Join). An Event containing an invitee’s new Log ID and Key Set, encrypted with the Key Set of the inviting Peer’s Log.
+(Join). An Event containing an invitee's new Log ID and Key Set, encrypted with the Key Set of the inviting Peer's Log.
 \end{definition}
@@
 For a Join to be successful, all Log Writers must receive a copy of the new Key Set so they can properly handle future Events in the new Log.
-Instead of encrypting a Join with the public key of each existing Writer, we can encrypt a single Join with the Key Set of the inviting Peer’s Log, which the other Writers also have.
+Instead of encrypting a Join with the public key of each existing Writer, we can encrypt a single Join with the Key Set of the inviting Peer's Log, which the other Writers also have.
 
@@
 \begin{enumerate}
-\item \label{Perf1}New Events are pushed directly to the Thread’s other Log Writers.
-\item \label{Perf2}New Events are pushed directly to the target Log’s Follower(s), who may not maintain their own Log.
-\item \label{Perf3} New Events are published over Libp2p’s gossip-based Pub/Sub infrastructure using TID as a topic, which provides potentially unknown Readers or Followers with an opportunity to consume Events in real-time.
+\item \label{Perf1}New Events are pushed directly to the Thread's other Log Writers.
+\item \label{Perf2}New Events are pushed directly to the target Log's Follower(s), who may not maintain their own Log.
+\item \label{Perf3} New Events are published over Libp2p's gossip-based Pub/Sub infrastructure using TID as a topic, which provides potentially unknown Readers or Followers with an opportunity to consume Events in real-time.
 \end{enumerate}
@@
 \item \label{Perf1}New Events may be pushed to web-based participants over a WebSocket.
-\item \label{Perf2}New Events may be pushed to the Thread’s other Log Writers via federated notification services like Apple Push Notification Service (APNS) [ref], Google Cloud Messaging (GCM) [ref], Firebase Cloud Messaging (FCM) [ref], and/or Windows Notification Service (WNS) [ref].
+\item \label{Perf2}New Events may be pushed to the Thread's other Log Writers via federated notification services like Apple Push Notification Service (APNS) [ref], Google Cloud Messaging (GCM) [ref], Firebase Cloud Messaging (FCM) [ref], and/or Windows Notification Service (WNS) [ref].
 \item \label{Perf3} New Events may trigger web-hooks [ref] enabling complex workflows (see Section X below).\\
@@
 
-Like many CQRS-based systems, the “public” API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store.
+Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store.
 A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store.
@@
 These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
-In the remainder of this section, we highlight a number of planned database abstractions to provide further utility to developers interested in building on Textile’s Threads protocol.\\
+In the remainder of this section, we highlight a number of planned database abstractions to provide further utility to developers interested in building on Textile's Threads protocol.\\
 
@@
 A key/value store built on top of Threads can map database `Put` operations into internal `KeyValueUpdated` events.
-These Events would then operate on a map-like view model, via an internal “projection“ (reducer function).
+These Events would then operate on a map-like view model, via an internal ``projection'' (reducer function).
 The key/value view could be persisted or kept in-memory, and would be directly queried via a corresponding `Get` operation.
@@
 The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, but exposes the underlying event sourced data for customizing as needed.
-Developers should not have to learn a whole new set of terms and tools to take advantage of Textile’s Threads capabilities.
+Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities.
 Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents:
@@
 Each database style will be implemented as a standalone software library, allowing application developers to pick and choose the solution most useful to the application at hand.
-Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these “example“ libraries.\\
+Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these ``example'' libraries.\\
 
@@
 For application developers familiar with common state management issues, tools such as Redux provide an opinionated framework for making state mutations predictable by imposing certain restrictions on how and when updates can happen.
-In fact, Redux builds on concepts from CRQS and ES itself [ref], and combines them with three core principals [ref]: 1) that the entire “state“ of an application should be stored in a single “store“, 2) where the only way to change the state is to emit an “action“ (or event), which 3) transforms the application state using a “pure“ (i.e., zero side effects) reducer function (projection).
-In practice, Redux doesn't care how a developer persists the application state, which means a Thread’s event store and read-only materialized views are excellent candidates.
+In fact, Redux builds on concepts from CRQS and ES itself [ref], and combines them with three core principals [ref]: 1) that the entire ``state' of an application should be stored in a single ``store'', 2) where the only way to change the state is to emit an ``action'' (or event), which 3) transforms the application state using a ``pure'' (i.e., zero side effects) reducer function (projection).
+In practice, Redux doesn't care how a developer persists the application state, which means a Thread's event store and read-only materialized views are excellent candidates.
 This focus on unidirectional data flow in perfectly in line with architecture proposed in this paper, and as such, a compatible Redux API will be built on top of Textile Threads.
@@
 At an abstract level, the Textile protocol provides a highly distributed framework for building shared, offline first, datastores that are fault tolerant, eventually consistent, and scalable.
-Any internal implementation details of a compliant Threads “client“ may use any number of well-established design patterns from the CQRS+ES (and related) literature to “extend“ the Threads protocol with additional features and controls.
-Some examples of extensions that are included by default in Textile’s Threads implementation are outlined in this section to provide some indication of the extensibility of Threads.
+Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS+ES (and related) literature to ``extend'' the Threads protocol with additional features and controls.
+Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility of Threads.
 
@@
 They can be used to rebuild the state of a view without having to query and re-play all previous events.
-When a Snapshot is available, a Thread Peer can rebuild the state of a given View by replaying any events generated since the latest Snapshot (with a corresponding reduction in the total number of events that need to be processed) using the View’s Reducer function.
-Multiple peers processing the same Log could create a Snapshot every 1000 events and be guaranteed to create the exact same Snapshot because each Peer’s Event counts are identical.
+When a Snapshot is available, a Thread Peer can rebuild the state of a given View by replaying any events generated since the latest Snapshot (with a corresponding reduction in the total number of events that need to be processed) using the View's Reducer function.
+Multiple peers processing the same Log could create a Snapshot every 1000 events and be guaranteed to create the exact same Snapshot because each Peer's Event counts are identical.
 
@@
 
-Textile’s Threads includes ACL management tooling based on a Role-based access control [ref] pattern, wherein individuals or groups are assigned roles which carry specific permissions.
+Textile's Threads includes ACL management tooling based on a Role-based access control [ref] pattern, wherein individuals or groups are assigned roles which carry specific permissions.
 Roles can be added and removed as needed.
@@
 Members of this role are able to verify Events and follow linkages.
-The Follow role is used to designate a “follower” peer for offline replication and/or backup.
+The Follow role is used to designate a ``follower'' peer for offline replication and/or backup.
 \end{definition}
@@
 \begin{definition}
-(Delete). Members of this role are able to delete Events, which implies access to Log Follow Keys. In practice, this means marking an older Event as “deleted”.
+(Delete). Members of this role are able to delete Events, which implies access to Log Follow Keys. In practice, this means marking an older Event as ``deleted''.
 \end{definition}
@@
 
- A Thread-level ACL has it’s own document ACL, which also applies to all other document ACLs:
+ A Thread-level ACL has it's own document ACL, which also applies to all other document ACLs:
 
@@
 Potential solutions currently being explored by Textile developers include key rotation at specific Event offsets [ref], and/or incorporating the Double Ratchet Algorithm [ref] for forward secrecy [ref].
-
-\section{Latex Examples}
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
-  \caption{A boat.}
-  \label{fig:boat1}
-\end{figure}
-
-\begin{quotation}
-A quote
-\end{quotation}
-
-\begin{figure*}
-  \includegraphics[width=\textwidth,height=4cm]{boat.jpg}
-  \caption{This is a full width boat.}
-  \label{fig:boat1}
-\end{figure*}
-
-Here is a citation models~\cite{gelenbe06}. The approach constructs
-
-here is a ref provides performance measures of interest to the SB and to the
-seller, is discussed in Section \ref{Performance} where we first
-discuss how the SB can behave in order to optimise outcomes that
-are in its best interest, and provide  numerical examples to
-illustrate the approach and the model predictions. We then explore
-how the SB can try to achieve balance and compete with the other
-bidders in Section \ref{KeepUp}. Finally Section
-\ref{Price-Dependent} generalises the analysis to the case where
-the bidding rates depend on the current price attained in the
-auction. Conclusions are drawn in Section \ref{Conclusions} where
-we also suggest further work.
 
@@
 \bibliographystyle{compj}
-% \bibliography{ModellingBidders}
-\input{whitepaper.bbl}
+\bibliography{whitepaper}
 

### 2019-09-20 draft.tex
all figures updated, new tables, lots of linkages
Signed-off-by: andrew <andrew@textile.io>

@@
 \usepackage{nicefrac}       % compact symbols for 1/2, etc.
+
+\usepackage{dblfloatfix}    % To enable figures at the bottom of page
+
+% trying to keep figures in the right sections instead of pushed to the end
+\usepackage[section]{placeins}
+
+% To create tables + create checkmarks
+\usepackage{booktabs}
+\usepackage{colortbl}
+\usepackage{multirow}% http://ctan.org/pkg/multirow
+
+\usepackage{array}
+\usepackage{tikz}
+\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
+
+\newcommand\Tstrut{\rule{0pt}{2.6ex}}       % "top" strut
+\newcommand\Bstrut{\rule[-0.9ex]{0pt}{0pt}} % "bottom" strut
+\newcommand{\TBstrut}{\Tstrut\Bstrut} % top&bottom struts
+
+% column centering special type
+\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}
+% End table requirements
+
+% Thread ID components fig requrements
+\usepackage{stackengine}
+% Thread ID equation setup 
+\def\textscrm#1{\textrm{\scriptsize#1}}
+\stackMath
+\def\rldp{1.3ex}
+\def\rlht{.8ex}
+\def\rlwd{.8pt}
+\setstackgap{L}{\rldp}
+\def\uvbarR#1{%
+  \def\stackalignment{r}\def\stacktype{S}\stackunder[-\rlwd]{%
+    \def\stackalignment{c}\def\stacktype{L}\stackunder{#1}{\rule{\rlwd}{\rlht}}%
+  }{\setbox0\hbox{$#1$}\rule{.5\wd0}{\rlwd}}%
+}
+\def\uvbarL#1{%
+  \def\stackalignment{l}\def\stacktype{S}\stackunder[-\rlwd]{%
+    \def\stackalignment{c}\def\stacktype{L}\stackunder{#1}{\rule{\rlwd}{\rlht}}%
+  }{\setbox0\hbox{$#1$}\rule{.5\wd0}{\rlwd}}%
+}
+\def\uvbar#1#2{%
+  \def\stacktype{S}\def\stackalignment{c}\def\useanchorwidth{T}\stackunder[0pt]{%
+    \def\stacktype{L}\setbox0\hbox{${}#1{}$}%
+    \stackunder[\rldp]{{}#1{}}{\rule{\wd0}{\rlwd}}%
+  }{\def\stacktype{S}\stackunder[1pt]{\rule{\rlwd}{\rlht}}{\scriptstyle#2}}%
+}
+\newcommand\ubar[2]{\uvbarR{\textrm{}}\uvbar{\textrm{#1}}{\textscrm{#2}}%
+  \uvbarL{\textrm{}}}
+\newcommand\phubar[2]{\uvbarR{\textrm{}}\uvbar{\phantom{\textrm{#1}}}{\textscrm{#2}}%
+  \uvbarL{\textrm{}}}
+
+% End thread Id requirements
 
@@
 
+% bolds "Example X" to match Figure X
+\usepackage[labelfont=bf]{caption}
+\DeclareCaptionType{example}
+
 \usepackage{trivfloat}
@@
 Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates \cite{fowlerEventSourcing}.
-The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see section  \ref{sec:ViewsProjections}).
+The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see Section  \ref{sec:ViewsProjections}).
 In other words, with minimal conformity, a single log can model multiple application states \cite{microsoftcorporationAzureApplicationArchitecture}.
@@
 IPLD is used to publish linked data (subject-predicate-object triples in the linked data world \cite{heathLinkedDataEvolving2011}) spread across different hosts, and where everything (entities, predicates, data sources) uses content addresses as unique identifiers.
-To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph\footnote{For reference, the Bitcoin Blockchain is a DAG, so is a Git version history.}.
+To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph\footnote{Other examples of Dags include the Bitcoin Blockchain or a Git version history.}.
 This allows all hash-linked data structures to be treated using a unified data model, analogous to linked data in the Semantic Web sense \cite{brendanobrienDeterministicQueryingDistributed2017}.
@@
 \subsubsection{Merkle-Clocks} \label{sec:merkleclocks}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/Divergent_Heads.png}
+  \caption{Divergent heads in a multi-writer Merkle-DAG }
+  \label{fig:merkledag}
+\end{figure}
 
@@
 
-On the downside, and as seen above (section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots):
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
-  \caption{Divergent heads in a multi-writer Merkle-DAG }
-  \label{fig:merkledag}
-\end{figure}
+On the downside, and as discussed above (section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots) (see Figure \ref{fig:merkledag})
 
@@
 Based on these observations, using a single Merkle-Clock to synchronize replicas can be problematic.
+
+% blocks the merkledag image from being rendered past here
+\FloatBarrier
 
@@
 All filesystems and databases have some notion of "access control." Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects [ref].
-An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object.
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
+An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object (see Figure /ref{tbl:ACL}).
+
+\begin{table*}
 \caption{Example Access Control List.}
-  \label{fig:boat1}
-\end{figure}
+    \centering
+    \begin{tabular} [t]{l>{\centering}p{0.2\linewidth}>{\centering}p{0.2\linewidth}>{\centering}p{0.2\linewidth}>{\centering\arraybackslash}p{0.2\linewidth}}
+\toprule
+&Create&Delete&Edit&Read\\
+\midrule
+John&\checkmark&\checkmark&\checkmark&\checkmark\\
+Jane&-&-&-&\checkmark\\
+Mary&-&-&\checkmark&\checkmark\\
+\bottomrule
+    \end{tabular}
+    \label{tbl:ACL}
+\end{table*}%
 
@@
 Threads offer data ownership and a multi-role data access architecture where owners can set independent permissions for writing, reading, and following data updates.
-Threads differ from previous solutions by extending on the multiaddress addressing scheme to allow PULL based replica synchronization in addition to more common PUSH based synchronization seen in decentralized protocols.
+Threads differ from previous solutions by extending on the multiaddress addressing scheme to allow pull-based replica synchronization in addition to more common push-based synchronization seen in decentralized protocols.
 The flexible event-based structure enables client applications to model advanced states through aggregates, views, and custom CRDTs.
@@
 
+% [!b] indicates that it will try to be on the bottom of the page
+\begin{figure}[!b]
+  \includegraphics[width=\linewidth]{figures/Event_Log.png}
+  \caption{A single-writer Merkle-Clock (Event Log).}
+  \label{fig:EventLog}
+\end{figure}
+
 In multi-writer systems, determining causal order in all replicas is particularity challenging.
 As previously discussed (section 2.2.4), a solution based on a Merkle-Clock is only partially ordered.
-That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy [ref]:
+That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy [ref].
 
@@
 Our solution to deal with divergent Merkle-Clocks is to institute a single-writer rule: A log can only be updated by a single replica or identity.
-An Event Log is a single-writer Merkle-Clock that can be totally ordered.
+An Event Log is a single-writer Merkle-Clock that can be totally ordered (Figure \ref{fig:EventLog}).
 Separate Event Logs can be composed into advanced structures, including CRDTs [Ref Enes].
 
-Practically speaking, for any given Log, Events are authored by a single IPFS Peer, or Writer.
+For any given Log, Events are authored by a single IPFS Peer, or Writer.
 
@@
 
-As seen above in section 2.3.4., a Merkle-Clock is simply a Merkle-DAG of Events:
+As seen above in Section \ref{sec:merkleclocks}., a Merkle-Clock is simply a Merkle-DAG of Events:
 
@@
 
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
-  \caption{A single-writer Merkle-Clock (Event Log).}
-  \label{fig:boat1}
-\end{figure}
-
-Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure X.X.) enabling log verification by readers (see Section  \ref{sec:KeysEncryption}).
-
-At a minimum, a node must link to its most immediate ancestor. However, links to older ancestors are often included as well to improve concurrency during traversal and verification [ref].
-
-TODO: Ref to appendix
-
-As seen above, an Event's actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
-
-\begin{example*}
+% blocks the EventLog image from being rendered past here
+\FloatBarrier
+
+\begin{example*}[!b]
 \centering
@@
 
+Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure X.X.) enabling log verification by readers (see Section  \ref{sec:KeysEncryption}).
+
+At a minimum, a node must link to its most immediate ancestor. However, links to older ancestors are often included as well to improve concurrency during traversal and verification [ref].
+
+TODO: Ref to appendix
+
+As seen above, an Event's actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
+
 Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses [ref].
@@
 
-Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log
-
-TODO: ref to embed
+Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log (Example {ex:Multiaddress}).
 
@@
 This ensures a conflict-free address list without additional complexity.
-Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
-This allows them to pull (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable.
+Some Followers may be in the business of replicating Logs (see Section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
+This allows them to pull (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable (Figure \ref{fig:Pulling}).
 
@@
 
-\begin{figure*}
+\begin{figure*}[!t]
 \centering
@@
 
-Textile Logs are designed to be shared, composed, and layered into datasets.
+Textile Logs are designed to be shared, composed, and layered into datasets (Figure \ref{fig:LogEncryption}).
 Therefore, Logs are encrypted by default in a manner that enables access control (see Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
@@
 
-A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pub/Sub-based synchronization. TIDs are defined with the following format:
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
-  \caption{The thread id formula.}
-  \label{fig:boat1}
-\end{figure}
+A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pub/Sub-based synchronization.
+TIDs are defined with the following format {see Figure \ref{fig:ThreadID}
+
+% [!b] bottom page
+\begin{figure*}[!b]
+\centering
+\caption{Components of a Thread Identity.}
+\par\vspace{.9\baselineskip}\leavevmode%
+\par{Thread ID = 0x62\quad}\ubar{0x01}{Version}\quad0x55\quad\ubar{0x539bc1dc03ee8cb5d478e41cc8a4546e}{Random Number} 
+\par\vspace{-.6\baselineskip}\leavevmode%
+\phantom{Thread ID = }\phubar{0x62}{Multibase}%
+  \phantom{\quad0x01\quad}\phubar{0x55}{Variant} \phantom{\quad7513b4dcf4efa291ffa92221659f19e2}
+\par\vspace{-.6\baselineskip}\leavevmode
+\label{fig:ThreadID}
+\end{figure*}
 
@@
 
-See Example \ref{ex:Identity}).
+See Example \ref{ex:Identity}a).
 \end{definition}
@@
 
-See Example \ref{ex:Identity}).
+See Example \ref{ex:Identity}b).
 \end{definition}
@@
 \begin{lstlisting}[language=json,firstnumber=1]
-# Raw identity. V1, 128 bit
+# (a) Raw identity. V1, 128 bit
 bafkxd5bjgi6k4zivuoyxo4ua4mzyy
 
-# ACL enabled identity. V1, 256 bit.
+# (b) ACL enabled identity. V1, 256 bit.
 bafyoiobghzefwlidfrwkqmzz2ka66zgmdmgeobw2mimktr5jivsavya
@@
 \item \label{Perf2} Events can be pulled from replicating Followers via HTTP, libp2p, RSS, Atom, etc.
-    a. In conjunction with push over WebSockets (2a above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events.
+    a. In conjunction with push over WebSockets (2a above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events (Figure \ref{fig:Pulling}).
 \item \label{Perf3} Writers and readers can receive new Events via a Pub/Sub subscription at the TID.
@@
 
-<insert diagram of pulling from a replicating Follower)
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/Pulling.png}
+  \caption{A pull-based request from a Follower.}
+  \label{fig:Pulling}
+\end{figure}
 
@@
 This type of Log replication can act as a data backup mechanism.
-It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (see section  \ref{sec:LogSync}).
+It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (see Section  \ref{sec:LogSync}).
 
@@
 
-\textbf{CRDTs} \\
+\textbf{CRDTs} \label{sec:TexCRDT} \\
 
@@
 Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition XXX) basis.
-Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see section 3.3.2 for details).
+Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see Section \ref{sec:TexCRDT}).
 ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible).
@@
 
-\begin{example}
+% [!b] moves it to the bottom of the page
+\begin{example}[!b]
 \centering
@@
 
-\subsubsection{Textile: The Thread \& Bot Network}
+\subsubsection{Textile: The Thread \& Bot Network} \label{sec:Bots}
 
@@
 
-\begin{example*}
+% [!p] forces it to just take the whole page
+\begin{example*}[!p]
 
@@
 \end{lstlisting}
-  \caption{This is a full width code block.}
+  \caption{Interfaces for an event body, header and node.}
 

### 2019-09-20 draft.tex
refs: moar refs - poorly chosen branch name
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

--- 2019-09-20 draft.tex
@@
 
-\subsubsection{Conflict-Free Replicated Data Types}
+\subsubsection{Conflict-Free Replicated Data Types}\label{sec:CRDTs}
 
@@
 The libp2p project provides a robust protocol communication stack.
-IPFS and a growing list of other projects [MISSING] are building on top of libp2p.
+IPFS and a growing list of other projects (Polkadot, Ethereum 2.0, Substrate, FileCoin, OpenBazzar, Keep, etc) are building on top of libp2p.
 Libp2p solves a number of challenges that are distinct to peer-to-peer (p2p) networks.
-A comprehensive coverage of networking issues in P2P systems is out of the scope for this paper, however, some core challenges that libp2p helps to address include NAT traversal, peer discovery and handshake protocols, and even encryption and transport security --- libp2p supports both unencrypted (e.g.
+A comprehensive coverage of networking issues in P2P systems is out of the scope for this paper, however, some core challenges that libp2p helps to address include network address translator \cite{srisureshIPNetworkAddress} (NAT) traversal, peer discovery and handshake protocols, and even encryption and transport security --- libp2p supports both unencrypted (e.g.
 TCP, UDP) and encrypted protocols (e.g.
 TLS, Noise) --- among others.
-Libp2p uses the concept of a multiaddress to address peers on a network, which essentially models network addresses as arbitrary encapsulations of protocols [ref, ref].
+Libp2p uses the concept of a multiaddress to address peers on a network, which essentially models network addresses as arbitrary encapsulations of protocols \cite{Multiaddr}.
 In addition to ``transport layer'' modules, libp2p provides several tools for sharing and/or disseminating data over a p2p network.
@@
 Pubsub is a standard messaging pattern where the publishers don't know who, if anyone, will subscribe to a given topic.
-Publishers send messages on a given topic or category, and Subscribers receive only messages on a give topic they are subscribed to.
-Libp2p's pubsub module can be configured to utilize a floodsub protocol --- which floods the network with messages, and peers are required to ignore messages they are not interested in --- or gossipsub --- which is a proximity-aware epidemic pubsub, where peers communicate with proximal peers, and messages can be routed more efficiently.
+\emph{Publishers} send messages on a given topic or category, and \emph{Subscribers} receive only messages on a give topic to which they are subscribed.
+Libp2p's pubsub module can be configured to utilize a \emph{floodsub} protocol --- which floods the network with messages, and peers are required to ignore messages they are not interested in --- or \emph{gossipsub} --- which is a proximity-aware epidemic pubsub, where peers communicate with proximal peers, and messages can be routed more efficiently.
 In those implementations, there is a benefit to using pubsub in that no direct connection between publishers and subscribers is required.
@@
 To solve some of those challenges, some systems introduce message echoing and other partial solutions.
-However, it makes more sense to use pubsub and libp2p as building blocks in systems that can effectively solve these issues, by choosing multi-modal communication strategies or leveraging tools such as deferred routing (e.g.
+However, it makes more sense to use pubsub and libp2p as \emph{building blocks} in systems that can effectively solve these issues, by choosing multi-modal communication strategies or leveraging tools such as deferred routing (e.g.
 inboxing) for greater tolerance of missed messages.
@@
 
-Pubsub and libp2p have so far dealt with the push transfer of data, but IPFS also offers a useful technology for hosting pull/request based data endpoints called, IPNS.
+Pubsub and libp2p have so far only dealt with \emph{push-based} transfer of data, but IPFS also offers a useful technology for hosting \emph{pull/request} based data endpoints called, IPNS.
 IPNS aims to address the challenge of mutable data within IPFS.
@@
 delete data) is permitted.
-Many decentralized networks like Secure Scuttlebutt [1] and Holochain [2] make use of agent-centric security.
+Many decentralized networks like Secure Scuttlebutt \cite{ScuttlebuttProtocolGuide} and Holochain \cite{ericharris-braunHolochainScalableAgentcentric2018} make use of agent-centric security.
 Each of these systems leverage cryptographic signatures to validate peer identities and messages.
@@
 
-All filesystems and databases have some notion of "access control." Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects [ref].
+All filesystems and databases have some notion of ``access control''.
+Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects \cite{shireyInternetSecurityGlossary2007}.
 An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object.
@@
   \caption{Example Access Control List.}
-  \label{fig:boat1}
+  \label{fig:ACL}
 \end{figure}
 
-Using ACLs in systems where identity is derived from various configurations of public-key infrastructure has been around for some time [1, 2].
+Using ACLs in systems where identity is derived from various configurations of public-key infrastructure has been around for some time \cite{herzbergAccessControlMeets2000}.
 Still, many existing database and communication protocols built on IPFS to date lack support for an ACL or only have primitive ACL support.
@@
 The ACL of an event-driven system is usually a list of access rules built from a series of events.
-For example, the two events, "grant Bob write access" and "revoke read access from Alice" would together result in a final ACL state where, Bob has read and write access, but Alice does not (see Section  \ref{sec:ViewsProjections}).
+For example, the two events, ``grant Bob write access'' and ``revoke read access from Alice'' would together result in a final ACL state where, Bob has read and write access, but Alice does not (see Section  \ref{sec:ViewsProjections}).
 
@@
 In multi-writer systems, determining causal order in all replicas is particularity challenging.
-As previously discussed (section 2.2.4), a solution based on a Merkle-Clock is only partially ordered.
-That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy [ref]:
+As previously discussed (section \ref{sec:merkleclocks}), a solution based on a Merkle-Clock is only partially ordered.
+That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:
 
@@
 
-Our solution to deal with divergent Merkle-Clocks is to institute a single-writer rule: A log can only be updated by a single replica or identity.
-An Event Log is a single-writer Merkle-Clock that can be totally ordered.
-Separate Event Logs can be composed into advanced structures, including CRDTs [Ref Enes].
-
-Practically speaking, for any given Log, Events are authored by a single IPFS Peer, or Writer.
+Our solution to deal with divergent Merkle-Clocks is to institute a \emph{single-writer rule}: A log can only be updated by a single replica or \emph{identity}.
+An Event Log is a \emph{single-writer Merkle-Clock} that can be totally ordered.
+\emph{Separate} Event Logs can be composed into advanced structures, including CRDTs \cite{enesSingleWriterPrincipleCRDT2017}.
+
+Practically speaking, for any given Log, Events are authored by a single IPFS Peer, or \emph{Writer}.
 
@@
 
-This single-writer setup is a core feature of Logs, and provides properties unique to the Threads protocol. For clarity, we can similarly define a Reader as any other Peer capable of reading a Log.
+This single-writer setup is a core feature of Logs, and provides properties unique to the Threads protocol.
+For clarity, we can similarly define a \emph{Reader} as any other Peer capable of reading a Log.
 
@@
 
-As seen above in section 2.3.4., a Merkle-Clock is simply a Merkle-DAG of Events:
-
-\begin{definition}
- (Event). An single node in a Merkle-Clock, stored on IPFS.
+As seen above in section \ref{sec:merkleclocks}, a Merkle-Clock is simply a Merkle-DAG of \emph{Events}:
+
+\begin{definition}
+ (Event). A single node in a Merkle-Clock, stored on IPFS.
 \end{definition}
@@
   \caption{A single-writer Merkle-Clock (Event Log).}
-  \label{fig:boat1}
+  \label{fig:EventLog}
 \end{figure}
 
-Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure X.X.) enabling log verification by readers (see Section  \ref{sec:KeysEncryption}).
-
-At a minimum, a node must link to its most immediate ancestor. However, links to older ancestors are often included as well to improve concurrency during traversal and verification [ref].
-
-TODO: Ref to appendix
+Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure \ref{fig:EventLog}) enabling log verification by readers (see Section  \ref{sec:KeysEncryption}).
+
+At a minimum, a node must link to its most immediate ancestor.
+However, links to older ancestors are often included as well to improve concurrency during traversal and verification \cite{meyerBamboo2019}.
+
+MISSING: Ref to appendix
 
@@
 
-Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses [ref].
-Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (see Example \ref{ex:Multiaddress}).
+Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses \cite{Multiaddr}.
+Here we introduce \emph{IPEL}, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (see Example \ref{ex:Multiaddress}).
 
@@
 
-Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table [ref] (DHT).
+Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table \cite{benetIPFSContentAddressed2014} (DHT).
 Instead, they are collected from Log Events (see Section  \ref{sec:UseCases}).
 This is in contrast to mutable data via IPNS for example, which requires querying the network (DHT) for updates.
-Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed like topology.
+Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed\footnote{This is similar to the append-only message feeds used in Secure Scuttlebutt’s global gossip network \cite{ScuttlebuttProtocolGuide}} like topology.
 Log addresses are recorded in an address book, similar to IPFS Peer address book:
+
+MISSING PEER ADDRESS BOOK?
 
@@
 
-\begin{definition}Follower).
+\begin{definition} (Follower).
 Log Writers can designate other IPFS Peers to ``follow'' a Log, potentially replicating and/or republishing Events.
@@
 
-Followers are represented as additional addresses, meaning that a Log address book may contain multiple multiaddresses for a single Log
-
-TODO: ref to embed
+Followers are represented as additional addresses, meaning that a Log address book may contain \emph{multiple} multiaddresses for a single Log .
+
+MISSING: ref to embed
 
@@
 Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
-This allows them to pull (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable.
+This allows them to \emph{pull} (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable.
 
@@
 
-The body, or content of an event, is encrypted by a Content Key.
+The body, or content of an event, is encrypted by a \emph{Content Key}.
 Content keys are generated per-content and never reused.
@@
 \item \label{Perf1}When broadcasting events to many possible recipients, a single-use symmetric key is generated per unique content body.
-\item \label{Perf2}When sending events to specific recipients, the recipient's public key can be used to restrict access from all others.
+\item \label{Perf2}When sending events to specific recipients, the recipient's public key can be used to restrict access from all others\footnote{much like private messages in Secure Scuttlebutt (https://www.scuttlebutt.nz/concepts/private-message)}.
 \end{enumerate}
@@
 If a single-use symmetric key is used for the Content key, it is necessary to distribute each new key to users by including it in the header of the Event Block.
-Therefore, the Event Block itself is further encrypted using a Read key.
+Therefore, the Event Block itself is further encrypted using a \emph{Read} key.
 The Read key is not distributed within the Log itself but is distributed to any peers to grant them access to the content of the Log.
@@
 Finally, the encrypted Event Block, its signature, and the IPLD linkage(s) from an Event to its antecedents are encrypted together using a Follow key.
-Follow keys allow Logs to be followed by peers on the network who do not have access to any content within the event, and can only see signatures and linkage(s) between Events.
+Follow keys allow Logs to be \emph{followed} by peers on the network who do not have access to any content within the event, and can only see signatures and linkage(s) between Events.
 
@@
 
-In Textile, the interface to Logs is managed as a Thread.
+In Textile, the \emph{interface} to Logs is managed as a Thread.
 As mentioned previously, a Thread is a collection of Logs on a given topic.
@@
 Threads provide the mechanism to combine multiple Logs from individual authors into singular shared states through the use of either cross-Log sequencing (e.g.
-using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock [ref]) or a CRDT.
+using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock \cite{kulkarniLogicalPhysicalClocks2014}) or a CRDT (see Section \ref{sec:CRDTs}).
 
@@
   \caption{The thread id formula.}
-  \label{fig:boat1}
+  \label{fig:ThreadId}
 \end{figure}
 
-TIDs share some similarities with UUIDs [ref] (version and variant) and IPFS-based CIDs and are multibase encoded [ref] for maximum forward-compatibility.
+TIDs share some similarities with UUIDs \cite{leachUniversallyUniqueIDentifier2005} (version and variant) and IPFS-based CIDs and are multibase encoded\footnote{https://github.com/multiformats/multibase} for maximum forward-compatibility.
 
@@
 
-Certain ID variants may be more appropriate than the others in specific use cases.
-For example, Textile provides an access-controlled Thread variant, which supports various collaborative structures --- e.g., social media feeds, shared documents, blogs, photo albums, etc.
+Certain ID \emph{variants} may be more appropriate than the others in specific use cases.
+For example, Textile provides an \emph{access-controlled} Thread variant, which supports various collaborative structures --- e.g., social media feeds, shared documents, blogs, photo albums, etc.
 
@@
 Log Writers, Readers, and Followers synchronize the state of their Logs by sending and receiving Events.
-Inspired by Git [ref], a reference to the latest Event in a Log is referred to as the Head (or sometimes the root).
+Inspired by Git\footnote{https://git-scm.com/}, a reference to the latest Event in a Log is referred to as the \emph{Head} (or sometimes the \emph{root}).
 When a new Event is received, Readers and Followers simply advance their Head reference.
 
-Regardless of the network protocol, Events are transported between Peers in a standardized Event Envelope:
+Regardless of the network protocol, Events are transported between Peers in a standardized \emph{Event Envelope}:
 
@@
 At this point, the Thread only exists on the Writer's machine.
-Whether for collaboration, reading, or following, the process of sharing a Thread with other Peers starts by authoring a special Event called an Invite, which contains a set of keys from all of the Thread's Logs, called a Key Set.
+Whether for collaboration, reading, or following, the process of sharing a Thread with other Peers starts by authoring a special Event called an \emph{Invite}, which contains a set of keys from all of the Thread's Logs, called a \emph{Key Set}.
 
@@
 Because the recipient does not yet have this Log's Key Set, the Event is encrypted with the recipient's public key.
-If the recipient accepts the Invite, they will author another special Event called a Join in a new Log of their own.
-
-\begin{definition}
-(Join). An Event containing an invitee's new Log ID and Key Set, encrypted with the Key Set of the inviting Peer's Log.
+If the recipient accepts the Invite, they will author another special Event called a \emph{Join} in a new Log of their own.
+
+\begin{definition}
+(Join). An Event containing an invitee's new Log ID and Key Set, encrypted with the Key Set of the \emph{inviting Peer's Log}.
 \end{definition}
@@
 \begin{enumerate}
-\item \label{Perf1}New Events are pushed directly to the Thread's other Log Writers.
+\item \label{Perf1}New Events are pushed\footnote{Here push means ``send to multiaddress(es)'', which may designate different protocols, e.g., p2p, HTTP, Bluetooth, etc.} directly to the Thread's other Log Writers.
 \item \label{Perf2}New Events are pushed directly to the target Log's Follower(s), who may not maintain their own Log.
@@
 
-Step 2 above allows for additional push mechanisms, as followers with public IP addresses become relays:
+Step 2 above allows for \emph{additional} push mechanisms, as followers with public IP addresses become relays:
 
@@
 \item \label{Perf1}New Events may be pushed to web-based participants over a WebSocket.
-\item \label{Perf2}New Events may be pushed to the Thread's other Log Writers via federated notification services like Apple Push Notification Service (APNS) [ref], Google Cloud Messaging (GCM) [ref], Firebase Cloud Messaging (FCM) [ref], and/or Windows Notification Service (WNS) [ref].
-\item \label{Perf3} New Events may trigger web-hooks [ref] enabling complex workflows (see Section X below).\\
+\item \label{Perf2}New Events may be pushed to the Thread's other Log Writers via federated notification services like Apple Push Notification Service (APNS), Google Cloud Messaging (GCM), Firebase Cloud Messaging (FCM), and/or Windows Notification Service (WNS).
+\item \label{Perf3} New Events may trigger web-hooks\footnote{https://en.wikipedia.org/wiki/Webhook} enabling complex workflows (see Section X below).\\
 \end{enumerate}
@@
 Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store.
-A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store.
-In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be materialized to local persistent storage).
+A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store\footnote{Though see section \ref{sec:SnapshotsCompaction} for a discussion of Snapshots, which may be used to speed up Views.}.
+In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be \emph{materialized} to local persistent storage).
 This simple subscription system provides a flexible framework for building complex View-driven application logic.
 
-\subsubsection{Default Views}
-
-The core Textile Threads interfaces are also exposed via interfaces compatible with existing datastores and/or communications systems.
+\subsubsection{Default Views}\label{sec:DefaultViews}
+
+The core Textile Threads interfaces are also exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
 These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
@@
 
-A key/value store built on top of Threads can map database `Put` operations into internal `KeyValueUpdated` events.
+A key/value store built on top of Threads can map database \texttt{Put} operations into internal \texttt{KeyValueUpdated} events.
 These Events would then operate on a map-like view model, via an internal ``projection'' (reducer function).
-The key/value view could be persisted or kept in-memory, and would be directly queried via a corresponding `Get` operation.
-Because Textile uses ES patterns, a `Del` operation would simply lead to an internal `KeyValueDeleted` event, marking the key/value pair for removal from the key/value materialized view.
+The key/value view could be persisted or kept in-memory, and would be directly queried via a corresponding \texttt{Get} operation.
+Because Textile uses ES patterns, a \texttt{Del} operation would simply lead to an internal \texttt{KeyValueDeleted} event, marking the key/value pair for removal from the key/value materialized view.
 
@@
 
-Where `Indexable` can be satisfied by any structure with a  `key` field and and `Query` is taken from the `go-datastore` interface library [ref] or similar.
-Tables, feeds, counters, and other simple stores can also be built on Threads, and may require ORM-based solutions to support persistence to certain backing stores.
+Where \texttt{Indexable} can be satisfied by any structure with a  \texttt{key} field and and \texttt{Query} is taken from the \texttt{go-datastore} interface library\footnote{https://github.com/ipfs/go-datastore} or similar.
+Tables, feeds, counters, and other simple stores can also be built on Threads, and may require ORM-based solutions to support persistence to certain backing stores\footnote{In-memory stores are almost always implementable using compositions of basic data structures.}.
 Each database style will be implemented as a standalone software library, allowing application developers to pick and choose the solution most useful to the application at hand.
@@
 Eventually consistent, CRDT-based stores can also be implemented on top of Threads.
-CRDT-based stores are particularly useful for managing Views of a document in a multi-peer collaborative editing environment (like GoogleDocs [ref] or similar).
-For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype [ref] that applies updates on a JSON document View model.
-Libraries such as Automerge [ref] provide useful examples of reducer function that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available.
-A practical example of using a JSON CRDT in Textile is given in section 3.4, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.\\
+CRDT-based stores are particularly useful for managing Views of a document in a multi-peer collaborative editing environment (like Google Docs or similar).
+For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype \cite{kleppmannConflictFreeReplicatedJSON2017} that applies updates on a JSON document View model.
+Libraries such as Automerge\footnote{https://github.com/automerge/automerge} provide useful examples of reducer function that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available.
+A practical example of using a JSON CRDT in Textile is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.\\
 
@@
 For application developers familiar with common state management issues, tools such as Redux provide an opinionated framework for making state mutations predictable by imposing certain restrictions on how and when updates can happen.
-In fact, Redux builds on concepts from CRQS and ES itself [ref], and combines them with three core principals [ref]: 1) that the entire ``state' of an application should be stored in a single ``store'', 2) where the only way to change the state is to emit an ``action'' (or event), which 3) transforms the application state using a ``pure'' (i.e., zero side effects) reducer function (projection).
+In fact, Redux builds on concepts from CRQS and ES itself \cite{reduxMotivation}, and combines them with three core principals: 1) that the entire ``state' of an application should be stored in a single ``store'', 2) where the only way to change the state is to emit an ``action'' (or event), which 3) transforms the application state using a ``pure'' (i.e., zero side effects) reducer function (projection).
 In practice, Redux doesn't care how a developer persists the application state, which means a Thread's event store and read-only materialized views are excellent candidates.
@@
 
-At an abstract level, the Textile protocol provides a highly distributed framework for building shared, offline first, datastores that are fault tolerant, eventually consistent, and scalable.
+At an abstract level, the Textile protocol provides a highly distributed framework for building shared, offline first, datastores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable.
 Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS+ES (and related) literature to ``extend'' the Threads protocol with additional features and controls.
@@
 
-\subsubsection{Snapshots and Compaction}
-
-Snapshots are simply the current state of a View at a given point in time.
+\subsubsection{Snapshots and Compaction}\label{sec:SnapshotsCompaction}
+
+Snapshots\footnote{The literature around snapshots and other CQRS and ES is somewhat confusing, we attempt to use the most common definitions here.} are simply the current state of a View at a given point in time.
 They can be used to rebuild the state of a view without having to query and re-play all previous events.
 When a Snapshot is available, a Thread Peer can rebuild the state of a given View by replaying any events generated since the latest Snapshot (with a corresponding reduction in the total number of events that need to be processed) using the View's Reducer function.
-Multiple peers processing the same Log could create a Snapshot every 1000 events and be guaranteed to create the exact same Snapshot because each Peer's Event counts are identical.
+Multiple peers processing the same Log could create a Snapshot every 1000 events and be guaranteed to create the exact same Snapshot because each Peer's Event counts are identical\footnote{Assuming any network partitions are only short-lived (i.e., that peers are able to share events consistently).}.
 
@@
 Compaction can similarly speed up re-hydration of state, and is useful when only the latest Event per Event type is required.
-In addition to being used as a starting point for initialization and recovery, compacted logs can be used to replace the Log from which they were created in order to free up local disk space.
+In addition to being used as a starting point for initialization and recovery, compacted logs can be used to \emph{replace} the Log from which they were created in order to free up local disk space.
 In Textile, compaction becomes much easier because a log only has one writer and conflicts are not possible.
@@
 There are two forms of access control possible in Threads, Sequence-level ACLs and Thread-level ACLs.
-Thread-level access control lists (ACLs) allow creators to specify who can follow, read, write, and delete Thread data.
-Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition XXX) basis.
-Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see section 3.3.2 for details).
+Thread-level access control lists (ACLs) allow creators to specify who can \emph{follow, read, write, and delete} Thread data.
+Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition \ref{ex:SequenceID}) basis.
+Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see section \ref{sec:DefaultViews} for details).
 ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible).
@@
 
-Textile's Threads includes ACL management tooling based on a Role-based access control [ref] pattern, wherein individuals or groups are assigned roles which carry specific permissions.
+Textile's Threads includes ACL management tooling based on a \emph{Role-based access control} \cite{sandhuRolebasedAccessControl1996} pattern, wherein individuals or groups are assigned roles which carry specific permissions.
 Roles can be added and removed as needed.
-Textile ACLs can make use of five distinct roles: No-access, Follow, Read, Write, and Delete.
+Textile ACLs can make use of five distinct roles\footnote{By default, Threads without access control operate similar to Secure Scuttlebutt (SSB; where every peer consumes what they want and writes what they want).}: \emph{No-access, Follow, Read, Write, and Delete}.
 
@@
 
-Below is a typical Thread-level ACL JSON document (where id is a unique id), which could be persisted to a local document store as part of a materialized view.
+Below is a typical Thread-level ACL JSON document (where \texttt{\_id} is a unique id), which could be persisted to a local document store as part of a materialized view.
 
@@
 
-The `default` key states the default role for all network peers.
-The `peers` map is where roles are delegated to specific peers.
-Here, `12D..dwaA6Qe` is likely the owner, `12D..dJT6nXY` is a designated follower, and `12D..P2c6ifo` has been given read access.
+The \texttt{default} key states the default role for all network peers.
+The \texttt{peers} map is where roles are delegated to specific peers.
+Here, \texttt{12D..dwaA6Qe} is likely the owner, \texttt{12D..dJT6nXY} is a designated follower, and \texttt{12D..P2c6ifo} has been given read access.
 
@@
 
-This means that only `12D..dwaA6Qe` is able to alter the access-control list.
+This means that only \texttt{12D..dwaA6Qe} is able to alter the access-control list.
 
@@
 In this paper, we described the challenges and considerations when attempting to create a protocol suitable for large-scale data storage, synchronization, and use in a distributed system.
-We identified six requirements for enabling user-siloed data: flexible data formats, efficient synchronization, conflict resolution, access-control, scalable storage, and network communication.
+We identified six requirements for enabling \emph{user-siloed} data: flexible data formats, efficient synchronization, conflict resolution, access-control, scalable storage, and network communication.
 We presented a novel solution to satisfy each of these six requirements.
@@
 The nested, or multi-layered, encryption combined with powerful ACL capabilities create new opportunities to build distributed services, or Bots, on a network of Textile peers.
-Based on the Follow key now available in Threads, Bots would be able to relay, replicate, or store data that is synchronized via real-time updates in a trust-less way [ref].
+Based on the Follow key now available in Threads, Bots would be able to relay, replicate, or store data that is synchronized via real-time updates in a \emph{trust-less} way.
 Bots can additionally enhance the IPFS network by providing a framework to build and deploy many kinds of trust-less or trusted services.
@@
 The use of a single Read and Follow key for an entire Log means that, should either of these keys be leaked via malicious (or other/accidental) means, there is no way to prevent a Peer with the leaked keys from listening to Events or traversing the Log history.
-Potential solutions currently being explored by Textile developers include key rotation at specific Event offsets [ref], and/or incorporating the Double Ratchet Algorithm [ref] for forward secrecy [ref].
+Potential solutions currently being explored by Textile developers include key rotation at specific Event offsets \cite{KeyRotation}, and/or incorporating the Double Ratchet Algorithm \cite{marlinspikeDoubleRatchetAlgorithm2016} for forward secrecy \cite{ungerSoKSecureMessaging2015}.
 
@@
 
-\section{Conclusions} \label{Conclusions}
-In this paper we have considered auctions in which bidders make
-offers that are sequentially increasing in value by a unit price
-in order to minimally surpass the previous highest bid, and
-modelled them as discrete state-space random processes in
-continuous time. Analytical solutions are obtained and measures
-that are of interest to the SB are derived.
-
-The measures that can be computed in this way include the SB's
-probability of winning the auction, its expected savings with
-respect to the maximum sum it is willing to pay, and the average
-time that the SB spends before it can make a purchase. An
-extension of the model that incorporates price-dependent
-behaviours of the agents has also been presented.
-
-The model allows us to quantitatively characterise intuitive and
-useful trade-offs between improving the SB's chances of buying a
-good quickly, and the price that it has to pay, in the presence of
-different levels of competition from the other bidders.
-
-There are interesting extensions and applications of these models
-that can be considered, such as the behaviour of bidders and
-sellers that may have time constraints for making a purchase, and
-the possibility of the SB's moving among different auctions so as
-to optimise measures which represent its self-interest. Another
-interesting area of study may be to examine bidders who are
-``rich'' and are willing to drive away rivals at any cost, and who
-may create different auction environments for bidders that have
-significantly different levels of wealth. Yet another area of
-interest concerns auctions where items are sold in batches of
-varying sizes, with prices which depend on the number of items
-that are being bought.
-
-\ack{This research was undertaken as part of the ALADDIN
-(Autonomous Learning Agents for Decentralised Data and Information
-Networks) project and is jointly funded by a BAE Systems and EPSRC
-(Engineering and Physical Research Council) strategic partnership
-(EP/C548051/1).}
+\ack{This was all Textile baby!}
 

### 2019-09-20 draft.tex

@@
 Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects \cite{shireyInternetSecurityGlossary2007}.
-An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object.
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
+An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object (see Figure /ref{tbl:ACL}).
+
+\begin{table*}
 \caption{Example Access Control List.}
-  \label{fig:ACL}
-\end{figure}
+    \centering
+    \begin{tabular} [t]{l>{\centering}p{0.2\linewidth}>{\centering}p{0.2\linewidth}>{\centering}p{0.2\linewidth}>{\centering\arraybackslash}p{0.2\linewidth}}
+\toprule
+&Create&Delete&Edit&Read\\
+\midrule
+John&\checkmark&\checkmark&\checkmark&\checkmark\\
+Jane&-&-&-&\checkmark\\
+Mary&-&-&\checkmark&\checkmark\\
+\bottomrule
+    \end{tabular}
+    \label{tbl:ACL}
+\end{table*}%
 
@@
 
+% [!b] indicates that it will try to be on the bottom of the page
+\begin{figure}[!b]
+  \includegraphics[width=\linewidth]{figures/Event_Log.png}
+  \caption{A single-writer Merkle-Clock (Event Log).}
+  \label{fig:EventLog}
+\end{figure}
+
 In multi-writer systems, determining causal order in all replicas is particularity challenging.
@@
 Our solution to deal with divergent Merkle-Clocks is to institute a \emph{single-writer rule}: A log can only be updated by a single replica or \emph{identity}.
-An Event Log is a \emph{single-writer Merkle-Clock} that can be totally ordered.
+An Event Log is a \emph{single-writer Merkle-Clock} that can be totally ordered (Figure \ref{fig:EventLog}).
 \emph{Separate} Event Logs can be composed into advanced structures, including CRDTs \cite{enesSingleWriterPrincipleCRDT2017}.
 
-Practically speaking, for any given Log, Events are authored by a single IPFS Peer, or \emph{Writer}.
+For any given Log, Events are authored by a single IPFS Peer, or \emph{Writer}.
 
@@
 
-As seen above in section \ref{sec:merkleclocks}, a Merkle-Clock is simply a Merkle-DAG of \emph{Events}:
+As seen above in Section \ref{sec:merkleclocks}, a Merkle-Clock is simply a Merkle-DAG of \emph{Events}:
 
@@
 
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
-  \caption{A single-writer Merkle-Clock (Event Log).}
-  \label{fig:EventLog}
-\end{figure}
-
-Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure \ref{fig:EventLog}) enabling log verification by readers (see Section  \ref{sec:KeysEncryption}).
-
-At a minimum, a node must link to its most immediate ancestor.
-However, links to older ancestors are often included as well to improve concurrency during traversal and verification \cite{meyerBamboo2019}.
-
-MISSING: Ref to appendix
-
-As seen above, an Event's actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
-
-\begin{example*}
+% blocks the EventLog image from being rendered past here
+\FloatBarrier
+
+\begin{example*}[!b]
 \centering
@@
 
-Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses \cite{Multiaddr}.
-Here we introduce \emph{IPEL}, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (see Example \ref{ex:Multiaddress}).
+Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure X.X.) enabling log verification by readers (see Section  \ref{sec:KeysEncryption}).
+
+At a minimum, a node must link to its most immediate ancestor. However, links to older ancestors are often included as well to improve concurrency during traversal and verification [ref].
+
+TODO: Ref to appendix
+
+As seen above, an Event's actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
+
+Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses [ref].
+Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (see Example \ref{ex:Multiaddress}).
 
@@
 
-Followers are represented as additional addresses, meaning that a Log address book may contain \emph{multiple} multiaddresses for a single Log .
-
-MISSING: ref to embed
+Followers are represented as additional addresses, meaning that a Log address book may contain \emph{multiple} multiaddresses for a single Log (Example {ex:Multiaddress}).
 
@@
 Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
-This allows them to \emph{pull} (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable.
+This allows them to \emph{pull} (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable (Figure \ref{fig:Pulling}).
 
@@
 
-A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pub/Sub-based synchronization. TIDs are defined with the following format:
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{boat.jpg}
-  \caption{The thread id formula.}
-  \label{fig:ThreadId}
-\end{figure}
+A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pub/Sub-based synchronization.
+TIDs are defined with the following format {see Figure \ref{fig:ThreadID}
+
+% [!b] bottom page
+\begin{figure*}[!b]
+\centering
+\caption{Components of a Thread Identity.}
+\par\vspace{.9\baselineskip}\leavevmode%
+\par{Thread ID = 0x62\quad}\ubar{0x01}{Version}\quad0x55\quad\ubar{0x539bc1dc03ee8cb5d478e41cc8a4546e}{Random Number} 
+\par\vspace{-.6\baselineskip}\leavevmode%
+\phantom{Thread ID = }\phubar{0x62}{Multibase}%
+  \phantom{\quad0x01\quad}\phubar{0x55}{Variant} \phantom{\quad7513b4dcf4efa291ffa92221659f19e2}
+\par\vspace{-.6\baselineskip}\leavevmode
+\label{fig:ThreadID}
+\end{figure*}
 
@@
 Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition \ref{ex:SequenceID}) basis.
-Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see section \ref{sec:DefaultViews} for details).
+Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see Section \ref{sec:TexCRDT}).
 ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible).

### 2019-09-20 draft.tex
cleanup: should be gtg in vscode now
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-\usepackage{pagecolor}
 %\copyrightyear{2009} \vol{00} \issue{0} \DOI{000}
@@
 
-% To create tables + create checkmarks
+% to make paragraph look a bit more like a subsubsubsection
+\makeatletter
+\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
+            {-2.5ex\@plus -1ex \@minus -.25ex}%
+            {1.25ex \@plus .25ex}%
+            {\normalfont\normalsize\bfseries}}
+\makeatother
+\setcounter{secnumdepth}{3} % how many sectioning levels to assign numbers to
+\setcounter{tocdepth}{3}    % how many sectioning levels to show in ToC
+
+% To create tables + create check-marks
 \usepackage{booktabs}
@@
 
-% Thread ID components fig requrements
+% Thread ID components fig requirements
 \usepackage{stackengine}
@@
 
+\usepackage{pagecolor}
+
 \begin{document}
@@
 There are a number of systems that utilize content addressing to access content and information [LIST SOME], and there is an active body of literature covering its design and implementation [MISSING].
-The Interplanetary File System (IPFS) --- which is a set of protocols to create a content-addressed, peer-to-peer filesystem \cite{benetIpfscontentAddressedVersioned2014} --- is one such system.
+The Interplanetary File System (IPFS) --- which is a set of protocols to create a content-addressed, peer-to-peer filesystem \cite{benetIPFSContentAddressed2014} --- is one such system.
 In IPFS, the address for any content is determined based on a cryptographic hash of the content itself.
@@
 
-As discussed above, IPFS uses the cryptographic hash of a given piece of content to define its content address (see \cite{benetIpfscontentAddressedVersioned2014} for details on this process).
+As discussed above, IPFS uses the cryptographic hash of a given piece of content to define its content address (see \cite{benetIPFSContentAddressed2014} for details on this process).
 However, in order to provide standards for accessing content-addressable data (on the web or elsewhere), it is necessary to define a common format or specification.
 In IPFS (and others \cite[e.g.,]{protocollabsFilecoinDecentralizedStorage2017} MORE NEEDED), this common data format is called Interplanetary Linked Data (IPLD)\footnote{https://ipld.io/}.
-As the name suggests, IPLD is based on principals of linked data \cite{timberners-leeLinkedData2009,bizerLinkedDataStory2011} with the added capabilities of a content-addressing storage network.
-
-IPLD is used to publish linked data (subject-predicate-object triples in the linked data world \cite{heathLinkedDataEvolving2011}) spread across different hosts, and where everything (entities, predicates, data sources) uses content addresses as unique identifiers.
+As the name suggests, IPLD is based on principals of linked data \cite{berners-leeLinkedData2009,bizerLinkedDataStory2011} with the added capabilities of a content-addressing storage network.
+
+IPLD is used to publish linked data (subject, predicate, object triples in the linked data world \cite{heathLinkedDataEvolving2011}) spread across different hosts, and where everything (e.g., entities, predicates, data sources) uses content addresses as unique identifiers.
 To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph\footnote{Other examples of Dags include the Bitcoin Blockchain or a Git version history.}.
@@
 \begin{enumerate}
-\item \label{Perf1}Broadcasting the Merkle-Clock requires broadcasting only the current root CID.
+\item Broadcasting the Merkle-Clock requires broadcasting only the current root CID.
 The whole Clock is unambiguously identified by the CID of its root and its full DAG can be walked down from it as needed.
-\item \label{Perf2}The immutable nature of a Merkle-DAG allows every replica to perform quick comparisons and fetch only those nodes that it does not already have.
-\item \label{Perf3} Merkle-DAG nodes are self-verified and immune to corruption and tampering. They can be fetched from any source willing to provide them, trusted or not.
-\item \label{Perf4} Identical nodes are de-duplicated by design: there can only be one unique representation for every event.
+\item The immutable nature of a Merkle-DAG allows every replica to perform quick comparisons and fetch only those nodes that it does not already have.
+\item Merkle-DAG nodes are self-verified and immune to corruption and tampering. They can be fetched from any source willing to provide them, trusted or not.
+\item Identical nodes are de-duplicated by design: there can only be one unique representation for every event.
 \end{enumerate}
@@
 
-Agent-centric security refers to the practice of maintaining data integrity without leveraging a central or blockchain-based consensus.
+Agent-centric security refers to the maintainance of data integrity without leveraging a central or blockchain-based consensus.
 The general approach is just to let the reader enforce permissions and perform validations, not the writer or some central authority.
-Agent-centric security is possible if the reader can reference local-only, tamper-free code or if the state can be used to determine whether a given operation (e.g.
-delete data) is permitted.
+Agent-centric security is possible if the reader can reference local-only, tamper-free code or if the state can be used to determine whether a given operation (e.g. delete data) is permitted.
 Many decentralized networks like Secure Scuttlebutt \cite{ScuttlebuttProtocolGuide} and Holochain \cite{ericharris-braunHolochainScalableAgentcentric2018} make use of agent-centric security.
@@
 Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table \cite{benetIPFSContentAddressed2014} (DHT).
-Instead, they are collected from Log Events (see Section  \ref{sec:UseCases}).
+Instead, they are collected from Log Events.
 This is in contrast to mutable data via IPNS for example, which requires querying the network (DHT) for updates.
@@
 \begin{enumerate}
-\item \label{Perf1}When broadcasting events to many possible recipients, a single-use symmetric key is generated per unique content body.
-\item \label{Perf2}When sending events to specific recipients, the recipient's public key can be used to restrict access from all others\footnote{much like private messages in Secure Scuttlebutt (https://www.scuttlebutt.nz/concepts/private-message)}.
+\item When broadcasting events to many possible recipients, a single-use symmetric key is generated per unique content body.
+\item When sending events to specific recipients, the recipient's public key can be used to restrict access from all others\footnote{Much like private messages in Secure Scuttlebutt (\url{https://www.scuttlebutt.nz/concepts/private-message})}.
 \end{enumerate}
@@
 \begin{definition}
-(Raw). This variant declares that consumers are not expected to make additional assumptions. This is the default variant.\\
-
-See Example \ref{ex:Identity}a).
+(Raw). This variant declares that consumers are not expected to make additional assumptions. This is the default variant (See Example \ref{ex:Identity}a).
 \end{definition}
@@
 (Access-Controlled): This variant declares that consumers should assume an access control list is composable from Log Events.
-The ACL represents a permissions rule set that must be applied when reading data (see Section  \ref{sec:AccessControl}).\\
-
-See Example \ref{ex:Identity}b).
+The ACL represents a permissions rule set that must be applied when reading data (see Section  \ref{sec:AccessControl} and Example \ref{ex:Identity}b).
 \end{definition}
@@
 Once a Peer has accepted an Invite, it will receive new Events from Log Writers.
-In cases where the invitee becomes a collaborator, i.e., a Writer, it is also responsible for sending its own Events out to the network.\\
-
-\textbf{Sending} \\
+In cases where the invitee becomes a collaborator --- i.e., a Writer --- it is also responsible for sending its own Events out to the network.
+
+\paragraph{Sending}
 
@@
 \begin{enumerate}
-\item \label{Perf1}New Events are pushed\footnote{Here push means ``send to multiaddress(es)'', which may designate different protocols, e.g., p2p, HTTP, Bluetooth, etc.} directly to the Thread's other Log Writers.
-\item \label{Perf2}New Events are pushed directly to the target Log's Follower(s), who may not maintain their own Log.
-\item \label{Perf3} New Events are published over Libp2p's gossip-based Pub/Sub infrastructure using TID as a topic, which provides potentially unknown Readers or Followers with an opportunity to consume Events in real-time.
+\item New Events are pushed\footnote{Here push means ``send to multiaddress(es)'', which may designate different protocols, e.g., p2p, HTTP, Bluetooth, etc.} directly to the Thread's other Log Writers.
+\item New Events are pushed directly to the target Log's Follower(s), who may not maintain their own Log.
+\item New Events are published over Libp2p's gossip-based Pub/Sub infrastructure using TID as a topic, which provides potentially unknown Readers or Followers with an opportunity to consume Events in real-time.
 \end{enumerate}
@@
 \begin{enumerate}
-\item \label{Perf1}New Events may be pushed to web-based participants over a WebSocket.
-\item \label{Perf2}New Events may be pushed to the Thread's other Log Writers via federated notification services like Apple Push Notification Service (APNS), Google Cloud Messaging (GCM), Firebase Cloud Messaging (FCM), and/or Windows Notification Service (WNS).
-\item \label{Perf3} New Events may trigger web-hooks\footnote{https://en.wikipedia.org/wiki/Webhook} enabling complex workflows (see Section X below).\\
+\item New Events may be pushed directly to web-based participants over a WebSocket.
+\item New Events may be pushed to the Thread's other Log Writers via federated notification services like Apple Push Notification Service (APNS), Google Cloud Messaging (GCM), Firebase Cloud Messaging (FCM), and/or Windows Notification Service (WNS).
+\item New Events may trigger web-hooks\footnote{https://en.wikipedia.org/wiki/Webhook}, which could enable many complex (e.g., IFTTT\footnote{\url{https://ifttt.com}}) workflows.
 \end{enumerate}
 
-\textbf{Receiving} \\
-
-Similar to sending, there are multiple paths to receiving new Events that together maximize connectivity between Peers who are often offline or unreachable.
+\paragraph{Receiving}
+
+There are multiple paths to receiving new Events, that together, maximize connectivity between Peers who are often offline or unreachable.
 
 \begin{enumerate}
-\item \label{Perf1}Log Writers can receive Events directly from the author.
-\item \label{Perf2} Events can be pulled from replicating Followers via HTTP, libp2p, RSS, Atom, etc.
+\item Log Writers can receive Events directly from the author.
+\item Events can be pulled from replicating Followers via HTTP, libp2p, RSS, Atom, etc.
     a. In conjunction with push over WebSockets (2a above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events (Figure \ref{fig:Pulling}).
-\item \label{Perf3} Writers and readers can receive new Events via a Pub/Sub subscription at the TID.
+\item Writers and readers can receive new Events via a Pub/Sub subscription at the TID.
 \end{enumerate}
@@
 These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
-In the remainder of this section, we highlight a number of planned database abstractions to provide further utility to developers interested in building on Textile's Threads protocol.\\
-
-\textbf{Databases} \\
+In the remainder of this section, we highlight a number of planned database abstractions to provide further utility to developers interested in building on Textile's Threads protocol.
+
+\paragraph{Databases}
 
@@
 \caption{The Document store interface.}
-\end{example} \label{ex:AddrBook}
+\end{example} \label{ex:DocStore}
 
@@
 Each database style will be implemented as a standalone software library, allowing application developers to pick and choose the solution most useful to the application at hand.
-Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these ``example'' libraries.\\
-
-\textbf{CRDTs} \label{sec:TexCRDT} \\
+Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these ``example'' libraries.
+
+\paragraph{CRDTs} \label{sec:TexCRDT}
 
@@
 Libraries such as Automerge\footnote{https://github.com/automerge/automerge} provide useful examples of reducer function that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available.
-A practical example of using a JSON CRDT in Textile is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.\\
-
-\textbf{Application State} \\
+A practical example of using a JSON CRDT in Textile is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.
+
+\paragraph{Application State}
 

### 2019-09-20 draft.tex
fix: all urls are properly formatted now
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 In IPFS, the address for any content is determined based on a cryptographic hash of the content itself.
-In practice, the IPFS Content IDentifier (CID) is a multihash, which is a self-describing ``protocol for differentiating outputs from various well-established cryptographic hash functions, addressing size + encoding considerations\footnote{https://multiformats.io/}''.
+In practice, the IPFS Content IDentifier (CID) is a multihash, which is a self-describing ``protocol for differentiating outputs from various well-established cryptographic hash functions, addressing size + encoding considerations\footnote{\url{https://multiformats.io/}}''.
 That addressing system confers several benefits to the network, including tamper resistance (we can be confident that a given piece of content has not been modified en route if its hash matches what we were expected/requested) and de-duplication (because the same content from different peers will produce the same hash address).
@@
 
-While content addressing doesn't tell you how to get a file, IPFS (via libp2p\footnote{https://libp2p.io/}) provides a system for moving files across the network.
+While content addressing doesn't tell you how to get a file, IPFS (via libp2p\footnote{\url{https://libp2p.io/}}) provides a system for moving files across the network.
 On the IPFS network, a client who wants specific content requests the CID from the network of IPFS hosts.
@@
 However, in order to provide standards for accessing content-addressable data (on the web or elsewhere), it is necessary to define a common format or specification.
-In IPFS (and others \cite[e.g.,]{protocollabsFilecoinDecentralizedStorage2017} MORE NEEDED), this common data format is called Interplanetary Linked Data (IPLD)\footnote{https://ipld.io/}.
+In IPFS (and others \cite[e.g.,]{protocollabsFilecoinDecentralizedStorage2017} MORE NEEDED), this common data format is called Interplanetary Linked Data (IPLD)\footnote{\url{https://ipld.io/}}.
 As the name suggests, IPLD is based on principals of linked data \cite{berners-leeLinkedData2009,bizerLinkedDataStory2011} with the added capabilities of a content-addressing storage network.
@@
 
-On the downside, and as discussed above (section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots) (see Figure \ref{fig:merkledag})
-
-In the figure above, two replicas (top row, bottom row) are attempting to write (left to right) events to the same Merkle-Clock.
+On the downside, and as discussed above (section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots).
+For example, in Figure \ref{fig:merkledag}, two replicas (top row, bottom row) are attempting to write (left to right) events to the same Merkle-Clock.
 After the first replica writes event A, the second writes event A' and properly links to A.
@@
 
-TIDs share some similarities with UUIDs \cite{leachUniversallyUniqueIDentifier2005} (version and variant) and IPFS-based CIDs and are multibase encoded\footnote{https://github.com/multiformats/multibase} for maximum forward-compatibility.
+TIDs share some similarities with UUIDs \cite{leachUniversallyUniqueIDentifier2005} (version and variant) and IPFS-based CIDs and are multibase encoded\footnote{\url{https://github.com/multiformats/multibase}} for maximum forward-compatibility.
 
@@
 Log Writers, Readers, and Followers synchronize the state of their Logs by sending and receiving Events.
-Inspired by Git\footnote{https://git-scm.com/}, a reference to the latest Event in a Log is referred to as the \emph{Head} (or sometimes the \emph{root}).
+Inspired by Git\footnote{\url{https://git-scm.com/}}, a reference to the latest Event in a Log is referred to as the \emph{Head} (or sometimes the \emph{root}).
 When a new Event is received, Readers and Followers simply advance their Head reference.
@@
 \item New Events may be pushed to the Thread's other Log Writers via federated notification services like Apple Push Notification Service (APNS), Google Cloud Messaging (GCM), Firebase Cloud Messaging (FCM), and/or Windows Notification Service (WNS).
-\item New Events may trigger web-hooks\footnote{https://en.wikipedia.org/wiki/Webhook}, which could enable many complex (e.g., IFTTT\footnote{\url{https://ifttt.com}}) workflows.
+\item New Events may trigger web-hooks\footnote{\url{https://en.wikipedia.org/wiki/Webhook}}, which could enable many complex (e.g., IFTTT\footnote{\url{https://ifttt.com}}) workflows.
 \end{enumerate}
@@
 
-Where \texttt{Indexable} can be satisfied by any structure with a \texttt{key} field and and \texttt{Query} is taken from the \texttt{go-datastore} interface library\footnote{https://github.com/ipfs/go-datastore} or similar.
+Where \texttt{Indexable} can be satisfied by any structure with a \texttt{key} field and and \texttt{Query} is taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
 Tables, feeds, counters, and other simple stores can also be built on Threads, and may require ORM-based solutions to support persistence to certain backing stores\footnote{In-memory stores are almost always implementable using compositions of basic data structures.}.
@@
 For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype \cite{kleppmannConflictFreeReplicatedJSON2017} that applies updates on a JSON document View model.
-Libraries such as Automerge\footnote{https://github.com/automerge/automerge} provide useful examples of reducer function that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available.
+Libraries such as Automerge\footnote{\url{https://github.com/automerge/automerge}} provide useful examples of reducer function that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available.
 A practical example of using a JSON CRDT in Textile is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.

### 2019-09-20 draft.tex
one more
Signed-off-by: andrew <andrew@textile.io>
Some additional cleanup and prep

@@
 % !TEX encoding = UTF-8 Unicode
-\documentclass{comjnl}
+\documentclass{textile}
 
@@
             {1.25ex \@plus .25ex}%
-            {\normalfont\normalsize\bfseries}}
+            {\normalfont\normalsize\itshape}}
 \makeatother
@@
 Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects \cite{shireyInternetSecurityGlossary2007}.
-An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object (see Figure /ref{tbl:ACL}).
-
-\begin{table*}
+An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object (see Figure \ref{tbl:ACL}).
+
+\begin{table}
 \caption{Example Access Control List.}
     \centering
-    \begin{tabular} [t]{l>{\centering}p{0.2\linewidth}>{\centering}p{0.2\linewidth}>{\centering}p{0.2\linewidth}>{\centering\arraybackslash}p{0.2\linewidth}}
+    \begin{tabular} [t]{l>{\centering}p{0.15\linewidth}>{\centering}p{0.15\linewidth}>{\centering}p{0.15\linewidth}>{\centering\arraybackslash}p{0.15\linewidth}}
 \toprule
@@
     \label{tbl:ACL}
-\end{table*}%
+\end{table}
 
@@
 
-\ack{This was all Textile baby!}
+\ack{This was all Lucas baby! Thanks Lucas}
 

### 2019-09-20 draft.tex
updates: more updates
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-Cryptographically linked events can also represent a clock.
+Cryptographically linked events can also represent a clock (See Section \ref{sec:merkleclocks}).
 One such example is called the Merkle-Clock \cite{sanjuanMerkleCRDTs2019}.
@@
 \end{quote}
-
-See Section \ref{sec:merkleclocks}.
 
@@
 
-Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses [ref].
-Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (see Example \ref{ex:Multiaddress}).
-
-In practice, to reach a Log via it's IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (see Example \ref{ex:Multiaddress}).
+Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses\cite{Multiaddr}.
+Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses.
+To reach a Log via it's IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (see Example \ref{ex:Multiaddress}).
 
@@
 Log addresses are recorded in an address book, similar to IPFS Peer address book:
-
-MISSING PEER ADDRESS BOOK?
 

### 2019-09-22 draft.tex
mods on mods
Signed-off-by: andrew <andrew@textile.io>

--- 2019-09-20 draft.tex
@@
 \usepackage{dblfloatfix}    % To enable figures at the bottom of page
+
+% keep footnotes from wrapping pages,
+\interfootnotelinepenalty=10000
 
@@
 
-\title[An event-sourced and distributed database with IPLD and materialized views]{An event-sourced and distributed database with IPLD and materialized views}
+\title[An event-sourced and decentralized database with IPLD and materialized views]{An event-sourced and decentralized database with IPLD and materialized views}
 \author{Pick}
@@
 An alternative, called user-siloed data is one area of research that if implemented successfully, could reverse the flow of value derived from personal data, from apps to users.
-In this paper, we investigate the data formats, and storage and transfer protocols necessary to build a system for user ownership of large-scale digital datasets.
+In this paper, we investigate the data formats, access control, and transfer protocols necessary to build a system for user ownership of large-scale digital datasets.
 The aim of the proposed system is to help power a new generation of web technologies.
@@
 
-We now describe some of the technologies and concepts motivating the design of our novel distributed database system.
-We highlight some of the advantages and lessons that can be learned from event-sourcing and discuss drawbacks to using these approaches in distributed systems.
+We now describe some of the technologies and concepts motivating the design of our novel decentralized database system.
+We highlight some of the advantages and lessons that can be learned from event-sourcing and discuss drawbacks to using these approaches in decentralized systems.
 We provide an overview of some important technologies related to IPFS that make it possible to rethink event sourcing in a decentralized network.
@@
 \midrule
+Jane&-&-&-&\checkmark\\
 John&\checkmark&\checkmark&\checkmark&\checkmark\\
-Jane&-&-&-&\checkmark\\
 Mary&-&-&\checkmark&\checkmark\\
@@
 \begin{definition}
-(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log's Read key (see Section  \ref{sec:KeysEncryption}).
+(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log's Read Key (see Section  \ref{sec:KeysEncryption}).
 \end{definition}
@@
 
-Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure X.X.) enabling log verification by readers (see Section  \ref{sec:KeysEncryption}).
+Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure \ref{fig:EventLog}) enabling log verification by readers (see Section  \ref{sec:KeysEncryption}).
 
@@
 
-TODO: Ref to appendix
-
-As seen above, an Event's actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
+As shown \ref{appx:EventNode}, an Event's actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
 
 Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses [ref].
-Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (see Example \ref{ex:Multiaddress}).
-
-In practice, to reach a Log via it's IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (see Example \ref{ex:Multiaddress}).
-
-\begin{example}
+Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (Example \ref{ex:Multiaddress}).
+
+In practice, to reach a Log via it's IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (Example \ref{ex:Multiaddress}).
+
+Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table \cite{benetIPFSContentAddressed2014} (DHT).
+Instead, they are collected from Log Events.
+This is in contrast to mutable data via IPNS for example, which requires querying the network (DHT) for updates.
+Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed\footnote{This is similar to the append-only message feeds used in Secure Scuttlebutt’s global gossip network \cite{ScuttlebuttProtocolGuide}} like topology.
+Log addresses are recorded in an address book, similar to IPFS Peer address book (Example \ref{ex:KeyBook}).
+
+Addresses can expire by specifying a time-to-live (TTL) value when adding or updating them in the address book, which allows for unresponsive addresses to eventually be removed.
+
+Modern, real-world networks consist of many mobile or otherwise sparsely connected computers (Peers).
+Therefore, datasets distributed across such networks can be thought of as highly partitioned.
+To ensure updates are available between mostly offline or otherwise disconnected Peers (like mobile devices), Textile Logs are designed with a built-in replication or follower mechanism.
+
+\begin{definition} (Follower).
+Log Writers can designate other IPFS Peers to ``follow'' a Log, potentially replicating and/or republishing Events.
+A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (see Section  \ref{sec:KeysEncryption}), but is not able to read the Log's contents.
+Followers should be server-based --- i.e., always online and behind a public IP address.\end{definition}
+
+Followers are represented as additional addresses, meaning that a Log address book may contain \emph{multiple} multiaddresses for a single Log (Example \ref{ex:Multiaddress}).
+
+In practice, Writers are solely responsible for announcing their Log's addresses.
+This ensures a conflict-free address list without additional complexity.
+Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
+This allows them to \emph{pull} (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable (Figure \ref{fig:Pulling}).
+
+\subsubsection{Keys \& Encryption} \label{sec:KeysEncryption}
+
+\begin{figure*}[!t]
+\centering
+\begin{minipage}{0.6\textwidth}
+  \includegraphics[width=\linewidth]{figures/Event_Log_With_Encryption.png}
+  \caption{The three layers of Log Event encryption.}
+  \label{fig:LogEncryption}
+  \end{minipage}
+\end{figure*}
+
+\begin{example}[!b]
 \begin{lstlisting}
@@
 }
-\end{lstlisting}
-\caption{AddrBook storing log addresses.}
-\end{example} \label{ex:AddrBook}
-
-Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table \cite{benetIPFSContentAddressed2014} (DHT).
-Instead, they are collected from Log Events.
-This is in contrast to mutable data via IPNS for example, which requires querying the network (DHT) for updates.
-Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed\footnote{This is similar to the append-only message feeds used in Secure Scuttlebutt’s global gossip network \cite{ScuttlebuttProtocolGuide}} like topology.
-Log addresses are recorded in an address book, similar to IPFS Peer address book:
-
-MISSING PEER ADDRESS BOOK?
-
-Addresses can expire by specifying a time-to-live (TTL) value when adding or updating them in the address book, which allows for unresponsive addresses to eventually be removed.
-
-Modern, real-world networks consist of many mobile or otherwise sparsely connected computers (Peers).
-Therefore, datasets distributed across such networks can be thought of as highly partitioned.
-To ensure updates are available between mostly offline or otherwise disconnected Peers (like mobile devices), Textile Logs are designed with a built-in replication or follower mechanism.
-
-\begin{definition} (Follower).
-Log Writers can designate other IPFS Peers to ``follow'' a Log, potentially replicating and/or republishing Events.
-A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (see Section  \ref{sec:KeysEncryption}), but is not able to read the Log's contents.
-Followers should be server-based --- i.e., always online and behind a public IP address.\end{definition}
-
-Followers are represented as additional addresses, meaning that a Log address book may contain \emph{multiple} multiaddresses for a single Log (Example {ex:Multiaddress}).
-
-In practice, Writers are solely responsible for announcing their Log's addresses.
-This ensures a conflict-free address list without additional complexity.
-Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
-This allows them to \emph{pull} (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable (Figure \ref{fig:Pulling}).
-
-\subsubsection{Keys \& Encryption} \label{sec:KeysEncryption}
-
-\begin{figure*}[!t]
-\centering
-\begin{minipage}{0.6\textwidth}
-  \includegraphics[width=\linewidth]{figures/Event_Log_With_Encryption.png}
-  \caption{The three layers of Log Event encryption.}
-  \label{fig:LogEncryption}
-  \end{minipage}
-\end{figure*}
-
-\begin{example}
-\begin{lstlisting}
 type KeyBook interface {
@@
 \end{lstlisting}
-\caption{A KeyBook storing log keys.}
-\end{example} \label{ex:KeyBook}
+\caption{The AddrBook interface for storing log addresses and the KeyBook interface for storing log keys.}
+\label{ex:KeyBook}
+\end{example}
 
@@
 The body, or content of an event, is encrypted by a \emph{Content Key}.
-Content keys are generated per-content and never reused.
-The Content key is distributed directly in the header of the Event Block.
+Content Keys are generated per-content and never reused.
+The Content Key is distributed directly in the header of the Event Block.
 We define the Content Key as follows,
@@
 (Content Key).
-The Content key is a variable-format key used to encrypt the body (content) of an event.
+The Content Key is a variable-format key used to encrypt the body (content) of an event.
 This key can be symmetric, asymmetric, or possibly non-existent in cases where encryption is not needed.
@@
 
-If a single-use symmetric key is used for the Content key, it is necessary to distribute each new key to users by including it in the header of the Event Block.
+If a single-use symmetric key is used for the Content Key, it is necessary to distribute each new key to users by including it in the header of the Event Block.
 Therefore, the Event Block itself is further encrypted using a \emph{Read} key.
-The Read key is not distributed within the Log itself but is distributed to any peers to grant them access to the content of the Log.
-
-\begin{definition}
- (Read Key). The Read key is a symmetric key created by the Log owner and used to encrypt the Content Key in each event.
-\end{definition}
-
-Finally, the encrypted Event Block, its signature, and the IPLD linkage(s) from an Event to its antecedents are encrypted together using a Follow key.
-Follow keys allow Logs to be \emph{followed} by peers on the network who do not have access to any content within the event, and can only see signatures and linkage(s) between Events.
-
-\begin{definition}
-(Follow Key). The Follow key is a symmetric key created by the Log owner and used to encrypt the entire event payload before adding the event to the Log.
-\end{definition}
-
-Much like the Log address book, Log keys are stored in a key book (see Example \ref{ex:KeyBook}).
+The Read Key is not distributed within the Log itself but is distributed to any peers to grant them access to the content of the Log.
+
+\begin{definition}
+ (Read Key). The Read Key is a symmetric key created by the Log owner and used to encrypt the Content Key in each event.
+\end{definition}
+
+Finally, the encrypted Event Block, its signature, and the IPLD linkage(s) from an Event to its antecedents are encrypted together using a Follow Key.
+Follow Keys allow Logs to be \emph{followed} by peers on the network who do not have access to any content within the event, and can only see signatures and linkage(s) between Events.
+
+\begin{definition}
+(Follow Key). The Follow Key is a symmetric key created by the Log owner and used to encrypt the entire event payload before adding the event to the Log.
+\end{definition}
+
+Much like the Log address book, Log keys are stored in a key book (Example \ref{ex:KeyBook}).
 
@@
 A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pub/Sub-based synchronization.
-TIDs are defined with the following format {see Figure \ref{fig:ThreadID}
+TIDs are defined with the format shown Figure \ref{fig:ThreadID}.
 
 % [!b] bottom page
-\begin{figure*}[!b]
+\begin{figure*}[!t]
 \centering
@@
 A new Thread is created by generating a TID and Log.
-The Log's creator is the Writer, meaning it has possession of the Log's Identity, Read, and Follow keys.
+The Log's creator is the Writer, meaning it has possession of the Log's Identity, Read, and Follow Keys.
 All of these keys are needed to compose Events.
@@
 \begin{definition}
-(Key Set). A set of keys for a Log. Depending on the context, a Key Set may contain the Follow and Read key, or just the Follow key. Encrypted with the recipient's public key.
+(Key Set). A set of keys for a Log. Depending on the context, a Key Set may contain the Follow and Read Key, or just the Follow Key. Encrypted with the recipient's public key.
 \end{definition}
@@
 Once a Peer has accepted an Invite, it will receive new Events from Log Writers.
-In cases where the invitee becomes a collaborator --- i.e., a Writer --- it is also responsible for sending its own Events out to the network.
+In cases where the invitee becomes a collaborator (i.e., a Writer) it is also responsible for sending its own Events out to the network.
 
@@
 \item Events can be pulled from replicating Followers via HTTP, libp2p, RSS, Atom, etc.
-    a. In conjunction with push over WebSockets (2a above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events (Figure \ref{fig:Pulling}).
+    a.
+In conjunction with push over WebSockets (seen in Step 2 of the additional push mechanisms above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events (Figure \ref{fig:Pulling}).
 \item Writers and readers can receive new Events via a Pub/Sub subscription at the TID.
@@
 
-The notion of the Follow key (see Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial.
+The notion of the Follow Key (see Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial.
 This allows any Peer in the network to be granted the responsibility of replicating data from another Peer without having read access to the data contained within the Log entries.
@@
 These Events would then operate on a map-like view model, via an internal ``projection'' (reducer function).
-The key/value view could be persisted or kept in-memory, and would be directly queried via a corresponding \texttt{Get} operation.
-Because Textile uses ES patterns, a \texttt{Del} operation would simply lead to an internal \texttt{KeyValueDeleted} event, marking the key/value pair for removal from the key/value materialized view.
+The key/value view can be persisted or kept in-memory, and would be directly queried via a corresponding \texttt{Get} operation.
+Because Textile uses ES patterns, a \texttt{Del} operation would simply lead to an internal \texttt{KeyValueDeleted} event, marking the key/value pair for removal from the key/value materialized view (Example \ref{ex:KVStore}).
 
@@
 \caption{The Key-Value store interface.}
-\end{example} \label{ex:KVStore}
+ \label{ex:KVStore}
+\end{example}
 
@@
 Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities.
-Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents:
+Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents (Example \ref{ex:DocStore}).
 
@@
 \caption{The Document store interface.}
-\end{example} \label{ex:DocStore}
+\label{ex:DocStore}
+\end{example}
 
@@
 For application developers familiar with common state management issues, tools such as Redux provide an opinionated framework for making state mutations predictable by imposing certain restrictions on how and when updates can happen.
-In fact, Redux builds on concepts from CRQS and ES itself \cite{reduxMotivation}, and combines them with three core principals: 1) that the entire ``state' of an application should be stored in a single ``store'', 2) where the only way to change the state is to emit an ``action'' (or event), which 3) transforms the application state using a ``pure'' (i.e., zero side effects) reducer function (projection).
+In fact, Redux builds on concepts from CRQS and ES itself \cite{reduxMotivation}, and combines them with three core principals:
+
+\begin{enumerate}
+\item The entire \emph{state} of an application should be stored in a single \emph{store}.
+\item The only way to change the state is to emit an \emph{action} (or event).
+\item transforms the application state using a \emph{pure} (i.e., zero side effects) reducer function (projection).
+\end{enumerate}
+
  In practice, Redux doesn't care how a developer persists the application state, which means a Thread's event store and read-only materialized views are excellent candidates.
@@
 
-Below is a typical Thread-level ACL JSON document (where \texttt{\_id} is a unique id), which could be persisted to a local document store as part of a materialized view.
-
-todo: fixidexampleabove
-
-todo: change code type below
+A typical Thread-level ACL JSON (see Example \ref{ex:AclJson}) can be persisted to a local document store as part of a materialized view.
 
@@
 \end{lstlisting}
-\caption{blahblah}
+\caption{ACL JSON document with \texttt{\_id} being the unique ID. }
+ \label{ex:AclJson}
 \end{minipage}
@@
 
- A Thread-level ACL has it's own document ACL, which also applies to all other document ACLs:
+ A Thread-level ACL has it's own document ACL, which also applies to all other document ACLs (see Example \ref{ex:ThreadAcl}).
 
@@
 \end{lstlisting}
-\caption{Blahblah}
+\caption{Thread and document ACL}
+\label{ex:ThreadAcl}
 \end{example}
@@
 The nested, or multi-layered, encryption combined with powerful ACL capabilities create new opportunities to build distributed services, or Bots, on a network of Textile peers.
-Based on the Follow key now available in Threads, Bots would be able to relay, replicate, or store data that is synchronized via real-time updates in a \emph{trust-less} way.
+Based on the Follow Key now available in Threads, Bots would be able to relay, replicate, or store data that is synchronized via real-time updates in a \emph{trust-less} way.
 Bots can additionally enhance the IPFS network by providing a framework to build and deploy many kinds of trust-less or trusted services.
@@
 
-The use of a single Read and Follow key for an entire Log means that, should either of these keys be leaked via malicious (or other/accidental) means, there is no way to prevent a Peer with the leaked keys from listening to Events or traversing the Log history.
+The use of a single Read and Follow Key for an entire Log means that, should either of these keys be leaked via malicious (or other/accidental) means, there is no way to prevent a Peer with the leaked keys from listening to Events or traversing the Log history.
 Potential solutions currently being explored by Textile developers include key rotation at specific Event offsets \cite{KeyRotation}, and/or incorporating the Double Ratchet Algorithm \cite{marlinspikeDoubleRatchetAlgorithm2016} for forward secrecy \cite{ungerSoKSecureMessaging2015}.
+
+\ack{This was all Lucas baby! Thanks Lucas}
+
+\nocite{*}
+
+\bibliographystyle{compj}
+\bibliography{whitepaper}
+
+\appendix
 
@@
 \begin{example*}[!p]
-
+\section{Event node interface} \label{appx:EventNode} % stuck it inside the example so it lands on the same full page
 \begin{lstlisting}
 // Node is the most basic component of a log.
-// Note: In practice, this is encrypted with the Follow key.
+// Note: In practice, this is encrypted with the Follow Key.
 type Node interface {
@@
 // Event represents the content of an update.
-// Note: In practice, this is encrypted with the Read key.
+// Note: In practice, this is encrypted with the Read Key.
 type Event interface {
@@
 \end{lstlisting} 
-  \caption{Interfaces for an event body, header and node.}
-
 \end{example*}
 
-\ack{This was all Lucas baby! Thanks Lucas}
-
-\nocite{*}
-
-\bibliographystyle{compj}
-\bibliography{whitepaper}
-
 \end{document}

### 2019-09-22 draft.tex

@@
 
-Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses [ref].
-Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses (Example \ref{ex:Multiaddress}).
+Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses\cite{Multiaddr}.
+Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses.
+To reach a Log via it's IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (see Example \ref{ex:Multiaddress}).
 

### 2019-09-23 draft.tex
update: ready for sharing?
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

--- 2019-09-20 draft.tex
@@
 One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data.
-Other projects have identified app silo-ed data as a problem [ref], and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}.
+Other projects have identified app silo-ed data as a problem [MISSING], and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}.
 However, none so far have provided a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data.
@@
 \begin{definition} (Command query responsibility segregation).
-Command query responsibility segregation or CQRS is a design pattern whereby reads and writes are separated into different models, using commands to write data, and queries to read data [ref].
+Command query responsibility segregation or CQRS is a design pattern whereby reads and writes are separated into different models, using commands to write data, and queries to read data \cite{martinfowlerCQRS2011}.
 \end{definition}
@@
 
-At a minimum, a node must link to its most immediate ancestor. However, links to older ancestors are often included as well to improve concurrency during traversal and verification [ref].
+At a minimum, a node must link to its most immediate ancestor.
+However, links to older ancestors are often included as well to improve concurrency during traversal and verification \cite{meyerBamboo2019}.
 
@@
 The use of well-established design patterns from the CQRS+ES (and related) literature means the Threads protocol can provide the foundation for many higher-level concepts and technologies.
-For instance, the public application programming interface (API) for a Thread provides several features that you would expect when operating on a dataset or table within a database.
-Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more.
-
 Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store.
@@
 
-\subsubsection{Default Views}\label{sec:DefaultViews}
-
-The core Textile Threads interfaces are also exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
+\subsubsection{Databse Views}\label{sec:DefaultViews}
+
+The public API for a Thread already provides several features that you would expect when operating on a dataset or table within a database.
+Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more.
+As such, it is not difficult to imagine a simplified API in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
+
+The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, while still maintaining the power and flexibility of CQRS + ES.
+Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities.
 These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
-In the remainder of this section, we highlight a number of planned database abstractions to provide further utility to developers interested in building on Textile's Threads protocol.
-
-\paragraph{Databases}
-
-A key/value store built on top of Threads can map database \texttt{Put} operations into internal \texttt{KeyValueUpdated} events.
-These Events would then operate on a map-like view model, via an internal ``projection'' (reducer function).
-The key/value view could be persisted or kept in-memory, and would be directly queried via a corresponding \texttt{Get} operation.
-Because Textile uses ES patterns, a \texttt{Del} operation would simply lead to an internal \texttt{KeyValueDeleted} event, marking the key/value pair for removal from the key/value materialized view.
+
+A database built on Threads would map database operations into Events.
+These Events would then mutate a View Model via its corresponding Reducer function, as per above.
+The actual View could be persisted or kept in-memory, and would be directly queried via a corresponding database-style APIs.
+Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Event, marking the database item for removal from the underlying materialized View.
+
+\paragraph{Key/Value}
+
+A key/value store built on Threads would ``map'' these database Events into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}.
+These Events would then mutate a map-like View Model.
+To complete the database-like API, the entire API could be encapsulated in a database sturcture that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).
 
@@
 type TextileKVStore interface {
-  Put(key string, value interface{}) error
-  Get(key string) (interface{}, error)
+  Put(key string, value Node) error
+  Get(key string) (Node, error)
   Del(key string) error
@@
 
-Similar abstractions could (and will) be used to implement additional database types and functions.
-The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, but exposes the underlying event sourced data for customizing as needed.
-Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities.
-Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents:
+\paragraph{No-SQL}
+
+Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents.
+The interface for such as store, again built using a materialized View, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
 
@@
 
-Where \texttt{Indexable} can be satisfied by any structure with a \texttt{key} field and and \texttt{Query} is taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
+\paragraph{Others}
+
+Similar abstractions could (and will) be used to implement additional database types and functions.
 Tables, feeds, counters, and other simple stores can also be built on Threads, and may require ORM-based solutions to support persistence to certain backing stores\footnote{In-memory stores are almost always implementable using compositions of basic data structures.}.
-Each database style will be implemented as a standalone software library, allowing application developers to pick and choose the solution most useful to the application at hand.
+Each database style would be implemented as a standalone wrapper/software library, allowing application developers to pick and choose the solution most useful to the application at hand.
 Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these ``example'' libraries.
@@
 A practical example of using a JSON CRDT in Textile is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.
-
-\paragraph{Application State}
-
-For application developers familiar with common state management issues, tools such as Redux provide an opinionated framework for making state mutations predictable by imposing certain restrictions on how and when updates can happen.
-In fact, Redux builds on concepts from CRQS and ES itself \cite{reduxMotivation}, and combines them with three core principals: 1) that the entire ``state' of an application should be stored in a single ``store'', 2) where the only way to change the state is to emit an ``action'' (or event), which 3) transforms the application state using a ``pure'' (i.e., zero side effects) reducer function (projection).
-In practice, Redux doesn't care how a developer persists the application state, which means a Thread's event store and read-only materialized views are excellent candidates.
-This focus on unidirectional data flow in perfectly in line with architecture proposed in this paper, and as such, a compatible Redux API will be built on top of Textile Threads.
 

### 2019-09-23 draft.tex

--- 2019-09-22 draft.tex
@@
 
-\subsubsection{Default Views}\label{sec:DefaultViews}
-
-The core Textile Threads interfaces are also exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
+\subsubsection{Databse Views}\label{sec:DefaultViews}
+
+The public API for a Thread already provides several features that you would expect when operating on a dataset or table within a database.
+Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more.
+As such, it is not difficult to imagine a simplified API in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
+
+The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, while still maintaining the power and flexibility of CQRS + ES.
+Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities.
 These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
-In the remainder of this section, we highlight a number of planned database abstractions to provide further utility to developers interested in building on Textile's Threads protocol.
-
-\paragraph{Databases}
-
-A key/value store built on top of Threads can map database \texttt{Put} operations into internal \texttt{KeyValueUpdated} events.
-These Events would then operate on a map-like view model, via an internal ``projection'' (reducer function).
-The key/value view can be persisted or kept in-memory, and would be directly queried via a corresponding \texttt{Get} operation.
-Because Textile uses ES patterns, a \texttt{Del} operation would simply lead to an internal \texttt{KeyValueDeleted} event, marking the key/value pair for removal from the key/value materialized view (Example \ref{ex:KVStore}).
+
+A database built on Threads would map database operations into Events.
+These Events would then mutate a View Model via its corresponding Reducer function, as per above.
+The actual View could be persisted or kept in-memory, and would be directly queried via a corresponding database-style APIs.
+Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Event, marking the database item for removal from the underlying materialized View.
+
+\paragraph{Key/Value}
+
+A key/value store built on Threads would ``map'' these database Events into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}.
+These Events would then mutate a map-like View Model.
+To complete the database-like API, the entire API could be encapsulated in a database sturcture that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).
 
@@
 
-Similar abstractions could (and will) be used to implement additional database types and functions.
-The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, but exposes the underlying event sourced data for customizing as needed.
-Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities.
-Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents (Example \ref{ex:DocStore}).
+\paragraph{No-SQL}
+
+Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents.
+The interface for such as store, again built using a materialized View, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
 

### 2019-09-23 draft.tex
cleanup: move to biblatex
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 \usepackage{amsmath}
+
+\usepackage[sorting=nyt,style=numeric-comp]{biblatex}
+\bibliography{whitepaper}
 
@@
 This concept can be expressed succinctly by the state machine approach \cite{schneiderImplementingFaulttolerantServices1990}: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state.
-This is a powerful concept baked into a simple structure, and is at the heart of many distributed database systems \cite{LogWhatEvery}.
+This is a powerful concept baked into a simple structure, and is at the heart of many distributed database systems \cite{jaykrepsLogWhatEvery2013}.
 
@@
 To support such requirements, developers have begun to utilize event sourcing and command query responsibility segregation (CQRS) patterns \cite{bettsExploringCQRSEvent2013}, relying on append only logs to support immutable state histories.
-Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate event sourcing and CQRS-based applications, including Event Store \cite{EventStore}, Apache Kafaka \cite{ApacheKafka} and Samza \cite{Samza}, among others (see \cite{kleppmannDesigningDataintensiveApplications2017}).
+Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate event sourcing and CQRS-based applications, including Event Store \cite{EventStore}, Apache Kafaka \cite{ApacheKafka} and Samza \cite{ApacheSamza}, among others (see \cite{kleppmannDesigningDataintensiveApplications2017}).
 
@@
 An event handler and corresponding reducer/fold function used to build and maintain a view from a set of (filtered) events.
-While projections may lead to the generation of new events, their reducer should be a pure function (see \cite{eventsouceEventSourcingBasics} and \cite{redux.js.orgGettingStartedRedux} for examples from existing systems).
+While projections may lead to the generation of new events, their reducer should be a pure function (see \cite{eventsourceEventSourcingBasics} and \cite{reduxGettingStartedRedux} for examples from existing systems).
 \end{definition}
@@
 
-The CAP theorem \cite{brewerRobustRobustDistributed2000,gilbertBrewerConjectureFeasibility2002a} states that a distributed database can guarantee only two of the following three promises at the same time: consistency (i.e., that every read receives the most recent write or an error), availability (i.e., that every request receives a (possibly out-of-date) non-error response, and partition tolerance (i.e., that the system continues to operate despite an arbitrary number of messages being dropped (or delayed) by the network).
+The CAP theorem \cite{brewerRobustDistributedSystems2000,gilbertBrewerConjectureFeasibility2002a} states that a distributed database can guarantee only two of the following three promises at the same time: consistency (i.e., that every read receives the most recent write or an error), availability (i.e., that every request receives a (possibly out-of-date) non-error response, and partition tolerance (i.e., that the system continues to operate despite an arbitrary number of messages being dropped (or delayed) by the network).
 As such, many distributed systems are now designed to provide availability and partition tolerance by trading consistency for eventual consistency.
@@
 TLS, Noise) --- among others.
-Libp2p uses the concept of a multiaddress to address peers on a network, which essentially models network addresses as arbitrary encapsulations of protocols \cite{Multiaddr}.
+Libp2p uses the concept of a multiaddress to address peers on a network, which essentially models network addresses as arbitrary encapsulations of protocols \cite{multiformatsMultiaddr}.
 In addition to ``transport layer'' modules, libp2p provides several tools for sharing and/or disseminating data over a p2p network.
@@
 Agent-centric security is possible if the reader can reference local-only, tamper-free code or if the state can be used to determine whether a given operation (e.g. delete data) is permitted.
-Many decentralized networks like Secure Scuttlebutt \cite{ScuttlebuttProtocolGuide} and Holochain \cite{ericharris-braunHolochainScalableAgentcentric2018} make use of agent-centric security.
+Many decentralized networks like Secure Scuttlebutt \cite{securescuttlebuttScuttlebuttProtocolGuide} and Holochain \cite{ericharris-braunHolochainScalableAgentcentric2018} make use of agent-centric security.
 Each of these systems leverage cryptographic signatures to validate peer identities and messages.
@@
 
-Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses\cite{Multiaddr}.
+Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses\cite{multiformatsMultiaddr}.
 Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses.
@@
 This is in contrast to mutable data via IPNS for example, which requires querying the network (DHT) for updates.
-Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed\footnote{This is similar to the append-only message feeds used in Secure Scuttlebutt’s global gossip network \cite{ScuttlebuttProtocolGuide}} like topology.
+Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed\footnote{This is similar to the append-only message feeds used in Secure Scuttlebutt’s global gossip network \cite{securescuttlebuttScuttlebuttProtocolGuide}} like topology.
 Log addresses are recorded in an address book, similar to IPFS Peer address book (Example \ref{ex:KeyBook}).
@@
 The use of a single Read and Follow Key for an entire Log means that, should either of these keys be leaked via malicious (or other/accidental) means, there is no way to prevent a Peer with the leaked keys from listening to Events or traversing the Log history.
-Potential solutions currently being explored by Textile developers include key rotation at specific Event offsets \cite{KeyRotation}, and/or incorporating the Double Ratchet Algorithm \cite{marlinspikeDoubleRatchetAlgorithm2016} for forward secrecy \cite{ungerSoKSecureMessaging2015}.
-
-\ack{This was all Lucas baby! Thanks Lucas}
+Potential solutions currently being explored by Textile developers include key rotation at specific Event offsets \cite{hashicorpKeyRotation}, and/or incorporating the Double Ratchet Algorithm \cite{marlinspikeDoubleRatchetAlgorithm2016} for forward secrecy \cite{ungerSoKSecureMessaging2015}.
+
+\ack{Thanks to the fantastic Textile community for helpful suggestions, feature requests, and reviews.}
 
@@
 
-\bibliographystyle{compj}
-\bibliography{whitepaper}
+\printbibliography
 

### 2019-09-23 draft.tex
fix: mention fundersgit status
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>
fix: minor code lineup cleanup
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 {
-  "_id": "bafykrq5i25vd64ghamtgus6lue74k", // TID
+  "_id": "bafykrq5i25vd64ghamtgus6lue74k",
   "default": "no-access",
@@
 {
-  "_id": "bafykrq5i25vd64ghamtgus6lue74k-acl", // TID-acl
+  "_id": "bafykrq5i25vd64ghamtgus6lue74k-acl",
   "default": "no-access",
@@
 
-\ack{Thanks to the fantastic Textile community for helpful suggestions, feature requests, and reviews.}
+\ack{Thanks to our funders and the excellent Textile community for helpful suggestions, feature requests, and reviews.}
 

### 2019-09-23 draft.tex
minor caps
Signed-off-by: andrew <andrew@textile.io>

--- 2019-09-23 draft.tex
@@
 
-On the downside, and as discussed above (section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots).
+On the downside, and as discussed above (Section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots).
 For example, in Figure \ref{fig:merkledag}, two replicas (top row, bottom row) are attempting to write (left to right) events to the same Merkle-Clock.
@@
 In multi-writer systems, determining causal order in all replicas is particularity challenging.
-As previously discussed (section \ref{sec:merkleclocks}), a solution based on a Merkle-Clock is only partially ordered.
+As previously discussed (Section \ref{sec:merkleclocks}), a solution based on a Merkle-Clock is only partially ordered.
 That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:

### 2019-09-23 draft.tex
gets rid of all "see Section"
Signed-off-by: andrew <andrew@textile.io>

@@
 Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates \cite{fowlerEventSourcing}.
-The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see Section  \ref{sec:ViewsProjections}).
+The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (Section  \ref{sec:ViewsProjections}).
 In other words, with minimal conformity, a single log can model multiple application states \cite{microsoftcorporationAzureApplicationArchitecture}.
@@
 The ACL of an event-driven system is usually a list of access rules built from a series of events.
-For example, the two events, ``grant Bob write access'' and ``revoke read access from Alice'' would together result in a final ACL state where, Bob has read and write access, but Alice does not (see Section  \ref{sec:ViewsProjections}).
+For example, the two events, ``grant Bob write access'' and ``revoke read access from Alice'' would together result in a final ACL state where, Bob has read and write access, but Alice does not (Section  \ref{sec:ViewsProjections}).
 
@@
 \begin{definition}
-(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log's Read Key (see Section  \ref{sec:KeysEncryption}).
+(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log's Read Key (Section  \ref{sec:KeysEncryption}).
 \end{definition}
@@
 
-Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure \ref{fig:EventLog}) enabling log verification by readers (see Section \ref{sec:KeysEncryption}).
+Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure \ref{fig:EventLog}) enabling log verification by readers (Section \ref{sec:KeysEncryption}).
 
@@
 Log Writers can designate other IPFS Peers to ``follow'' a Log, potentially replicating and/or republishing Events.
-A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (see Section  \ref{sec:KeysEncryption}), but is not able to read the Log's contents.
+A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (Section  \ref{sec:KeysEncryption}), but is not able to read the Log's contents.
 Followers should be server-based --- i.e., always online and behind a public IP address.\end{definition}
@@
 This ensures a conflict-free address list without additional complexity.
-Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
+Some Followers may be in the business of replicating Logs (Section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
 This allows them to \emph{pull} (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable (Figure \ref{fig:Pulling}).
@@
 Textile Logs are designed to be shared, composed, and layered into datasets (Figure \ref{fig:LogEncryption}).
-Therefore, Logs are encrypted by default in a manner that enables access control (see Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
+Therefore, Logs are encrypted by default in a manner that enables access control (Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
 
@@
 Threads provide the mechanism to combine multiple Logs from individual authors into singular shared states through the use of either cross-Log sequencing (e.g.
-using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock \cite{kulkarniLogicalPhysicalClocks2014}) or a CRDT (see Section \ref{sec:CRDTs}).
+using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock \cite{kulkarniLogicalPhysicalClocks2014}) or a CRDT (Section \ref{sec:CRDTs}).
 
@@
 (Access-Controlled): This variant declares that consumers should assume an access control list is composable from Log Events.
-The ACL represents a permissions rule set that must be applied when reading data (see Section  \ref{sec:AccessControl} and Example \ref{ex:Identity}b).
+The ACL represents a permissions rule set that must be applied when reading data (Section  \ref{sec:AccessControl} and Example \ref{ex:Identity}b).
 \end{definition}
@@
 An Event containing a mapping of Log IDs to Key Sets, which can be used to join a Thread.
-Threads backed by an access control list (see Section  \ref{sec:AccessControl}) will also include the current ACL for the Thread in an Invite.
+Threads backed by an access control list (Section  \ref{sec:AccessControl}) will also include the current ACL for the Thread in an Invite.
 This enables Peers to invite others to only read or follow a Thread, instead of becoming a full-on collaborator, i.e., a new Log Writer.
@@
 
-The notion of the Follow Key (see Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial.
+The notion of the Follow Key (Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial.
 This allows any Peer in the network to be granted the responsibility of replicating data from another Peer without having read access to the data contained within the Log entries.
 This type of Log replication can act as a data backup mechanism.
-It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (see Section  \ref{sec:LogSync}).
+It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (Section  \ref{sec:LogSync}).
 
@@
 In practice, Snapshots are written to their own Event log and stored locally.
-They can potentially be synced (see Section \ref{sec:LogSync}) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time).
+They can potentially be synced (Section \ref{sec:LogSync}) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time).
 They can similarly be used for initializing state during recovery.
@@
 Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition \ref{ex:SequenceID}) basis.
-Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see Section \ref{sec:TexCRDT}).
+Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (Section \ref{sec:TexCRDT}).
 ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible).

### 2019-09-23 draft.tex
adding more missing citations
Signed-off-by: andrew <andrew@textile.io>

@@
 One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data.
-Other projects have identified app silo-ed data as a problem [MISSING], and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}.
+Other projects have identified app silo-ed data as a problem \cite{bernersLee_readwrite_2013,de_montjoye_trusted_2012}, and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}.
 However, none so far have provided a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data.
@@
 Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates \cite{fowlerEventSourcing}.
-The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (Section  \ref{sec:ViewsProjections}).
+The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see Section  \ref{sec:ViewsProjections}).
 In other words, with minimal conformity, a single log can model multiple application states \cite{microsoftcorporationAzureApplicationArchitecture}.
@@
 A review of the types of possible CRDTs is beyond the scope of this paper, however, it is important to note their role in eventually consistent systems and how they relate to clock-based event ordering.
-See for example \cite{sanjuanMerkleCRDTs2019} and [MISSING] for informative reviews of these types of data structures.
+See for example \cite{enesSingleWriterPrincipleCRDT2017,sanjuanMerkleCRDTs2019} for informative reviews of these types of data structures.
 
@@
 
-There are a number of systems that utilize content addressing to access content and information [LIST SOME], and there is an active body of literature covering its design and implementation [MISSING].
+There are a number of systems that utilize content addressing to access content and information (e.g.
+Git, IPFS, Perkeep, Tahoe-LAFS), and there is an active body of literature covering its design and implementation \cite{benetIPFSContentAddressed2014,selimi_tahoeLafs_2014,rhea_fast_2008}.
 The Interplanetary File System (IPFS) --- which is a set of protocols to create a content-addressed, peer-to-peer filesystem \cite{benetIPFSContentAddressed2014} --- is one such system.
@@
 
-On the downside, and as discussed above (Section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots).
+On the downside, and as discussed above (section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots).
 For example, in Figure \ref{fig:merkledag}, two replicas (top row, bottom row) are attempting to write (left to right) events to the same Merkle-Clock.
@@
 The ACL of an event-driven system is usually a list of access rules built from a series of events.
-For example, the two events, ``grant Bob write access'' and ``revoke read access from Alice'' would together result in a final ACL state where, Bob has read and write access, but Alice does not (Section  \ref{sec:ViewsProjections}).
+For example, the two events, ``grant Bob write access'' and ``revoke read access from Alice'' would together result in a final ACL state where, Bob has read and write access, but Alice does not (see Section  \ref{sec:ViewsProjections}).
 
@@
 In multi-writer systems, determining causal order in all replicas is particularity challenging.
-As previously discussed (Section \ref{sec:merkleclocks}), a solution based on a Merkle-Clock is only partially ordered.
+As previously discussed (section \ref{sec:merkleclocks}), a solution based on a Merkle-Clock is only partially ordered.
 That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:
@@
 \begin{definition}
-(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log's Read Key (Section  \ref{sec:KeysEncryption}).
+(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log's Read Key (see Section  \ref{sec:KeysEncryption}).
 \end{definition}
@@
 
-Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure \ref{fig:EventLog}) enabling log verification by readers (Section \ref{sec:KeysEncryption}).
+Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure \ref{fig:EventLog}) enabling log verification by readers (see Section \ref{sec:KeysEncryption}).
 
@@
 Log Writers can designate other IPFS Peers to ``follow'' a Log, potentially replicating and/or republishing Events.
-A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (Section  \ref{sec:KeysEncryption}), but is not able to read the Log's contents.
+A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (see Section  \ref{sec:KeysEncryption}), but is not able to read the Log's contents.
 Followers should be server-based --- i.e., always online and behind a public IP address.\end{definition}
@@
 This ensures a conflict-free address list without additional complexity.
-Some Followers may be in the business of replicating Logs (Section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
+Some Followers may be in the business of replicating Logs (see section  \ref{sec:Bots}), in which case Writers will announce the additional Log address to Readers.
 This allows them to \emph{pull} (or subscribe to push-based) Events from the Follower's Log address when the Writer is offline or unreachable (Figure \ref{fig:Pulling}).
@@
 Textile Logs are designed to be shared, composed, and layered into datasets (Figure \ref{fig:LogEncryption}).
-Therefore, Logs are encrypted by default in a manner that enables access control (Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
+Therefore, Logs are encrypted by default in a manner that enables access control (see Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
 
@@
 Threads provide the mechanism to combine multiple Logs from individual authors into singular shared states through the use of either cross-Log sequencing (e.g.
-using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock \cite{kulkarniLogicalPhysicalClocks2014}) or a CRDT (Section \ref{sec:CRDTs}).
+using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock \cite{kulkarniLogicalPhysicalClocks2014}) or a CRDT (see Section \ref{sec:CRDTs}).
 
@@
 (Access-Controlled): This variant declares that consumers should assume an access control list is composable from Log Events.
-The ACL represents a permissions rule set that must be applied when reading data (Section  \ref{sec:AccessControl} and Example \ref{ex:Identity}b).
+The ACL represents a permissions rule set that must be applied when reading data (see Section  \ref{sec:AccessControl} and Example \ref{ex:Identity}b).
 \end{definition}
@@
 An Event containing a mapping of Log IDs to Key Sets, which can be used to join a Thread.
-Threads backed by an access control list (Section  \ref{sec:AccessControl}) will also include the current ACL for the Thread in an Invite.
+Threads backed by an access control list (see Section  \ref{sec:AccessControl}) will also include the current ACL for the Thread in an Invite.
 This enables Peers to invite others to only read or follow a Thread, instead of becoming a full-on collaborator, i.e., a new Log Writer.
@@
 
-The notion of the Follow Key (Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial.
+The notion of the Follow Key (see Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial.
 This allows any Peer in the network to be granted the responsibility of replicating data from another Peer without having read access to the data contained within the Log entries.
 This type of Log replication can act as a data backup mechanism.
-It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (Section  \ref{sec:LogSync}).
+It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (see Section  \ref{sec:LogSync}).
 
@@
 In practice, Snapshots are written to their own Event log and stored locally.
-They can potentially be synced (Section \ref{sec:LogSync}) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time).
+They can potentially be synced (see Section \ref{sec:LogSync}) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time).
 They can similarly be used for initializing state during recovery.
@@
 Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition \ref{ex:SequenceID}) basis.
-Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (Section \ref{sec:TexCRDT}).
+Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (see Section \ref{sec:TexCRDT}).
 ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible).

### 2019-09-23 draft.tex
change: incremental commit
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-Replication of Thread data can happen by duplicating/syncing Log data to other Peers.
-
 The notion of the Follow Key (Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial.
@@
 
-\subsection{Thread Views} \label{sec:threadviews}
-
-The use of well-established design patterns from the CQRS+ES (and related) literature means the Threads protocol can provide the foundation for many higher-level concepts and technologies.
-Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store.
-A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store\footnote{Though see section \ref{sec:SnapshotsCompaction} for a discussion of Snapshots, which may be used to speed up Views.}.
+\section{Thread Internals}
+
+To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems, as well as ideas and designs from Flux\footnote{http://facebook.github.io/flux}/Redux\footnote{Flux/Redux build on concepts from CRQS and ES themselves, and Redux is arguably an implementation of the Flux application architecture (i.e., Flux is a pattern, Redux is a library).} \cite{reduxMotivation} and domain driven design [\dots] (DDD).
+
+One more linking sentence \dots
+
+\subsection{Overview}
+
+For application developers familiar with common application state management issues, tools such as Flux/Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
+If you take a look at the architecture of a Flux (or Redux) application, you'll notice several core similaries to CQRS and ES systems (see Figure MISSING).
+In both cases, the focus is on unidirectional data flows that build downstream views from atomic updates in the form of events (or actions).
+
 In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be \emph{materialized} to local persistent storage).
@@
 
+\subsection{Actions}
+
+Both Flux and Redux use `Actions`, which are essentially events.
+Actions cause predictable updates to state via `Reducer` functions in both Flux and Redux, and in both cases, the `Store` is the current state of the application (to which we can send queries or subscribe to changes).
+The differences arise in how they two handle actions: in Flux, `Actions` are dispatched from a singleton `Dispatcher`, which then calls a set of `Reducer` functions that mutate a set of `Stores`.
+Conversely in Redux, `Actions` are dispatched to a singleton `Store`, which then applies a pure (in the functional sense) `Reducer` (which can be composed of multiple pure reducers) to replace the application state with a new immutable state.
+
+\subsubsection{Action Creators}
+
+Both Flux\footnote{\url{https://facebook.github.io/flux/docs/in-depth-overview}} and Redux\footnote{\url{https://redux.js.org/basics/actions}} have the concept of `Action` creators, which are simple functions that create `Actions`.
+In Redux, these create and return a new `Action`, in Flux, they generally also dispatch said `Action` (Redux folks call these bound `Action` creators).
+As you might have noticed, if designed around a domain, Action creators provide bounded context that can be (roughly) compared to an aggregate root (i.e., a cluster of objects, that create a clear reference to the root aggregate).
+
+\subsection{Dispatcher}
+
+Dispatcher is used to store actions and dispatch events to stores.
+It would be a Datastore, which would be responsible for persisting to its own internal Event Store (it essentially is both the Event Store and the Event Bus/Dispatcher).
+It is based on a `TxDatastore``, so that Actions` are only ever `Dispatch`ed if they were successfully stored in the Event Store.
+Additionally, all ``side effects'' should happen within the `Transaction`.
+This is a big reason for the use of Flux rather than Redux architecture\dots because we need some way to consistently and transactionally both store and transmit actions.
+The Dispatcher is actually responsible for running a `Store`'s `Reducer`, once that `Store` has registered with the Dispatcher.
+Its quite simple, and ends up being a pretty tiny class/object.
+
+While a `Store`'s `Reducer` can actually `Dispatch` new `Actions` along the way, this is generally discouraged due to the potential for race conditions and infinite loops.
+Having said that, borrowing concepts from ACID databases and some Redux middlewares, it is possible to avoid these issues by using `Transactions` to some degree.
+Speaking of `Transactions`, if done right, this gets us essentially what you get using Redux ``Sagas'', but with a much simpler interface (basically no interface, the `Transactions` are all hidden inside the `Dispatcher`), which is also really cool.
+Another useful pattern here which works nicely with `Transactions` is batched `Actions` (which is the preferred way\footnote{\url{https://twitter.com/dan_abramov/status/656074974533459968}} to send multiple actions in Redux).
+This works by grouping Actions into meta `Actions` (`BatchedActions`), and then running their `Reducers` in series.
+An example of where this is useful is something like image resizing.
+If you `Dispatch` an `ImageAdded` `Action`, you'd ideally also batch that with a `ImageResized` `Action` or two along the way.
+
+Commands are Actions, the Event Store is your Dispatcher, Projections are Reducers, Materialized Views are Stores, and Queries can be called from Views.
+
+Ok, so that's pretty nice, but what about on the read side?
+Here's where we have to borrow some ideas and terminology from a combination of CQRS and the Resolve library (because Flux and Redux don't really have a read side, just state).
+
+\subsection{Read Models}
+
+Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of views, which are used to optimize queries on the local event store.
+
+This focus on uni-directional data flows that produce immutable state updates via pure functions provides an intuitive framework for predictable state management (just like it says on the box).
+And it is really nice on the client side.
+However, it doesn't scale very well to large sets of data if you are updating your whole state tree each time a new event comes in.
+So how might we use this nice front-end pattern on the backend?
+Others have also thought about this, and they [settled on](https://medium.com/resolvejs/resolve-redux-backend-ebcfc79bbbea)...
+you guessed it, CQRS+ES+DDD.
+
+If we stick with the concept of dispatching `Actions` (Flux and Redux do it differently, but that's an implementation detail), on the write side, then based on these `Actions`, the read side updates the read models, and these models provide data to the queries.
+In practice, a `ReadStore` is defined by a projection function (`Reducer`) and a set of query resolver functions (`Resolvers`).
+
+The `Reducer` functions build a `ReadStore`'s state based on incoming `Actions`.
+Query `Resolvers` use data from the accumulated state to answer specific queries.
+`Reducers` should be ``pure'' functions (where possible), though in practice they aren't really pure, because we are mutating (persistent) state here.
+So basically, we have a Flux `Store` implementation here.
+The `Store`(s) hold/define their own `Reducer` and underlying state.
+Note that we may have multiple `Stores` here (like in Flux).
+
+We might also want to think about `ViewStore`s, which would be more like a Redux `Store`, in that they would be smaller, read-only (immutable), would be built on the fly, and defined by a pure reducer function.
+There are some additional nice things about `ViewStore`s (such as `Snapshots`) that could be easily baked in.
+The nice thing about them is they are very much in line with ES and Redux.
+
+Like in Flux, `Stores` (as they are outlined above) have several specific properties:
+
+Cache data\\
+Expose public getters to access data (never have public setters)\\
+Respond to specific `Actions` from the `Dispatcher`\\
+Always emit a change when their data changes\\
+Only emit changes during a `Dispatch`\\
+
+The other thing about `Stores` is their set of `Resolver` methods.
+These are just pre-defined query functions (ideally the underlying storage is optimized to support these queries) that exist on the `Store` instance (the public APIs so to speak).
+The interesting thing here is that these can similarly be defined around a domain to produce essentially domain aggregates like in DDD.
+
+`ViewStores` are a special kind of `ReadStore`.
+They are queried based on aggregate ID (maybe?), and regenerated on the fly from past events.
+`ViewStores` are like Redux state.
+They should only ever replace their internal state, not mutate it.
+
 \subsubsection{Databse Views}\label{sec:DefaultViews}
+
+For example, what if your domain was a database? In which case, our earlier thoughts about database wrappers might become something like this:
 
@@
 
-A database built on Threads would map database operations into Events.
-These Events would then mutate a View Model via its corresponding Reducer function, as per above.
-The actual View could be persisted or kept in-memory, and would be directly queried via a corresponding database-style APIs.
-Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Event, marking the database item for removal from the underlying materialized View.
+A database built on Threads would map database operations into \texttt{Actions}.
+Here, the \emph{domain} is the databse.
+These \texttt{Actions} would then mutate a \texttt{ReadModel} via its corresponding \texttt{Reducer} function, as per above.
+The \texttt{Store} could be persisted or kept in-memory (as a \texttt{ViewModel}), and would be directly queried via a corresponding database-style \texttt{Resolver} metods.
+Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} \texttt{Action}, marking the database item for removal from the \texttt{ReadStore}.
 
@@
 
-A key/value store built on Threads would ``map'' these database Events into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}.
-These Events would then mutate a map-like View Model.
-To complete the database-like API, the entire API could be encapsulated in a database sturcture that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).
+A key/value store built on Threads would ``map'' these database \texttt{Actions} into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}.
+These \texttt{Actions} would then mutate a map-like \texttt{ReadModel}.
+To complete the database-like API, the \texttt{ActionCreator} and \texttt{ReadModel} methods outlined in Section MISSING could be encapsulated in a database structure that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).
 
@@
 Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents.
-The interface for such as store, again built using a materialized View, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
+The interface for such as store, again built using a materialized \texttt{ReadStore}, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
 
@@
 A practical example of using a JSON CRDT in Textile is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.
-
-\paragraph{Application State}
-
-For application developers familiar with common state management issues, tools such as Redux provide an opinionated framework for making state mutations predictable by imposing certain restrictions on how and when updates can happen.
-In fact, Redux builds on concepts from CRQS and ES itself \cite{reduxMotivation}, and combines them with three core principals:
-
-\begin{enumerate}
-\item The entire \emph{state} of an application should be stored in a single \emph{store}.
-\item The only way to change the state is to emit an \emph{action} (or event).
-\item transforms the application state using a \emph{pure} (i.e., zero side effects) reducer function (projection).
-\end{enumerate}
 

### 2019-09-23 draft.tex
shifting of figs
Signed-off-by: andrew <andrew@textile.io>
remove "see Definition"
Signed-off-by: andrew <andrew@textile.io>

--- 2019-09-23 draft.tex
@@
 How a system provides eventual consistency is often decided based on the intended use of the system.
-Two well-documented categories of solutions include logs (see Definition  \ref{def:Logs}), sequences of deterministic changes, and CRDTs (see Definition section  \ref{def:CRDT}).
+Two well-documented categories of solutions include logs (Definition  \ref{def:Logs}), sequences of deterministic changes, and CRDTs (Definition  \ref{def:CRDT}).
 Logs work best when there is only a single-writer or complete event causality is known.
@@
 
-CRDTs (see Definition \ref{def:CRDT}) are one way to achieve strong eventual consistency, where once all replicas have received the same events, they will arrive at the same final state, \emph{regardless of ordering}\footnote{Though non-commutative CRDTs may require a specific ordering of events in certain cases \cite{sanjuanMerkleCRDTs2019}}.
+CRDTs (Definition \ref{def:CRDT}) are one way to achieve strong eventual consistency, where once all replicas have received the same events, they will arrive at the same final state, \emph{regardless of ordering}\footnote{Though non-commutative CRDTs may require a specific ordering of events in certain cases \cite{sanjuanMerkleCRDTs2019}}.
 A review of the types of possible CRDTs is beyond the scope of this paper, however, it is important to note their role in eventually consistent systems and how they relate to clock-based event ordering.
@@
 
-\begin{example*}
+\begin{example*}[!b]
 \centering
@@
 Thread-level access control lists (ACLs) allow creators to specify who can \emph{follow, read, write, and delete} Thread data.
-Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition \ref{ex:SequenceID}) basis.
+Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (Definition \ref{ex:SequenceID}) basis.
 Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (Section \ref{sec:TexCRDT}).

### 2019-09-23 draft.tex
fixes alignment
Signed-off-by: andrew <andrew@textile.io>
update: rough first draft
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

--- 2019-09-23 draft.tex
@@
 
-\subsection{Thread Views} \label{sec:threadviews}
-
-The use of well-established design patterns from the CQRS+ES (and related) literature means the Threads protocol can provide the foundation for many higher-level concepts and technologies.
-Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store.
-A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store\footnote{Though see section \ref{sec:SnapshotsCompaction} for a discussion of Snapshots, which may be used to speed up Views.}.
-In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be \emph{materialized} to local persistent storage).
-This simple subscription system provides a flexible framework for building complex View-driven application logic.
-
-\subsubsection{Databse Views}\label{sec:DefaultViews}
-
-The public API for a Thread already provides several features that you would expect when operating on a dataset or table within a database.
-Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more.
-As such, it is not difficult to imagine a simplified API in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
+\section{Thread Internals}
+
+To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems, as well as ideas and designs from Flux\footnote{http://facebook.github.io/flux}/Redux\footnote{Flux/Redux build on concepts from CRQS and ES themselves, and Redux is arguably an implementation of the Flux application architecture (i.e., Flux is a pattern, Redux is a library).} \cite{reduxMotivation} and domain driven design [\dots] (DDD).
+In the following sections, we outline the internal structure of a local Threads ``client'' at a relatively high-level, and outline how it can be used to build intuitive developer-facing APIs to make adopting and using Threads ``the right choice'' for a wide range of developers.
+
+\subsection{Overview}
+
+For application developers familiar with common application state management issues, tools such as Flux/Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
+If you take a look at the architecture of a Flux (or Redux) application, you'll notice several core similaries to CQRS and ES systems (see Figure \dots).
+In both cases, the focus is on unidirectional data flows that build downstream views from atomic updates in the form of events (or actions).
+
+We adopt a similar flow in Threads, such that views are updated by subscribing to specific events (or actions as in Flux/Redux), and applying a reducer function that is used to update the view state (which may be \emph{materialized} to local persistent storage).
+This simple subscription system provides a flexible framework for building complex event-driven application logic.
+In the remdainder of this section, we outline the specifics of the Textile Threads internals, focusing on three core components: Actions, the Dispatcher, and the Read/View Stores.
+
+\subsubsection{Actions}
+
+Like with Flux/Redux, to add an Event to the internal Threads system, we use \emph{Actions}.
+Actions cause predictable updates to state via \emph{Reducer} functions.
+In practice, Actions are \emph{dispatched} from a singleton \emph{Dispatcher}, which then stores the Event in an \emph{Event Store}, and calls a set of pure (in the functional sense) Reducer functions that mutate a set of downstream \emph{Read/View Stores}.
+
+In addition to dispatching 'raw' Actions, application developers are able to build and interact with ction \emph{Creators}, which are simple functions (or groups of functions) that create (and possibly dispatch) new Actions.
+For those familiar with DDD, if built around a specific \emph{domain}, Action Creators provide bounded context that can be (roughly) compared to an aggregate root (i.e., a cluster of objects, that create a clear reference to the root aggregate).
+This is a powerful design feature that we will take advantage of in section \dots.
+
+\subsubsection{Dispatcher}
+
+In order to actually \emph{store} events and dispatch actions to downstream views, a Dispatcher is used.
+The dispatcher wraps an embedded Event Store, which is responsible for persisting Events.
+This means a Dispatcher is both an Event Store an Event Bus/Dispatcher.
+In practice, the internal Event Store supports ACID transactions [\dots], so that Actions are only ever dispatched if they were successfully persisted to the Event Store.
+
+All ``side effects'' of a given Action happen within a `Transaction`, in order to ensure consistently of both storage (Event Store) and actions.
+Transactions provide a similar experience to Redux ``Sagas'' [\dots], but with a simpler interface.
+Once an Event has been persisted in the internal Event Store, the Dispatcher is responsible for running a `Store`'s `Reducer` callback.
+In order to make this possible, Stores must \emph{registered} with the Dispatcher.
+
+Actions can also be batch dispatched, which is the preferred way to dispatch multiple Actions at a time.
+This works by grouping Actions into meta Actions, and then running their Reducers in series.
+An example of where this might be useful is when grouping Actions that regularly occur together, such as image processing pipelines.
+
+\subsubsection{Read Models}
+
+The Dispatcher, combined with Actions and Creators provide a useful interface for the write side of the Threads CQRS framework.
+Like many CQRS-based systems, the read side revolves around the concept of views.
+Here, we borrow some ideas and terminology from a combination of Flux/Redux and CQRS, but with a slightly more abstracted API layer, similar to DDD.
+
+As Actions as dispatched on the write side, they are reacted to on the read side via Reducers, causing updates to Read Models, which provide interfaces for queries and accessing state.
+In practice, a ReadStore is defined by a Reducer function and a set of query Resolver functions.
+
+The Reducer functions build a ReadStore's state based on incoming Actions.
+Query Resolvers use data from the accumulated state to answer specific queries.
+Reducers should be ``pure'' functions (where possible)\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}.
+This is very similar to a Flux Store, and it is possible to have multiple Stores here, each specifying their own model and Reducer function.
+
+Similarly, Threads defines a ViewStore, which are a special kind of ReadStore, akin to a Redux Store.
+In practice, they are smaller, read-only (immutable) models, that are built on-the-fly, and defined by a pure Reducer function.
+Threads use ViewStores to define the ACL for a given Thread.
+
+Like in Flux, Stores (as they are outlined above) have several specific properties:
+
+\begin{itemize}
+  \item Cache data
+  \item Expose public getters to access data (never have public setters)
+  \item Respond to specific `Actions` from the `Dispatcher`
+  \item Always emit a change when their data changes
+  \item Only emit changes during a `Dispatch`
+\end{itemize}
+
+To query Stores, we use their set of Resolver methods.
+These are just pre-defined query functions\footnote{Ideally the underlying storage is optimized to support these queries} that exist on the Store instance (the public APIs so to speak).
+The interesting thing here is that these can similarly be defined around a domain to produce essentially domain aggregates in the DDD sense.
+
+\subsection{Abstractions}\label{sec:DefaultViews}
+
+Building on the idea of bounded context and aggregate roots from DDD, Threads provides several default abstractions based on the idea that a database provides a useful context within which to structure our Events/Actions.
+Indeed, the components of a Thread already provide several features that you would expect when operating on a dataset, or table within a database: each Thread is defined by a unique ID, and provides facilities for access control and permissions, networking, sharing, invites, and more.
+As such, it is not difficult to imagine a simplified API in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems --- database domains.
 
@@
 
-A database built on Threads would map database operations into Events.
-These Events would then mutate a View Model via its corresponding Reducer function, as per above.
-The actual View could be persisted or kept in-memory, and would be directly queried via a corresponding database-style APIs.
-Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Event, marking the database item for removal from the underlying materialized View.
+A database built on Threads would map database operations into \texttt{Actions}.
+Here, the \emph{domain} is the databse.
+These \texttt{Actions} would then mutate a \texttt{ReadModel} via its corresponding \texttt{Reducer} function, as per above.
+The \texttt{Store} could be persisted or kept in-memory (as a \texttt{ViewModel}), and would be directly queried via corresponding database-style \texttt{Resolver} metods.
+Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} \texttt{Action}, marking the database item for removal from the \texttt{ReadStore}.
 
@@
 
-A key/value store built on Threads would ``map'' these database Events into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}.
-These Events would then mutate a map-like View Model.
-To complete the database-like API, the entire API could be encapsulated in a database sturcture that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).
+A key/value store built on Threads would ``map'' these database \texttt{Actions} into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}.
+These \texttt{Actions} would then mutate a map-like \texttt{ReadModel}.
+To complete the database-like API, the \texttt{ActionCreator} and \texttt{ReadModel} methods outlined in Section \dots could be encapsulated in a database structure that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).
 
@@
 
-Eventually consistent, CRDT-based stores can also be implemented on top of Threads.
-CRDT-based stores are particularly useful for managing Views of a document in a multi-peer collaborative editing environment (like Google Docs or similar).
-For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype \cite{kleppmannConflictFreeReplicatedJSON2017} that applies updates on a JSON document View model.
-Libraries such as Automerge\footnote{\url{https://github.com/automerge/automerge}} provide useful examples of reducer function that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available.
-A practical example of using a JSON CRDT in Textile is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.
-
-\paragraph{Application State}
-
-For application developers familiar with common state management issues, tools such as Redux provide an opinionated framework for making state mutations predictable by imposing certain restrictions on how and when updates can happen.
-In fact, Redux builds on concepts from CRQS and ES itself \cite{reduxMotivation}, and combines them with three core principals:
-
-\begin{enumerate}
-\item The entire \emph{state} of an application should be stored in a single \emph{store}.
-\item The only way to change the state is to emit an \emph{action} (or event).
-\item transforms the application state using a \emph{pure} (i.e., zero side effects) reducer function (projection).
-\end{enumerate}
-
-In practice, Redux doesn't care how a developer persists the application state, which means a Thread's event store and read-only materialized views are excellent candidates.
-This focus on unidirectional data flow in perfectly in line with architecture proposed in this paper, and as such, a compatible Redux API will be built on top of Textile Threads.
+Eventually consistent, CRDT-based stores can also be implemented on top of Threads' Event-driven architecture.
+CRDT-based Stores are particularly useful for managing views of a document in a multi-peer collaborative editing environment (like Google Docs or similar).
+For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype \cite{kleppmannConflictFreeReplicatedJSON2017} that merges updates to a JSON document in a ReadModels Reduder function.
+Libraries such as Automerge\footnote{\url{https://github.com/automerge/automerge}} provide useful examples of reducer functions that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available [\dots].
+A practical example of using a JSON CRDT in Threads is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document defined as a ReadModel, with queries defined for an access-controlled Thread implementation.
 
@@
 
-At an abstract level, the Textile protocol provides a highly distributed framework for building shared, offline first, datastores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable.
-Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS+ES (and related) literature to ``extend'' the Threads protocol with additional features and controls.
-Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility of Threads.
+At an abstract level, the Textile protocol provides a highly distributed framework for building shared, offline first, Stores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable.
+Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS and ES (and related) literature to ``extend'' the Threads protocol with additional features and controls.
+Indeed, by designing our system around Actions, a Dispartcher, and generic ReadStores, we aim to make easy to extend Threads in many different ways.
+Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility this design affords.
 
@@
 
-Snapshots\footnote{The literature around snapshots and other CQRS and ES is somewhat confusing, we attempt to use the most common definitions here.} are simply the current state of a View at a given point in time.
-They can be used to rebuild the state of a view without having to query and re-play all previous events.
-When a Snapshot is available, a Thread Peer can rebuild the state of a given View by replaying any events generated since the latest Snapshot (with a corresponding reduction in the total number of events that need to be processed) using the View's Reducer function.
+Snapshots\footnote{The literature around snapshots and other CQRS and ES is somewhat confusing, we attempt to use the most common definitions here.} are simply the current state of a Store at a given point in time.
+They can be used to rebuild the state of a ViewStore without having to query and re-play all previous events.
+When a Snapshot is available, a Thread Peer can rebuild the state of a given ViewStore by replaying any events generated since the latest Snapshot (with a corresponding reduction in the total number of events that need to be processed) using the Stores's Reducer function.
 Multiple peers processing the same Log could create a Snapshot every 1000 events and be guaranteed to create the exact same Snapshot because each Peer's Event counts are identical\footnote{Assuming any network partitions are only short-lived (i.e., that peers are able to share events consistently).}.
@@
 They can potentially be synced (Section \ref{sec:LogSync}) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time).
-They can similarly be used for initializing state during recovery.
-
-Compaction can similarly speed up re-hydration of state, and is useful when only the latest Event per Event type is required.
-In addition to being used as a starting point for initialization and recovery, compacted logs can be used to \emph{replace} the Log from which they were created in order to free up local disk space.
-In Textile, compaction becomes much easier because a log only has one writer and conflicts are not possible.
-In practice, Log compaction is a local-only operation (i.e., other Peers do not need to be aware that Compaction was performed).
+They can similarly be used for initializing a Store's during recovery.
+
+Compaction can similarly speed up re-hydration of a Stores's state, and is useful when only the latest Action of a given type is required.
+In addition to being used as a starting point for initialization and recovery, compacted Stores can be used to \emph{replace} the Store from which they were created in order to free up local disk space.
+In practice, Compaction is a local-only operation (i.e., other Peers do not need to be aware that Compaction was performed).
 
@@
 Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition \ref{ex:SequenceID}) basis.
-Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (Section \ref{sec:TexCRDT}).
-ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible).
+Both types of ACLs are implemented as JSON CRDT ViewStores (see Section \ref{sec:TexCRDT}).
+ACLs implemented as JSON ViewStores provide two advantages over static or external ACL rules (although static and external ACLs are also possible).
 First, ACLs are fully mutable, allowing developers to create advanced rules for collaboration with any combination of readers, writers, and followers.

### 2019-09-23 draft.tex

@@
 \phantom{Thread ID = }\phubar{0x62}{Multibase}%
-  \phantom{\quad0x01\quad}\phubar{0x55}{Variant} \phantom{\quad7513b4dcf4efa291ffa92221659f19e2}
-\par\vspace{-.6\baselineskip}\leavevmode
+  \phantom{\quad0x01\quad}\phubar{0x55}{Variant} \phantom{\quad0x539bc1dc03ee8cb5d478e41cc8a4546}
 \label{fig:ThreadID}

### 2019-09-23 draft.tex
update: cleanup, refs, better text
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems, as well as ideas and designs from Flux\footnote{http://facebook.github.io/flux}/Redux\footnote{Flux/Redux build on concepts from CRQS and ES themselves, and Redux is arguably an implementation of the Flux application architecture (i.e., Flux is a pattern, Redux is a library).} \cite{reduxMotivation} and domain driven design [\dots] (DDD).
-In the following sections, we outline the internal structure of a local Threads ``client'' at a relatively high-level, and outline how it can be used to build intuitive developer-facing APIs to make adopting and using Threads ``the right choice'' for a wide range of developers.
+To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems, as well as ideas and designs from Flux \cite{facebookFluxInDepthOverview2019} and Redux\footnote{Flux/Redux build on concepts from CRQS and ES themselves, and Redux is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
+In this section, we outline the internal structure of a local Threads ``client'' at a relatively high-level, and outline how it can be used to build intuitive developer-facing APIs to make adopting and using Threads ``the right choice'' for a wide range of developers.
 
@@
 
-For application developers familiar with common application state management issues, tools such as Flux/Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
-If you take a look at the architecture of a Flux (or Redux) application, you'll notice several core similaries to CQRS and ES systems (see Figure \dots).
+For developers familiar with common application state management issues, tools such as Flux/Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
+If you take a look at the architecture of a Flux (or Redux) application, you'll notice several core similaries to CQRS and ES systems (see \cite{facebookFluxInDepthOverview2019} for example).
 In both cases, the focus is on unidirectional data flows that build downstream views from atomic updates in the form of events (or actions).
 
-We adopt a similar flow in Threads, such that views are updated by subscribing to specific events (or actions as in Flux/Redux), and applying a reducer function that is used to update the view state (which may be \emph{materialized} to local persistent storage).
+We adopt a similar flow in Threads (see Figure \dots), such that views are updated by subscribing to specific events (or actions as in Flux/Redux), and applying a reducer function that is used to update the view state (which may be \emph{materialized} to local persistent storage).
 This simple subscription system provides a flexible framework for building complex event-driven application logic.
-In the remdainder of this section, we outline the specifics of the Textile Threads internals, focusing on three core components: Actions, the Dispatcher, and the Read/View Stores.
-
-\subsubsection{Actions}
+In the following, we outline the specifics of the Textile Threads internals, focusing on three core components: Actions, the Dispatcher, and Read/View Stores.
+
+\subsubsection{Actions}\label{sec:actions}
 
@@
 Actions cause predictable updates to state via \emph{Reducer} functions.
-In practice, Actions are \emph{dispatched} from a singleton \emph{Dispatcher}, which then stores the Event in an \emph{Event Store}, and calls a set of pure (in the functional sense) Reducer functions that mutate a set of downstream \emph{Read/View Stores}.
+In practice, Actions are \emph{dispatched} from a singleton \emph{Dispatcher} (Section \ref{sec:dispatcher}), which then stores the Event in an \emph{Event Store}, and calls a set of pure (in the functional sense) Reducer functions that mutate a set of downstream \emph{Read/View Stores}.
 
@@
 For those familiar with DDD, if built around a specific \emph{domain}, Action Creators provide bounded context that can be (roughly) compared to an aggregate root (i.e., a cluster of objects, that create a clear reference to the root aggregate).
-This is a powerful design feature that we will take advantage of in section \dots.
-
-\subsubsection{Dispatcher}
+This is a powerful design feature that we will take advantage of in section \ref{sec:AccessControl}.
+
+\subsubsection{Dispatcher}\label{sec:dispatcher}
 
@@
 This means a Dispatcher is both an Event Store an Event Bus/Dispatcher.
-In practice, the internal Event Store supports ACID transactions [\dots], so that Actions are only ever dispatched if they were successfully persisted to the Event Store.
-
-All ``side effects'' of a given Action happen within a `Transaction`, in order to ensure consistently of both storage (Event Store) and actions.
-Transactions provide a similar experience to Redux ``Sagas'' [\dots], but with a simpler interface.
-Once an Event has been persisted in the internal Event Store, the Dispatcher is responsible for running a `Store`'s `Reducer` callback.
+In practice, the internal Event Store supports ACID transactions \cite{haerderPrinciplesTransactionorientedDatabase1983}, so that Actions are only ever dispatched if they were successfully persisted to the Event Store.
+
+All ``side effects'' of a given Action happen within a \emph{Transaction}, in order to ensure consistently of both storage (Event Store) and actions.
+Transactions provide a similar experience to Redux ``Sagas''\footnote{\url{https://redux-saga.js.org/}}, but with a simpler interface.
+Once an Event has been persisted in the internal Event Store, the Dispatcher is responsible for running a Store's Reducer callback.
 In order to make this possible, Stores must \emph{registered} with the Dispatcher.
@@
 
-\subsubsection{Read Models}
+\subsubsection{Read Models}\label{sec:readModels}
 
@@
 
-The Reducer functions build a ReadStore's state based on incoming Actions.
-Query Resolvers use data from the accumulated state to answer specific queries.
-Reducers should be ``pure'' functions (where possible)\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}.
-This is very similar to a Flux Store, and it is possible to have multiple Stores here, each specifying their own model and Reducer function.
-
-Similarly, Threads defines a ViewStore, which are a special kind of ReadStore, akin to a Redux Store.
-In practice, they are smaller, read-only (immutable) models, that are built on-the-fly, and defined by a pure Reducer function.
-Threads use ViewStores to define the ACL for a given Thread.
-
-Like in Flux, Stores (as they are outlined above) have several specific properties:
+Like in Flux, Stores (as they are defined here) have several specific properties:
 
@@
 
+Threads also define \emph{ViewStores}, which are a special kind of ReadStore, akin to a Redux Store.
+In practice, they are smaller, read-only (immutable) models, that are built on-the-fly, and defined by a pure Reducer function.
+Textile Threads use ViewStores to define the ACL for a given Thread.
+
+\paragraph{Reducers}
+
+The Reducer functions build a ReadStore's state based on incoming Actions.
+Query Resolvers use data from the accumulated state to answer specific queries.
+Reducers should be ``pure'' functions (where possible)\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}.
+This is very similar to a Flux Store, and it is possible to have multiple Stores here, each specifying their own model and Reducer function.
+
+\paragraph{Resolvers}
+
 To query Stores, we use their set of Resolver methods.
@@
 
-A database built on Threads would map database operations into \texttt{Actions}.
+A database built on Threads would map database operations into Actions.
 Here, the \emph{domain} is the databse.
-These \texttt{Actions} would then mutate a \texttt{ReadModel} via its corresponding \texttt{Reducer} function, as per above.
-The \texttt{Store} could be persisted or kept in-memory (as a \texttt{ViewModel}), and would be directly queried via corresponding database-style \texttt{Resolver} metods.
-Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} \texttt{Action}, marking the database item for removal from the \texttt{ReadStore}.
+These Actions would then mutate a ReadModel via its corresponding Reducer function, as per above.
+The Store could be persisted or kept in-memory (as a ViewModel), and would be directly queried via corresponding database-style Resolver metods.
+Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Action, marking the database item for removal from the ReadStore.
 
@@
 
-A key/value store built on Threads would ``map'' these database \texttt{Actions} into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}.
-These \texttt{Actions} would then mutate a map-like \texttt{ReadModel}.
-To complete the database-like API, the \texttt{ActionCreator} and \texttt{ReadModel} methods outlined in Section \dots could be encapsulated in a database structure that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).
+A key/value store built on Threads would ``map'' these database Actions into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}.
+These Actions would then mutate a map-like ReadModel.
+To complete the database-like API, the Action Creator and ReadModel methods outlined in Section \ref{sec:actions} could be encapsulated in a database structure that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).
 

### 2019-09-23 draft.tex
update: adds defs, cleans redux terminology
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 We propose Threads, a protocol and decentralized database that runs on IPFS meant to help decouple apps from user-data.
-The Threads protocol is designed based on an underlying, CQRS-inspired event-sourcing system for synchronizing data across collaborating peers on a network.
-Threads offer data ownership and a multi-role data access architecture where owners can set independent permissions for writing, reading, and following data updates.
-Threads differ from previous solutions by extending on the multiaddress addressing scheme to allow pull-based replica synchronization in addition to more common push-based synchronization seen in decentralized protocols.
-The flexible event-based structure enables client applications to model advanced states through aggregates, views, and custom CRDTs.
+Inspired by Event Sourcing, CQRS, and Flux, Threads is a protocol for creating and synchronizing state across collaborating peers/replicas on a network.
+Threads offer a multi-layered encryption and data access architecture that enables datasets with independent roles for writing, reading, and following changes.
+By extending on the multiaddress addressing scheme, Threads differs from previous solutions by allowing PULL based replica synchronization in addition to PUSH based synchronization that is common in distributed protocols.
+The flexible event-based structure enables client applications to create advanced applications states, including materialized views, and custom CRDTs.
 
@@
 
-To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems, as well as ideas and designs from Flux \cite{facebookFluxInDepthOverview2019} and Redux\footnote{Flux/Redux build on concepts from CRQS and ES themselves, and Redux is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
+To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems, as well as ideas and designs from Flux \cite{facebookFluxInDepthOverview2019} and Redux\footnote{Redux builds on concepts from CRQS and ES itself, and is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
 In this section, we outline the internal structure of a local Threads ``client'' at a relatively high-level, and outline how it can be used to build intuitive developer-facing APIs to make adopting and using Threads ``the right choice'' for a wide range of developers.
@@
 
-For developers familiar with common application state management issues, tools such as Flux/Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
-If you take a look at the architecture of a Flux (or Redux) application, you'll notice several core similaries to CQRS and ES systems (see \cite{facebookFluxInDepthOverview2019} for example).
+For developers familiar with common application state management issues, tools such as Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
+If you take a look at the architecture of such an application, you'll notice several core similaries to CQRS and ES systems (see \cite{facebookFluxInDepthOverview2019} and/or \cite{abramovCaseFlux2015}).
 In both cases, the focus is on unidirectional data flows that build downstream views from atomic updates in the form of events (or actions).
 
-We adopt a similar flow in Threads (see Figure \dots), such that views are updated by subscribing to specific events (or actions as in Flux/Redux), and applying a reducer function that is used to update the view state (which may be \emph{materialized} to local persistent storage).
+We adopt a similar flow in Threads (see Figure \dots), such that views are updated by subscribing to specific events (or actions), and applying a reducer function that is used to update the view state (which may be \emph{materialized} to local persistent storage).
 This simple subscription system provides a flexible framework for building complex event-driven application logic.
@@
 
-Like with Flux/Redux, to add an Event to the internal Threads system, we use \emph{Actions}.
+Like with Flux, to add an Event to the internal Threads system, we use \emph{Actions}.
 Actions cause predictable updates to state via \emph{Reducer} functions.
 In practice, Actions are \emph{dispatched} from a singleton \emph{Dispatcher} (Section \ref{sec:dispatcher}), which then stores the Event in an \emph{Event Store}, and calls a set of pure (in the functional sense) Reducer functions that mutate a set of downstream \emph{Read/View Stores}.
+
+\begin{definition}
+  (Action).
+Actions are simple structures that are used to send data from an external application to an internal Event Store.
+They are the only source of information for the Store, and contain (at a minimum) a \textnormal{\texttt{Type}}, a \textnormal{\texttt{Data}} payload, and a \textnormal{\texttt{Timestamp}}.
+\end{definition}
 
@@
 
+\begin{definition}
+  (Dispatcher).
+A Dispatcher is used to both persist Events to the local Event Store, as well as broadcast Action payloads to registered Store Reducers.
+Only one Dispatcher per application is ever required.
+\end{definition}
+
 All ``side effects'' of a given Action happen within a \emph{Transaction}, in order to ensure consistently of both storage (Event Store) and actions.
@@
 Once an Event has been persisted in the internal Event Store, the Dispatcher is responsible for running a Store's Reducer callback.
-In order to make this possible, Stores must \emph{registered} with the Dispatcher.
+In order to make this possible, Stores must \emph{register} their Reducer with the Dispatcher.
 
@@
 The Dispatcher, combined with Actions and Creators provide a useful interface for the write side of the Threads CQRS framework.
-Like many CQRS-based systems, the read side revolves around the concept of views.
-Here, we borrow some ideas and terminology from a combination of Flux/Redux and CQRS, but with a slightly more abstracted API layer, similar to DDD.
-
-As Actions as dispatched on the write side, they are reacted to on the read side via Reducers, causing updates to Read Models, which provide interfaces for queries and accessing state.
+Like many CQRS-based systems, the read side revolves around the concept of views (or Stores).
+As Actions are dispatched on the write side, they are reacted to on the read side via Reducers, causing updates to Read/View Models, which provide interfaces for queries and accessing state.
 In practice, a ReadStore is defined by a Reducer function and a set of query Resolver functions.
-
-Like in Flux, Stores (as they are defined here) have several specific properties:
-
-\begin{itemize}
-  \item Cache data
-  \item Expose public getters to access data (never have public setters)
-  \item Respond to specific `Actions` from the `Dispatcher`
-  \item Always emit a change when their data changes
-  \item Only emit changes during a `Dispatch`
-\end{itemize}
+This is very similar to a Redux Store, though it is possible to have \emph{multiple} Stores as in the more general Flux pattern, which is a departure of the ``single source of truth'' in Redux.
+
+\begin{definition}
+  (Reducer).
+Reducers are ``pure'' functions (where possible)\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}, and are written to modify a ReadStore's state based on incoming Actions.
+\end{definition}
+
+\begin{definition}
+  (Resolver).
+Resolvers are simply pre-defined query functions\footnote{Ideally the underlying storage is optimized to support these queries} that exist on the Store instance (i.e., the Store's API).
+The interesting thing here is that these can similarly be defined around a domain to produce essentially domain aggregates in the DDD sense.
+\end{definition}
 
@@
 Textile Threads use ViewStores to define the ACL for a given Thread.
-
-\paragraph{Reducers}
-
-The Reducer functions build a ReadStore's state based on incoming Actions.
-Query Resolvers use data from the accumulated state to answer specific queries.
-Reducers should be ``pure'' functions (where possible)\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}.
-This is very similar to a Flux Store, and it is possible to have multiple Stores here, each specifying their own model and Reducer function.
-
-\paragraph{Resolvers}
-
-To query Stores, we use their set of Resolver methods.
-These are just pre-defined query functions\footnote{Ideally the underlying storage is optimized to support these queries} that exist on the Store instance (the public APIs so to speak).
-The interesting thing here is that these can similarly be defined around a domain to produce essentially domain aggregates in the DDD sense.
 

### 2019-09-23 draft.tex
update: add readstore definition
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 In practice, a ReadStore is defined by a Reducer function and a set of query Resolver functions.
-This is very similar to a Redux Store, though it is possible to have \emph{multiple} Stores as in the more general Flux pattern, which is a departure of the ``single source of truth'' in Redux.
+This is very similar to a Redux Store, though it is possible to have \emph{multiple} Stores as in the more general Flux pattern.
 
@@
 Textile Threads use ViewStores to define the ACL for a given Thread.
+
+\begin{definition}
+  (ReadStore).
+Stores in Threads provide flexible APIs for persisting and accessing state.
+The are defined by a Reducer function for handling incoming Actions, and a set of query Resolver functions for handling queries.
+\end{definition}
 

### 2019-09-24 draft.tex
update: add ref and broadcaster info
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

--- 2019-09-23 draft.tex
@@
 
-To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems, as well as ideas and designs from Flux \cite{facebookFluxInDepthOverview2019} and Redux\footnote{Redux builds on concepts from CRQS and ES itself, and is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
+To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems (e.g., \cite{ereminReduxInspiredBackend2019}), as well as ideas and designs from Flux \cite{facebookFluxInDepthOverview2019}, Redux\footnote{Redux builds on concepts from CRQS and ES itself, and is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
 In this section, we outline the internal structure of a local Threads ``client'' at a relatively high-level, and outline how it can be used to build intuitive developer-facing APIs to make adopting and using Threads ``the right choice'' for a wide range of developers.
@@
 As Actions are dispatched on the write side, they are reacted to on the read side via Reducers, causing updates to Read/View Models, which provide interfaces for queries and accessing state.
-In practice, a ReadStore is defined by a Reducer function and a set of query Resolver functions.
-This is very similar to a Redux Store, though it is possible to have \emph{multiple} Stores as in the more general Flux pattern, which is a departure of the ``single source of truth'' in Redux.
+Read Stores are also responsible for notifying downstream consumers (application logic) of changes to their state via \emph{broadcast channels}.
+In practice, a Store is defined by a Reducer function, a set of query Resolver functions \cite{ereminReduxInspiredBackend2019}, and an Event Bus (or Broadcast channel).
+This similar in some respects to a Redux Store, though it is possible to have \emph{multiple} Stores as in the more general Flux pattern.
 
@@
   (Reducer).
-Reducers are ``pure'' functions (where possible)\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}, and are written to modify a ReadStore's state based on incoming Actions.
+Reducers are ``pure'' functions (where possible)\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}, and are written to modify a Store's state based on incoming Actions.
 \end{definition}
@@
 
-Threads also define \emph{ViewStores}, which are a special kind of ReadStore, akin to a Redux Store.
-In practice, they are smaller, read-only (immutable) models, that are built on-the-fly, and defined by a pure Reducer function.
-Textile Threads use ViewStores to define the ACL for a given Thread.
+\begin{definition}
+  (Store).
+Stores in Threads provide flexible APIs for persisting and accessing state.
+They are defined by a Reducer function for handling incoming Actions, and a set of query Resolver functions for handling queries.
+They are additionally responsible for notifying downstream comsumers of updates to their state via a Broadcaster.
+\end{definition}
 
@@
 The Store could be persisted or kept in-memory (as a ViewModel), and would be directly queried via corresponding database-style Resolver metods.
-Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Action, marking the database item for removal from the ReadStore.
+Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Action, marking the database item for removal from the Store.
 
@@
 Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents.
-The interface for such as store, again built using a materialized \texttt{ReadStore}, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
+The interface for such as store, again built using a materialized Store, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
 
@@
 Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS and ES (and related) literature to ``extend'' the Threads protocol with additional features and controls.
-Indeed, by designing our system around Actions, a Dispartcher, and generic ReadStores, we aim to make easy to extend Threads in many different ways.
+Indeed, by designing our system around Actions, a Dispartcher, and generic Stores, we aim to make easy to extend Threads in many different ways.
 Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility this design affords.

### 2019-09-26 draft.tex
update: [wip] changes to struct + internals
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-\subsubsection{CQRS, Event Sourcing, and Logs}
+\subsubsection{CQRS, Event Sourcing, and Logs}\label{sec:cqrs}
 
@@
 We propose Threads, a protocol and decentralized database that runs on IPFS meant to help decouple apps from user-data.
-Inspired by Event Sourcing, CQRS, and Flux, Threads is a protocol for creating and synchronizing state across collaborating peers/replicas on a network.
+Inspired by Event Sourcing, CQRS, and Redux, Threads is a protocol for creating and synchronizing state across collaborating peers/replicas on a network.
 Threads offer a multi-layered encryption and data access architecture that enables datasets with independent roles for writing, reading, and following changes.
@@
 In multi-writer systems, determining causal order in all replicas is particularity challenging.
-As previously discussed (Section \ref{sec:merkleclocks}), a solution based on a Merkle-Clock is only partially ordered.
+As previously discussed (Section \ref{sec:merkleclocks}), a multi-writer solution, such as \texttt{ipfs-log}\footnote{\url{https://github.com/orbitdb/ipfs-log}}, creates a Merkle-Clock that is only partially ordered.
 That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:
@@
 
-In many cases, the total order of an event sequence is meaningful, and this type of deterministic merge strategy in not sufficient.
+In many cases, the total order of an event sequence is meaningful, and this type of deterministic merge strategy is not sufficient.
 A git merge highlights one such case, in which additional information is sometimes needed in order to resolve a line conflict.
 Ledger based transactions present a similar challenge where the absolute order of transactions between multiple peers really does matter.
+
+Multi-writer systems based on linked log events [MISSING] with relaxed conssitency [CAP] (which is common in p2p systems) can lead to situtaitons in which merges are imppossible due to computational or network-based restrictions.
+For example, any one missing event may cause a branching or split, leading to (effectively) independent event 'trees'.
 
@@
 \emph{Separate} Event Logs can be composed into advanced structures, including CRDTs \cite{enesSingleWriterPrincipleCRDT2017}.
+
+in multi-writer logs we solve this by having linked-data in single-writer logs
+
+in multi-writer systems, conflicts arise along the way
+in order to proceed, we need to define ways to deal with conflits
+with imperfect information (which is common) this inevitably leads to situations where we need to decide how to reconsile the conflict
+this might include a merge, etc
+
+in single writer systems, conflict resolution is deferred to a point at which a decision is required
+this means that imperfect information may be supplemented along the way, without causing conflucts
+it also means that applications can choose conflict resolution strategies specific to the task at hand
+if using crdt, conflicts are irrelevant
+in some instances, an additional clock may be required, such as vector or bloom clock to deal with total ordering
+finally, even merges are possible if this is the desired confluct resolution strategy
 
@@
 Inspired by Git\footnote{\url{https://git-scm.com/}}, a reference to the latest Event in a Log is referred to as the \emph{Head} (or sometimes the \emph{root}).
-When a new Event is received, Readers and Followers simply advance their Head reference.
+When a new Event is received, Readers and Followers simply advance their Head reference for the given Log\footnote{This is similar to how a system such as OrbitDB [MISSING] works, except we are tracking \emph{multiple} Heads (one per Log), rather than a single Head.}.
 
@@
 
-\section{Thread Internals}
-
-To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems (e.g., \cite{ereminReduxInspiredBackend2019}), as well as ideas and designs from Flux \cite{facebookFluxInDepthOverview2019}, Redux\footnote{Redux builds on concepts from CRQS and ES itself, and is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
-In this section, we outline the internal structure of a local Threads ``client'' at a relatively high-level, and outline how it can be used to build intuitive developer-facing APIs to make adopting and using Threads ``the right choice'' for a wide range of developers.
+\section{Thread Internals}\label{sec:internals}
+
+Previous sections have outlined and discussed the core features of the Textile Threads protocol.
+However, we have not yet discussed dealing with Log Events in practice.
+In this section, we provide a relatively high-level discussion of the internal implementation details of a Threads compatible Event Store.
+The Event Store outlined here takes advantage of ideas from several existing CQRS and ES systems (e.g., \cite{ereminReduxInspiredBackend2019}), as well as concepts and designs from Flux \cite{facebookFluxInDepthOverview2019}, Redux\footnote{Redux builds on concepts from CRQS and ES itself, and is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
+Following this discussion of Threads \emph{internals}, in Section \ref{sec:interfaces} we outline how it can be used to build intuitive \emph{public} (developer-facing) APIs to make adopting and using Threads ``the right choice'' for a wide range of developers.
 
@@
 
-We adopt a similar flow in Threads (see Figure \dots), such that views are updated by subscribing to specific events (or actions), and applying a reducer function that is used to update the view state (which may be \emph{materialized} to local persistent storage).
-This simple subscription system provides a flexible framework for building complex event-driven application logic.
-In the following, we outline the specifics of the Textile Threads internals, focusing on three core components: Actions, the Dispatcher, and Read/View Stores.
-
-\subsubsection{Actions}\label{sec:actions}
-
-Like with Flux, to add an Event to the internal Threads system, we use \emph{Actions}.
-Actions cause predictable updates to state via \emph{Reducer} functions.
-In practice, Actions are \emph{dispatched} from a singleton \emph{Dispatcher} (Section \ref{sec:dispatcher}), which then stores the Event in an \emph{Event Store}, and calls a set of pure (in the functional sense) Reducer functions that mutate a set of downstream \emph{Read/View Stores}.
-
-\begin{definition}
-  (Action).
-Actions are simple structures that are used to send data from an external application to an internal Event Store.
-They are the only source of information for the Store, and contain (at a minimum) a \textnormal{\texttt{Type}}, a \textnormal{\texttt{Data}} payload, and a \textnormal{\texttt{Timestamp}}.
-\end{definition}
-
-In addition to dispatching 'raw' Actions, application developers are able to build and interact with ction \emph{Creators}, which are simple functions (or groups of functions) that create (and possibly dispatch) new Actions.
-For those familiar with DDD, if built around a specific \emph{domain}, Action Creators provide bounded context that can be (roughly) compared to an aggregate root (i.e., a cluster of objects, that create a clear reference to the root aggregate).
-This is a powerful design feature that we will take advantage of in section \ref{sec:AccessControl}.
+% [!b] indicates that it will try to be on the bottom of the page
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/Architecture.png}
+  \caption{Architectural diagram for internal Event Store implementation.}
+  \label{fig:Architecture}
+\end{figure}
+
+We adopt a similar flow in Threads (see Figure \ref{fig:Architecture}).
+Like any ES-based system, Threads are built on \emph{Events}.
+Events are similar to actions in a Flux-based system, and are used to produce \emph{predictable} updates to downstream state.
+Similarly to a DDD-based pattern, to add an Event to the internal Threads system, we use \emph{Commands}, which are dispatched from a singleton \emph{Dispatcher} as Events.
+The Dispatcher then stores the derived Event in an \emph{Event Store}, and calls a set of registered \emph{Reducer} functions that mutate a (set of) downstream view \emph{Stores} defined by a corresponding set of view \emph{Models}, all within a single \emph{Transaction}.
+This unidirectional, transaction-based system, provides a flexible framework for building complex event-driven application logic.
+
+\subsubsection{Events \& Commands}\label{sec:commands}
+
+Like most CQRS + ES based systems, \emph{Events} are at the heart of Threads.
+Every update to local and shared (i.e., across peers) application state happens via Events (see also Section \ref{sec:cqrs}).
+Events are similar to actions in Redux, and are used to describe ``changes to an application state'' \cite{fowlerEventSourcing} (e.g., a photo was added, an item was added to a shopping cart, etc).
+Related, \emph{Commands} are simple structures that are used to send data from a \emph{local application} to an internal Event Store.
+They are generally called via an \emph{Event Creator}, which, if built around a specific \emph{domain}, provide bounded context that can be (roughly) compared to an aggregate root in DDD \cite{evansDomaindrivenDesignTackling2004a}.
 
@@
 
-In order to actually \emph{store} events and dispatch actions to downstream views, a Dispatcher is used.
-The dispatcher wraps an embedded Event Store, which is responsible for persisting Events.
-This means a Dispatcher is both an Event Store an Event Bus/Dispatcher.
-In practice, the internal Event Store supports ACID transactions \cite{haerderPrinciplesTransactionorientedDatabase1983}, so that Actions are only ever dispatched if they were successfully persisted to the Event Store.
-
-\begin{definition}
-  (Dispatcher).
-A Dispatcher is used to both persist Events to the local Event Store, as well as broadcast Action payloads to registered Store Reducers.
+In order to actually persist and dispatch Events to downstream view Models, a \emph{Dispatcher} is used.
+As shown in Figure \ref{fig:Architecture}, the Dispatcher is at the center of the ES system.
+All Events must go through the singleton Dispatcher, whether these be from local Event Creators, or Network Peers.
+The Dispatcher is responsible for ensuring that incoming Events are persisted to the Event Store (the ``source of truth'' for the system), as well as dispatched to downstream view Models by way of a set of Reducer functions.
+
+% \begin{definition}
+%   (Dispatcher).
+A Dispatcher is used to both persist Events to the local Event Store, as well as broadcast Event payloads to registered Store Reducers.
 Only one Dispatcher per application is ever required.
-\end{definition}
-
-All ``side effects'' of a given Action happen within a \emph{Transaction}, in order to ensure consistently of both storage (Event Store) and actions.
-Transactions provide a similar experience to Redux ``Sagas''\footnote{\url{https://redux-saga.js.org/}}, but with a simpler interface.
-Once an Event has been persisted in the internal Event Store, the Dispatcher is responsible for running a Store's Reducer callback.
-In order to make this possible, Stores must \emph{register} their Reducer with the Dispatcher.
-
-Actions can also be batch dispatched, which is the preferred way to dispatch multiple Actions at a time.
-This works by grouping Actions into meta Actions, and then running their Reducers in series.
-An example of where this might be useful is when grouping Actions that regularly occur together, such as image processing pipelines.
-
-\subsubsection{Read Models}\label{sec:readModels}
-
-The Dispatcher, combined with Actions and Creators provide a useful interface for the write side of the Threads CQRS framework.
-Like many CQRS-based systems, the read side revolves around the concept of views (or Stores).
-As Actions are dispatched on the write side, they are reacted to on the read side via Reducers, causing updates to Read/View Models, which provide interfaces for queries and accessing state.
-Read Stores are also responsible for notifying downstream consumers (application logic) of changes to their state via \emph{broadcast channels}.
-In practice, a Store is defined by a Reducer function, a set of query Resolver functions \cite{ereminReduxInspiredBackend2019}, and an Event Bus (or Broadcast channel).
-This similar in some respects to a Redux Store, though it is possible to have \emph{multiple} Stores as in the more general Flux pattern.
-
-\begin{definition}
-  (Reducer).
-Reducers are ``pure'' functions (where possible)\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}, and are written to modify a Store's state based on incoming Actions.
-\end{definition}
-
-\begin{definition}
-  (Resolver).
-Resolvers are simply pre-defined query functions\footnote{Ideally the underlying storage is optimized to support these queries} that exist on the Store instance (i.e., the Store's API).
-The interesting thing here is that these can similarly be defined around a domain to produce essentially domain aggregates in the DDD sense.
-\end{definition}
-
-\begin{definition}
-  (Store).
-Stores in Threads provide flexible APIs for persisting and accessing state.
+% \end{definition}
+
+\subsubsection{Transactions}\label{sec:transactions}
+
+All Reducer function calls (and ``side effects'') due to a given Event happen within a single \emph{Transaction} \cite{haerderPrinciplesTransactionorientedDatabase1983}, in order to ensure consistently of both storage (Event Store) and Models.
+Transactions provide a similar experience to Redux \emph{Sagas}\footnote{\url{https://redux-saga.js.org/}}, but with a simpler interface.
+Once an Event has been persisted in the internal Event Store, the Dispatcher is responsible for running a view Model's Reducer callback.
+In order to make this possible, Models must \emph{register} their Reducer with the Dispatcher.
+
+Events can also be \emph{batch dispatched}, which is the preferred way to dispatch multiple Events at a time.
+This works by grouping Events, and then running their Reducers in series.
+An example of where this is useful is when grouping externally derived \emph{Peer-Events}, which can be dispatched directly from the local Peer Host.
+
+\subsubsection{View Models/Stores}\label{sec:views}
+
+As in many CQRS-based systems, as Events are dispatched on the \emph{write side}, they are reacted to on the \emph{read side} via Reducers, causing updates to view Models, which provide interfaces for queries and accessing persisted view \emph{Stores} (state).
+View Stores are then responsible for notifying downstream consumers (application logic) of changes to their state via a \emph{Broadcaster} (i.e., event emitter).
+They are similar in some respects to a Redux Store, though it is possible to have \emph{multiple} view Models/Stores as in the more general Flux pattern.
+
+A view Model is generally defined by custom update logic (i.e., a Reducer), a (possibly ORM\footnote{Object-relational mapping}-based) view Store for persistence, an Event Bus (or Broadcaster) for notifying downstream consumers, and a set of query/\emph{Resolver} \cite{ereminReduxInspiredBackend2019} functions.
+In practice, a view Model may \emph{also} wrap the Event Creator application logic that is used to generate upstream Events.
+This provides an intuitive, singular access point to the \emph{local} system, while also leaving room for updates via \emph{external} peer-events.
+
+\subsubsection{External Events}\label{sec:external}
+
+In addition to locally derived Events (such as via application logic and user interactions), Threads are designed so that Peers may collaborate in a given Thread via Events.
+Events generated by other, network peers are called Peer Events, and they enter the system via a Peer Host.
+The Peer Host is responsible for dealing with incoming Events (be they push or pull).
+These Events are no different from locally-derived Events, though in practice, the Peer Host is required to validate incoming Events before they are dispatched to the system via the Dispatcher.
+
+% \begin{definition}
+%   (Reducer).
+Reducers are ``pure'' (in the functional sense) functions\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}, and are written to modify a Store's state based on incoming Actions.
+% \end{definition}
+
+% \begin{definition}
+%   (Resolver).
+Resolvers are simply pre-defined query functions that exist on the Store instance (i.e., the Store's API).
+The interesting thing here is that these can similarly be defined around a domain to produce domain aggregates (in the DDD sense).
+% \end{definition}
+
+% \begin{definition}
+%   (Store).
+Stores provide flexible APIs for persisting and accessing state.
 They are defined by a Reducer function for handling incoming Actions, and a set of query Resolver functions for handling queries.
 They are additionally responsible for notifying downstream comsumers of updates to their state via a Broadcaster.
-\end{definition}
-
-\subsection{Abstractions}\label{sec:DefaultViews}
+% \end{definition}
+
+\section{Thread Interfaces}\label{sec:interfaces}
+
+To make Threads as easy to adopt and use as possible, we've
 
@@
 
-A database built on Threads would map database operations into Actions.
-Here, the \emph{domain} is the databse.
-These Actions would then mutate a ReadModel via its corresponding Reducer function, as per above.
-The Store could be persisted or kept in-memory (as a ViewModel), and would be directly queried via corresponding database-style Resolver metods.
-Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Action, marking the database item for removal from the Store.
-
-\paragraph{Key/Value}
-
-A key/value store built on Threads would ``map'' these database Actions into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}.
-These Actions would then mutate a map-like ReadModel.
-To complete the database-like API, the Action Creator and ReadModel methods outlined in Section \ref{sec:actions} could be encapsulated in a database structure that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).
+\subsection{Databases}\label{sec:databases}
+
+Even higher-level database abstractions could also be built on Threads.
+For example, a key/value store built on Threads would ``map'' key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del} to an internal (i.e., private) view Model with similarly defined Event Creators and Resolver functions.
+These Events would then mutate the internal (map-like) view Model effectively encapsulating the entire Event Store in a database structure that satisfies a given interface (see Figure \ref{ex:KVStore} for example).
 
@@
 
-\paragraph{No-SQL}
-
 Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents.
-The interface for such as store, again built using a materialized Store, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
+The interface for such as store, again built using a ``wrapped'' view Model, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
 
@@
 
-\paragraph{Others}
-
 Similar abstractions could (and will) be used to implement additional database types and functions.
-Tables, feeds, counters, and other simple stores can also be built on Threads, and may require ORM-based solutions to support persistence to certain backing stores\footnote{In-memory stores are almost always implementable using compositions of basic data structures.}.
+Tables, feeds, counters, and other simple stores can also be built on Threads.
 Each database style would be implemented as a standalone wrapper/software library, allowing application developers to pick and choose the solution most useful to the application at hand.
@@
 
-\paragraph{CRDTs} \label{sec:TexCRDT}
-
-Eventually consistent, CRDT-based stores can also be implemented on top of Threads' Event-driven architecture.
+\subsection{CRDTs} \label{sec:TexCRDT}
+
+Eventually consistent, CRDT-based structures can also be implemented on top of Threads' Event-driven architecture.
 CRDT-based Stores are particularly useful for managing views of a document in a multi-peer collaborative editing environment (like Google Docs or similar).
-For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype \cite{kleppmannConflictFreeReplicatedJSON2017} that merges updates to a JSON document in a ReadModels Reduder function.
+For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype \cite{kleppmannConflictFreeReplicatedJSON2017} that merges updates to a JSON document in a view Model's Reduder function.
 Libraries such as Automerge\footnote{\url{https://github.com/automerge/automerge}} provide useful examples of reducer functions that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available [\dots].
-A practical example of using a JSON CRDT in Threads is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document defined as a ReadModel, with queries defined for an access-controlled Thread implementation.
+A practical example of using a JSON CRDT in Threads is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document defined as a default view Model, with interfaces defined for an access-controlled Threads implementation.
 

### 2019-09-26 draft.tex
updates: internals section ready for review
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-\subsubsection{Variants}
+\subsubsection{Variants}\label{sec:variants}
 
@@
 The Event Store outlined here takes advantage of ideas from several existing CQRS and ES systems (e.g., \cite{ereminReduxInspiredBackend2019}), as well as concepts and designs from Flux \cite{facebookFluxInDepthOverview2019}, Redux\footnote{Redux builds on concepts from CRQS and ES itself, and is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
-Following this discussion of Threads \emph{internals}, in Section \ref{sec:interfaces} we outline how it can be used to build intuitive \emph{public} (developer-facing) APIs to make adopting and using Threads ``the right choice'' for a wide range of developers.
+Following this discussion of Threads \emph{internals}, in Section \ref{sec:interfaces} we outline how it can be used to build intuitive developer-facing application programing interfaces (APIs) to make adopting and using Threads ``the right choice'' for a wide range of developers.
 
@@
 
-To make Threads as easy to adopt and use as possible, we've
-
-Building on the idea of bounded context and aggregate roots from DDD, Threads provides several default abstractions based on the idea that a database provides a useful context within which to structure our Events/Actions.
-Indeed, the components of a Thread already provide several features that you would expect when operating on a dataset, or table within a database: each Thread is defined by a unique ID, and provides facilities for access control and permissions, networking, sharing, invites, and more.
-As such, it is not difficult to imagine a simplified API in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems --- database domains.
-
-The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, while still maintaining the power and flexibility of CQRS + ES.
+To make Threads as easy to adopt and use as possible, Textile has designed a public (developer facing) API on top of the Threads internals that simplifies dealing with events and data, while still maintaining the power and flexibility of CQRS and ES.
 Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities.
 These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
+Inspired by tools such as MondoDB [REF], Parse [REF], and Realm [REF], as well as the idea of bounded context and aggregate roots from DDD, we provide simple abstractions and associated framework for performing distributed database operations as if they were local.
+
+Indeed, the components of a Thread already provide several features that you would expect when operating on a dataset or table within a database: each Thread is defined by a unique ID, and provides facilities for access control and permissions, networking, sharing, invites, and more.
+To illustrate how these underlying components can we combined to produce a simple API with minimal configuration and boilerplate, consider the following example, in which we provide pseudo-code for a hypothetical Photos application.
+
+\subsection{Illustative Example}\label{sec:example}
+
+For any application, the first step would be to import the software development kit (SDK) for the given platform or system; however, for the sake of brevity we do not cover this process here.
+Instead we focus on the higher-level logic required to create and interact with a shared Thread.
+
+To create a useful application, developers start with view \texttt{Models}, as in Example \ref{ex:models}.
+A Model is essentially the public API for the underlying view Models from Section \ref{sec:views}).
+Building on this, developers might create a new Thread for a user to store \texttt{Contact} information, as well as their mobile phone's camera roll photos, as in Example \ref{ex:stores}.
+This would create a new view Store ``under-to-hood'', to be populated by any corresponding update Events.
+All Model instances (returned via \texttt{AddModel}) have several special methods and properties specific to the Threads API, several of which we will explore here.
+
+\begin{example}
+  \begin{minipage}{.45\textwidth}
+    \begin{lstlisting}
+Photo = NewModel({
+  _id: UUID,
+  thumbnail: Buffer,
+  original: Buffer,
+});
+
+Contact = NewModel({
+  _id: UUID,
+  name: {type: String, index: true},
+  avatar: Photo,
+});
+
+// Photos may be grouped into messages.
+Message = NewModel({
+  _id: UUID,
+  author: Contact,
+  body: String,
+  photos: [Photo],
+})
+    \end{lstlisting}
+  \end{minipage}
+  \caption{Create a photo entity and some way to represent a photo's author.}
+  \label{ex:models}
+\end{example}
+
+\begin{example}
+  \begin{minipage}{.45\textwidth}
+    \begin{lstlisting}
+AddressBook = NewDocumentStore("AddressBook")
+// This store should take Contacts.
+AddressBook.AddModel("Contact", Contact)
+// If needed, additional models can be added...
+
+// Create another store for camera roll photos.
+CameraRoll = NewDocumentStore("CameraRoll")
+// This store should take photos.
+CameraRoll.AddModel("Photo", Photo)
+
+// Create another store for a shared album.
+Dogs = NewDocumentStore("Dogs")
+Dogs.AddModel("Message", Message)
+
+// Messages can also be nested.
+Message.AddModel("Message", Message)
+    \end{lstlisting}
+  \end{minipage}
+  \caption{Create address book and camera roll stores. This will create new Threads.}
+  \label{ex:stores}
+\end{example}
+
+The next step is to actually \emph{create} and add a \texttt{Message} object to a shared album.
+In example \ref{ex:adding}, a message instance is created via the custom \texttt{Message} class, and then added to the shared \texttt{Dogs} album.
+Behind the scenes, the Model (which is providing an Event Creator interface) is internally responsible for transforming the \texttt{Add} \emph{Command} to an Event.
+
+\begin{example}
+  \begin{minipage}{.45\textwidth}
+    \begin{lstlisting}
+// Create a message with a photo.
+MyMessage = Message.Create({
+  author: <author_id>,
+  body: "This is Lucas.",
+  photos: [{
+    thumbnail: <buffer>,
+    original: <buffer>
+  }]
+})
+// Now it can be added to Dogs "album".
+Dogs.Add(MyMessage)
+    \end{lstlisting}
+  \end{minipage}
+  \caption{Adding data to a shared Thread.}
+  \label{ex:adding}
+\end{example}
+
+An example Event is given in Example \ref{ex:event}, and is the result of a new Message Event.
+Now, any updates to an existing Message instance will automatically generate the required underlying update Events.
+For example, Example \ref{ex:updates} shows the body text of the previous example being updated, and saved (committed) to the Thread.
+Behind the scenes, this event will be added to the User's local Log, and pushed to any peers identified in the Thread's ACL document (see Sections \ref{sec:variants} and \ref{sec:TexCRDT}).
+In practice, the Model generates another Event that carries the diff and a document identifier.
+
+\begin{example}
+  \begin{minipage}{.45\textwidth}
+    \begin{lstlisting}
+{
+  "body": {
+    "data": {
+      "author_id": <author_id>,
+      "body": "This is Lucas.",
+      "photos": [{
+        "thumbnail": <buffer>,
+        "original": <buffer>
+      }]
+    }
+  },
+  "header": {
+    "time": 1569434034737,
+    "key": "215bs...1DXJ"
+  }
+}
+    \end{lstlisting}
+  \end{minipage}
+\caption{A new Message Event.}
+\label{ex:event}
+\end{example}
+
+\begin{example}
+  \begin{minipage}{.45\textwidth}
+    \begin{lstlisting}
+MyMessage.body = "Actually, this is Fido."
+MyMessage.save()
+
+// New Event with diff and document id
+{
+  "body": {
+    "data": {
+      "doc_id": MyMessage._id,
+      "body": "Actually, this is Fido."
+    }
+  },
+  "header": {
+    "time": 1569434035737,
+    "key": "iJMfqWy...1qfJyc29RS"
+  }
+}
+    \end{lstlisting}
+  \end{minipage}
+\caption{Message updates are persisted and transmitted automatically.}
+\label{ex:updates}
+\end{example}
+
+Additional core Threads functionality is also enabled from this higher-level API.
+Example \ref{ex:others} demonstrates several features common in a Threads-based workflow, including queries, creating invites and changing permissions/access control (see also Section \ref{sec:AccessControl}), as well as subscribing to updates and changes at various levels of the Threads API.
+These subscriptions would enable downstream consumers (views, front-end stores, etc.) to receive updates as changes to the Thread are made via underlying Events.
+
+\begin{example}
+  \begin{minipage}{.45\textwidth}
+    \begin{lstlisting}
+// Query for message, select only thumbnail.
+Dogs.findOne({ "_id": MyMessage._id }, "thumbnail")
+
+// Every (Event increments version tag
+MyMessage.version
+
+// All doc changes in Dogs Thread
+Dogs.subscribe()
+// All changes to all Messages in all Threads
+Message.subscribe() 
+// Changes specific to this document
+MyMessage.subscribe()
+
+// Create invite Event IFF User has permission
+Dogs.grant(<peer_id>, <role>)
+
+// Alter ACL OR create invite if needed/allowed
+MyMessage.grant(<peer_id>, <role>)
+    \end{lstlisting}
+  \end{minipage}
+\caption{Additional Threads-based API functionality.}
+\label{ex:others}
+\end{example}
+
+\subsection{Modules}\label{sec:modules}
+
+One of Textile's stated goals is to allow individuals to better capture the value of their data while still enabling developers to build new interfaces and experiences on top of said data.
+A key piece of this goal is to provide \emph{inter-application} data access to the \emph{same underlying user(-sioled) data}.
+In order to meet this goal, it is necessary for developers to be using the same data structure and conventions when building their apps.
+In conjunction with community developers, Textile will provide a number of \emph{Modules} designed to wrap a given domain (e.g., Photos) into a singular software package to facilitate this.
+This way, developers need only agree on the given data Module in order to provide seamless cross- and even inter-application experiences.
+For example, any developer looking to provide a view on top of a user's Photos (perhaps their phone's camera roll) may utilize the Photos Module (which may be designed as in the example above).
+They may also extend this Module, to provide additional functionality.
+
+In building on top of an existing Module, developers ensure other application developers are also able to interact with the data produced by their app.
+This is a powerful feature, as it enables tigher coupling between applications, and it allows for smaller apps that focus on a very specific user experience (say, filters on Photos).
+However, it is also useful because it provides a \emph{logically centralized}, platform-based developer experience, without the actual centralized infrastructure.
+APIs for Photos, Messages/Chat, Music, Video, Storage, etc are all possible, extensible, and available to all developers.
+This is a powerful concept, but it is also flexible.
+For application developers working on very specific or niche apps with less need for cross-application usability, Modules are not needed, and they can instead focus on custom Models.
+However, it is likely that developers who build on openly available 'standard' Modules will provide a more useful experience for their users.
 
@@
 
-Even higher-level database abstractions could also be built on Threads.
-For example, a key/value store built on Threads would ``map'' key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del} to an internal (i.e., private) view Model with similarly defined Event Creators and Resolver functions.
-These Events would then mutate the internal (map-like) view Model effectively encapsulating the entire Event Store in a database structure that satisfies a given interface (see Figure \ref{ex:KVStore} for example).
+With such a simple interface to such a complex underlying process, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
+For example, a key-value store built on Threads would ``map'' key-value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del} to an internal (i.e., private) Model as in the previous section, with similarly defined methods.
+The generated Events would then mutate the internal (map-like) view Model effectively encapsulating the entire Event Store in a database structure that satisfies a given interface (see Figure \ref{ex:KVStore} for example).
+These too would be distributed as Modules, making it easy for developers to swap in/substitute existing backend infrastructure.
 
@@
 
-Snapshots\footnote{The literature around snapshots and other CQRS and ES is somewhat confusing, we attempt to use the most common definitions here.} are simply the current state of a Store at a given point in time.
-They can be used to rebuild the state of a ViewStore without having to query and re-play all previous events.
-When a Snapshot is available, a Thread Peer can rebuild the state of a given ViewStore by replaying any events generated since the latest Snapshot (with a corresponding reduction in the total number of events that need to be processed) using the Stores's Reducer function.
-Multiple peers processing the same Log could create a Snapshot every 1000 events and be guaranteed to create the exact same Snapshot because each Peer's Event counts are identical\footnote{Assuming any network partitions are only short-lived (i.e., that peers are able to share events consistently).}.
-
-In practice, Snapshots are written to their own Event log and stored locally.
-They can potentially be synced (Section \ref{sec:LogSync}) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time).
-They can similarly be used for initializing a Store's during recovery.
-
-Compaction can similarly speed up re-hydration of a Stores's state, and is useful when only the latest Action of a given type is required.
+Snapshots\footnote{The literature around snapshots and other CQRS and ES terms is somewhat confusing, we attempt to use the most common definitions here.} are simply the current state of a Store at a given point in time.
+They can be used to rebuild the state of a view Store without having to query and re-play all previous Events.
+When a Snapshot is available, a Thread Peer can rebuild the state of a given view Store/Model by replaying any Events generated since the latest Snapshot (with a corresponding reduction in the total number of Events that need to be processed) using the Model's Reducer function.
+Multiple Peers processing the same Log could create a Snapshot every 1000 Events and be guaranteed to create the exact same Snapshot because each Peer's Event counts are identical\footnote{Assuming any network partitions are only short-lived (i.e., that peers are able to share events consistently).}.
+
+In practice, Snapshots are written to their own internal Event Store and stored locally.
+They can potentially be synced (Section \ref{sec:LogSync}) to other Peers as a form of data backup or to optimize state initialization when a new Peer starts participating in a shared Thread (saving disk space, bandwidth, and time).
+They can similarly be used for initializing a local view Store during recovery.
+
+Compaction can similarly speed up re-hydration of a Stores's state, and is useful when only the latest Event of a given type is required.
 In addition to being used as a starting point for initialization and recovery, compacted Stores can be used to \emph{replace} the Store from which they were created in order to free up local disk space.
@@
 One of the most important properties of a shared data model is the ability to apply access control rules.
-There are two forms of access control possible in Threads, Sequence-level ACLs and Thread-level ACLs.
+There are two forms of access control possible in Threads, Entity-level ACLs and Thread-level ACLs.
 Thread-level access control lists (ACLs) allow creators to specify who can \emph{follow, read, write, and delete} Thread data.
-Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition \ref{ex:SequenceID}) basis.
-Both types of ACLs are implemented as JSON CRDT ViewStores (see Section \ref{sec:TexCRDT}).
-ACLs implemented as JSON ViewStores provide two advantages over static or external ACL rules (although static and external ACLs are also possible).
+Similarly, Entity-level ACLs provide more granular control to Thread-writers on a per-Entity (see Definition \ref{ex:SequenceID}) basis.
+Both types of ACLs are implemented as JSON CRDTs (see Section \ref{sec:TexCRDT}) wrapped in a custom view Model (see Section \ref{sec:interfaces}).
+ACLs implemented as JSON Models provide two advantages over static or external ACL rules (although static and external ACLs are also possible).
 First, ACLs are fully mutable, allowing developers to create advanced rules for collaboration with any combination of readers, writers, and followers.
-Second, because ACLs are JSON documents, ACLs can specify their own editing rules (i.e.
+Second, because ACLs are essentially mutable JSON documents, they can specify their \emph{own editing rules} (i.e.
 allowing multiple Thread participants to modify the ACL) in a self-referencing way.
@@
 \begin{definition}
-(Sequence).
-An Event Sequence is a series of ordered events referring to a specific entity or object.
-For example, an ACL JSON document is a single entity made up of a sequence of Thread Events.
-A Sequence might have a unique UUID (see Example \ref{ex:SequenceID}) as in the example below.
+(Entity).
+An Entity is made of of a series of ordered Events referring to a specific entity or object.
+For example, an ACL JSON document is a single entity made up of a sequence of Thread Events that describe a JSON document.
+An Entity might have a unique UUID (see Example \ref{ex:SequenceID}) which can be referenced across/within Event updates.
 \end{definition}
@@
 
-A typical Thread-level ACL JSON (see Example \ref{ex:AclJson}) can be persisted to a local document store as part of a materialized view.
+A typical Thread-level ACL JSON (see Example \ref{ex:AclJson}) can be persisted to a local Event Store as part of the flow described in Section \ref{sec:internals}.
+See also Section \ref{sec:interfaces}, and in particular, Example \ref{ex:others} for the public API for editing ACL definitions.
 

### 2019-09-26 draft.tex
update: clarify single-writer benefits + orbitdb ref
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-\subsection{ Event Logs}
+\subsection{Event Logs}
+
+In multi-writer systems, conflicts arise as disparate peers end up producing conflicting state changes, or changes that end up out of sync.
+In order to proceed, there must be some way to deal with these conflits.
+As mentioned previously (see Section \ref{sec:merkleclocks}), one way to avoid conflicts to some degree is to determine the total order of events in the shared log.
+In multi-writer systems, determining this causal order in all replicas is particularity challenging.
+In some cases (e.g, \texttt{ipfs-log} \cite{markroberthendersonOrbitDBFieldManual2019}), a Merkle-Clock can be used to induce ordering.
+However, this is only \emph{partially ordered}.
+That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:
+
+\begin{quote}
+A total order can be useful \dots and could be obtained, for example, by considering concurrent events to be equal.
+Similarly, a strict total order could be built by sorting concurrent events by the CID or their nodes or by any other arbitrary user-defined strategy based on additional information attached to the clock nodes (data-layer conflict resolution).
+\end{quote}
+
+In many cases, the total order of an event sequence is meaningful, and this type of deterministic merge strategy is not sufficient.
+A git merge highlights one such case, in which additional information is sometimes needed in order to resolve a line conflict.
+Ledger based transactions present a similar challenge where the absolute order of transactions between multiple peers really does matter.
 
@@
 
-In multi-writer systems, determining causal order in all replicas is particularity challenging.
-As previously discussed (Section \ref{sec:merkleclocks}), a multi-writer solution, such as \texttt{ipfs-log}\footnote{\url{https://github.com/orbitdb/ipfs-log}}, creates a Merkle-Clock that is only partially ordered.
-That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:
-
-\begin{quote}
-A total order can be useful \dots and could be obtained, for example, by considering concurrent events to be equal.
-Similarly, a strict total order could be built by sorting concurrent events by the CID or their nodes or by any other arbitrary user-defined strategy based on additional information attached to the clock nodes (data-layer conflict resolution).
-\end{quote}
-
-In many cases, the total order of an event sequence is meaningful, and this type of deterministic merge strategy is not sufficient.
-A git merge highlights one such case, in which additional information is sometimes needed in order to resolve a line conflict.
-Ledger based transactions present a similar challenge where the absolute order of transactions between multiple peers really does matter.
-
-Multi-writer systems based on linked log events [MISSING] with relaxed conssitency [CAP] (which is common in p2p systems) can lead to situtaitons in which merges are imppossible due to computational or network-based restrictions.
-For example, any one missing event may cause a branching or split, leading to (effectively) independent event 'trees'.
-
-Our solution to deal with divergent Merkle-Clocks is to institute a \emph{single-writer rule}: A log can only be updated by a single replica or \emph{identity}.
-An Event Log is a \emph{single-writer Merkle-Clock} that can be totally ordered (Figure \ref{fig:EventLog}).
+Clearly, multi-writer systems with relaxed conssitency \cite{brewerRobustDistributedSystems2000,gilbertBrewerConjectureFeasibility2002a} and imperfect information (e.g., caused by computational or network issues, both of which are common in p2p systems) will almost inevitably lead to situations where we need to decide how to reconsile a conflict.
+Some systems come with conflict resolution strategies ``built in'' (e.g., \texttt{ipfs-log} uses a built-in CRDT).
+However, this is not ideal, because it forces developers/users into a very specific usage pattern.
+It is particularly limiting in situations where developers might like to institute their own CRDT or domain-specific conflict resolution strategies.
+
+Our solution to dealing with log conflicts (i.e., divergent Merkle-Clocks) is to institute a \emph{single-writer rule}: A Log can only be updated by a single replica or \emph{identity}.
+An Event Log is then a \emph{single-writer Merkle-Clock} that can be totally ordered (Figure \ref{fig:EventLog}).
 \emph{Separate} Event Logs can be composed into advanced structures, including CRDTs \cite{enesSingleWriterPrincipleCRDT2017}.
 
-in multi-writer logs we solve this by having linked-data in single-writer logs
-
-in multi-writer systems, conflicts arise along the way
-in order to proceed, we need to define ways to deal with conflits
-with imperfect information (which is common) this inevitably leads to situations where we need to decide how to reconsile the conflict
-this might include a merge, etc
-
-in single writer systems, conflict resolution is deferred to a point at which a decision is required
-this means that imperfect information may be supplemented along the way, without causing conflucts
-it also means that applications can choose conflict resolution strategies specific to the task at hand
-if using crdt, conflicts are irrelevant
-in some instances, an additional clock may be required, such as vector or bloom clock to deal with total ordering
-finally, even merges are possible if this is the desired confluct resolution strategy
+This presents a novel take on multi-writer systems \cite{enesSingleWriterPrincipleCRDT2017}, such that conflict resolution is deferred to a point at which a decision is actually \emph{required}.
+This means that imperfect information may be supplemented along the way (i.e., at a later time), without causing conflicts in the mean time.
+It also means that applications can choose conflict resolution strategies specific to the task at hand.
+For example, if using a downstream CRDT, ordering is actually irrelevant, and can be ignored completely.
+Alternatively, an additional clock may be required to ensure consistent ordering, such as vector or bloom clock (see Section \ref{sec:LogicalClocks}).
+Finally, even manual merge-type strategies are possible if this is the desired conflict resolution strategy.
 
@@
 With such a simple interface to such a complex underlying process, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
+Here we draw inspiration from similar projects (e.g., OrbitDB \cite{markroberthendersonOrbitDBFieldManual2019}) which have made it much easier for developers familiar with centralized database systems to make the move to decentralized systems such as Threads.
 For example, a key-value store built on Threads would ``map'' key-value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del} to an internal (i.e., private) Model as in the previous section, with similarly defined methods.

### 2019-09-26 draft.tex
update: adds footnotes for db examples
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
-Inspired by tools such as MondoDB [REF], Parse [REF], and Realm [REF], as well as the idea of bounded context and aggregate roots from DDD, we provide simple abstractions and associated framework for performing distributed database operations as if they were local.
+Inspired by tools such as MondoDB\footnote{\url{http://www.mongodb.com}}, Parse\footnote{\url{https://parseplatform.org}}, and Realm\footnote{\url{https://realm.io}}, as well as the idea of bounded context and aggregate roots from DDD, we provide simple abstractions and associated framework for performing distributed database operations as if they were local.
 

### 2019-09-26 draft.tex
updates: sander nits + various typos
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
   \includegraphics[width=\linewidth]{figures/Hash_Request.png}
-  \caption{The crypographic hash of content is used to make a request to the network of IPFS peers.
-Using the built-in routing machanisms and the distributed hash table, a peer(s) hosting the requested content is identified and content is returned.
+  \caption{The cryptographic hash of content is used to make a request to the network of IPFS peers.
+Using the built-in routing mechanisms and the distributed hash table, a peer(s) hosting the requested content is identified and content is returned.
 }
@@
 Libp2p solves a number of challenges that are distinct to peer-to-peer (p2p) networks.
-A comprehensive coverage of networking issues in P2P systems is out of the scope for this paper, however, some core challenges that libp2p helps to address include network address translator \cite{srisureshIPNetworkAddress} (NAT) traversal, peer discovery and handshake protocols, and even encryption and transport security --- libp2p supports both unencrypted (e.g.
+A comprehensive coverage of networking issues in P2P systems is out of the scope for this paper, however, some core challenges that libp2p helps to address include network address translator \cite{srisureshIPNetworkAddress} (NAT) traversal, peer discovery and handshake protocols, and even encryption and transport security --- libp2p supports both un-encrypted (e.g.
 TCP, UDP) and encrypted protocols (e.g.
@@
 
-Agent-centric security refers to the maintainance of data integrity without leveraging a central or blockchain-based consensus.
+Agent-centric security refers to the maintenance of data integrity without leveraging a central or blockchain-based consensus.
 The general approach is just to let the reader enforce permissions and perform validations, not the writer or some central authority.
@@
 
-All filesystems and databases have some notion of ``access control''.
+All file-systems and databases have some notion of ``access control''.
 Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects \cite{shireyInternetSecurityGlossary2007}.
@@
 In multi-writer systems, conflicts arise as disparate peers end up producing conflicting state changes, or changes that end up out of sync.
-In order to proceed, there must be some way to deal with these conflits.
+In order to proceed, there must be some way to deal with these conflicts.
 As mentioned previously (see Section \ref{sec:merkleclocks}), one way to avoid conflicts to some degree is to determine the total order of events in the shared log.
@@
 
-Clearly, multi-writer systems with relaxed conssitency \cite{brewerRobustDistributedSystems2000,gilbertBrewerConjectureFeasibility2002a} and imperfect information (e.g., caused by computational or network issues, both of which are common in p2p systems) will almost inevitably lead to situations where we need to decide how to reconsile a conflict.
+Clearly, multi-writer systems with relaxed consistency \cite{brewerRobustDistributedSystems2000,gilbertBrewerConjectureFeasibility2002a} and imperfect information (e.g., caused by computational or network issues, both of which are common in p2p systems) will almost inevitably lead to situations where we need to decide how to reconcile a conflict.
 Some systems come with conflict resolution strategies ``built in'' (e.g., \texttt{ipfs-log} uses a built-in CRDT).
@@
 For example, if using a downstream CRDT, ordering is actually irrelevant, and can be ignored completely.
-Alternatively, an additional clock may be required to ensure consistent ordering, such as vector or bloom clock (see Section \ref{sec:LogicalClocks}).
+Alternatively, an additional clock may be required to ensure consistent ordering, such as a vector or Bloom clock (see Section \ref{sec:LogicalClocks}).
 Finally, even manual merge-type strategies are possible if this is the desired conflict resolution strategy.
@@
 In this section, we provide a relatively high-level discussion of the internal implementation details of a Threads compatible Event Store.
-The Event Store outlined here takes advantage of ideas from several existing CQRS and ES systems (e.g., \cite{ereminReduxInspiredBackend2019}), as well as concepts and designs from Flux \cite{facebookFluxInDepthOverview2019}, Redux\footnote{Redux builds on concepts from CRQS and ES itself, and is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
+The Event Store outlined here takes advantage of ideas from several existing CQRS and ES systems (e.g., \cite{ereminReduxInspiredBackend2019}), as well as concepts and designs from Flux \cite{facebookFluxInDepthOverview2019}, Redux\footnote{Redux builds on concepts from CQRS and ES itself, and is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
 Following this discussion of Threads \emph{internals}, in Section \ref{sec:interfaces} we outline how it can be used to build intuitive developer-facing application programing interfaces (APIs) to make adopting and using Threads ``the right choice'' for a wide range of developers.
@@
 For developers familiar with common application state management issues, tools such as Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
-If you take a look at the architecture of such an application, you'll notice several core similaries to CQRS and ES systems (see \cite{facebookFluxInDepthOverview2019} and/or \cite{abramovCaseFlux2015}).
+If you take a look at the architecture of such an application, you'll notice several core similarities to CQRS and ES systems (see \cite{facebookFluxInDepthOverview2019} and/or \cite{abramovCaseFlux2015}).
 In both cases, the focus is on unidirectional data flows that build downstream views from atomic updates in the form of events (or actions).
@@
 They are defined by a Reducer function for handling incoming Actions, and a set of query Resolver functions for handling queries.
-They are additionally responsible for notifying downstream comsumers of updates to their state via a Broadcaster.
+They are additionally responsible for notifying downstream consumers of updates to their state via a Broadcaster.
 % \end{definition}
@@
 
-Indeed, the components of a Thread already provide several features that you would expect when operating on a dataset or table within a database: each Thread is defined by a unique ID, and provides facilities for access control and permissions, networking, sharing, invites, and more.
+Indeed, the components of a Thread already provide several features that you would expect when operating on a dataset or table within a database: each Thread is defined by a unique ID, and provides facilities for access control and permissions, networking, and more.
 To illustrate how these underlying components can we combined to produce a simple API with minimal configuration and boilerplate, consider the following example, in which we provide pseudo-code for a hypothetical Photos application.
 
-\subsection{Illustative Example}\label{sec:example}
+\subsection{Illustrative Example}\label{sec:example}
 
@@
 MyMessage.body = "Actually, this is Fido."
-MyMessage.save()
+MyMessage.Save()
 
@@
 // Query for message, select only thumbnail.
-Dogs.findOne({ "_id": MyMessage._id }, "thumbnail")
+Dogs.FindOne({ "_id": MyMessage._id }, "thumbnail")
 
 // Every (Event increments version tag
-MyMessage.version
+MyMessage.Version()
 
@@
 // All changes to all Messages in all Threads
-Message.subscribe() 
+Message.Subscribe() 
 // Changes specific to this document
-MyMessage.subscribe()
+MyMessage.Subscribe()
 
 // Create invite Event IFF User has permission
-Dogs.grant(<peer_id>, <role>)
+Dogs.Grant(<peer_id>, <role>)
 
 // Alter ACL OR create invite if needed/allowed
-MyMessage.grant(<peer_id>, <role>)
+MyMessage.Grant(<peer_id>, <role>)
     \end{lstlisting}
@@
 One of Textile's stated goals is to allow individuals to better capture the value of their data while still enabling developers to build new interfaces and experiences on top of said data.
-A key piece of this goal is to provide \emph{inter-application} data access to the \emph{same underlying user(-sioled) data}.
+A key piece of this goal is to provide \emph{inter-application} data access to the \emph{same underlying user(-siloed) data}.
 In order to meet this goal, it is necessary for developers to be using the same data structure and conventions when building their apps.
@@
 In building on top of an existing Module, developers ensure other application developers are also able to interact with the data produced by their app.
-This is a powerful feature, as it enables tigher coupling between applications, and it allows for smaller apps that focus on a very specific user experience (say, filters on Photos).
+This is a powerful feature, as it enables tighter coupling between applications, and it allows for smaller apps that focus on a very specific user experience (say, filters on Photos).
 However, it is also useful because it provides a \emph{logically centralized}, platform-based developer experience, without the actual centralized infrastructure.
@@
 For application developers working on very specific or niche apps with less need for cross-application usability, Modules are not needed, and they can instead focus on custom Models.
-However, it is likely that developers who build on openly available 'standard' Modules will provide a more useful experience for their users.
+However, it is likely that developers who build on openly available \emph{standard} Modules will provide a more useful experience for their users.
 
@@
 Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS and ES (and related) literature to ``extend'' the Threads protocol with additional features and controls.
-Indeed, by designing our system around Actions, a Dispartcher, and generic Stores, we aim to make easy to extend Threads in many different ways.
+Indeed, by designing our system around Actions, a Dispartcher, and generic Stores, we aim to make it easy to extend Threads in many different ways.
 Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility this design affords.

### 2019-09-26 draft.tex

@@
 
-% [!b] bottom page
-\begin{figure*}[!t]
+\begin{figure*}[th]
 \centering
@@
 
-\begin{example*}[!b]
+\begin{example}[hb]
 \centering
@@
   \label{ex:Identity}
-\end{example*}
+\end{example}
 
@@
 
-\section{Thread Internals}\label{sec:internals}
-
-Previous sections have outlined and discussed the core features of the Textile Threads protocol.
+\section{Threads Internals}\label{sec:internals}
+
+Previous sections have discussed the core features of the Textile Threads protocol.
 However, we have not yet discussed dealing with Log Events in practice.
-In this section, we provide a relatively high-level discussion of the internal implementation details of a Threads compatible Event Store.
+In this section, we provide a description of a Threads-compatible Event Store implementation.
 The Event Store outlined here takes advantage of ideas from several existing CQRS and ES systems (e.g., \cite{ereminReduxInspiredBackend2019}), as well as concepts and designs from Flux \cite{facebookFluxInDepthOverview2019}, Redux\footnote{Redux builds on concepts from CQRS and ES itself, and is arguably an implementation of the Flux application architecture.} \cite{reduxMotivation} and domain driven design \cite{evansDomaindrivenDesignTackling2004a} (DDD)\footnote{\url{https://dddcommunity.org}}.
@@
 
-For developers familiar with common application state management issues, tools such as Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
+Application state management tools such as Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
 If you take a look at the architecture of such an application, you'll notice several core similarities to CQRS and ES systems (see \cite{facebookFluxInDepthOverview2019} and/or \cite{abramovCaseFlux2015}).
@@
 
-% [!b] indicates that it will try to be on the bottom of the page
 \begin{figure}
@@
 Events are similar to actions in a Flux-based system, and are used to produce \emph{predictable} updates to downstream state.
-Similarly to a DDD-based pattern, to add an Event to the internal Threads system, we use \emph{Commands}, which are dispatched from a singleton \emph{Dispatcher} as Events.
+Similarly to a DDD-based pattern, to add an Event to the internal Threads system, we use \emph{Event Creators}, which dispatch Events to the system via a singleton \emph{Dispatcher}.
 The Dispatcher then stores the derived Event in an \emph{Event Store}, and calls a set of registered \emph{Reducer} functions that mutate a (set of) downstream view \emph{Stores} defined by a corresponding set of view \emph{Models}, all within a single \emph{Transaction}.
@@
 
-\subsubsection{Events \& Commands}\label{sec:commands}
+\subsubsection{Events \& Creators}\label{sec:creators}
 
@@
 Events are similar to actions in Redux, and are used to describe ``changes to an application state'' \cite{fowlerEventSourcing} (e.g., a photo was added, an item was added to a shopping cart, etc).
-Related, \emph{Commands} are simple structures that are used to send data from a \emph{local application} to an internal Event Store.
-They are generally called via an \emph{Event Creator}, which, if built around a specific \emph{domain}, provide bounded context that can be (roughly) compared to an aggregate root in DDD \cite{evansDomaindrivenDesignTackling2004a}.
+Related, \emph{Event Creators} are used to send Events from a \emph{local application} to an internal Event Store.
+If built around a specific \emph{domain}, Creators provide bounded context that can be roughly compared to an aggregate root in DDD \cite{evansDomaindrivenDesignTackling2004a}.
 
@@
 All Reducer function calls (and ``side effects'') due to a given Event happen within a single \emph{Transaction} \cite{haerderPrinciplesTransactionorientedDatabase1983}, in order to ensure consistently of both storage (Event Store) and Models.
-Transactions provide a similar experience to Redux \emph{Sagas}\footnote{\url{https://redux-saga.js.org/}}, but with a simpler interface.
+Transactions are similar to Redux \emph{Sagas}\footnote{\url{https://redux-saga.js.org/}} in terms of outcome, but with stronger consistency guarantees.
 Once an Event has been persisted in the internal Event Store, the Dispatcher is responsible for running a view Model's Reducer callback.
@@
 
-Events can also be \emph{batch dispatched}, which is the preferred way to dispatch multiple Events at a time.
+% Events can also be \emph{batch dispatched}, which is the preferred way to dispatch multiple Events at a time.
 This works by grouping Events, and then running their Reducers in series.
@@
 
-As in many CQRS-based systems, as Events are dispatched on the \emph{write side}, they are reacted to on the \emph{read side} via Reducers, causing updates to view Models, which provide interfaces for queries and accessing persisted view \emph{Stores} (state).
+As in many CQRS-based systems, Events are dispatched on the write side, and are \emph{reacted to} on the read side.
+Reactions happen via Reducer fuctions, which cause updates to view Models, which in turn provide interfaces for queries and accessing persisted view \emph{Stores} (state).
 View Stores are then responsible for notifying downstream consumers (application logic) of changes to their state via a \emph{Broadcaster} (i.e., event emitter).
@@
 
-In addition to locally derived Events (such as via application logic and user interactions), Threads are designed so that Peers may collaborate in a given Thread via Events.
+In addition to locally derived Events (i.e., from application logic and user interactions), Threads are designed so that Peers may collaborate in a given Thread via Events.
 Events generated by other, network peers are called Peer Events, and they enter the system via a Peer Host.
@@
 
-To make Threads as easy to adopt and use as possible, Textile has designed a public (developer facing) API on top of the Threads internals that simplifies dealing with events and data, while still maintaining the power and flexibility of CQRS and ES.
+To make Threads as easy to adopt and use as possible, Textile has designed a developer facing API on top of the Threads internals that simplifies dealing with events and data, while still maintaining the power and flexibility of CQRS and ES.
 Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities.
 These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
-Inspired by tools such as MondoDB\footnote{\url{http://www.mongodb.com}}, Parse\footnote{\url{https://parseplatform.org}}, and Realm\footnote{\url{https://realm.io}}, as well as the idea of bounded context and aggregate roots from DDD, we provide simple abstractions and associated framework for performing distributed database operations as if they were local.
+Inspired by tools such as MondoDB\footnote{\url{http://www.mongodb.com}}, Parse\footnote{\url{https://parseplatform.org}}, and Realm\footnote{\url{https://realm.io}}, as well as the idea of bounded context and aggregate roots from DDD, we provide simple abstractions for performing distributed database operations as if they were local.
 
@@
 
-For any application, the first step would be to import the software development kit (SDK) for the given platform or system; however, for the sake of brevity we do not cover this process here.
+% For any application, the first step would be to import the software development kit (SDK) for the given platform or system; however, for the sake of brevity we do not cover this process here.
 Instead we focus on the higher-level logic required to create and interact with a shared Thread.
@@
 Building on this, developers might create a new Thread for a user to store \texttt{Contact} information, as well as their mobile phone's camera roll photos, as in Example \ref{ex:stores}.
-This would create a new view Store ``under-to-hood'', to be populated by any corresponding update Events.
-All Model instances (returned via \texttt{AddModel}) have several special methods and properties specific to the Threads API, several of which we will explore here.
+This would create a new view Store ``under-to-hood'' (with corresponding indexes, etc), to be mutated by incoming Events.
 
@@
 // Create another store for a shared album.
-Dogs = NewDocumentStore("Dogs")
-Dogs.AddModel("Message", Message)
+MyDogsAlbum = NewDocumentStore("Dogs")
+MyDogsAlbum.AddModel("Message", Message)
 
@@
 The next step is to actually \emph{create} and add a \texttt{Message} object to a shared album.
-In example \ref{ex:adding}, a message instance is created via the custom \texttt{Message} class, and then added to the shared \texttt{Dogs} album.
-Behind the scenes, the Model (which is providing an Event Creator interface) is internally responsible for transforming the \texttt{Add} \emph{Command} to an Event.
+In example \ref{ex:adding}, a message instance is created via the custom \texttt{Message} class, and then added to the shared \texttt{Dogs} Thread (which represents an album here).
+Behind the scenes, the Model (which is providing an Event Creator interface) is internally responsible for dispatching the Event through the local Dispatcher.
 
@@
 // Now it can be added to Dogs "album".
-Dogs.Add(MyMessage)
+MyDogsAlbum.Add(MyMessage)
     \end{lstlisting}
@@
 
-Additional core Threads functionality is also enabled from this higher-level API.
+All instances and models in Threads have several special methods and properties specific to the Threads API, several of which we will explore here.
 Example \ref{ex:others} demonstrates several features common in a Threads-based workflow, including queries, creating invites and changing permissions/access control (see also Section \ref{sec:AccessControl}), as well as subscribing to updates and changes at various levels of the Threads API.
@@
 One of Textile's stated goals is to allow individuals to better capture the value of their data while still enabling developers to build new interfaces and experiences on top of said data.
-A key piece of this goal is to provide \emph{inter-application} data access to the \emph{same underlying user(-siloed) data}.
+A key piece of this goal is to provide \emph{inter-application} data access to the \emph{same underlying user-siloed data}.
 In order to meet this goal, it is necessary for developers to be using the same data structure and conventions when building their apps.
 In conjunction with community developers, Textile will provide a number of \emph{Modules} designed to wrap a given domain (e.g., Photos) into a singular software package to facilitate this.
-This way, developers need only agree on the given data Module in order to provide seamless cross- and even inter-application experiences.
+This way, developers need only agree on the given data Module in order to provide seamless inter-application experiences.
 For example, any developer looking to provide a view on top of a user's Photos (perhaps their phone's camera roll) may utilize the Photos Module (which may be designed as in the example above).
@@
 In building on top of an existing Module, developers ensure other application developers are also able to interact with the data produced by their app.
-This is a powerful feature, as it enables tighter coupling between applications, and it allows for smaller apps that focus on a very specific user experience (say, filters on Photos).
-However, it is also useful because it provides a \emph{logically centralized}, platform-based developer experience, without the actual centralized infrastructure.
+This enables tighter coupling between applications, and it allows for smaller apps that can focus on a very specific user experience (say, filters on Photos).
+Furthermore, it provides a \emph{logically centralized}, platform-like developer experience, without the actual centralized infrastructure.
 APIs for Photos, Messages/Chat, Music, Video, Storage, etc are all possible, extensible, and available to all developers.
 This is a powerful concept, but it is also flexible.
-For application developers working on very specific or niche apps with less need for cross-application usability, Modules are not needed, and they can instead focus on custom Models.
-However, it is likely that developers who build on openly available \emph{standard} Modules will provide a more useful experience for their users.
+For application developers working on very specific or niche apps with less need for inter-application usability, Modules are not needed, and they can instead focus on custom Models.
+However, it is likely that developers who build on openly available \emph{standard} Modules will provide a more useful experience for their users, and will benefit from the \emph{network effects} \cite{shapiroInformationRulesStrategic1998} produced by many interoperable apps.
 
@@
 
-With such a simple interface to such a complex underlying process, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
+With these interface simplifications, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
 Here we draw inspiration from similar projects (e.g., OrbitDB \cite{markroberthendersonOrbitDBFieldManual2019}) which have made it much easier for developers familiar with centralized database systems to make the move to decentralized systems such as Threads.
@@
 
-At an abstract level, the Textile protocol provides a highly distributed framework for building shared, offline first, Stores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable.
+At an abstract level, the Textile protocol provides a distributed framework for building shared, offline first, Stores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable.
 Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS and ES (and related) literature to ``extend'' the Threads protocol with additional features and controls.
-Indeed, by designing our system around Actions, a Dispartcher, and generic Stores, we aim to make it easy to extend Threads in many different ways.
+Indeed, by designing our system around Events, a Dispatcher, and generic Stores, we make it easy to extend Threads in many different ways.
 Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility this design affords.
@@
 They can be used to rebuild the state of a view Store without having to query and re-play all previous Events.
-When a Snapshot is available, a Thread Peer can rebuild the state of a given view Store/Model by replaying any Events generated since the latest Snapshot (with a corresponding reduction in the total number of Events that need to be processed) using the Model's Reducer function.
+When a Snapshot is available, a Thread Peer can rebuild the state of a given view Store/Model by replaying only Events generated since the latest Snapshot using the Model's Reducer function.
 Multiple Peers processing the same Log could create a Snapshot every 1000 Events and be guaranteed to create the exact same Snapshot because each Peer's Event counts are identical\footnote{Assuming any network partitions are only short-lived (i.e., that peers are able to share events consistently).}.
@@
 
-Compaction can similarly speed up re-hydration of a Stores's state, and is useful when only the latest Event of a given type is required.
-In addition to being used as a starting point for initialization and recovery, compacted Stores can be used to \emph{replace} the Store from which they were created in order to free up local disk space.
-In practice, Compaction is a local-only operation (i.e., other Peers do not need to be aware that Compaction was performed).
+Compaction is a local-only operation (i.e., other Peers do not need to be aware that Compaction was performed) performed on an Event Store to free up local disk space.
+As a result, it can speed up re-hydration of a downstream Stores's state by reducing the number of Events that need to be processed.
+Compaction is useful when only the latest Event of a given type is required.
 

### 2019-09-27 draft.tex
draft: add blurb about redux front-end connection
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
+\subsubsection{Tighter Coupling with Front End Models}\label{sec:coupling}
+
+Implementing Threads internals (see Section \ref{sec:internals}) using similar patterns to common frontend workflows (e.g., Redux) presents opportunities for tighter coupling between ``backend'' logic and fontend views.
+This is a major advantage of tools such as reSolve\footnote{\url{https://reimagined.github.io/resolve/}}, where ``system changes can be reflected immediately [on the frontend], without the need to re-query the backend'' \cite{ereminReduxInspiredBackend2019}.
+Textile developers will develop frameworks to more directly expose the internals of Threads to frontend SDKs, making it possible to sync application state across and between apps and services on the Textile network.
+
 \ack{Thanks to our funders and the excellent Textile community for helpful suggestions, feature requests, and reviews.}

### 2019-09-27 draft.tex
updated el section
Signed-off-by: andrew <andrew@textile.io>

--- 2019-09-26 draft.tex
@@
 In order to proceed, there must be some way to deal with these conflicts.
-As mentioned previously (see Section \ref{sec:merkleclocks}), one way to avoid conflicts to some degree is to determine the total order of events in the shared log.
-In multi-writer systems, determining this causal order in all replicas is particularity challenging.
 In some cases (e.g, \texttt{ipfs-log} \cite{markroberthendersonOrbitDBFieldManual2019}), a Merkle-Clock can be used to induce ordering.
-However, this is only \emph{partially ordered}.
-That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:
+This approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy and therefor still requires, data-layer conflict resolution \cite{sanjuanMerkleCRDTs2019}:
 
@@
 
-In many cases, the total order of an event sequence is meaningful, and this type of deterministic merge strategy is not sufficient.
-A git merge highlights one such case, in which additional information is sometimes needed in order to resolve a line conflict.
-Ledger based transactions present a similar challenge where the absolute order of transactions between multiple peers really does matter.
+Therefore, solutions such as \texttt{ipfs-log} include a built-in CRDT to manage conflicts not resolved by the Merkle-Clock.
+This approach works for many cases, but the use of a deterministic merge strategy can be insufficient in cases where complicated data structures or long chains of missed events need to be reconciled.
+A git merge highlights one such example, in which a predetermined merge strategy can be used, but is not often the best choice in practice.
+Furthermore, a multi-writer log using a linked data format (e.g., a Merkle-Dag) in imperfect networking or storage environments can lead to suboptimal states where it may exceedingly costly (networking, storage, and computation) to regain consistency.
+
+A promising approach is one that can leverage the benefits of both a Merkle-Clock for events from any one peer and a less constrained ordering mechanism to combine events from all peers.
+In this more ideal setup, developers can more freely institute their own CRDTs or domain-specific conflict resolution strategies.
+Additionally, it more naturally supports use-cases where all peers contributing to a dataset may not be interested in following or replicating the events of all other peers.
 
@@
 % [!b] indicates that it will try to be on the bottom of the page
-\begin{figure}[!b]
+\begin{figure}[!t]
   \includegraphics[width=\linewidth]{figures/Event_Log.png}
@@
 \end{figure}
-
-Clearly, multi-writer systems with relaxed consistency \cite{brewerRobustDistributedSystems2000,gilbertBrewerConjectureFeasibility2002a} and imperfect information (e.g., caused by computational or network issues, both of which are common in p2p systems) will almost inevitably lead to situations where we need to decide how to reconcile a conflict.
-Some systems come with conflict resolution strategies ``built in'' (e.g., \texttt{ipfs-log} uses a built-in CRDT).
-However, this is not ideal, because it forces developers/users into a very specific usage pattern.
-It is particularly limiting in situations where developers might like to institute their own CRDT or domain-specific conflict resolution strategies.
 

### 2019-09-27 draft.tex
edit fixes
Signed-off-by: andrew <andrew@textile.io>

@@
 
-In multi-writer systems, conflicts arise as disparate peers end up producing conflicting state changes, or changes that end up out of sync.
+In multi-writer systems, conflicts arise as disparate peers end up producing disconnected state changes, or changes that end up out of sync.
 In order to proceed, there must be some way to deal with these conflicts.
 In some cases (e.g, \texttt{ipfs-log} \cite{markroberthendersonOrbitDBFieldManual2019}), a Merkle-Clock can be used to induce ordering.
-This approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy and therefor still requires, data-layer conflict resolution \cite{sanjuanMerkleCRDTs2019}:
+This approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:
 
@@
 Therefore, solutions such as \texttt{ipfs-log} include a built-in CRDT to manage conflicts not resolved by the Merkle-Clock.
-This approach works for many cases, but the use of a deterministic merge strategy can be insufficient in cases where complicated data structures or long chains of missed events need to be reconciled.
+This approach works for many cases, but the use of a deterministic resolution strategy can be insufficient in cases where complicated data structures or long chains of missed events need to be reconciled.
 A git merge highlights one such example, in which a predetermined merge strategy can be used, but is not often the best choice in practice.

### 2019-09-27 draft.tex
clarify
Signed-off-by: andrew <andrew@textile.io>

@@
 Therefore, solutions such as \texttt{ipfs-log} include a built-in CRDT to manage conflicts not resolved by the Merkle-Clock.
-This approach works for many cases, but the use of a deterministic resolution strategy can be insufficient in cases where complicated data structures or long chains of missed events need to be reconciled.
+This approach works for many cases, but the use of a deterministic resolution strategy can be insufficient in cases with complicated data structures or complicated network topologies.
 A git merge highlights one such example, in which a predetermined merge strategy can be used, but is not often the best choice in practice.

### 2019-09-27 draft.tex
nits: minor typos
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-Therefore, solutions such as \texttt{ipfs-log} include a built-in CRDT to manage conflicts not resolved by the Merkle-Clock.
+Solutions such as \texttt{ipfs-log} include a built-in CRDT to manage conflicts not resolved by the Merkle-Clock.
 This approach works for many cases, but the use of a deterministic resolution strategy can be insufficient in cases with complicated data structures or complicated network topologies.
 A git merge highlights one such example, in which a predetermined merge strategy can be used, but is not often the best choice in practice.
-Furthermore, a multi-writer log using a linked data format (e.g., a Merkle-Dag) in imperfect networking or storage environments can lead to suboptimal states where it may exceedingly costly (networking, storage, and computation) to regain consistency.
+Furthermore, a multi-writer log using a linked data format (e.g., a Merkle-Dag) in imperfect networking or storage environments can lead to suboptimal states where it may be exceedingly costly (networking, storage, and computation) to regain consistency.
 

### 2019-09-27 draft.tex
updates
Signed-off-by: andrew <andrew@textile.io>

@@
 
-\title[An event-sourced and decentralized database with IPLD and materialized views]{An event-sourced and decentralized database with IPLD and materialized views}
+\title[A protocol and event-sourced database for decentralized user-siloed data]{A protocol and event-sourced database for decentralized user-siloed data}
 \author{Pick}
@@
 \author{Sutula}
+\affiliation{www.textile.io} \email{research@textile.io}
+\author{?Others?}
+\affiliation{other?} \email{research@textile.io}
 \author{Hill}
@@
 This provides an intuitive, singular access point to the \emph{local} system, while also leaving room for updates via \emph{external} peer-events.
+In practice, the Store is a lightweight interface that can be implemented by one of many database management systems (\ref{appx:DBMS}).
 
@@
 \appendix 
-
 % [!p] forces it to just take the whole page
@@
 
+\onecolumn
+\section{Database Management Systems (DBMS)} \label{appx:DBMS}
+
+\begin{enumerate}
+	\item Relational
+    \begin{enumerate}
+        \item MariaDB \url{https://mariadb.org/}
+    	\item PostgreSQL \url{https://www.postgresql.org/}
+        \item SQLite \url{https://www.sqlite.org}
+    \end{enumerate}
+    \item Key value stores
+    \begin{enumerate}
+        \item Dynamo \url{https://aws.amazon.com/dynamodb/}
+    	\item LevelDB \url{https://github.com/google/leveldb}
+        \item Redis \url{https://redis.io/}
+    \end{enumerate}
+    \item Document stores
+    \begin{enumerate}
+        \item CouchDB \url{http://couchdb.apache.org/}
+    	\item MongoDB \url{https://www.mongodb.com/}
+        \item RethinkDB \url{https://rethinkdb.com/}
+    \end{enumerate}
+\end{enumerate}
+
 \end{document}

### 2019-09-27 draft.tex
updates
Signed-off-by: andrew <andrew@textile.io>
more
Signed-off-by: andrew <andrew@textile.io>

@@
 \affiliation{www.textile.io} \email{research@textile.io}
-\author{?Others?}
+\author{?Gozalishivili others?}
 \affiliation{other?} \email{research@textile.io}
@@
 This provides an intuitive, singular access point to the \emph{local} system, while also leaving room for updates via \emph{external} peer-events.
-In practice, the Store is a lightweight interface that can be implemented by one of many database management systems (\ref{appx:DBMS}).
+In practice, the Store is a lightweight interface that can be implemented by one of many database management systems (\ref{appx:DBMS}) providing end users/developers the greatest flexibility.
 

### 2019-09-27 draft.tex
hyp
Signed-off-by: andrew <andrew@textile.io>
title fix
Signed-off-by: andrew <andrew@textile.io>
review: incremental updates
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-\title[A protocol and event-sourced database for decentralized user-siloed data]{A protocol and event-sourced database for decentralized user-siloed data}
+\title[Textile Threads]{A protocol and event-sourced database for decentralized user-siloed data}
 \author{Pick}
@@
 Compared to their predecessors, modern cloud-based apps and services provide an extremely high level of convenience.
-Users can entirely forget about the risk of data loss and enjoy seamless access to their apps across multiple devices.
+Users can completely forget about data management, and enjoy seamless access to their apps across multiple devices.
 This convenience is now expected, but has come at the cost of additional consequences for users.
-One such consequence is that many of the same development patterns that bring convenience (e.g., single sign-on, minimal encryption, centralized servers, and databases) also enable or even require data hoarding by the apps.
-While collecting large amounts of users' data can create value for the companies building apps (e.g., app telemetry, predictive tools, or even new revenue streams), that value flows mostly in one direction; Apps collect, process and benefit from a user's private data, but users rarely have access to the original data or the new insights that come from it.
-Additionally,  the data may not be readily accessible to other apps and services.
-This is app-siloed data.
-
-While companies collecting user data may not be itself a significant problem, it has been shown that over time company incentives often shift from providing value to users to extracting value from users \cite{dixonWhyDecentralizationMatters2018}.
-When this incentive shift happens for companies that have also been hoarding user data, that data may be a new source of value or even revenue for the company.
+One such consequence is that many of the same development patterns that bring convenience (e.g., single sign-on, minimal encryption, centralized servers and databases) also enable, or even require data hoarding by apps.
+While collecting large amounts of users' data can create value for the companies building apps (e.g., app telemetry, predictive tools, or even new revenue streams), that value flows mostly in one direction: apps collect, process and benefit from a user's private data, but users rarely have access to the original data or the new insights that come from it.
+Additionally, the data is generally not accessible to other apps and services.
+This is \emph{app-siloed} data.
+
+While companies collecting user data may not itself be a significant problem, it has been shown that, over time company incentives often shift from \emph{providing} value to users, to \emph{extracting} value from users \cite{dixonWhyDecentralizationMatters2018}.
+When this incentive shift happens for companies that have also been hoarding user data, that data may be a new source of value/revenue for the company.
 It may be possible to stop this trend through extreme privacy measures, government intervention/legislation, or by boycotting companies that collect any form of data.
 Ideally, there is an alternative approach that allows individuals to capture the value of their data and still allow developers to build new interfaces and experience on top of this data.
-This approach is called user-siloed data and it fundamentally separates apps from their user's data.
+This approach is called \emph{user-siloed} data and it fundamentally separates apps from their users' data.
 
 One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data.
-Other projects have identified app silo-ed data as a problem \cite{bernersLee_readwrite_2013,de_montjoye_trusted_2012}, and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}.
-However, none so far have provided a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data.
-This is the key to building great applications on user-siloed data.
+Other projects have identified app-siloed data as a problem \cite{bernersLee_readwrite_2013,de_montjoye_trusted_2012}, and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}.
+However, none so far have addressed the problem directly, by providing a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data.
 
 In this paper, we study existing technologies that could be used to build a network for user-siloed data.
-We outline six challenge areas: flexible data format, efficient synchronization, conflict resolution, access-control, scalable storage, and network communication.
-Based on our findings, we propose a novel architecture for event sourcing with Interplanetary Linked Data (IPLD), designed to store, share, and host user-sioled datasets at scale.
-Our proposed design leverages new and existing protocols to solve major challenges involved with building a secure and distributed network for user data while at the same time providing the flexibility and scalability required by today's applications.
-
-\section{Background}
-\label{sec:Background}
+We outline six challenge areas: flexible data formats, efficient synchronization, conflict resolution, access-control, scalable storage, and network communication.
+Based on our findings, we propose a novel architecture for event sourcing (ES) with Interplanetary Linked Data (IPLD), that is designed to store, share, and host user-siloed datasets at scale.
+Our proposed design leverages new and existing protocols to solve major challenges with building a secure and distributed network for user data while at the same time providing the flexibility and scalability required by today's apps.
+
+\section{Background}\label{sec:Background}
 
 We now describe some of the technologies and concepts motivating the design of our novel decentralized database system.
-We highlight some of the advantages and lessons that can be learned from event-sourcing and discuss drawbacks to using these approaches in decentralized systems.
-We provide an overview of some important technologies related to IPFS that make it possible to rethink event sourcing in a decentralized network.
-Finally, we cover challenges of security and access control on an open and decentralized network and discuss how they are used in popular databases external to IPFS.
+We highlight some of the advantages and lessons learned from event sourcing (ES) and discuss drawbacks to using these approaches in decentralized systems.
+We provide an overview of some important technologies related to the Interplanetary File System (IPFS) that make it possible to rethink ES in a decentralized network.
+Finally, we cover challenges to security and access control on an open and decentralized network, and discuss how they are used in popular databases outside the IPFS context.
 
 \subsection{Data Synchronization}
-
-\subsubsection{CQRS, Event Sourcing, and Logs}\label{sec:cqrs}
 
@@
 To model realistic systems, this type of framework often requires complex techniques for mapping data between domain models and database tables, where the same data model is used to both query and update a database.
-A powerful alternative is to use a set of append only logs to model the state of an object simply by applying its change sequence in the correct order.
+A powerful alternative is to use a set of append-only logs to model the state of an object simply by applying its change sequence in the correct order.
 This concept can be expressed succinctly by the state machine approach \cite{schneiderImplementingFaulttolerantServices1990}: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state.
@@
 
-In modern applications it is critical to have reliable mechanisms for publishing updates and events (i.e., to support event-driven architectures), scalability (optimized write and read operations), forward-compatible application updates (e.g., code changes, retroactive events, and others), auditing systems, etc.
-To support such requirements, developers have begun to utilize event sourcing and command query responsibility segregation (CQRS) patterns \cite{bettsExploringCQRSEvent2013}, relying on append only logs to support immutable state histories.
-Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate event sourcing and CQRS-based applications, including Event Store \cite{EventStore}, Apache Kafaka \cite{ApacheKafka} and Samza \cite{ApacheSamza}, among others (see \cite{kleppmannDesigningDataintensiveApplications2017}).
-
-\begin{definition} (Command query responsibility segregation).
+\subsubsection{CQRS, Event Sourcing, and Logs}\label{sec:cqrs}
+
+For most apps, it is critical to have reliable mechanisms for publishing updates and events (i.e., to support event-driven architectures), scalability (optimized write and read operations), forward-compatible application updates (e.g., code changes, retroactive events), auditing systems, etc.
+To support such requirements, developers have begun to utilize event sourcing and command query responsibility segregation (CQRS) patterns \cite{bettsExploringCQRSEvent2013}, relying on append-only logs to support immutable state histories.
+Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate ES and CQRS-based designs, including Event Store \cite{EventStore}, Apache Kafaka \cite{ApacheKafka} and Samza \cite{ApacheSamza}, among others \cite{kleppmannDesigningDataintensiveApplications2017}.
+
+\begin{definition}
+  (CQRS).
 Command query responsibility segregation or CQRS is a design pattern whereby reads and writes are separated into different models, using commands to write data, and queries to read data \cite{martinfowlerCQRS2011}.
@@
 
-\begin{definition} (Event sourcing). Event sourcing (ES) is a design pattern for persisting the state of an application as an append-only log of state-changing events. \end{definition}
-
-ES is particularly useful in contexts where system components are distributed or decentralized, such as in decentralized applications (DApps).
-This is because the design of systems that use event sourcing often necessitate infrastructure for brokering messages between loosely coupled software components and services.
-This feature will become increasingly important as we outline background concepts related to data synchronization in the following sections.
+\begin{definition}
+  (ES). Event sourcing or ES is a design pattern for persisting the state of an application as an append-only log of state-changing events.
+\end{definition}
 
@@
     \end{enumerate}
-    \item Key value stores
+    \item Key-value stores
     \begin{enumerate}

### 2019-09-27 draft.tex
update: additional nits
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 
-\title[Textile Threads]{A protocol and event-sourced database for decentralized user-siloed data}
+\title[Textile Threads]{A protocol \& event-sourced database for decentralized user-siloed data}
 \author{Pick}
@@
 
-A key principal of ES and append only logs is that all changes to application state are stored as a sequence of events.
+A key principal of ES and append-only logs is that all changes to application state are stored as a sequence of events.
 Because any given state is simply the result of a series of atomic updates, the log can be used to reconstruct past states or process retroactive updates \cite{fowlerEventSourcing}.
-The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (Section  \ref{sec:ViewsProjections}).
+The same principal means a log can be viewed as a mechanism to support an infinite number of valid state interpretations (see Section  \ref{sec:ViewsProjections}).
 In other words, with minimal conformity, a single log can model multiple application states \cite{microsoftcorporationAzureApplicationArchitecture}.
@@
 \subsubsection{Views \& Projections} \label{sec:ViewsProjections}
-
-\begin{definition}  (View).
-A (typically highly de-normalized) read-only model of the data.
-Views are tailored to the interfaces and display requirements of the application, which helps to maximize both display and query performance.
-Views that are backed by a database or filesystem-optimized access are referred to as materialized views.
-\end{definition}
-
-\begin{definition} (Projection).
-An event handler and corresponding reducer/fold function used to build and maintain a view from a set of (filtered) events.
-While projections may lead to the generation of new events, their reducer should be a pure function (see \cite{eventsourceEventSourcingBasics} and \cite{reduxGettingStartedRedux} for examples from existing systems).
-\end{definition}
 
@@
 It allows developers to define views into the underlying data that are best suited for the use case or user interface they are building.
-Multiple views can be built from the same underlying event log, and they can be quite different from one another.
-
-Views themselves are enabled by projections\footnote{Terminology in this section may differ from some other examples of ES and CQRS patterns, but reflects the underlying architecture and designs that the Textile team will elaborate on in \ref{sec:ThreadsProtocol}}.
-Projections can be thought of as transformations or reducers that are applied to each event in a stream of events.
-They update the data backing the views, be this in memory, or persisted to a database.
+Multiple (potentially very different) views can be built from the same underlying event log.
+
+\begin{definition}
+  (View).
+A (typically highly de-normalized) read-only model of event data.
+Views are tailored to the requirements of the application, which helps to maximize display and query performance.
+Views that are backed by a database or filesystem-optimized access are referred to as materialized views.
+\end{definition}
+
+\begin{definition}
+  (Projection).
+An event handler and corresponding reducer/fold function used to build and maintain a view from a set of (filtered) events.
+While projections may lead to the generation of new events, their reducer should be a pure function.
+\end{definition}
+
+Views themselves are enabled by projections\footnote{Terminology in this section may differ from some other examples of ES and CQRS patterns, but reflects the underlying architecture and designs that the Textile team will elaborate on in Section \ref{sec:ThreadsProtocol}}, which can be thought of as transformations that are applied to each event in a stream.
+They update the data backing the views, be this in memory or persisted to a database.
 In a distributed setting, it may be necessary for projections to define and operate as eventually consistent data structures, to ensure all peers operating on the same stream of events have a consistent representation of the data.
@@
 
-The CAP theorem \cite{brewerRobustDistributedSystems2000,gilbertBrewerConjectureFeasibility2002a} states that a distributed database can guarantee only two of the following three promises at the same time: consistency (i.e., that every read receives the most recent write or an error), availability (i.e., that every request receives a (possibly out-of-date) non-error response, and partition tolerance (i.e., that the system continues to operate despite an arbitrary number of messages being dropped (or delayed) by the network).
-As such, many distributed systems are now designed to provide availability and partition tolerance by trading consistency for eventual consistency.
-Eventual consistency allows state replicas to diverge temporarily, but eventually arrive back to the same state again.
+The CAP theorem \cite{brewerRobustDistributedSystems2000,gilbertBrewerConjectureFeasibility2002a} states that a distributed database can guarantee only two of the following three promises at the same time: consistency (i.e., that every read receives the most recent write or an error), availability (i.e., that every request receives a [possibly out-of-date] non-error response), and partition tolerance (i.e., that the system continues to operate despite an arbitrary number of messages being dropped [or delayed] by the network).
+As such, many distributed systems are now designed to provide availability and partition tolerance by trading consistency for \emph{eventual} consistency.
+Eventual consistency allows state replicas to diverge temporarily, but eventually arrive back to the same state.
 While an active area of research, designing systems with provable eventual consistency guarantees remains challenging \cite{shapiroComprehensiveStudyConvergent2011a,almeidaDeltaStateReplicated2018}.
 
-\begin{definition} (CRDT).
+\begin{definition}
+  (CRDT).
 A conflict-free replicated data type (CRDT) assures eventual consistency through optimistic replication (i.e.
 all new updates are allowed) and eventual merging.
-CRDTs rely on data structures that are mathematically guaranteed to resolve concurrent updates the same way regardless of the order those events were received.
+CRDTs rely on data structures that are mathematically guaranteed to resolve concurrent updates the same way regardless of the order in which those events were received.
 \end{definition} \label{def:CRDT}
@@
 Two well-documented categories of solutions include logs (Definition  \ref{def:Logs}), sequences of deterministic changes, and CRDTs (Definition  \ref{def:CRDT}).
-Logs work best when there is only a single-writer or complete event causality is known.
 In many distributed systems with multiple log writers, a minimum requirement for synchronization through logs is that the essential order of events is respected and can be determined \cite{schwarzDetectingCausalRelationships1994,katzInterleavingSetTemporal1990}.
 For these cases, logical clocks are a useful tool for eventual consistency and total ordering \cite{kulkarniLogicalPhysicalClocks2014}.
-However, some scenarios (e.g., temporarily missing events, or ambiguous order) can force a replica into a state that cannot be later resolved without costly recalculation.
-In specific cases, CRDTs can provide an alternative to log-based consensus (see below).
+However, some scenarios (e.g., temporarily missing events or ambiguous order) can force a replica into a state that cannot be later resolved without costly recalculation.
+In specific cases, CRDTs can provide an alternative to log-based consensus (see Section \ref{sec:CRDTs}).
 
@@
 
-In a distributed system, where multiple machines, each with an independent clock, are creating events, local timestamps can't be used to construct global event causality.
+In a distributed system with multiple machines (each with an independent clock) each creating events, local timestamps can't be used to construct global event causality.
 Machine clocks are never perfectly synchronized \cite{lamportTimeClocksOrdering1978}, meaning that one machine's concept of "now" is not necessarily the same as another machine's.

### 2019-09-27 draft.tex

@@
 How a system provides eventual consistency is often decided based on the intended use of the system.
-Two well-documented categories of solutions include logs (Definition  \ref{def:Logs}), sequences of deterministic changes, and CRDTs (Definition  \ref{def:CRDT}).
-In many distributed systems with multiple log writers, a minimum requirement for synchronization through logs is that the essential order of events is respected and can be determined \cite{schwarzDetectingCausalRelationships1994,katzInterleavingSetTemporal1990}.
+Two well-documented categories of solutions include logs (Definition  \ref{def:Logs}) and CRDTs (Definition  \ref{def:CRDT}).
+
+A common minimum requirement for log synchronization across multiple peers is that the essential order of events is respected and/or can be determined \cite{schwarzDetectingCausalRelationships1994,katzInterleavingSetTemporal1990}.
 For these cases, logical clocks are a useful tool for eventual consistency and total ordering \cite{kulkarniLogicalPhysicalClocks2014}.
@@
 
-In a distributed system with multiple machines (each with an independent clock) each creating events, local timestamps can't be used to construct global event causality.
-Machine clocks are never perfectly synchronized \cite{lamportTimeClocksOrdering1978}, meaning that one machine's concept of "now" is not necessarily the same as another machine's.
+In a distributed system with multiple peers (each with an independent clock) creating events, local timestamps can't be used to determine ``global'' event causality.
+Machine clocks are never perfectly synchronized \cite{lamportTimeClocksOrdering1978}, meaning that one peer's concept of ``now'' is not necessarily the same as another.
 Machine speed, network speed, and other factors compound the issue.
 For this reason, simple wall-clock time does not provide a sufficient notion of order in a distributed system.
-Alternatives to wall-clock time exist to help distributed systems achieve eventual consistency.
-Examples include various logical clocks (Lamport \cite{lamportTimeClocksOrdering1978} Schwartz \cite{schwarzDetectingCausalRelationships1994}, and Bloom \cite{ramabajaBloomClock2019}, Hybrid variants \cite{kulkarniLogicalPhysicalClocks2014}, etc), which use counter-based time-stamping to provide partial ordering.
-
-Cryptographically linked events can also represent a clock (See Section \ref{sec:merkleclocks}).
-One such example is called the Merkle-Clock \cite{sanjuanMerkleCRDTs2019}.
-The Merkle-Clock relies on properties of a Merkle-DAG to provide strict partial ordering between events, which has limitations \cite[sec.
+Alternatives to wall-clock time exist to help achieve eventual consistency.
+Examples include various logical clocks (Lamport \cite{lamportTimeClocksOrdering1978} Schwartz \cite{schwarzDetectingCausalRelationships1994}, Bloom \cite{ramabajaBloomClock2019}, and Hybrid variants \cite{kulkarniLogicalPhysicalClocks2014}, etc.), which use ``counter''-based time-stamps to provide partial ordering.
+
+Cryptographically linked events can also represent a clock (see Section \ref{sec:merkleclocks}).
+One such example is called the Merkle-Clock \cite{sanjuanMerkleCRDTs2019}, which relies on properties of a Merkle-DAG to provide strict partial ordering between events.
+This approach does have its limitations however \cite[sec.
 4.3]{sanjuanMerkleCRDTs2019}:
@@
 
-In distributed models that use clock-based ordering as described above, a replica can arrive in a state that will require conflict resolution through consensus or rollback.
-Those operations can be expensive, having adverse effects on the scalability of distributed systems, especially ones attempting to synchronize an application state.
-
 CRDTs (Definition \ref{def:CRDT}) are one way to achieve strong eventual consistency, where once all replicas have received the same events, they will arrive at the same final state, \emph{regardless of ordering}\footnote{Though non-commutative CRDTs may require a specific ordering of events in certain cases \cite{sanjuanMerkleCRDTs2019}}.
@@
 
-\subsection{Content-based addressing}
-
-Internet application architecture today is often designed as a system of clients (users) communicating to endpoints (hosts).
-Communication between clients and endpoints (servers) happens via the TCP/IP protocol stack and therefore largely relies on a mechanism referred to as location-based addressing.
+\subsection{Content-based Addressing}
+
+Internet application architecture today is often designed as a system of clients (users) communicating to endpoints (hosts or servers).
+Communication between clients and endpoints usually happens via the TCP/IP protocol stack and depends on \emph{location-based addressing}.
 Location-based addressing, where the client makes a request that is routed to a specific endpoint based on prior knowledge (e.g., the domain name or IP address), works relatively well for many use-cases.
-However, there are many reasons why addressing content by location is problematic, such as duplication of storage, inefficient use of bandwidth, link rot, centralized control, and authentication issues.
-An alternative to location addressing, called content addressing may provide a solution to those problems.
-Content addressing is where the content itself is used to create an address and retrieve that content from a network \cite{mortContentBasedAddressing2012}.
-
-\subsubsection{IPFS \& Content-based addressing}
-
-There are a number of systems that utilize content addressing to access content and information (e.g.
-Git, IPFS, Perkeep, Tahoe-LAFS), and there is an active body of literature covering its design and implementation \cite{benetIPFSContentAddressed2014,selimi_tahoeLafs_2014,rhea_fast_2008}.
-The Interplanetary File System (IPFS) --- which is a set of protocols to create a content-addressed, peer-to-peer filesystem \cite{benetIPFSContentAddressed2014} --- is one such system.
-In IPFS, the address for any content is determined based on a cryptographic hash of the content itself.
-In practice, the IPFS Content IDentifier (CID) is a multihash, which is a self-describing ``protocol for differentiating outputs from various well-established cryptographic hash functions, addressing size + encoding considerations\footnote{\url{https://multiformats.io/}}''.
-That addressing system confers several benefits to the network, including tamper resistance (we can be confident that a given piece of content has not been modified en route if its hash matches what we were expected/requested) and de-duplication (because the same content from different peers will produce the same hash address).
-Additionally, IPFS content addresses are immutable and universally unique.
+However, there are many reasons why addressing content by location is problematic, such as duplication of storage, inefficient use of bandwidth, invalid/dead links (link rot), centralized control, and authentication issues.
+
+An alternative to location addressing, called \emph{content-based addressing}, may provide a solution to many of the problems associated with location-based addressing.
+Content-based addressing is where the content itself is used to create an address which is then used to retrieve said content from the network \cite{mortContentBasedAddressing2012}.
+
+\subsubsection{IPFS \& Content-based Addressing}
+
+There are a number of systems that utilize content-based addressing to access content and information (e.g.
+Git, IPFS, Perkeep, Tahoe-LAFS, etc), and there is an active body of literature covering its design and implementation \cite{benetIPFSContentAddressed2014,selimi_tahoeLafs_2014,rhea_fast_2008}.
+The Interplanetary File System (IPFS) --- which is a set of protocols to create a content-addressed, peer-to-peer (P2P) filesystem --- is one such system \cite{benetIPFSContentAddressed2014}.
+In IPFS, the address for any piece of content is determined based on a cryptographic hash of the content itself.
+In practice, an IPFS Content IDentifier (CID) is a multihash, which is a self-describing ``protocol for differentiating outputs from various well-established cryptographic hash functions, addressing size [and] encoding considerations'' \cite{protocollabsMultihash}.
+That addressing system confers several benefits to the network, including tamper resistance (i.e., a given piece of content has not been modified en route if its hash matches what we were expecting/requested) and de-duplication (i.e., the same content from different peers will produce the same hash address).
+Additionally, IPFS content-based addresses are immutable and universally unique.
 
@@
   \caption{The cryptographic hash of content is used to make a request to the network of IPFS peers.
-Using the built-in routing mechanisms and the distributed hash table, a peer(s) hosting the requested content is identified and content is returned.
-}
+Using the built-in routing mechanisms and a distributed hash table (DHT), peers hosting the requested content are identified and content is returned.}
   \label{fig:contentaddressing}
@@
 
-While content addressing doesn't tell you how to get a file, IPFS (via libp2p\footnote{\url{https://libp2p.io/}}) provides a system for moving files across the network.
+While content-based addressing doesn't dictate to a peer \emph{how} to get a piece of content, IPFS (via libp2p\footnote{\url{https://libp2p.io/}}) does provide a system for moving content across the network.
 On the IPFS network, a client who wants specific content requests the CID from the network of IPFS hosts.
 The client's request is routed to the first host capable of fulfilling the request (i.e., the first host that is actively storing the content behind the given CID).
-The IPFS network can be seen as a distributed file system, with all of the benefits that come with this type of file system design.
+The IPFS network can be seen as a distributed file system, with many of the benefits that come with this type of system design.
 
@@
 
-As discussed above, IPFS uses the cryptographic hash of a given piece of content to define its content address (see \cite{benetIPFSContentAddressed2014} for details on this process).
+As discussed previously, IPFS uses the cryptographic hash of a given piece of content to define its content-based address (see \cite{benetIPFSContentAddressed2014} for details on this process).
 However, in order to provide standards for accessing content-addressable data (on the web or elsewhere), it is necessary to define a common format or specification.
-In IPFS (and others \cite[e.g.,]{protocollabsFilecoinDecentralizedStorage2017} MORE NEEDED), this common data format is called Interplanetary Linked Data (IPLD)\footnote{\url{https://ipld.io/}}.
-As the name suggests, IPLD is based on principals of linked data \cite{berners-leeLinkedData2009,bizerLinkedDataStory2011} with the added capabilities of a content-addressing storage network.
-
-IPLD is used to publish linked data (subject, predicate, object triples in the linked data world \cite{heathLinkedDataEvolving2011}) spread across different hosts, and where everything (e.g., entities, predicates, data sources) uses content addresses as unique identifiers.
-To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph\footnote{Other examples of Dags include the Bitcoin Blockchain or a Git version history.}.
+In IPFS and other systems [\cite[e.g.,][]{protocollabsFilecoinDecentralizedStorage2017} \textcolor{red}{MORE NEEDED}], this common data format is called Interplanetary Linked Data (IPLD)\footnote{\url{https://ipld.io/}}.
+As the name suggests, IPLD is based on principals of linked data \cite{berners-leeLinkedData2009,bizerLinkedDataStory2011} with the added capabilities of a content-based addressing storage network.
+
+IPLD is used to represent linked data that is spread across different ``hosts'', such that everything (e.g., entities, predicates, data sources) \cite{heathLinkedDataEvolving2011} uses content-based addresses as unique identifiers.
+To form its structure, IPLD implements a Merkle DAG, or directed acyclic graph\footnote{Other examples of DAGs include the Bitcoin blockchain or a Git version history.}.
 This allows all hash-linked data structures to be treated using a unified data model, analogous to linked data in the Semantic Web sense \cite{brendanobrienDeterministicQueryingDistributed2017}.
@@
 
-A Merkle-Clock is a Merkle-DAG that represents a sequence of events.
-In other words, a Merkle-Clock is an append-only log \cite{sanjuanMerkleCRDTs2019}.
+A Merkle-Clock is a Merkle-DAG that represents a sequence of events, or a log \cite{sanjuanMerkleCRDTs2019}.
 When implemented on IPFS (or an equivalent network where content can be cryptographically addressed and fetched), Merkle-Clocks provide a number of benefits for data synchronization between replicas \cite[sec.
@@
 \begin{enumerate}
-\item Broadcasting the Merkle-Clock requires broadcasting only the current root CID.
-The whole Clock is unambiguously identified by the CID of its root and its full DAG can be walked down from it as needed.
-\item The immutable nature of a Merkle-DAG allows every replica to perform quick comparisons and fetch only those nodes that it does not already have.
+  \item Sharing the Merkle-Clock can be done using only the \emph{root} CID. The whole Clock is unambiguously identified by the CID of its root, and its full structure can be traversed as needed.
+  \item The immutable nature of a Merkle-DAG allows every replica to perform quick comparisons, and fetch only those nodes (leaves) that it is missing.
   \item Merkle-DAG nodes are self-verified and immune to corruption and tampering. They can be fetched from any source willing to provide them, trusted or not.
@@
 
-On the downside, and as discussed above (Section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots).
-For example, in Figure \ref{fig:merkledag}, two replicas (top row, bottom row) are attempting to write (left to right) events to the same Merkle-Clock.
+On the downside (see also Section \ref{sec:LogicalClocks}), a Merkle-Clock cannot order divergent heads (or roots).
+For example, in Figure \ref{fig:merkledag}, two replicas (left and right columns) are attempting to write (top to bottom) events to the same Merkle-Clock.
 After the first replica writes event A, the second writes event A' and properly links to A.
-At that point, the two replicas stop receiving events from one another.
-To a third replica that does continue to receive events, there would now be two independent heads, 1 and 1'.
+At that point, perhaps the two replicas stop receiving events from one another.
+To a third replica (not pictured) that does continue to receive events, there would now be two independent heads, 1 and 1'.
 For the third replica, resolving these two logs of events may be costly (many updates happened since the last common node) or impossible (parts of the chain may not be available on the network).
-See Section  \ref{sec:EventualConsistency}.
 
 In order to reduce the likelihood of divergent heads, all replicas should be perfectly connected and be able to fetch all events and linkages in the Merkle-Clock.
-On real networks with many often offline replicas (mobile and IoT devices, laptops, etc.), these conditions are rarely met.
-Based on these observations, using a single Merkle-Clock to synchronize replicas can be problematic.
+On real-world networks with many often replicas that are often offline (mobile and Internet of things (IoT) devices, laptops, etc.), these conditions are rarely met, making the use of a single Merkle-Clock to synchronize replicas problematic.
 
@@
 So far we have primarily discussed the mechanics of creating or linking content in a series of updates.
-Now we will overview some common networking tools for connecting distributed peers who aim to maintain replicas of a shared database.
+Now we will overview some common networking tools for connecting distributed peers who aim to maintain replicas of a shared state.
 This could be any decentralized network of interacting entities (e.g., cloud servers, IoT devices, botnets, sensor networks, mobile apps, etc) collectively updating a shared state.
-IPFS contains a collection of protocols and systems to help address the networking needs required by different use-cases.
-That is to say no matter what type of device we are talking about --- be it a phone, desktop computer, browser, or Internet-enabled appliance --- it should be able to communicate with other devices.
-However, each of the networking approaches described below comes with strengths and weaknesses when used to synchronize data.
+IPFS contains a collection of protocols and systems to help address the networking needs of different use-cases and devices --- be it a phone, desktop computer, browser, or Internet-enabled appliance.
 
@@
 The libp2p project provides a robust protocol communication stack.
-IPFS and a growing list of other projects (Polkadot, Ethereum 2.0, Substrate, FileCoin, OpenBazzar, Keep, etc) are building on top of libp2p.
-Libp2p solves a number of challenges that are distinct to peer-to-peer (p2p) networks.
+IPFS and a growing list of other projects (e.g., Polkadot, Ethereum 2.0, Substrate, FileCoin, OpenBazzar, Keep, etc) are building on top of libp2p.
+Libp2p solves a number of challenges that are distinct to P2P networks.
 A comprehensive coverage of networking issues in P2P systems is out of the scope for this paper, however, some core challenges that libp2p helps to address include network address translator \cite{srisureshIPNetworkAddress} (NAT) traversal, peer discovery and handshake protocols, and even encryption and transport security --- libp2p supports both un-encrypted (e.g.
-TCP, UDP) and encrypted protocols (e.g.
-TLS, Noise) --- among others.
-Libp2p uses the concept of a multiaddress to address peers on a network, which essentially models network addresses as arbitrary encapsulations of protocols \cite{multiformatsMultiaddr}.
-In addition to ``transport layer'' modules, libp2p provides several tools for sharing and/or disseminating data over a p2p network.
+TCP, UDP) and encrypted (e.g.
+TLS, Noise) protocols --- among others.
+Libp2p uses the concept of a multiaddress to address peers on a network, which essentially models network addresses as arbitrary encapsulations of protocols \cite{protocollabsMultiaddr}.
+In addition to ``transport layer'' modules, libp2p provides several tools for sharing and/or disseminating data over a P2P network.
 
@@
 
-One of the most commonly used p2p distribution layers built on libp2p, is its Pubsub (or publish-subscribe) system.
+One of the most commonly used P2P distribution layers built on libp2p, is its Pubsub (or publish-subscribe) system.
 Pubsub is a standard messaging pattern where the publishers don't know who, if anyone, will subscribe to a given topic.
 \emph{Publishers} send messages on a given topic or category, and \emph{Subscribers} receive only messages on a give topic to which they are subscribed.
-Libp2p's pubsub module can be configured to utilize a \emph{floodsub} protocol --- which floods the network with messages, and peers are required to ignore messages they are not interested in --- or \emph{gossipsub} --- which is a proximity-aware epidemic pubsub, where peers communicate with proximal peers, and messages can be routed more efficiently.
-In those implementations, there is a benefit to using pubsub in that no direct connection between publishers and subscribers is required.
-
-Another benefit to using pubsub is the ability to publish topical sequences of updates to multiple recipients.
+Libp2p's Pubsub module can be configured to utilize a \emph{floodsub} protocol --- which floods the network with messages, and peers are required to ignore messages in which they are not interested --- or \emph{gossipsub} --- which is a proximity-aware epidemic Pubsub system, where peers communicate with proximal peers, and messages can be routed more efficiently.
+In those implementations, there is a benefit to using Pubsub in that no direct connection between publishers and subscribers is required.
+
+Another benefit to using Pubsub is the ability to publish topical sequences of updates to multiple recipients.
 Like libp2p, encryption is a separate concern and often added in steps prior to data transmission.
-However, like libp2p, pubsub doesn't offer any simple solutions for transferring encryption keys (beyond public keys), synchronizing datasets across peers (i.e.
+However, also like libp2p, Pubsub doesn't offer any simple solutions for transferring encryption keys (beyond public keys), synchronizing datasets across peers (i.e.
 they aren't databases), or enforcing any measures for access control (e.g.
 anyone subscribed to a topic can also author updates on that topic).
-To solve some of those challenges, some systems introduce message echoing and other partial solutions.
-However, it makes more sense to use pubsub and libp2p as \emph{building blocks} in systems that can effectively solve these issues, by choosing multi-modal communication strategies or leveraging tools such as deferred routing (e.g.
+To solve some of these challenges, some systems introduce message echoing and other partial solutions.
+However, it makes more sense to use Pubsub and libp2p as \emph{building blocks} in systems that can effectively solve these issues, such as using multi-modal communication strategies or leveraging tools such as deferred routing (e.g.
 inboxing) for greater tolerance of missed messages.
@@
 
-Pubsub and libp2p have so far only dealt with \emph{push-based} transfer of data, but IPFS also offers a useful technology for hosting \emph{pull/request} based data endpoints called, IPNS.
+Our discussion of Pubsub and libp2p has so far only dealt with \emph{push-based} transfer of data; but IPFS also offers a useful technology for hosting \emph{pull-/request-}based data endpoints based on an Interplanetary Name System (IPNS).
 IPNS aims to address the challenge of mutable data within IPFS.
-IPNS relies on a global namespace (shared by participating IPFS peers) based on Public Key Infrastructure.
+It relies on a global namespace (shared by all participating IPFS peers) based on Public Key Infrastructure (PKI).
 By using IPNS, a content creator generates a new address in the global namespace and points that address to an endpoint (e.g.
 a CID).
-Using their private key, a content creator can update what static route the IPNS address refers to.
-IPNS isn't only useful for creating static addresses that can point to content addresses, IPNS is compatible with a naming system external from IPFS, such as DNS, onion, or bit addresses.
-However, many use-cases that require highly mutable data, require rapid availability of updates, or want flexible multi-party access control may not yet be suitable for IPNS.
-Taken together, libp2p, pubsub, IPNS, and IPFS more generally provide a useful toolkit for building robust abstractions to deliver fast, scalable, data synchronization in the decentralized network.
+Using their private key, a content creator can update the static route to which the IPNS address refers.
+But IPNS isn't only useful for creating static addresses pointing to content-based addresses; it is also compatible with external naming systems such as DNS, Onion, or bit addresses.
+
+Unfortunately, many use-cases that require highly mutable data, also require rapid availability of updates, or need flexible multi-party access control, which is not currently viable using IPNS alone.
+However, taken together, libp2p, pubsub, IPNS, and IPFS more generally provide a useful toolkit for building robust abstractions to deliver fast, scalable, data synchronization on a decentralized network.
 
@@
 
-\subsubsection{Identity}
-
-IPFS is an implementation of public key infrastructure, where every node on the IPFS network has a key-pair.
+IPFS is an implementation of PKI, where every node on the network has a key-pair.
 In addition to using the key-pair for secure communication between nodes, IPFS also uses the key-pair as the basis for identity.
-Specifically, when a new IPFS node is created, a new key-pair is generated, and then the public key is transformed into the nodes Peer ID.
-
-\subsubsection{Agent-centric security}
+Specifically, when a new IPFS node is created, a new key-pair is generated, and this public key is used to generate the node's Peer IDentity (Peer ID).
+
+\subsubsection{Agent-centric Security}
 
 Agent-centric security refers to the maintenance of data integrity without leveraging a central or blockchain-based consensus.
-The general approach is just to let the reader enforce permissions and perform validations, not the writer or some central authority.
-Agent-centric security is possible if the reader can reference local-only, tamper-free code or if the state can be used to determine whether a given operation (e.g. delete data) is permitted.
-Many decentralized networks like Secure Scuttlebutt \cite{securescuttlebuttScuttlebuttProtocolGuide} and Holochain \cite{ericharris-braunHolochainScalableAgentcentric2018} make use of agent-centric security.
+The general approach is to let the reader enforce permissions and perform validations, not the writer or some central authority.
+Agent-centric security is possible if the reader can reference local-only, tamper-free code or if the local system state can be used to determine whether a given operation (e.g., delete operation) is permitted.
+Many decentralized networks, such as Secure Scuttlebutt \cite{securescuttlebuttScuttlebuttProtocolGuide} and Holochain \cite{ericharris-braunHolochainScalableAgentcentric2018}, make use of agent-centric security.
 Each of these systems leverage cryptographic signatures to validate peer identities and messages.
@@
 Many make use of an access-control list (ACL), which is a list of permissions attached to an object or group of objects \cite{shireyInternetSecurityGlossary2007}.
-An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete the object (see Figure \ref{tbl:ACL}).
+An ACL determines which users or processes can access an object and whether a particular user or process with access can modify or delete an object (see Figure \ref{tbl:ACL}).
 
@@
 
-Using ACLs in systems where identity is derived from various configurations of public-key infrastructure has been around for some time \cite{herzbergAccessControlMeets2000}.
+Using ACLs in systems where identity is derived from various configurations of PKI has been around for some time \cite{herzbergAccessControlMeets2000}.
 Still, many existing database and communication protocols built on IPFS to date lack support for an ACL or only have primitive ACL support.
@@
 Even more, many systems deploy an all-or-none security model, where those with access to a database have complete access, including write capabilities.
-ACLs should be mutable over time and permission to modify an ACL should also be recorded in an ACL.
+Ideally, ACLs are mutable over time, and permission to modify an ACL should also be recorded in an ACL.
 
 Event-driven systems (e.g., event sourcing) often make use of ACLs with some distinct properties.
-The ACL of an event-driven system is usually a list of access rules built from a series of events.
-For example, the two events, ``grant Bob write access'' and ``revoke read access from Alice'' would together result in a final ACL state where, Bob has read and write access, but Alice does not (Section  \ref{sec:ViewsProjections}).
-
-\section{The Threads protocol}\label{sec:ThreadsProtocol}
+The ACL of an ES-based system is usually a list of access rules built from a series of events.
+For example, the two events, ``grant Bob write access'' and ``revoke read access from Alice'' would together result in a final ACL state where, Bob has read and write access, but Alice does not.
+
+\section{The Threads Protocol}\label{sec:ThreadsProtocol}
 
 We propose Threads, a protocol and decentralized database that runs on IPFS meant to help decouple apps from user-data.
-Inspired by Event Sourcing, CQRS, and Redux, Threads is a protocol for creating and synchronizing state across collaborating peers/replicas on a network.
+Inspired by event sourcing and object-based database abstractions, Threads is a protocol for creating and synchronizing state across collaborating peers on a network.
 Threads offer a multi-layered encryption and data access architecture that enables datasets with independent roles for writing, reading, and following changes.
-By extending on the multiaddress addressing scheme, Threads differs from previous solutions by allowing PULL based replica synchronization in addition to PUSH based synchronization that is common in distributed protocols.
-The flexible event-based structure enables client applications to create advanced applications states, including materialized views, and custom CRDTs.
-
-Threads are topic-based collections of single-writer logs.
+By extending on the multiaddress addressing scheme, Threads differs from previous solutions by allowing \emph{pull}-based replica synchronization in addition to \emph{push}-based synchronization that is common in distributed protocols.
+The flexible event-based structure enables client applications to derive advanced applications states, including queriable materialized views, and custom CRDTs.
+
+In essence, Threads are topic-based collections of single-writer logs.
 Taken together, these logs represent the current ``state'' of an object or dataset.
-The basic units of Threads, namely Logs and Events, provide a framework for users to create, store, and transmit data in a P2P distributed network.
+The basic units of Threads --- Logs and Events --- provide a framework for developers to create, store, and transmit data in a P2P distributed network.
 By structuring the underlying architecture in specific ways, this framework can be deployed to solve many of the problems discussed above.
@@
 This approach works for many cases, but the use of a deterministic resolution strategy can be insufficient in cases with complicated data structures or complicated network topologies.
-A git merge highlights one such example, in which a predetermined merge strategy can be used, but is not often the best choice in practice.
-Furthermore, a multi-writer log using a linked data format (e.g., a Merkle-Dag) in imperfect networking or storage environments can lead to suboptimal states where it may be exceedingly costly (networking, storage, and computation) to regain consistency.
-
-A promising approach is one that can leverage the benefits of both a Merkle-Clock for events from any one peer and a less constrained ordering mechanism to combine events from all peers.
-In this more ideal setup, developers can more freely institute their own CRDTs or domain-specific conflict resolution strategies.
-Additionally, it more naturally supports use-cases where all peers contributing to a dataset may not be interested in following or replicating the events of all other peers.
+A git merge highlights one such example, in which a predetermined merge strategy could be used, but is not often the best choice in practice.
+Furthermore, a multi-writer log using a linked data format (e.g., a Merkle-DAG) in imperfect networking or storage environments can lead to states where it is prohibitively difficult (e.g., due to networking, storage, and/or computational costs) to regain consistency.
+
+A promising approach to dealing with this is to leverage the benefits of both a Merkle-Clock for events from any one peer, and a less constrained ordering mechanism to combine events from all peers.
+In this case, developers can more freely institute their own CRDTs or domain-specific conflict resolution strategies.
+Additionally, it naturally supports use-cases where all peers contributing to a dataset may not be interested in following or replicating the events of all other peers (e.g., in a Pubsub-based system).
 
@@
 
-% [!b] indicates that it will try to be on the bottom of the page
 \begin{figure}[!t]
@@
 Our solution to dealing with log conflicts (i.e., divergent Merkle-Clocks) is to institute a \emph{single-writer rule}: A Log can only be updated by a single replica or \emph{identity}.
-An Event Log is then a \emph{single-writer Merkle-Clock} that can be totally ordered (Figure \ref{fig:EventLog}).
-\emph{Separate} Event Logs can be composed into advanced structures, including CRDTs \cite{enesSingleWriterPrincipleCRDT2017}.
-
-This presents a novel take on multi-writer systems \cite{enesSingleWriterPrincipleCRDT2017}, such that conflict resolution is deferred to a point at which a decision is actually \emph{required}.
-This means that imperfect information may be supplemented along the way (i.e., at a later time), without causing conflicts in the mean time.
-It also means that applications can choose conflict resolution strategies specific to the task at hand.
-For example, if using a downstream CRDT, ordering is actually irrelevant, and can be ignored completely.
+An Event Log is then a \emph{single-writer Merkle-Clock} that can be totally ordered (Figure \ref{fig:EventLog}), and \emph{separate} Event Logs can be composed into advanced structures, including CRDTs \cite{enesSingleWriterPrincipleCRDT2017}.
+
+This presents a novel take on multi-writer systems \cite{enesSingleWriterPrincipleCRDT2017}, such that conflict resolution is \emph{deferred} to a point at which a decision is actually required.
+This means that imperfect information may be supplemented along the way without causing conflicts in the mean time.
+It also means that apps can choose conflict resolution strategies specific to the task at hand.
+For example, if using a downstream CRDT, ordering is actually irrelevant and can be ignored completely.
 Alternatively, an additional clock may be required to ensure consistent ordering, such as a vector or Bloom clock (see Section \ref{sec:LogicalClocks}).
@@
 
+\begin{definition}
+(Writer). The single IPFS Peer capable of writing to an Event Log.
+\end{definition}
+
+\begin{definition}
+(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log's Read Key (Section  \ref{sec:KeysEncryption}).
+\end{definition}
+
+\begin{definition}
+ (Event). A single node in a Merkle-Clock, stored on IPFS.
+\end{definition}
+
 For any given Log, Events are authored by a single IPFS Peer, or \emph{Writer}.
-
-\begin{definition}
-(Writer). The single IPFS Peer capable of writing to an Event Log.
-\end{definition}
-
 This single-writer setup is a core feature of Logs, and provides properties unique to the Threads protocol.
 For clarity, we can similarly define a \emph{Reader} as any other Peer capable of reading a Log.
-
-\begin{definition}
-(Reader). Any Peer capable of reading a Log. Practically speaking, this means any Peer with the Log's Read Key (Section  \ref{sec:KeysEncryption}).
-\end{definition}
-
-As seen above in Section \ref{sec:merkleclocks}, a Merkle-Clock is simply a Merkle-DAG of \emph{Events}:
-
-\begin{definition}
- (Event). A single node in a Merkle-Clock, stored on IPFS.
-\end{definition}
+Related, a Merkle-Clock (see Section \ref{sec:merkleclocks}) is simply a Merkle-DAG of \emph{Events}.
 
@@
 
-Together with a cryptographic signature, an Event is written to a log with an additional node (see Figure \ref{fig:EventLog}) enabling log verification by readers (Section \ref{sec:KeysEncryption}).
-
-At a minimum, a node must link to its most immediate ancestor.
+Together with a cryptographic signature, an Event is written to a log with an additional Node (see Figure \ref{fig:EventLog}) enabling Log verification by Readers (Section \ref{sec:KeysEncryption}).
+At a minimum, a Node must link to its most immediate ancestor.
 However, links to older ancestors are often included as well to improve concurrency during traversal and verification \cite{meyerBamboo2019}.
 
-As shown \ref{appx:EventNode}, an Event's actual content, or body, is contained in a separate node. This allows Events to carry any arbitrary node structure, from complex directories to raw bytes.
-
-Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses\cite{multiformatsMultiaddr}.
+As shown in \ref{appx:EventNode}, an Event's actual content (or body), is contained in a separate Node. This allows Events to carry any arbitrary Node structure, from complex directories to raw bytes.
+
+Much like IPFS Peers, Logs are identified on the network with addresses, or more specifically, with multiaddresses \cite{protocollabsMultiaddr}.
 Here we introduce IPEL, or Interplanetary Event Log, as a new protocol tag to be used when composing Log multiaddresses.
@@
 
-In practice, to reach a Log via it's IPEL multiaddress, it must be encapsulated in an IPFS Peer multiaddress (Example \ref{ex:Multiaddress}).
-
-Unlike peer multiaddresses, Log addresses are not stored in the global IPFS distributed hash table \cite{benetIPFSContentAddressed2014} (DHT).
+Unlike peer multiaddresses, Log addresses are not stored in the global IPFS DHT \cite{benetIPFSContentAddressed2014}.
 Instead, they are collected from Log Events.
@@
 Instead, updates are requested directly from the (presumably trusted) peers that produced them, resulting in a hybrid of content-addressed Events arranged over a data-feed\footnote{This is similar to the append-only message feeds used in Secure Scuttlebutt’s global gossip network \cite{securescuttlebuttScuttlebuttProtocolGuide}} like topology.
-Log addresses are recorded in an address book, similar to IPFS Peer address book (Example \ref{ex:KeyBook}).
-
-Addresses can expire by specifying a time-to-live (TTL) value when adding or updating them in the address book, which allows for unresponsive addresses to eventually be removed.
+Log addresses are recorded in an address book, similar to an IPFS Peer address book (see Example \ref{ex:KeyBook}).
+Addresses can also expire by specifying a time-to-live (TTL) value when adding or updating them in the address book, which allows for unresponsive addresses to eventually be removed.
 
@@
 Therefore, datasets distributed across such networks can be thought of as highly partitioned.
-To ensure updates are available between mostly offline or otherwise disconnected Peers (like mobile devices), Textile Logs are designed with a built-in replication or follower mechanism.
-
-\begin{definition} (Follower).
+To ensure updates are available between mostly offline or otherwise disconnected Peers, Textile Logs are designed with a built-in replication or \emph{Follower} mechanism.
+Followers are represented as additional addresses, meaning that a Log address book may contain \emph{multiple} multiaddresses for a single Log (see Example \ref{ex:Multiaddress}).
+
+\begin{definition}
+  (Follower).
 Log Writers can designate other IPFS Peers to ``follow'' a Log, potentially replicating and/or republishing Events.
 A Follower is capable of receiving Log updates and traversing linkages via the Follow Key (Section  \ref{sec:KeysEncryption}), but is not able to read the Log's contents.
-Followers should be server-based --- i.e., always online and behind a public IP address.\end{definition}
-
-Followers are represented as additional addresses, meaning that a Log address book may contain \emph{multiple} multiaddresses for a single Log (Example \ref{ex:Multiaddress}).
+Followers should be server-based --- i.e., always online and behind a public IP address.
+\end{definition}
 
@@
 Textile Logs are designed to be shared, composed, and layered into datasets (Figure \ref{fig:LogEncryption}).
-Therefore, Logs are encrypted by default in a manner that enables access control (Section  \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
+As such, they are encrypted by default in a manner that enables access control (Section \ref{sec:AccessControl}) and the Follower mechanism discussed in the previous section.
+Much like the Log address book, Log \emph{keys} are stored in a key book (Example \ref{ex:KeyBook}).
 
@@
 
-The body, or content of an event, is encrypted by a \emph{Content Key}.
-Content Keys are generated per-content and never reused.
+The body, or content of an Event, is encrypted by a \emph{Content Key}.
+Content Keys are generated for each piece of content and never reused.
 The Content Key is distributed directly in the header of the Event Block.
-We define the Content Key as follows,
 
@@
   (Content Key).
-The Content Key is a variable-format key used to encrypt the body (content) of an event.
+The Content Key is a variable-format key used to encrypt the body (content) of an Event.
 This key can be symmetric, asymmetric, or possibly non-existent in cases where encryption is not needed.
-One of two common encryption choices will typically be used per event,
+\end{definition}
+
+One of two common encryption choices will typically be used for the Content Key of an Event:
   \begin{enumerate}
     \item When broadcasting events to many possible recipients, a single-use symmetric key is generated per unique content body.
-\item When sending events to specific recipients, the recipient's public key can be used to restrict access from all others\footnote{Much like private messages in Secure Scuttlebutt (\url{https://www.scuttlebutt.nz/concepts/private-message})}.
+    \item When sending events to specific recipients, the recipient's public key can be used to restrict access from all others\footnote{Much like private messages in Secure Scuttlebutt.}.
   \end{enumerate}
-\end{definition}
 
@@
 Therefore, the Event Block itself is further encrypted using a \emph{Read} key.
-The Read Key is not distributed within the Log itself but is distributed to any peers to grant them access to the content of the Log.
+The Read Key is not distributed within the Log itself but via a separate (secure) channel to all Peers who require access to the content of the Log.
 
@@
 Finally, the encrypted Event Block, its signature, and the IPLD linkage(s) from an Event to its antecedents are encrypted together using a Follow Key.
-Follow Keys allow Logs to be \emph{followed} by peers on the network who do not have access to any content within the event, and can only see signatures and linkage(s) between Events.
-
-\begin{definition}
-(Follow Key). The Follow Key is a symmetric key created by the Log owner and used to encrypt the entire event payload before adding the event to the Log.
-\end{definition}
-
-Much like the Log address book, Log keys are stored in a key book (Example \ref{ex:KeyBook}).
+Follow Keys allow Logs to be \emph{followed} by peers on the network who do not have access to any content within the event.
+Followers can only see signatures and linkage(s) between Events.
+
+\begin{definition}
+(Follow Key). The Follow Key is a symmetric key created by the Log owner and used to encrypt the entire Event payload before adding the Event to the Log.
+\end{definition}
 
@@
 
-In Textile, the \emph{interface} to Logs is managed as a Thread.
-As mentioned previously, a Thread is a collection of Logs on a given topic.
-Threads are an event-sourced, distributed database, and can be used to maintain a single, collaboratively edited, watched, or hosted dataset across multiple Peers.
-Threads provide the mechanism to combine multiple Logs from individual authors into singular shared states through the use of either cross-Log sequencing (e.g.
+In Textile, the \emph{interface} to Logs is managed as a Thread, which is a collection of Logs on a given topic.
+Threads are an event sourced, distributed database, and can be used to maintain a single, collaboratively edited, followed, or hosted dataset across multiple Peers.
+Threads provide the mechanism to combine multiple Logs from individual Writers into singular shared states through the use of either cross-Log sequencing (e.g.
 using a Bloom Clock, Merkle-Clock, or Hybrid Logical Clock \cite{kulkarniLogicalPhysicalClocks2014}) or a CRDT (Section \ref{sec:CRDTs}).
@@
 
-A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pub/Sub-based synchronization.
+A unique Thread IDentity (TID) is used to group together Logs which compose a single dataset and as a topic identifier within Pubsub-based synchronization.
 TIDs are defined with the format shown Figure \ref{fig:ThreadID}.
@@
 TIDs share some similarities with UUIDs \cite{leachUniversallyUniqueIDentifier2005} (version and variant) and IPFS-based CIDs and are multibase encoded\footnote{\url{https://github.com/multiformats/multibase}} for maximum forward-compatibility.
+Base32 encoding is used by default, but any multibase-supported string encoding may be used.
 
@@
 
-Base32 encoding is used by default. However, any multibase-supported string encoding may be used.
-
 \begin{definition}
@@
 
-Certain ID \emph{variants} may be more appropriate than the others in specific use cases.
+Certain TID \emph{variants} may be more appropriate than others in specific contexts.
 For example, Textile provides an \emph{access-controlled} Thread variant, which supports various collaborative structures --- e.g., social media feeds, shared documents, blogs, photo albums, etc.
@@
 
-\subsubsection{Log Synchronization} \label{sec:LogSync}
+\subsubsection{Log Synchronization}\label{sec:LogSync}
 
@@
 Inspired by Git\footnote{\url{https://git-scm.com/}}, a reference to the latest Event in a Log is referred to as the \emph{Head} (or sometimes the \emph{root}).
-When a new Event is received, Readers and Followers simply advance their Head reference for the given Log\footnote{This is similar to how a system such as OrbitDB [MISSING] works, except we are tracking \emph{multiple} Heads (one per Log), rather than a single Head.}.
-
-Regardless of the network protocol, Events are transported between Peers in a standardized \emph{Event Envelope}:
-
-\begin{definition}
-(Event Envelope). An over-the-wire message containing an Event and the sender's signature of the Event.
-\end{definition}
-
+When a new Event is received, Readers and Followers simply advance their Head reference for the given Log.
+This is similar to how a system such as OrbitDB \cite{markroberthendersonOrbitDBFieldManual2019} works, except we are tracking \emph{multiple} Heads (one per Log), rather than a single Head.
+
+Regardless of the network protocol, Events are transported between Peers in a standardized \emph{Event Envelope}.
 A new Thread is created by generating a TID and Log.
@@
 \begin{definition}
+  (Event Envelope). An over-the-wire message containing an Event and the sender's signature of the Event.
+\end{definition}
+
+\begin{definition}
   (Invite).
 An Event containing a mapping of Log IDs to Key Sets, which can be used to join a Thread.
-Threads backed by an access control list (Section  \ref{sec:AccessControl}) will also include the current ACL for the Thread in an Invite.
-This enables Peers to invite others to only read or follow a Thread, instead of becoming a full-on collaborator, i.e., a new Log Writer.
-\end{definition}
-
-\begin{definition}
-(Key Set). A set of keys for a Log. Depending on the context, a Key Set may contain the Follow and Read Key, or just the Follow Key. Encrypted with the recipient's public key.
+Threads backed by an ACL (Section  \ref{sec:AccessControl}) will also include the current ACL for the Thread in an Invite.
+This enables Peers to invite others to only read or follow a Thread, instead of becoming a full-on collaborator (i.e., a new Log Writer).
+\end{definition}
+
+\begin{definition}
+  (Key Set). A set of keys for a Log. Depending on the context, a Key Set may contain the Follow and Read Key, or just the Follow Key. A Key Set is encrypted with the recipient's public key.
 \end{definition}
@@
 Instead of encrypting a Join with the public key of each existing Writer, we can encrypt a single Join with the Key Set of the inviting Peer's Log, which the other Writers also have.
-
 Once a Peer has accepted an Invite, it will receive new Events from Log Writers.
@@
 
-Sending is performed in multiple phases because, invariably, some Thread participants will be offline or unresponsive.
+Sending is performed in multiple phases because, invariably, some Thread participants will be offline or unresponsive:
 
 \begin{enumerate}
-\item New Events are pushed\footnote{Here push means ``send to multiaddress(es)'', which may designate different protocols, e.g., p2p, HTTP, Bluetooth, etc.} directly to the Thread's other Log Writers.
+  \item New Events are pushed\footnote{Here push means ``send to multiaddress(es)'', which may designate different protocols, e.g., P2P, HTTP, Bluetooth, etc.} directly to the Thread's other Log Writers.
   \item New Events are pushed directly to the target Log's Follower(s), who may not maintain their own Log.
-\item New Events are published over Libp2p's gossip-based Pub/Sub infrastructure using TID as a topic, which provides potentially unknown Readers or Followers with an opportunity to consume Events in real-time.
+  \item New Events are published over gossip-based Pubsub using TID as a topic, which provides potentially unknown Readers or Followers with an opportunity to consume Events in real-time.
 \end{enumerate}
@@
   \item New Events may be pushed to the Thread's other Log Writers via federated notification services like Apple Push Notification Service (APNS), Google Cloud Messaging (GCM), Firebase Cloud Messaging (FCM), and/or Windows Notification Service (WNS).
-\item New Events may trigger web-hooks\footnote{\url{https://en.wikipedia.org/wiki/Webhook}}, which could enable many complex (e.g., IFTTT\footnote{\url{https://ifttt.com}}) workflows.
+  \item New Events may trigger web-hooks, which could enable many complex (e.g., IFTTT\footnote{\url{https://ifttt.com}}) workflows.
 \end{enumerate}
@@
 
-There are multiple paths to receiving new Events, that together, maximize connectivity between Peers who are often offline or unreachable.
+There are multiple paths to receiving new Events, that together maximize connectivity between Peers who are often offline or unreachable.
 
 \begin{enumerate}
-\item Log Writers can receive Events directly from the author.
-\item Events can be pulled from replicating Followers via HTTP, libp2p, RSS, Atom, etc.
-    a.
-In conjunction with push over WebSockets (seen in Step 2 of the additional push mechanisms above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events (Figure \ref{fig:Pulling}).
+  \item Log Writers can receive Events directly from the Writer.
+  \item Events can be pulled from Followers via HTTP, libp2p, RSS, Atom, etc.
+
+    \begin{enumerate}
+      \item In conjunction with push over WebSockets (seen in Step 2 of the additional push mechanisms above), this method provides web-based Readers and Followers with a reliable mechanism for receiving Log Events (Figure \ref{fig:Pulling}).
+    \end{enumerate}
   \item Writers and readers can receive new Events via a Pub/Sub subscription at the TID.
@@
 The notion of the Follow Key (Section  \ref{sec:KeysEncryption}) makes duplicating all Log Events trivial.
-This allows any Peer in the network to be granted the responsibility of replicating data from another Peer without having read access to the data contained within the Log entries.
+This allows any Peer on the network to be granted the responsibility of replicating data from another Peer without having read access to the raw Log entries.
 This type of Log replication can act as a data backup mechanism.
-It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service is granted read access to the Log Events (Section  \ref{sec:LogSync}).
+It can also be used to build services that react to Log Events, potentially pushing data to disparate, non-Textile systems, especially if the replication service \emph{is} granted read access to the Log Events (Section  \ref{sec:LogSync}).
 
@@
 Application state management tools such as Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen.
-If you take a look at the architecture of such an application, you'll notice several core similarities to CQRS and ES systems (see \cite{facebookFluxInDepthOverview2019} and/or \cite{abramovCaseFlux2015}).
+The architecture of such an application shares several core features with CQRS and ES systems (see \cite{facebookFluxInDepthOverview2019} and/or \cite{abramovCaseFlux2015}).
 In both cases, the focus is on unidirectional data flows that build downstream views from atomic updates in the form of events (or actions).
@@
 
-Like most CQRS + ES based systems, \emph{Events} are at the heart of Threads.
+Like most CQRS and ES based systems, \emph{Events} are at the heart of Threads.
 Every update to local and shared (i.e., across peers) application state happens via Events (see also Section \ref{sec:cqrs}).
-Events are similar to actions in Redux, and are used to describe ``changes to an application state'' \cite{fowlerEventSourcing} (e.g., a photo was added, an item was added to a shopping cart, etc).
+Events are used to describe ``changes to an application state'' \cite{fowlerEventSourcing} (e.g., a photo was added, an item was added to a shopping cart, etc).
 Related, \emph{Event Creators} are used to send Events from a \emph{local application} to an internal Event Store.
@@
 As shown in Figure \ref{fig:Architecture}, the Dispatcher is at the center of the ES system.
-All Events must go through the singleton Dispatcher, whether these be from local Event Creators, or Network Peers.
+All Events must go through the singleton Dispatcher, whether these be from local Event Creators or Network Peers.
 The Dispatcher is responsible for ensuring that incoming Events are persisted to the Event Store (the ``source of truth'' for the system), as well as dispatched to downstream view Models by way of a set of Reducer functions.
@@
 As in many CQRS-based systems, Events are dispatched on the write side, and are \emph{reacted to} on the read side.
-Reactions happen via Reducer fuctions, which cause updates to view Models, which in turn provide interfaces for queries and accessing persisted view \emph{Stores} (state).
+Reactions happen via Reducer functions, which cause updates to view Models, which in turn provide interfaces for queries and accessing persisted view \emph{Stores} (state).
 View Stores are then responsible for notifying downstream consumers (application logic) of changes to their state via a \emph{Broadcaster} (i.e., event emitter).
@@
 In practice, a view Model may \emph{also} wrap the Event Creator application logic that is used to generate upstream Events.
-This provides an intuitive, singular access point to the \emph{local} system, while also leaving room for updates via \emph{external} peer-events.
+This provides an intuitive, singular access point to the \emph{local} system, while also leaving room for updates via \emph{external} Peer Events.
 In practice, the Store is a lightweight interface that can be implemented by one of many database management systems (\ref{appx:DBMS}) providing end users/developers the greatest flexibility.
@@
 The Peer Host is responsible for dealing with incoming Events (be they push or pull).
-These Events are no different from locally-derived Events, though in practice, the Peer Host is required to validate incoming Events before they are dispatched to the system via the Dispatcher.
+These Events are no different from locally-derived Events, though in practice the Peer Host is required to validate incoming Events before they are dispatched to the system via the Dispatcher.
 
@@
 To make Threads as easy to adopt and use as possible, Textile has designed a developer facing API on top of the Threads internals that simplifies dealing with events and data, while still maintaining the power and flexibility of CQRS and ES.
-Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities.
-These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
+Developers should not have to learn a whole new set of terms and tools to take advantage of Threads' capabilities.
+These simple, public-facing APIs will be comfortable to application developers looking to leverage distributed systems that leverage user-siloed data, with minimal configuration and maximum interoperability.
 Inspired by tools such as MondoDB\footnote{\url{http://www.mongodb.com}}, Parse\footnote{\url{https://parseplatform.org}}, and Realm\footnote{\url{https://realm.io}}, as well as the idea of bounded context and aggregate roots from DDD, we provide simple abstractions for performing distributed database operations as if they were local.
@@
 Indeed, the components of a Thread already provide several features that you would expect when operating on a dataset or table within a database: each Thread is defined by a unique ID, and provides facilities for access control and permissions, networking, and more.
-To illustrate how these underlying components can we combined to produce a simple API with minimal configuration and boilerplate, consider the following example, in which we provide pseudo-code for a hypothetical Photos application.
+To illustrate how these underlying components can be combined to produce a simple API with minimal configuration and boilerplate, consider the following example in which we provide pseudo-code for a hypothetical Photos app.
 
@@
 A typical Thread-level ACL JSON (see Example \ref{ex:AclJson}) can be persisted to a local Event Store as part of the flow described in Section \ref{sec:internals}.
-See also Section \ref{sec:interfaces}, and in particular, Example \ref{ex:others} for the public API for editing ACL definitions.
+See also Section \ref{sec:interfaces}, and in particular Example \ref{ex:others} for the public API for editing ACL definitions.
 
@@
 Here, \texttt{12D..dwaA6Qe} is likely the owner, \texttt{12D..dJT6nXY} is a designated follower, and \texttt{12D..P2c6ifo} has been given read access.
-
 A Thread-level ACL has it's own document ACL, which also applies to all other document ACLs (see Example \ref{ex:ThreadAcl}).
@@
 
-In this paper, we described the challenges and considerations when attempting to create a protocol suitable for large-scale data storage, synchronization, and use in a distributed system.
+In this paper, we described the challenges and considerations when creating a protocol suitable for large-scale data storage, synchronization, and use in a distributed system.
 We identified six requirements for enabling \emph{user-siloed} data: flexible data formats, efficient synchronization, conflict resolution, access-control, scalable storage, and network communication.
-We presented a novel solution to satisfy each of these six requirements.
-We have introduced a solution that extends on IPFS and prior research done by Textile and others, which we term Threads.
+We have introduced a solution to these requirements that extends on IPFS and prior research done by Textile and others, which we term Threads.
 Threads are a novel data architecture that builds upon a collection of protocols to deliver a scalable and robust storage system for end-user data.
@@
 We show that the flexible core structure of single-writer append-only logs can be used to compose higher-order structures such as Threads, Views, and/or CRDTs.
-In particular, we show that through the design of specific default Views, we can support important features such as access control lists.
+In particular, we show that through the design of specific default view Models, we can support important features such as access control lists and common and/or complex data models.
 The Threads protocol described here is flexible enough to derive numerous specific database types (e.g.
@@
 The cryptography used throughout Threads will help shift the data ownership model from apps to users.
-
-TABLE
 
@@
 
-\ack{Thanks to our funders and the excellent Textile community for helpful suggestions, feature requests, and reviews.}
+\ack{Thanks to our funders and the excellent Textile community for helpful suggestions, feature requests, and reviews.
+The work presented here has also benefitted from numerous discussions with members of the decentralized web community.}
 

### 2019-09-29 draft.tex
abstract: cleans up
Signed-off-by: andrew <andrew@textile.io>

@@
 As the Internet expands, the division between a person's digital and physical existence continues to blur.
-An emerging issue of concern is that the digital part of a person's life is mainly being captured by apps and stored away from the person's control by the companies building those apps.
-An alternative, called user-siloed data is one area of research that if implemented successfully, could reverse the flow of value derived from personal data, from apps to users.
+An emerging issue of concern is that the digital part of a person's life is being stored away from the person's control by companies behind apps and services on the Internet.
+An alternative architecture for data on the web, called user-siloed data, aims to reverse the flow of data and its derived value so that users, not apps capture it.
 In this paper, we investigate the data formats, access control, and transfer protocols necessary to build a system for user ownership of large-scale digital datasets.
-The aim of the proposed system is to help power a new generation of web technologies.
-Our solution combines a novel use of event sourcing with Interplanetary Linked Data (IPLD) to provide a distributed, scalable, and flexible database solution for decentralized applications.
+The proposed system aims to help power a new generation of web technologies.
+Our solution combines a novel use of event sourcing, Interplanetary Linked Data (IPLD), and access control to provide a distributed, scalable, and flexible database solution for decentralized applications.
 \end{abstract}
@@
 
-\subsubsection{Enhanced Log Security}
+\subsubsection{Textile: The Thread \& Bot Network} \label{sec:Bots}
 
@@
 
-\subsubsection{Textile: The Thread \& Bot Network} \label{sec:Bots}
+\subsubsection{Enhanced Log Security}
 

### 2019-09-29 draft.tex
more edits
Signed-off-by: andrew <andrew@textile.io>

@@
 
-While companies collecting user data may not itself be a significant problem, it has been shown that, over time company incentives often shift from \emph{providing} value to users, to \emph{extracting} value from users \cite{dixonWhyDecentralizationMatters2018}.
-When this incentive shift happens for companies that have also been hoarding user data, that data may be a new source of value/revenue for the company.
+While the fact that companies collect user data may not itself be a significant problem, problems may arise if over time a company's incentives shift from \emph{providing} value to users, to \emph{extracting} value from users \cite{dixonWhyDecentralizationMatters2018}.
+When this incentive shift happens, companies that have been creating data-silos may treat that data as new source of value or revenue for the company.
 It may be possible to stop this trend through extreme privacy measures, government intervention/legislation, or by boycotting companies that collect any form of data.
@@
 Other projects have identified app-siloed data as a problem \cite{bernersLee_readwrite_2013,de_montjoye_trusted_2012}, and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}.
-However, none so far have addressed the problem directly, by providing a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data.
+However, none so far have addressed the problem of \emph{how data should be collected}, by providing a sufficiently interoperable protocol for scalable storage, transmission, and use of user-siloed application data.
 
@@
 We provide an overview of some important technologies related to the Interplanetary File System (IPFS) that make it possible to rethink ES in a decentralized network.
-Finally, we cover challenges to security and access control on an open and decentralized network, and discuss how they are used in popular databases outside the IPFS context.
+Finally, we cover challenges to security and access control on an open and decentralized network, and discuss how they are designed in popular database management (DBMS) systems outside the IPFS context.
 
@@
 
-When developing large-scale software systems, it is common to store data in a relational database management system (RDBMS).
-To model realistic systems, this type of framework often requires complex techniques for mapping data between domain models and database tables, where the same data model is used to both query and update a database.
-A powerful alternative is to use a set of append-only logs to model the state of an object simply by applying its change sequence in the correct order.
+To model realistic systems, apps often need to map data between domain models and database tables, where the same data model is used to both query and update a database.
+To solve \emph{synchronization} it is often helpful to handle updates on just the database and then provide the query and interfaces only later in a DBMS.
+One powerful approach to synchronization is to use a set of append-only logs to model the state of an object simply by applying its change sequence in the correct order.
 This concept can be expressed succinctly by the state machine approach \cite{schneiderImplementingFaulttolerantServices1990}: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state.
@@
 
-With these interface simplifications, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
+With these interface simplifications, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores or DBMS.
 Here we draw inspiration from similar projects (e.g., OrbitDB \cite{markroberthendersonOrbitDBFieldManual2019}) which have made it much easier for developers familiar with centralized database systems to make the move to decentralized systems such as Threads.
@@
 Each database style would be implemented as a standalone wrapper/software library, allowing application developers to pick and choose the solution most useful to the application at hand.
-Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these ``example'' libraries.
+Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these \emph{reference} libraries.
 
@@
 
-At an abstract level, the Textile protocol provides a distributed framework for building shared, offline first, Stores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable.
-Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS and ES (and related) literature to ``extend'' the Threads protocol with additional features and controls.
+The Textile protocol provides a distributed framework for building shared, offline first, Stores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable.
+Any internal implementation details of a compliant Threads \emph{client} may use any number of well-established design patterns from the CQRS and ES (and related) literature to \emph{extend} the Threads protocol with additional features and controls.
 Indeed, by designing our system around Events, a Dispatcher, and generic Stores, we make it easy to extend Threads in many different ways.
-Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility this design affords.
+Some extensions included by Textile's Threads implementation are outlined in this section to provide some understanding of the extensibility this design affords.
 
@@
 We show that the flexible core structure of single-writer append-only logs can be used to compose higher-order structures such as Threads, Views, and/or CRDTs.
-In particular, we show that through the design of specific default view Models, we can support important features such as access control lists and common and/or complex data models.
+In particular, we show that through the design of specific default view Models, we can support important features such as access control lists and common, specialized, or complex data models.
 The Threads protocol described here is flexible enough to derive numerous specific database types (e.g.
@@
 
-\subsubsection{Textile: The Thread \& Bot Network} \label{sec:Bots}
-
-Threads change the relationship between a user, their data, and the services they connect with that data.
-The nested, or multi-layered, encryption combined with powerful ACL capabilities create new opportunities to build distributed services, or Bots, on a network of Textile peers.
-Based on the Follow Key now available in Threads, Bots would be able to relay, replicate, or store data that is synchronized via real-time updates in a \emph{trust-less} way.
-Bots can additionally enhance the IPFS network by providing a framework to build and deploy many kinds of trust-less or trusted services.
-Examples include simple archival, caching and republishing Bots.
-Other examples include payment, re-encryption, or bridges to Web 2.0 services to offer decentralized access to Web 2.0.
-
 \subsubsection{Enhanced Log Security}
@@
 This is a major advantage of tools such as reSolve\footnote{\url{https://reimagined.github.io/resolve/}}, where ``system changes can be reflected immediately [on the frontend], without the need to re-query the backend'' \cite{ereminReduxInspiredBackend2019}.
-Textile developers will develop frameworks to more directly expose the internals of Threads to frontend SDKs, making it possible to sync application state across and between apps and services on the Textile network.
+Textile developers will create frameworks to more directly expose the internals of Threads to frontend SDKs (or DMBS, see \ref{appx:DBMS}), making it possible to sync application state across and between apps and services on the IPFS network.
+
+\subsubsection{Textile: The Thread \& Bot Network} \label{sec:Bots}
+
+Threads change the relationship between a user, their data, and the services they connect with that data.
+The nested, or multi-layered, encryption combined with powerful ACL capabilities create new opportunities to build distributed services, or Bots, in the network of IPFS peers.
+Based on the Follow Key now available in Threads, Bots can relay, replicate, or store data that is synchronized via real-time updates in a \emph{trust-less}, partially trusted, or fully-trusted way.
+Bots can additionally enhance the IPFS network by providing a framework to build and deploy many new kinds of services available over HTTP or P2P.
+Services could include simple data archival, caching and republishing, translation, data conversion, and more.
+Advanced examples could include payment, re-encryption, or bridges to Web 2.0 services to offer decentralized access to Web 2.0.
 

### 2019-09-30 draft.tex
update date
Signed-off-by: andrew <andrew@textile.io>
adding Draft to page title
Signed-off-by: andrew <andrew@textile.io>
fix for carson
Signed-off-by: andrew <andrew@textile.io>

@@
 
-\title[Textile Threads]{A protocol \& event-sourced database for decentralized user-siloed data}
+\title[Draft - Textile Threads]{A protocol \& event-sourced database for decentralized user-siloed data}
 \author{Pick}
@@
 
-\shortauthors{www.textile.io}
-
-\revised{18 September 2019}
+\shortauthors{draft - www.textile.io}
+
+\revised{29 September 2019}
 
@@
 
-One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data.
+One of the most exciting aspects of user-siloed data is the ability to build data-driven apps and services while users remain in control of their own data.
 Other projects have identified app-siloed data as a problem \cite{bernersLee_readwrite_2013,de_montjoye_trusted_2012}, and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}.
-However, none so far have addressed the problem of \emph{how data should be collected}, by providing a sufficiently interoperable protocol for scalable storage, transmission, and use of user-siloed application data.
+However, none so far have addressed the problem of \emph{how data should be collected to make it extensible}, nor have they provided a sufficiently interoperable protocol for scalable storage, transmission, and use of user-siloed application data.
 

### 2019-09-30 draft.tex
fix fluff
Signed-off-by: andrew <andrew@textile.io>

@@
 An alternative architecture for data on the web, called user-siloed data, aims to reverse the flow of data and its derived value so that users, not apps capture it.
-In this paper, we investigate the data formats, access control, and transfer protocols necessary to build a system for user ownership of large-scale digital datasets.
+In this paper, we investigate the data formats, access control, and transfer protocols necessary to build a system for user ownership of data.
 The proposed system aims to help power a new generation of web technologies.

### 2019-09-30 draft.tex
more edits
Signed-off-by: andrew <andrew@textile.io>

@@
 For example, a key-value store built on Threads would ``map'' key-value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del} to an internal (i.e., private) Model as in the previous section, with similarly defined methods.
-The generated Events would then mutate the internal (map-like) view Model effectively encapsulating the entire Event Store in a database structure that satisfies a given interface (see Figure \ref{ex:KVStore} for example).
+The generated Events would then mutate the internal map-like view Model effectively encapsulating the entire Event Store in a database structure that satisfies a given interface (see Figure \ref{ex:KVStore} for example).
 These too would be distributed as Modules, making it easy for developers to swap in/substitute existing backend infrastructure.
@@
 
-Similar abstractions could (and will) be used to implement additional database types and functions.
+Similar abstractions will be used to implement additional database types and functions.
 Tables, feeds, counters, and other simple stores can also be built on Threads.

### 2019-09-30 draft.tex
udate: cmd -> local event, auto-date
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

@@
 \usepackage{dblfloatfix}    % To enable figures at the bottom of page
+
+\usepackage[useregional]{datetime2}
 
@@
 
-\revised{29 September 2019}
+\revised{\today}
 
@@
 We adopt a similar flow in Threads (see Figure \ref{fig:Architecture}).
-Like any ES-based system, Threads are built on \emph{Events}.
+Like any CQRS/ES-based system, Threads are built on \emph{Events}.
 Events are similar to actions in a Flux-based system, and are used to produce \emph{predictable} updates to downstream state.
@@
 The Dispatcher then stores the derived Event in an \emph{Event Store}, and calls a set of registered \emph{Reducer} functions that mutate a (set of) downstream view \emph{Stores} defined by a corresponding set of view \emph{Models}, all within a single \emph{Transaction}.
-This unidirectional, transaction-based system, provides a flexible framework for building complex event-driven application logic.
+This unidirectional, transaction-based system provides a flexible framework for building complex event-driven application logic.
 
@@
 
-Like most CQRS and ES based systems, \emph{Events} are at the heart of Threads.
-Every update to local and shared (i.e., across peers) application state happens via Events (see also Section \ref{sec:cqrs}).
+\emph{Events} are at the heart of Threads --- every update to local and shared (i.e., across peers) state happens via Events (see also Section \ref{sec:cqrs}).
 Events are used to describe ``changes to an application state'' \cite{fowlerEventSourcing} (e.g., a photo was added, an item was added to a shopping cart, etc).
-Related, \emph{Event Creators} are used to send Events from a \emph{local application} to an internal Event Store.
+Related, \emph{Event Creators} are used to send Events from a \emph{local} application to an internal Event Store.
 If built around a specific \emph{domain}, Creators provide bounded context that can be roughly compared to an aggregate root in DDD \cite{evansDomaindrivenDesignTackling2004a}.
@@
 
-In order to actually persist and dispatch Events to downstream view Models, a \emph{Dispatcher} is used.
+In order to persist and dispatch Events to downstream view Models, a \emph{Dispatcher} is used.
 As shown in Figure \ref{fig:Architecture}, the Dispatcher is at the center of the ES system.
-All Events must go through the singleton Dispatcher, whether these be from local Event Creators or Network Peers.
+All Events must go through the singleton Dispatcher, whether these be from local Event Creators or remote Peers.
 The Dispatcher is responsible for ensuring that incoming Events are persisted to the Event Store (the ``source of truth'' for the system), as well as dispatched to downstream view Models by way of a set of Reducer functions.
@@
 
-\subsubsection{External Events}\label{sec:external}
+\subsubsection{Remote Events}\label{sec:external}
 
 In addition to locally derived Events (i.e., from application logic and user interactions), Threads are designed so that Peers may collaborate in a given Thread via Events.
-Events generated by other, network peers are called Peer Events, and they enter the system via a Peer Host.
+Events generated by other, network peers are called \emph{Remote Events}, and they enter the system via a Peer Host.
 The Peer Host is responsible for dealing with incoming Events (be they push or pull).
-These Events are no different from locally-derived Events, though in practice the Peer Host is required to validate incoming Events before they are dispatched to the system via the Dispatcher.
-
-% \begin{definition}
-%   (Reducer).
-Reducers are ``pure'' (in the functional sense) functions\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}, and are written to modify a Store's state based on incoming Actions.
-% \end{definition}
-
-% \begin{definition}
-%   (Resolver).
-Resolvers are simply pre-defined query functions that exist on the Store instance (i.e., the Store's API).
-The interesting thing here is that these can similarly be defined around a domain to produce domain aggregates (in the DDD sense).
-% \end{definition}
-
-% \begin{definition}
-%   (Store).
-Stores provide flexible APIs for persisting and accessing state.
-They are defined by a Reducer function for handling incoming Actions, and a set of query Resolver functions for handling queries.
-They are additionally responsible for notifying downstream consumers of updates to their state via a Broadcaster.
-% \end{definition}
+These Events are no different from \emph{Local Events}, though in practice the Peer Host is required to validate Remote Events before they are dispatched to the internal system.
 

### 2021-03-31 whitepaper/old_draft.tex
feat: updates to use new pandoc and features
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

--- 2019-09-30 draft.tex
@@
 To support such requirements, developers have begun to utilize event sourcing and command query responsibility segregation (CQRS) patterns \cite{bettsExploringCQRSEvent2013}, relying on append-only logs to support immutable state histories.
-Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate ES and CQRS-based designs, including Event Store \cite{EventStore}, Apache Kafaka \cite{ApacheKafka} and Samza \cite{ApacheSamza}, among others \cite{kleppmannDesigningDataintensiveApplications2017}.
+Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate ES and CQRS-based designs, including Event Store \cite{EventStore}, Apache Kafka \cite{ApacheKafka} and Samza \cite{ApacheSamza}, among others \cite{kleppmannDesigningDataintensiveApplications2017}.
 
